### 2023-02-13 paper/markdown/manuscript.Rmd
Import and explorative data analysis
Missingness, characteristic of the study cohort, comparison of included/excluded individuals, numeric variable distribution tests and consistency of the mental health battery tools (McDonald's omega)

---
title: "Three distinct patterns of mental health response after a sport accident"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author: 
  - Katharina Hüfner, MD:
      institute: [psych]
      email: Katharina.Huefner@i-med.ac.at
      correspondence: true
  - Hanna Salvotti, MD:
      institute: [psych]
  - Piotr Tymoszuk, PhD:
      institute: [daas]

institute:
  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria

keywords: "sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: ptsd_biblio.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600)

set_flextable_defaults(font.family = 'Cambria', 
                       font.size = 10)

```

\newpage

# Abstract

\newpage

# Keywords

sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth

\newpage

# Introduction

\newpage

# Methods

\newpage

# Results

\newpage

# Discussion

# Conclusion

\newpage

# Acknowledgments

# Author's contribution

# Conflict of interest

# Data and code availability

An R data (RDa) file with anonymized patient data will be made available upon request to the corresponding author. 
The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.

\newpage

# Tables

```{r tab-cohort-demo, tab.cap = 'Demographic and socioeconomic characteristic of the study cohort. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(tables$cohort_demo) %>% 
  flextable::width(1:2, width = 5.5, unit = 'cm') %>% 
  theme_vanilla

```

\newpage

```{r tab-cohort-accident, tab.cap = 'Characteristic of the sport accident, injury, psychological management and accident consequences. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(tables$cohort_accident) %>% 
  flextable::width(1, width = 5.5, unit = 'cm') %>% 
  flextable::width(1:2, width = 7, unit = 'cm') %>% 
  footnote(2, 1, 
           value = as_paragraph('ski/snowdoard: alpine skiing, snowboarding and cross-country skiing\nsledding: sledding or bobsled\nmountain: hiking, climbing and skitouring\nbiking: mountainbike, tour and road cycling'), 
           part = 'body', 
           ref_symbols = 'a') %>% 
  footnote(8, 1, 
           value = as_paragraph('AIS: acute injury severity score'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-cohort-mental, tab.cap = 'Mental health characteristic of the study participants at survey completion. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(tables$cohort_mental) %>% 
  flextable::width(1:2, width = 5.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; PCL-5 DSM-5: PTSD Checklist for DSM-5; PTGI: post-traumatic growth inventory'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

# Figures

```{r fig-consort, fig.width = mm_inch(figures$consort$w), fig.height = mm_inch(figures$consort$h), fig.cap = 'Flow diagram of the analysis inclusion process.'}

figures$consort$plot

```

__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process.__

\newpage

# References

### 2023-02-13 paper/markdown/supplementary_material.Rmd
Import and explorative data analysis
Missingness, characteristic of the study cohort, comparison of included/excluded individuals, numeric variable distribution tests and consistency of the mental health battery tools (McDonald's omega)

---
title: "Three distinct patterns of mental health response after a sport accident"
subtitle: "Supplementary Material"
author: "Psychiatry Study Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: ptsd_biblio.bib
csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 9)

```

\newpage

# Supplementary Methods

\newpage

# Supplementary Tables

```{r tab-analysis-vars, tab.cap = 'Variables used in the analysis pipeline. The table is available in a supplementary Excel file.'}

suppl_tables$analysis_vars[1, 1] %>% 
  mutate(Section = '') %>% 
  set_names(' ') %>% 
  flextable

```

```{r tab-mental-battery, tab.cap = 'Mental health assessment battery. The table is available in a supplementary Excel file.'}

suppl_tables$mental_battery[1, 1] %>% 
  mutate(Section = '') %>% 
  set_names(' ') %>% 
  flextable

```

\newpage

```{r tab-incl-excl, tab.cap = 'Significant and near signifcant (unadjusted p < 0.05) differences between individuals excluded from analysis and analyzed study participants. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tables$incl_excl) %>% 
  flextable::width(1, width = 3.2, unit = 'cm') %>% 
  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
  flextable::width(4, width = 2.4, unit = 'cm') %>% 
  flextable::width(5, width = 2.3, unit = 'cm') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03c/\u00B2 test with Cramer V effec size statistic. P values were corrected for multiple testing with the false discovery method (FDR)'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(3:4, 1, 
           value = as_paragraph('AIS: acute injury severity score'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  footnote(7:10, 1, 
           value = as_paragraph('PTGI: post-traumatic growth inventory'), 
           part = 'body', 
           ref_symbols = 'c') %>% 
  footnote(11, 1, 
           value = as_paragraph('PSS4: 4-item perceived stress scale'), 
           part = 'body', 
           ref_symbols = 'd') %>% 
  footnote(12, 1, 
           value = as_paragraph('SOC-9L: Leipzig 9-item sense of soherence scale'), 
           part = 'body', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

\newpage

# Supplementary Figures

\newpage

# References

### 2023-02-13 exec.R

@@ -1,0 +1,7 @@
+# Launches the entire pipeline
+
+  library(soucer)
+  
+  print(source_all(c('import.R', 
+                     'exploration.R', 
+                     'paper.R')))

### 2023-02-13 exploration scripts/bias.R

@@ -1,0 +1,132 @@
+# Comparison of the individuals included in the analysis 
+# and those excluded by any reason
+
+  insert_head()
+  
+# container ------
+  
+  excl <- list()
+  
+# parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## an analysis table containing the study variables for the excluded
+  ## and included individuals
+  
+  excl$analysis_tbl <- 
+    list(included = ptsd$dataset, 
+         excluded = ptsd$cleared %>% 
+           filter(!ID %in% ptsd$dataset$ID)) %>% 
+    map(select, 
+        ID, 
+        all_of(eda_globals$variables$variable)) %>% 
+    compress(names_to = 'analysis_status') %>% 
+    mutate(analysis_status = factor(analysis_status, 
+                                    c('included', 'excluded')))
+  
+# Descriptive stats -------
+  
+  insert_msg('Descriptive stats')
+  
+  excl$desc_stats <- excl$analysis_tbl %>% 
+    explore(variables = eda_globals$variables$variable, 
+            split_factor = 'analysis_status', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for differences between the analysis groups ------
+  
+  insert_msg('Testing for differences by analysis status')
+  
+  ## Mann-Whitney or Chi-squared test
+  ## multiple variables are not normally distributed
+  ##
+  ## working with safely(): some variables are so missing in the excluded 
+  ## set that no testing is possible
+  ##
+  ## FDR correction
+  
+  excl$test <- 
+    list(variables = eda_globals$variables$variable, 
+         types = eda_globals$variables$types) %>% 
+    future_pmap(safely(compare_variables), 
+                excl$analysis_tbl, 
+                split_factor = 'analysis_status', 
+                what = 'eff_size', 
+                ci = FALSE, 
+                exact = FALSE, 
+                pub_styled = TRUE, 
+                .options = furrr_options(seed = TRUE)) %>% 
+    map_dfr(~.x$result) %>% 
+    re_adjust %>% 
+    mutate(plot_cap = paste(eff_size, significance, sep = ', '), 
+           y_lab = ifelse(test == 'Chi-squared test', 
+                          '% of strata', 
+                          exchange(variable, 
+                                   dict = ptsd$var_lexicon, 
[…]

### 2023-02-13 exploration scripts/cohort.R

@@ -1,0 +1,118 @@
+# Cohort characteristic in a tabular form
+# 
+
+  insert_head()
+  
+# container list ------
+  
+  cohort <- list()
+
+# descriptive stats -----
+  
+  insert_msg('Descriptive stats')
+  
+  cohort$desc_stats <- ptsd$dataset %>% 
+    explore(variables = eda_globals$variables$variable, 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# ready to use tables: demographic, accident and mental health features -----
+ 
+  insert_msg('Ready-to-use tables')
+  
+  ## variable categories 
+  
+  cohort$var_class <- 
+    list(demo = c('obs_time', 
+                  'sex', 
+                  'age', 
+                  'residence_alpine_region', 
+                  'education', 
+                  'employment_status', 
+                  'sport_profession', 
+                  'trauma_risk_profession', 
+                  'household_income_class', 
+                  'smoking_status', 
+                  'drug_status', 
+                  'somatic_comorbidity', 
+                  'somatic_comorbidity_type', 
+                  'psych_comorbidity', 
+                  'traumatic_event', 
+                  'prime_trauma_event', 
+                  'prime_trauma_event_past'), 
+         accident = c('prior_accident', 
+                      'sport_type', 
+                      'accident_alone', 
+                      'accident_culprit', 
+                      'accident_injured_persons', 
+                      'accident_rescue', 
+                      'accident_rescue_mode', 
+                      'injury_sev_strata', 
+                      'injury_head', 
+                      'injury_face', 
+                      'injury_neck', 
+                      'injury_chest', 
+                      'injury_abdomen', 
+                      'injury_spine', 
+                      'injury_upper_limbs', 
+                      'injury_lower_limbs', 
+                      'injury_external_other', 
+                      'psych_support_post_accident', 
+                      'psych_therapy_post_accident', 
+                      'accident_aftermath', 
+                      'same_sport_type_post_accident', 
+                      'caution_post_accident'), 
+         mental = c(## gneral measures of mental health and QoL
+                    'pss4_total', 
+                    'gad7_total', 
+                    'gad7_total_class', 
+                    'phq9_total', 
+                    'phq9_total_class', 
+                    'phq_events_total', 
+                    'phq_events_total_class', 
+                    'eurohis_total', 
[…]

### 2023-02-13 exploration scripts/consistency.R

@@ -1,0 +1,287 @@
+# Consistency of the assessment battery components measured by
+# McDonald's omega
+
+  insert_head()
+  
+# container -----
+  
+  cons <- list()
+  
+# parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# globals ------
+  
+  insert_msg('Globals')
+  
+  ## a list with variables
+  
+  cons$var_list <- list(dsm5_total = c('dsm5_B', 
+                                       'dsm5_C', 
+                                       'dsm5_D', 
+                                       'dsm5_E'), 
+                        ptgi_total = c('ptgi_fctI', 
+                                       'ptgi_fctII', 
+                                       'ptgi_fctIII', 
+                                       'ptgi_fctIV', 
+                                       'ptgi_fctV'), 
+                        rs13_total = paste0('rs13_q', 1:13), 
+                        cage_total = paste0('cage_q', 1:4), 
+                        pss4_total = paste0('pss4_q', 1:4), 
+                        phq9_total = paste0('phqd_psych_satisf_q', 1:9), 
+                        gad7_total = paste0('phqd_psych_satisf_q', 10:16), 
+                        phqd_panic_total = paste0('phqd_panic_q', 1:4), 
+                        phq_events_total = paste0('phqd_events_q', 1:13), 
+                        eurohis_total = paste0(c('eurohis_qol', 
+                                                 'eurohis_health', 
+                                                 'eurohis_energy', 
+                                                 'eurohis_finances', 
+                                                 'eurohis_activity', 
+                                                 'eurohis_selfesteem', 
+                                                 'eurohis_relationship', 
+                                                 'eurohis_housing')))
+  
+  ## variable labels
+  
+  cons$var_labs <- cons$var_list %>% 
+    map(~ifelse(stri_detect(.x, regex = '_q\\d+$'), 
+                stri_extract(.x, regex = 'q\\d+$') %>% 
+                  stri_replace(fixed = 'q', replacement = 'Item '), 
+                exchange(.x,  
+                         dict = ptsd$var_lexicon, 
+                         value = 'label') %>% 
+                  stri_replace(fixed = ' score', replacement = '')))
+  
[…]

### 2023-02-13 exploration scripts/distribution.R

@@ -1,0 +1,85 @@
+# Investigating distribution of the numeric variables: 
+# (1) normality by Shapiro-Wilk test
+# (2) compatibility with the Poisson distribution (Mean vs Variance)
+
+  insert_head()
+  
+# container list -----
+  
+  distr <- list()
+
+# normality check with Shapiro-wilk test -----
+  
+  insert_msg('Normality check')
+  
+  ## table
+  
+  distr$norm_results <- explore(ptsd$dataset, 
+                                what = 'normality', 
+                                variables = eda_globals$numeric_vars, 
+                                pub_styled = TRUE)
+  
+  ## summary plot
+  
+  distr$norm_summ_plot <- explore(ptsd$dataset, 
+                                  variables = eda_globals$numeric_vars, 
+                                  what = 'normality', 
+                                  pub_styled = FALSE) %>% 
+    exchange(variable = 'variable', 
+             dict = ptsd$var_lexicon, 
+             value = 'label') %>% 
+    ggplot(aes(x = stat, 
+               y = reorder(variable, stat))) + 
+    geom_bar(stat = 'identity', 
+             color = 'black', 
+             fill = 'steelblue') + 
+    geom_vline(xintercept = 0.9, 
+               linetype = 'dashed') + 
+    globals$common_theme + 
+    theme(axis.title.y = element_blank()) + 
+    labs(title = 'Normality of variable distribution', 
+         subtitle = 'Shapiro-Wilk test, normality cutoff W > 0.9', 
+         x = 'W statistic')
+  
+# Mean and variance comparison -----
+  
+  insert_msg('Mean and variance')
+  
+  ## table
+  
+  distr$mvr_tbl <- explore(ptsd$dataset, 
+                           variables = eda_globals$numeric_vars, 
+                           what = 'list', 
+                           pub_styled = FALSE) %>% 
+    map(~.x$statistic) %>% 
+    map(~tibble(mean = .x$value[1], 
+                sd = .x$value[2])) %>% 
+    map2_dfr(., names(.), 
+             ~mutate(.x, 
+                     variable = .y, 
+                     var = sd^2, 
+                     mvr = mean/var)) %>% 
+    select(variable, mean, sd, var, mvr)
+  
+  ## plotting the Mean-Variance Ratios
+
+  distr$mvr_plot <- distr$mvr_tbl %>% 
+    exchange(variable = 'variable', 
+             dict = ptsd$var_lexicon, 
[…]

### 2023-02-13 exploration scripts/missingness.R

@@ -1,0 +1,73 @@
+# Missing answers per participant and variable in the analysis dataset
+
+  insert_head()
+  
+# container ------
+  
+  miss <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  miss$analysis_mtx <- 
+    ptsd$dataset[c('ID', eda_globals$variables$variable)] %>% 
+    column_to_rownames('ID') %>% 
+    as.matrix
+  
+  miss$analysis_mtx <- is.na(miss$analysis_mtx)
+  
+# record and variable missingness: missing answers per participant ----
+  
+  insert_msg('Missingness per participant and variable')
+  
+  miss$tables[c('variable', 'record')] <- 
+    list(miss$analysis_mtx, t(miss$analysis_mtx)) %>%
+    map(colSums) %>% 
+    map2(., c('variable', 'ID'), 
+         ~compress(.x, names_to = .y, values = 'n')) %>% 
+    map2(., c(nrow(miss$analysis_mtx), 
+              ncol(miss$analysis_mtx)), 
+         ~mutate(.x, 
+                 n_total = .y, 
+                 fraction = n/n_total, 
+                 percent = fraction * 100))
+  
+# plots with percentages of missing entries -------
+  
+  insert_msg('Plotting the percentages of missing entries')
+  
+  miss$plots <- 
+    list(x = miss$tables[c('variable', 'record')], 
+         y = c('Missing entries per variable', 
+               'Missing entries per participant'), 
+         z = c('steelblue', 'coral3'), 
+         w = c('variable', 'ID')) %>% 
+    pmap(function(x, y, z, w, v) x %>% 
+           filter(percent != 0) %>% 
+           ggplot(aes(x = percent, 
+                      y = reorder(.data[[w]], percent))) + 
+           geom_bar(stat = 'identity', 
+                    color = 'black', 
+                    fill = z) + 
+           globals$common_theme + 
+           theme(axis.title.y = element_blank()) + 
+           labs(title = y, 
+                subtitle = c(paste('variables: n =', 
+                                   ncol(miss$analysis_mtx)), 
+                             paste('participants: n =', 
+                                   nrow(miss$analysis_mtx))) %>% 
+                  paste(collapse = ', '), 
+                x = '% of missing entries'))
+  
+  ## labels for variables
+  
+  miss$plots$variable <- miss$plots$variable + 
+    scale_y_discrete(labels = colnames(miss$analysis_mtx) %>% 
[…]

### 2023-02-13 exploration.R

@@ -1,0 +1,72 @@
+# Exploratory data analysis: 
+# cohort characteristic, numeric variable distribution, distribution plots
+# Export of the formatted data tables
+
+# tools ------
+
+  library(plyr)
+  library(tidyverse)
+  library(rlang)
+  library(stringi)
+  library(exda)
+  library(soucer)
+  library(furrr)
+  library(writexl)
+  library(rstatix)
+  library(trafo)
+  library(psych)
+  library(clustTools)
+  
+  insert_head()
+  
+  source_all('./tools/tools.R', 
+             message = TRUE, 
+             crash = TRUE)
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  eda_globals <- list()
+  
+  eda_globals$variables <- ptsd$var_lexicon %>% 
+    filter(type %in% c('characteristic', 'response')) %>% 
+    mutate(type = ifelse(format == 'numeric', 'violin', 'stack'), 
+           y_lab = ifelse(format == 'numeric', 
+                          exchange(variable, 
+                                   dict = ptsd$var_lexicon, 
+                                   value = 'axis_lab'), 
+                          '% of strata'), 
+           plot_title = stri_capitalize(exchange(variable, 
+                                                 dict = ptsd$var_lexicon, 
+                                                 value = 'label')), 
+           plot_subtitle = exchange(variable, 
+                                    dict = ptsd$var_lexicon, 
+                                    value = 'description'), 
+           types = ifelse(type == 'violin', 'wilcoxon_r', 'cramer_v'), 
+           types2 = ifelse(type == 'violin', 'kruskal_etasq', 'cramer_v')) %>% 
+    select(variable, type, y_lab, plot_title, plot_subtitle, types, types2)
+  
+  eda_globals$numeric_vars <- eda_globals$variables %>% 
+    filter(type == 'violin') %>% 
+    .$variable
+  
+  eda_globals$factor_vars <- eda_globals$variables %>% 
+    filter(type != 'violin') %>% 
+    .$variable
+  
+# launching the exploration and export scripts -----
+  
+  insert_msg('Launching the exploration scripts')
+  
+  c('./exploration scripts/missingness.R', 
+    './exploration scripts/cohort.R', 
+    './exploration scripts/distribution.R', 
+    './exploration scripts/consistency.R', 
+    './exploration scripts/bias.R') %>% 
+    source_all(message = TRUE, 
+               crash = TRUE)
+  
+# END -----
+  
+  insert_tail()

### 2023-02-13 import.R

@@ -1,0 +1,808 @@
+# This script imports the data of the CovILD study
+
+# toolbox ----
+
+  library(readxl)
+  library(foreign)
+  library(soucer)
+  library(stringi)
+  library(plyr)
+  library(tidyverse)
+  library(rlang)
+  library(trafo)
+
+  source_all('./tools/tools.R', 
+             message = TRUE, crash = TRUE)
+
+  insert_head()
+  
+# data containers ----
+  
+  ptsd <- list()
+  globals <- list()
+  
+# globals ----
+
+  ## graphics
+  
+  globals$corr_colors <- c('negative' = 'steelblue4', 
+                           'positive' = 'firebrick4', 
+                           'ns' = 'gray60')
+  
+  globals$common_text <- element_text(size = 8, 
+                                      face = 'plain', 
+                                      color = 'black')
+  
+  globals$common_margin <- ggplot2::margin(t = 4, l = 3, r = 2, unit = 'mm')
+  
+  globals$common_theme <- theme_classic() + 
+    theme(axis.text = globals$common_text, 
+          axis.title = globals$common_text, 
+          plot.title = element_text(size = 8, 
+                                    face = 'bold', 
+                                    color = 'black', 
+                                    hjust = 0), 
+          plot.subtitle = globals$common_text, 
+          plot.tag = element_text(size = 8, 
+                                  face = 'plain', 
+                                  color = 'black', 
+                                  hjust = 0), 
+          plot.tag.position = 'bottom', 
+          legend.text = globals$common_text, 
+          legend.title = globals$common_text, 
+          strip.text = globals$common_text,
+          strip.background = element_rect(fill = 'gray95', color = 'gray80'), 
+          plot.margin = globals$common_margin, 
+          panel.grid.major = element_line(color = 'gray90'))
+  
+  ## cluster colors
+  
+  globals$clust_colors <- c('neutral' = 'cadetblue3',
+                            'PTG' = 'darkolivegreen3', 
+                            'PTSD' = 'coral4')
+  
+  ## injury regions
+  
+  globals$injury_vars <- c('injury_head', 
+                           'injury_face', 
+                           'injury_neck', 
+                           'injury_upper_limbs', 
+                           'injury_chest', 
+                           'injury_spine', 
+                           'injury_abdomen', 
+                           'injury_lower_limbs', 
+                           'injury_external_other')
+  
+  globals$injury_labs <- c('Head', 
+                           'Face', 
+                           'Neck', 
[…]

### 2023-02-13 paper scripts/figures.R

@@ -1,0 +1,61 @@
+# Report figures 
+
+  insert_head()
+  
+# container list ------
+  
+  figures <- list()
+  
+# globals ------
+  
+  insert_msg('Globals setup')
+
+ # report_fig$radial_panel <- 
+ #   plot_grid(part_clust$radial_plot + 
+  #              theme(plot.tag = element_blank(), 
+  #                    legend.position = 'none'), 
+   #           plot_grid(get_legend(part_clust$radial_plot), 
+    #                    ggdraw() + 
+     #                     draw_text(part_clust$n_tag %>% 
+     #                                 stri_replace_all(fixed = ', ', 
+      #                                                 replacement = '\n'), 
+      #                              size = 8, 
+      #                              hjust = 0, 
+      #                              x = 0.2), 
+     #                   nrow = 2), 
+     #         ncol = 2, 
+     #         rel_widths = c(0.7, 0.3))
+    
+# Figure 1: flow diagram --------
+  
+  insert_msg('Figure 1: flow diagram')
+  
+  figures$consort <- 
+    plot_grid(ggdraw() + 
+                draw_image('./aux files/consort.png')) %>% 
+    as_figure(label = 'figure_1_flow_diagram', 
+              ref_name = 'consort', 
+              caption = paste('Flow diagram of the analysis', 
+                              'inclusion process.'), 
+              w = 90, 
+              h = 2166/2106 * 90)
+  
+# Saving the figures on the disc -------
+  
+  insert_msg('Saving the figures on the disc')
+  
+  figures$radial_panel <- NULL
+  
+  figures <- compact(figures)
+  
+  figures %>% 
+   walk(pickle, 
+         path = './paper/figures', 
+         format = 'pdf', 
+         device = cairo_pdf)
+  
+# END -----
+  
+  insert_tail
+  
+  

### 2023-02-13 paper scripts/render.R

@@ -1,0 +1,24 @@
+# Renders the manuscript and the supplementary material
+
+  insert_head()
+  
+# supplementary material ------
+  
+  insert_msg('Rendering the supplements')
+  
+  render('./paper/markdown/supplementary_material.Rmd', 
+         output_format = word_document2(number_sections = FALSE, 
+                                        reference_docx = 'ms_template.docx'), 
+         output_dir = './paper')
+
+# paper -----
+  
+  insert_msg('Rendering the paper')
+
+  render('./paper/markdown/manuscript.Rmd', 
+         output_format = my_word(), 
+         output_dir = './paper')
+  
+# END -----
+  
+  insert_tail()

### 2023-02-13 paper scripts/tables.R

@@ -1,0 +1,130 @@
+# Manuscript and supplement tables 
+
+  insert_head()
+  
+# container list ----
+  
+  tables <- list()
+  suppl_tables <- list()
+  
+# Tables 1 - 3: demographic and socioeconomic characteristic of the cohort ------
+  
+  insert_msg('Table 1 - 3: study characteristic')
+  
+  tables[c('cohort_demo', 
+           'cohort_accident', 
+           'cohort_mental')] <- 
+    cohort$result_tbl[c("demo", "accident", "mental")] %>% 
+    list(x = ., 
+         label = c('table_1_cohort_demography', 
+                   'table_2_cohort_accident', 
+                   'table_3_cohort_mental_health'), 
+         ref_name = c('cohort_demo', 
+                      'cohort_accident', 
+                      'cohort_mental'), 
+         caption = c(paste('Demographic and socioeconomic characteristic', 
+                           'of the study cohort.', 
+                           'Numeric variables are presented as medians', 
+                           'with interquartile ranges (IQR). Categorical', 
+                           'variables are presented as percentages', 
+                           'and counts within the complete', 
+                           'observation set.'), 
+                     paste('Characteristic of the sport accident,', 
+                           'injury, psychological management', 
+                           'and accident consequences.', 
+                           'Numeric variables are presented as medians', 
+                           'with interquartile ranges (IQR). Categorical', 
+                           'variables are presented as percentages', 
+                           'and counts within the complete', 
+                           'observation set.'), 
+                     paste('Mental health characteristic of the study', 
+                           'participants at survey completion.', 
+                           'Numeric variables are presented as medians', 
+                           'with interquartile ranges (IQR). Categorical', 
+                           'variables are presented as percentages', 
+                           'and counts within the complete', 
+                           'observation set.'))) %>% 
+    pmap(mdtable)
+  
+# Supplementary Table S1: variables used in the analysis pipeline -----
+  
+  insert_msg('Table S1: variables used in the analysis pipeline')
+  
+  suppl_tables$analysis_vars <- ptsd$var_lexicon %>% 
+    filter(variable %in% eda_globals$variables$variable) %>% 
+    transmute(Section = section, 
+              Variable = label, 
[…]

### 2023-02-13 paper.R

@@ -1,0 +1,35 @@
+# Report-generating scripts
+
+# tools -----
+
+  library(plyr)
+  library(tidyverse)
+  library(exda)
+  library(cowplot)
+  library(figur)
+  library(writexl)
+  library(knitr)
+  library(rmarkdown)
+  library(bookdown)
+  library(flextable)
+  library(soucer)
+  library(rmdformats)
+  library(trafo)
+
+  insert_head()
+  
+  source_all('./tools/tools.R', 
+             message = TRUE, crash = TRUE)
+  
+# Report scripts -----
+  
+  insert_msg('Launching the report scripts')
+  
+  c('./paper scripts/tables.R', 
+    './paper scripts/figures.R', 
+    './paper scripts/render.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+
+# END -----
+  
+  insert_tail()

### 2023-02-13 tools/tools.R

@@ -1,0 +1,330 @@
+# A medley of functional project tools
+
+  library(plyr)
+  library(tidyverse)
+  library(rlang)
+  library(glue)
+  library(ggtext)
+  library(stringi)
+  library(rstatix)
+  library(rmarkdown)
+  library(bookdown)
+  library(knitr)
+  
+# data import and transformation -----
+  
+  recode_var <- function(data, 
+                         old_var, 
+                         new_var, 
+                         ID_var = 'Passwort', 
+                         trans_fun = 'my_identity', 
+                         args = NULL) {
+    
+    ## renames a variable, if time and id variable names provided, they're included
+    ## in  the output as well
+    ## the transformation function enables
+    ## args specifies the arguments to the transforming function
+    
+    trans_call <- call2(trans_fun, 
+                        data[[old_var]], 
+                        !!!args)
+    
+    new_data <- data %>% 
+      transmute(!!ID_var := .data[[ID_var]], 
+                !!new_var := eval(trans_call))
+
+  }
+  
+  repl_string <- function(vector,
+                          regex, 
+                          replacement = '', 
+                          out_format = c('character', 'numeric', 'factor', 'date'), ...) {
+    
+    ## removes the specified regex from the text vectos
+    
+    out_format <- match.arg(out_format[1], 
+                            c('character', 'numeric', 'factor', 'date'))
+    
+    vector <- as.character(vector) %>% 
+      stri_replace(regex = regex, 
+                   replacement = replacement)
+    
+    switch(out_format, 
+           character = as.character(vector), 
+           numeric = as.numeric(vector), 
+           factor = factor(vector), 
+           date = as.Date(as.character(vector, 
+                                       tryFormats = c("%Y-%m-%d", 
+                                                      "%Y/%m/%d", 
+                                                      "%d.%m.%Y"))))
+    
+  }
+  
+  rm_spaces <- function(vector, 
+                        out_format = c('character', 'numeric', 'factor', 'date'), ...) {
+    
+    ## removes extra trailing spaces
+    
+    out_format <- match.arg(out_format[1], 
+                            c('character', 'numeric', 'factor', 'date'))
+    
+    repl_string(vector = vector, 
[…]

### 2023-02-14 paper/markdown/manuscript.Rmd
Effects of age, gender and mental illness
Association of demographic, socioeconomic, accident-related and mental health parameters with age, gender and mental illness in the study cohort

@@
 
-flextable::flextable(tables$cohort_demo) %>% 
-  flextable::width(1:2, width = 5.5, unit = 'cm') %>% 
+tables$cohort_demo %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
+  flextable::width(1, width = 5.5, unit = 'cm') %>% 
+  flextable::width(2, width = 7.5, unit = 'cm') %>% 
   theme_vanilla
@@
 
-flextable::flextable(tables$cohort_accident) %>% 
+tables$cohort_accident %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
   flextable::width(1, width = 5.5, unit = 'cm') %>% 
-  flextable::width(1:2, width = 7, unit = 'cm') %>% 
+  flextable::width(2, width = 7.5, unit = 'cm') %>% 
   footnote(2, 1, 
@@
 
-flextable::flextable(tables$cohort_mental) %>% 
-  flextable::width(1:2, width = 5.5, unit = 'cm') %>% 
+tables$cohort_mental %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
+  flextable::width(1, width = 5.5, unit = 'cm') %>% 
+  flextable::width(2, width = 7.5, unit = 'cm') %>% 
   footnote(1, 1, 
-           value = as_paragraph('PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; PCL-5 DSM-5: PTSD Checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+           value = as_paragraph('PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder'),
            part = 'header', 

### 2023-02-14 paper/markdown/supplementary_material.Rmd
Effects of age, gender and mental illness
Association of demographic, socioeconomic, accident-related and mental health parameters with age, gender and mental illness in the study cohort

@@
 
-```{r tab-incl-excl, tab.cap = 'Significant and near signifcant (unadjusted p < 0.05) differences between individuals excluded from analysis and analyzed study participants.
+```{r tab-incl-excl, tab.cap = 'Significant and near significant (non-adjusted p < 0.05) differences between individuals excluded from analysis and analyzed study participants.
 Numeric variables are presented as medians with interquartile ranges (IQR).
@@
 
-flextable::flextable(suppl_tables$incl_excl) %>% 
+suppl_tables$incl_excl %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
   flextable::width(1, width = 3.2, unit = 'cm') %>% 
@@
            ref_symbols = 'a') %>% 
-  footnote(3:4, 1, 
+  footnote(3, 1, 
            value = as_paragraph('AIS: acute injury severity score'), 
@@
            ref_symbols = 'b') %>% 
-  footnote(7:10, 1, 
+  footnote(6:9, 1, 
            value = as_paragraph('PTGI: post-traumatic growth inventory'), 
@@
            ref_symbols = 'c') %>% 
-  footnote(11, 1, 
+  footnote(10, 1, 
            value = as_paragraph('PSS4: 4-item perceived stress scale'), 
@@
            ref_symbols = 'd') %>% 
-  footnote(12, 1, 
-           value = as_paragraph('SOC-9L: Leipzig 9-item sense of soherence scale'), 
+  footnote(11, 1, 
+           value = as_paragraph('SOC-9L: Leipzig 9-item sense of coherence scale'), 
            part = 'body', 
@@
 
+```{r tab-age, tab.cap = 'Significant and near significant (false discovery rate-adjusted p < 0.1) demographic, socioeconomic, accident-related and mental health factors associated with age.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+suppl_tables$age %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
+  flextable::width(1, width = 3.2, unit = 'cm') %>% 
+  flextable::width(2:4, width = 4.3, unit = 'cm') %>% 
+  flextable::width(5, width = 2.3, unit = 'cm') %>% 
+  flextable::width(6, width = 2.1, unit = 'cm') %>% 
+  footnote(1, 5:6, 
+           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic; categorical variables: \u03c/\u00B2 test with Cramer V effec size statistic.
+P values were corrected for multiple testing with the false discovery method (FDR)'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(5, 1, 
+           value = as_paragraph('PTGI: post-traumatic growth inventory'), 
+           part = 'body', 
+           ref_symbols = 'b') %>%
+  footnote(6, 1, 
+           value = as_paragraph('SOC-9L: Leipzig 9-item sense of coherence scale'), 
+           part = 'body', 
+           ref_symbols = 'c') %>%
+  footnote(7, 1, 
+           value = as_paragraph('PHQ: patient health questionnaire'), 
+           part = 'body', 
+           ref_symbols = 'd') %>% 
+  footnote(8, 1, 
+           value = as_paragraph('EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale'), 
+           part = 'body', 
+           ref_symbols = 'e') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-gender, tab.cap = 'Significant and near significant (false discovery rate-adjusted p < 0.1) demographic, socioeconomic, accident-related and mental health factors associated with gender.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+suppl_tables$gender %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
+  flextable::width(1, width = 3.2, unit = 'cm') %>% 
+  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
+  flextable::width(4, width = 2.5, unit = 'cm') %>% 
+  flextable::width(5, width = 2.3, unit = 'cm') %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03c/\u00B2 test with Cramer V effec size statistic.
+P values were corrected for multiple testing with the false discovery method (FDR)'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(3, 1, 
+           value = as_paragraph('PCL-5 DSM-5: post-traumatic syndrome disorder checklist for DSM-5'), 
+           part = 'body', 
+           ref_symbols = 'b') %>% 
+  footnote(4:7, 1, 
+           value = as_paragraph('PTGI: post-traumatic growth inventory'), 
+           part = 'body', 
+           ref_symbols = 'c') %>% 
+  theme_vanilla()
+
+```
+
+\newpage
+
+```{r tab-mental, tab.cap = 'Significant and near significant (false discovery rate-adjusted p < 0.1) demographic, socioeconomic, accident-related and mental health factors associated with mental illness.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+suppl_tables$mental %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
+  flextable::width(1, width = 3.2, unit = 'cm') %>% 
+  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
+  flextable::width(4, width = 2.5, unit = 'cm') %>% 
+  flextable::width(5, width = 2.3, unit = 'cm') %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03c/\u00B2 test with Cramer V effec size statistic.
+P values were corrected for multiple testing with the false discovery method (FDR)'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(5:6, 1, 
+           value = as_paragraph('PTSD: post-traumatic syndrome disorder; PCL-5 DSM-5: PTSD checklist for DSM-5'), 
+           part = 'body', 
+           ref_symbols = 'b') %>% 
+  footnote(7, 1, 
+           value = as_paragraph('EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale'), 
+           part = 'body', 
+           ref_symbols = 'c') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
 # Supplementary Figures
+
+```{r fig-parts-mental-cohort, fig.width = mm_inch(suppl_figures$parts_mental_cohort$w), fig.height = mm_inch(suppl_figures$parts_mental_cohort$h), fig.cap = 'Injured body regions and detailed scoring of quality of life, post-traumatic syndrome disorder and post-traumatic growth in the study cohort.'}
+
+suppl_figures$parts_mental_cohort$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-parts-mental-cohort).
+Injured body regions and detailed scoring of quality of life, post-traumatic syndrome disorder and post-traumatic growth in the study cohort.__
+
+_For categorical variables, percentages of complete observations are presented as bars.
+Numeric variables are presented in violin plots with red diamonds denoting medians, interquartile ranges presented as red whiskers and single observations depicted as points.
+The number of complete cases is displayed in the plot captions._
+_(A) Distribution of injured body regions.
+(B) Scoring of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8).
+(C) Scoring of the post-traumatic syndrome disorder (PTSD) clusters with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD clusters.
+(E) Scoring of the factors of the post-traumatic growth with the post-traumatic growth inventory (PTGI) tool._
+
+\newpage
+
+```{r fig-age-mental, fig.width = mm_inch(suppl_figures$age_mental$w), fig.height = mm_inch(suppl_figures$age_mental$h), fig.cap = 'Association of measures of mental health after the accident with participant age.'}
+
+suppl_figures$age_mental$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-age-mental). Association of measures of mental health after the accident with participant's age.__
+
+_Study participants were stratified by age as young adults ($\leq$ 30 years), middle-aged (31 - 65 years) and elderly (> 65 years).
+Differences in demographic, socioeconomic, accident-related and mental health variables between the age strata were assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+P values were corrected for multiple testing with false discovery rate method.
+Scores of panic (PHQ: patient health questionnaire), resilience (RS-13: 13-item resilience scale), reduced sense of coherence (SOC-9L: Leipzig 9-item sense of coherence scale), diminished life quality (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) and of post-traumatic growth (PTGI: post-traumatic growth inventory) are presented in violin plots.
+Single observations are depicted as points.
+Red diamond with whiskers represent medians with interquartile ranges.
+Effect sizes and p values are indicated in the plot captions or in the plot Y axes.
+Numbers of complete observations are presented in the plot X axes or plot legends._
+
+\newpage
+
+```{r fig-gender-mental, fig.width = mm_inch(suppl_figures$gender_mental$w), fig.height = mm_inch(suppl_figures$gender_mental$h), fig.cap = 'Association of measures of mental health after the accident with participant gender.'}
+
+suppl_figures$gender_mental$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-gender-mental). Association of measures of mental health after the accident with participant's gender.__
+
+_Differences in scoring of post-traumatic syndrome (PTSD) disorder clusters (PCL-5 DSM-5: PTSD checklist for DSM-5) and post-traumatic growth factors (PTGI: post-traumatic growth inventory) between the genders were assessed by Mann-Whitney test with r effect size statistic.
+P values were corrected for multiple testing with false discovery rate method.
+Scores are presented in violin plots.
+Single observations are depicted as points.
+Red diamond with whiskers represent medians with interquartile ranges.
+Effect sizes and p values are indicated in the plot Y axes.
+Numbers of complete observations are presented in the plot legend._
+
+\newpage
+
+```{r fig-mental-illness-base, fig.width = mm_inch(suppl_figures$mental_illness_base$w), fig.height = mm_inch(suppl_figures$mental_illness_base$h), fig.cap = 'Readouts of anxiety, depression, stress, loss of sense of coherence and of resilience in participants with and without mental illness.'}
+
+suppl_figures$mental_illness_base$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-mental-illness-base). Readouts of anxiety, depression, stress, loss of sense of coherence and of resilience in participants with and without mental illness.__
+
+_Differences in symptoms of anxiety (GAD-7: 7-item general anxiety disorder scale), depression (PHQ-9: patient health questionnaire), somatization (PHQ-15), panic (PHQ-panic module), stress (PSS4: 4-item perceived stress scale), loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) and of resilience in participants with and without mental illness.
+For categorical variables, statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+For numeric variables, statistical significance was investigated by Mann-Whitney test with r effect size statistic.
+P values were corrected for multiple testing with false discovery rate (FDR) method.
+Distribution of categorical variable levels is presented in stack plots.
+Numerical variables are presented in violin plots with single observations depicted as points and red diamond with whiskers representing medians with interquartile ranges.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of complete observations are indicated in the plot X axes.
+_
+
+\newpage
+
+```{r fig-mental-illness-ext, fig.width = mm_inch(suppl_figures$mental_illness_ext$w), fig.height = mm_inch(suppl_figures$mental_illness_ext$h), fig.cap = 'Signs of diminished quality of life and symptoms of post-traumatic syndrome disorder in participants with and without mental illness.'}
+
+suppl_figures$mental_illness_ext$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-mental-illness-ext). Signs of diminished quality of life and symptoms of post-traumatic syndrome disorder in participants with and without mental illness.__
+
+_Differences in (A) scoring of diminished quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale), and (B) scoring of post-traumatic syndrome (PTSD) disorder clusters and frequency PTSD cluster positivity (PCL-5 DSM-5: PTSD checklist for DSM-5) and were compared between participants with and without mental illness.
+For categorical variables, statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+For numeric variables, statistical significance was investigated by Mann-Whitney test with r effect size statistic.
+P values were corrected for multiple testing with false discovery rate (FDR) method.
+Frequencies of PTSD cluster-positive participants are presented in bar plots.
+Numeric scores are presented in violin plots with single observations depicted as points and red diamond with whiskers representing medians with interquartile ranges.
+Effect sizes and p values are indicated in the plot Y axes.
+Numbers of complete observations are presented in the plot legend._
 

### 2023-02-14 exec.R

@@ -4,4 +4,5 @@
   
   print(source_all(c('import.R', 
                      'exploration.R', 
+                     'factor.R', 
                      'paper.R')))

### 2023-02-14 exploration scripts/cohort.R

@@ -24,8 +24,8 @@
   
   cohort$var_class <- 
     list(demo = c('obs_time', 
-                  'sex', 
                   'age', 
+                  'sex', 
                   'residence_alpine_region', 
                   'education', 
                   'employment_status', 
@@ -35,7 +35,7 @@
                   'smoking_status', 
                   'drug_status', 
                   'somatic_comorbidity', 
-                  'somatic_comorbidity_type', 
+                  #'somatic_comorbidity_type', 
                   'psych_comorbidity', 
                   'traumatic_event', 
                   'prime_trauma_event', 
@@ -46,19 +46,8 @@
                       'accident_culprit', 
                       'accident_injured_persons', 
                       'accident_rescue', 
-                      'accident_rescue_mode', 
                       'injury_sev_strata', 
-                      'injury_head', 
-                      'injury_face', 
-                      'injury_neck', 
-                      'injury_chest', 
-                      'injury_abdomen', 
-                      'injury_spine', 
-                      'injury_upper_limbs', 
-                      'injury_lower_limbs', 
-                      'injury_external_other', 
                       'psych_support_post_accident', 
-                      'psych_therapy_post_accident', 
                       'accident_aftermath', 
                       'same_sport_type_post_accident', 
                       'caution_post_accident'), 
@@ -71,37 +60,16 @@
                     'phq_events_total', 
                     'phq_events_total_class', 
                     'eurohis_total', 
-                    'eurohis_qol', 
-                    'eurohis_health', 
-                    'eurohis_energy', 
-                    'eurohis_finances', 
-                    'eurohis_activity', 
-                    'eurohis_selfesteem', 
-                    'eurohis_relationship', 
-                    'eurohis_housing', 
                     'soc9l_total', 
                     'rs13_total', 
                     'rs13_total_class', 
                     'cage_total_class', 
                     ## accident-specific measures
-                    'unwilling_flashback', 
+                    #'unwilling_flashback', 
                     'flashback_frequency', 
                     'dsm5_total', 
-                    'dsm5_B', 
-                    'dsm5_B_class', 
[…]

### 2023-02-14 exploration.R

@@ -11,12 +11,11 @@
   library(exda)
   library(soucer)
   library(furrr)
-  library(writexl)
   library(rstatix)
   library(trafo)
   library(psych)
   library(clustTools)
-  
+
   insert_head()
   
   source_all('./tools/tools.R', 

### 2023-02-14 factor scripts/age.R

@@ -1,0 +1,112 @@
+# Analysis for age classes 
+# (young adults: <= 30, middle-aged: <= 65, elderly: > 65)
+
+  insert_head()
+  
+# container -----
+  
+  age <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  age$variables <- fct_globals$variables$variable
+  
+  age$variables <- age$variables[age$variables != 'age']
+  
+  ## analysis table
+  
+  age$analysis_tbl <- ptsd$dataset %>% 
+    mutate(age = cut(age, 
+                     c(-Inf, 30, 65, Inf), 
+                     c('young', 'middle-aged', 'elderly'))) %>% 
+    select(ID, age, all_of(age$variables))
+  
+  ## statistical test type
+  
+  age$test_type <- 
+    age$analysis_tbl[age$variables] %>% 
+    map_lgl(is.numeric)
+  
+  age$test_type <- ifelse(age$test_type, 
+                          'kruskal_eta', 'cramer_v')
+  
+# Descriptive statistic ------
+  
+  insert_msg('Descriptive statistics')
+  
+  age$desc_stats <- age$analysis_tbl %>% 
+    explore(variables = age$variables, 
+            split_factor = 'age', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for differences between the age groups ------
+  
+  insert_msg('Testing for differences between the age strata')
+  
+  age$test <- age$analysis_tbl %>%
+    compare_variables(variables = age$variables, 
+                      split_factor = 'age', 
+                      what = 'eff_size', 
+                      types = age$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      adj_method = 'BH', 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('age'))) %>% 
+    format_fct_test
+  
+# Significant and near significant differences -------
+  
+  insert_msg('Significant and near significant differences')
+  
+  age$top_factors <- age$test %>% 
+    filter(p_adjusted < 0.1) %>% 
+    .$variable
+
+# Single variable plots ------
+  
+  insert_msg('Single variable plots')
+  
+  age$plots <- 
+    plot_fct(data = age$analysis_tbl, 
[…]

### 2023-02-14 factor scripts/gender.R

@@ -1,0 +1,101 @@
+# Effects of gender
+
+  insert_head()
+  
+# container -----
+  
+  gender <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  gender$variables <- fct_globals$variables$variable
+  
+  gender$variables <- gender$variables[gender$variables != 'sex']
+
+  ## statistical test type
+  
+  gender$test_type <- 
+    ptsd$dataset[gender$variables] %>% 
+    map_lgl(is.numeric)
+  
+  gender$test_type <- ifelse(gender$test_type, 
+                             'wilcoxon_r', 'cramer_v')
+  
+# Descriptive statistic ------
+  
+  insert_msg('Descriptive statistics')
+  
+  gender$desc_stats <- ptsd$dataset %>% 
+    explore(variables = gender$variables, 
+            split_factor = 'sex', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for differences between the genders ------
+  
+  insert_msg('Testing for differences between the gender strata')
+  
+  gender$test <- ptsd$dataset %>%
+    compare_variables(variables = gender$variables, 
+                      split_factor = 'sex', 
+                      what = 'eff_size', 
+                      types = gender$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      adj_method = 'BH', 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('ptsd'))) %>% 
+    format_fct_test
+  
+# Significant and near significant differences -------
+  
+  insert_msg('Significant and near significant differences')
+  
+  gender$top_factors <- gender$test %>% 
+    filter(p_adjusted < 0.1) %>% 
+    .$variable
+  
+# Single variable plots ------
+  
+  insert_msg('Single variable plots')
+  
+  gender$plots <- 
+    plot_fct(data = ptsd$dataset, 
+             test_data = gender$test, 
+             factor = 'sex', 
+             fill_scale = scale_fill_brewer(palette = 'Oranges'))
+  
+# Specific plots: PTSD, PTG and QoL scores ------
+  
+  insert_msg('Plot panels')
+  
+  gender$panels <- 
+    plot_fct_panels(data = ptsd$dataset, 
+                    test_data = gender$test, 
[…]

### 2023-02-14 factor scripts/mental_illness.R

@@ -1,0 +1,151 @@
+# Effects of mental illness
+
+  insert_head()
+  
+# container ------
+  
+  mental <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  mental$variables <- fct_globals$variables$variable
+  
+  mental$variables <- 
+    mental$variables[mental$variables != 'psych_comorbidity']
+  
+  ## statistical test type
+  
+  mental$test_type <- 
+    ptsd$dataset[mental$variables] %>% 
+    map_lgl(is.numeric)
+  
+  mental$test_type <- ifelse(mental$test_type, 
+                             'wilcoxon_r', 'cramer_v')
+
+# Descriptive statistic ------
+  
+  insert_msg('Descriptive statistics')
+  
+  mental$desc_stats <- ptsd$dataset %>% 
+    explore(variables = mental$variables, 
+            split_factor = 'psych_comorbidity', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for differences between the mental illness strata ------
+  
+  insert_msg('Testing for differences between the mental illness strata')
+  
+  mental$test <- ptsd$dataset %>%
+    compare_variables(variables = mental$variables, 
+                      split_factor = 'psych_comorbidity', 
+                      what = 'eff_size', 
+                      types = mental$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      adj_method = 'BH', 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('ptsd'))) %>% 
+    format_fct_test
+  
+# Significant and near significant differences -------
+  
+  insert_msg('Significant and near significant differences')
+  
+  mental$top_factors <- mental$test %>% 
+    filter(p_adjusted < 0.1) %>% 
+    .$variable
+  
+# Single variable plots ------
+  
+  insert_msg('Single variable plots')
+  
+  mental$plots <- 
+    plot_fct(data = ptsd$dataset, 
+             test_data = mental$test, 
+             factor = 'psych_comorbidity', 
+             fill_scale = scale_fill_brewer(palette = 'Greys'))
+  
+# Specific plots: PTSD, PTG and QoL scores ------
+  
+  insert_msg('Plot panels')
+  
[…]

### 2023-02-14 factor.R

@@ -1,0 +1,43 @@
+# Association of age, gender, mental illness, injury severity 
+# and psychological support with accident circumstances and mental health
+
+# tools -----
+
+  library(plyr)
+  library(tidyverse)
+  library(rlang)
+  library(stringi)
+  library(exda)
+  library(soucer)
+  library(furrr)
+  library(rstatix)
+  library(trafo)
+  library(clustTools)
+  
+  insert_head()
+  
+  source_all('./tools/tools.R', 
+             message = TRUE, 
+             crash = TRUE)
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  fct_globals <- list()
+  
+  fct_globals$variables <- ptsd$var_lexicon %>% 
+    filter(type %in% c('characteristic', 'response'))
+  
+# analysis scripts --------
+  
+  insert_msg('Analysis scripts')
+  
+  c('./factor scripts/age.R', 
+    './factor scripts/gender.R', 
+    './factor scripts/mental_illness.R') %>% 
+    source_all(message = TRUE, encoding = TRUE)
+  
+# END ------
+  
+  insert_tail()

### 2023-02-14 paper scripts/figures.R

@@ -6,26 +6,6 @@
   
   figures <- list()
   
-# globals ------
-  
-  insert_msg('Globals setup')
-
- # report_fig$radial_panel <- 
- #   plot_grid(part_clust$radial_plot + 
-  #              theme(plot.tag = element_blank(), 
-  #                    legend.position = 'none'), 
-   #           plot_grid(get_legend(part_clust$radial_plot), 
-    #                    ggdraw() + 
-     #                     draw_text(part_clust$n_tag %>% 
-     #                                 stri_replace_all(fixed = ', ', 
-      #                                                 replacement = '\n'), 
-      #                              size = 8, 
-      #                              hjust = 0, 
-      #                              x = 0.2), 
-     #                   nrow = 2), 
-     #         ncol = 2, 
-     #         rel_widths = c(0.7, 0.3))
-    
 # Figure 1: flow diagram --------
   
   insert_msg('Figure 1: flow diagram')
@@ -43,11 +23,7 @@
 # Saving the figures on the disc -------
   
   insert_msg('Saving the figures on the disc')
-  
-  figures$radial_panel <- NULL
-  
-  figures <- compact(figures)
-  
+
   figures %>% 
    walk(pickle, 
          path = './paper/figures', 

### 2023-02-14 paper scripts/supplementary.R

@@ -1,0 +1,206 @@
+# supplementary figures
+
+  insert_head()
+  
+# container ------
+  
+  suppl_figures <- list()
+  
+# Figure S1: injured parts and mental health details -----
+  
+  insert_msg('Figure S1: injured parts and mental health, cohort')
+  
+  suppl_figures$parts_mental_cohort <- 
+    list(cohort$injury$plot, 
+         cohort$mental_details$plots$eurohis, 
+         cohort$mental_details$plots$dsm, 
+         cohort$ptsd_cluster$plot, 
+         cohort$mental_details$plots$ptgi) %>% 
+    plot_grid(plotlist = ., 
+              ncol = 2, 
+              align = 'hv', 
+              axis = 'tblr', 
+              labels = c('A', 'B', 'C', '', 'D'), 
+              label_size = 10, 
+              rel_heights = c(1.4, 1, 1.1)) %>% 
+    as_figure(label = 'figure_s1_parts_mental_cohort', 
+              ref_name = 'parts_mental_cohort', 
+              caption = paste('Injured body regions and detailed scoring', 
+                              'of quality of life, post-traumatic syndrome', 
+                              'disorder and post-traumatic growth', 
+                              'in the study cohort.'), 
+              w = 180, 
+              h = 210)
+  
+# Figure S2: mental features associated with participant's age ------
+  
+  insert_msg('Figure S2: age and mental health')
+  
+  ## upper panel: panic, resilience and sense of coherence
+  
+  suppl_figures$age_mental$upper <- age$plots[c("phqd_panic_total", 
+                                                "rs13_total", 
+                                                "soc9l_total")] %>% 
+    map(~.x + theme(legend.position = 'none')) %>% 
+    plot_grid(plotlist = ., 
+              ncol = 3, 
+              align = 'hv', 
+              axis = 'tblr')
+  
+  ## bottom panel: panels of plots for QoL and PTG
+  
+  suppl_figures$age_mental$bottom <- 
+    plot_grid(age$panels$eurohis + 
+                theme(legend.position = 'none', 
+                      plot.subtitle = element_blank()), 
+              plot_grid(age$panels$ptgi + 
+                          theme(legend.position = 'none', 
+                                plot.subtitle = element_blank()), 
+                        get_legend(age$panels[[1]] + 
+                                     scale_fill_brewer(labels = age$analysis_tbl %>% 
+                                                         label_n(age), 
+                                                       palette = 'Blues') + 
[…]

### 2023-02-14 paper scripts/tables.R

@@ -79,8 +79,8 @@
   suppl_tables$incl_excl <- excl$result_tbl %>% 
     mdtable(label = 'table_s3_included_excluded', 
             ref_name = 'incl_excl', 
-            caption = paste('Significant and near signifcant', 
-                            '(unadjusted p < 0.05) differences between', 
+            caption = paste('Significant and near significant', 
+                            '(non-adjusted p < 0.05) differences between', 
                             'individuals excluded from analysis', 
                             'and analyzed study participants.', 
                             'Numeric variables are presented as medians', 
@@ -88,7 +88,82 @@
                             'variables are presented as percentages', 
                             'and counts within the complete', 
                             'observation set.'))
+  
+# Supplementary Table S4: effects of age ------
+  
+  insert_msg('Table S4: effects of age')
+  
+  suppl_tables$age <- age$result_tbl %>% 
+    mdtable(label = 'table_s4_age', 
+            ref_name = 'age', 
+            caption = paste('Significant and near significant', 
+                            '(false discovery rate-adjusted p < 0.1)', 
+                            'demographic, socioeconomic, accident-related', 
+                            'and mental health factors', 
+                            'associated with age.', 
+                            'Numeric variables are presented as medians', 
+                            'with interquartile ranges (IQR). Categorical', 
+                            'variables are presented as percentages', 
+                            'and counts within the complete', 
+                            'observation set.'))
 
+# Supplementary Table S5: effects of gender ------
+  
+  insert_msg('Table S5: effects of gender')
+  
+  suppl_tables$gender <- gender$result_tbl %>% 
+    mdtable(label = 'table_s5_gender', 
+            ref_name = 'gender', 
+            caption = paste('Significant and near significant', 
+                            '(false discovery rate-adjusted p < 0.1)', 
+                            'demographic, socioeconomic, accident-related', 
+                            'and mental health factors', 
+                            'associated with gender.', 
+                            'Numeric variables are presented as medians', 
+                            'with interquartile ranges (IQR). Categorical', 
+                            'variables are presented as percentages', 
+                            'and counts within the complete', 
+                            'observation set.'))
+  
+# Supplementary Table S6: effects of mental illness -------
+  
+  insert_msg('Table S6: Effect of mental illness')
+  
+  ## filtering out variables presented already in the supplementary figure
+  
+  suppl_tables$mental <- mental$result_tbl %>% 
[…]

### 2023-02-14 paper.R

@@ -27,6 +27,7 @@
   
   c('./paper scripts/tables.R', 
     './paper scripts/figures.R', 
+    './paper scripts/supplementary.R', 
     './paper scripts/render.R') %>% 
     source_all(message = TRUE, crash = TRUE)
 

### 2023-02-14 tools/tools.R

@@ -171,6 +171,18 @@
   }
   
   pss2date <- function(x) as.Date(x/86400, origin = "1582-10-14")
+  
+# plot axis labels with n numbers ------
+  
+  label_n <- function(data, split_factor, sep = '\nn = ') {
+    
+    split_expr <- enexpr(split_factor)
+    
+    counts <- count(data, !!split_expr)
+    
+    map2_chr(counts[[1]], counts[[2]], paste, sep = sep)
+    
+  }
 
 # variable:label translation, color setup -----
 
@@ -182,6 +194,212 @@
     
     return(colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)] %>% 
              sample(size = color_no))
+    
+  }
+  
+# Frequency computation and plots ------
+  
+  count_binary <- function(data) {
+    
+    ## counting of binary factors
+    
+    data <-  data %>% 
+      map_dfc(~as.numeric(.x) - 1) %>% 
+      filter(complete.cases(.))
+    
+    data %>% 
+      colSums %>% 
+      compress(names_to = 'variable',
+               values_to = 'n') %>% 
+      mutate(n_total = nrow(data), 
+             fraction = n/n_total, 
+             percent = fraction * 100)
+    
+  }
+  
+  plot_freq_bars <- function(data, 
+                             freq_var = 'percent', 
+                             cat_var = 'variable', 
+                             split_factor = NULL, 
+                             plot_title = NULL, 
+                             plot_subtitle = NULL, 
+                             x_lab = '% of cohort', 
+                             bar_color = 'steelblue', 
+                             fill_scale = scale_fill_brewer(), 
+                             show_freqs = TRUE, ...) {
+    
+    ## plots a simple bar plot of frequencies
+    
+    if(is.null(split_factor)) {
+      
+      freq_plot <- data %>% 
+        ggplot(aes(x = .data[[freq_var]], 
+                   y = .data[[cat_var]])) + 
+        geom_bar(stat = 'identity', 
+                 color = 'gray20', 
+                 fill = bar_color)
+      
+    } else {
+      
+      freq_plot <- data %>% 
+        ggplot(aes(x = .data[[freq_var]], 
+                   y = .data[[cat_var]], 
+                   fill = .data[[split_factor]])) + 
+        geom_bar(stat = 'identity', 
+                 color = 'gray20', 
+                 position = position_dodge(0.9)) + 
+        fill_scale
+      
+    }
+    
+    freq_plot <- freq_plot + 
+      globals$common_theme + 
[…]

### 2023-02-24 paper/markdown/manuscript.Rmd
Analysis pipeline, Manuscript, Suppelements

--- 2023-02-13 paper/markdown/manuscript.Rmd
@@
 ---
-title: "Three distinct patterns of mental health response after a sport accident"
+title: "Three distinct patterns of mental response after a sport accident"
 subtitle: "Manuscript"
@@
   - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria
+  - daas: Data Analysis as a Service Tirol, Innsbruck, Austria
 
@@
 
+__Background:__ Long-term psychological consequences of sport accidents and sport accident victims at risk of mental deterioration are poorly characterized. 
+We sought to identify patterns of mental response following leisure time sport accidents.
+
+__Methods:__ A cross-sectional survey study was conducted with 307 individuals treated at the Innsbruck University Hospital (Austria) following a non-professional sport accident. 
+Measures of anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic syndrome (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data.
+Patterns of mental readouts were defined by semi-supervised clustering and factors affecting the cluster assignment were identified by statistical testing and conditional random forest machine learning.
+
+__Results:__ Participants were assigned to three mental clusters characterized by a neutral mental response, post-traumatic growth and symptoms of post-traumatic syndrome disorder, respectively. 
+Overall differences in demographic, socioeconomic, injury- and accident-related variables between the clusters were low-to-moderate. 
+The post-traumatic growth and post-traumatic syndrome clusters were characterized by relatively frequent hospitalization, surgery, somatic illness and female gender. 
+The most predictive factors for the post-traumatic syndrome cluster were more cautious behavior and flashbacks during sport activity, low income, mental disorders, somatic consequences of the accident and subjective need for psychological support.
+
+__Conclusion:__ A subset of non-professional sport accident victims develops signs of mental health problems including symptoms of post-traumatic syndrome disorder. 
+Patients with low socioeconomic status, somatic and mental conditions, and hospitalization represent a vulnerable group which may benefit most from a psychological intervention.
+
 \newpage
@@
 
+## Study population
+
+_Hanna and Katharina: could you please provide a description?_
+
+## Procedures
+
+The psychometric battery (__Supplementary Table S2__) consisted of German versions of assessment tools for 
+anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
+depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
+panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
+persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
+resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
+loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
+loss of quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
+post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
+post-traumatic syndrome disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
+The tools displayed good-to-excellent consistency as measured by McDonald's $\omega$ > 0.8 [@McDonald1999] (__Supplementary Table S3__).
+
+Symptoms of anxiety were defined as GAD-7 $geq$ 10, symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], significant somatization/persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002].
+Resilient coping classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].
+
+Separate scores were computed for each domain of the PTGI tool (I: relations, II: possibilities, III: personal strength, IV: spiritual strength, V: life appreciation) with each item scores as 0: none, 1: very little, 2: little, 3: moderate, 4: great, 5: extremely great.
+In addition, the total PTGI score was calculated as the sum of all items [@Tedeschi1996].
+
+Separate scores were calculated for domains B, C, D and E of the PCL-5 DSM-5 tool along with the total score being the sum of all items. 
+Participants positive for the B domain or C domain were identified by at least one item scored with 'moderate', participants positive for the D or E domain were identified by at least two item scored with 'moderate'.
+Significant PTSD symptoms where assumed in participants screened positive for at least one of the PCL-5 DSM-5 domains [@Bovin2016].
+
+Traumatic events prior to the sport accident were assessed with the DIA-X tool [@Maercker2005]. 
+Flashbacks during sport activity, 
+were surveyed as a single yes/no item. 
+Flashbacks frequency was assessed in the following categories: none, more than one per month and more than one per year. 
+Smoking was surveyed as a single yes/no question. 
+Substance abuse was investigated with the CAGE tool [@OBrien2008]. 
+Data on hospital treatment, hospitalization and surgery were extracted from electronic patient's records. 
+Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].
+
+Additional information on study variables and their stratification schemes are presented in __Supplementary Methods__ and __Supplementary Table S1__.
+
+The study participants were assigned to the training and test subsets with the 2:1 size ratio with a random procedure aiming at minimal differences in accident year, age, gender, somatic and mental illness rates, frequencies of prior traumatic events [@Maercker2005] and distribution of injury severity grades [@Gennarelli2006] (__Figure 1__).
+The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as investigated by comparing Hopkins metrics of clustering tendencies [@Kassambara2020] in random subsamples of the entire dataset of varying size (__Supplementary Methods__).
+
+## Analysis endpoints
+
+The primary analysis endpoint was identification of patterns of mental response by clustering of leisure time sport accident victims in respect to measures of anxiety, depression, panic, somatization, quality of life, post-traumatic growth and post-traumatic symptom disorder by clustering analysis.
+The secondary analysis endpoints were demographic, socioeconomic and clinical characteristic of the mental response clusters and development of a classification procedure to discern individuals at risk of deteriorated mental health following a sport accident.
+
+## Statistical analysis
+
+Statistical analysis was done with R version 4.2.0. 
+Numeric variables are displayed as medians with interquartile ranges. 
+Categorical variables are presented as percentages and counts within the complete observation set. 
+The majority of numeric variables were non-normally distributed (Shapiro-Wilk test). 
+Hence, differences in numeric variables between analysis groups were analyzed by Mann-Whitney test with r effect size statistic or Kruskal-Wallis test with $\eta^2$ effect size statistic [@Kassambara2021].
+Differences in distribution of categorical variables were assessed by $\chi^2$ test with Cramer V effect size statistic [@Kassambara2021].
+
+The training subset was clustered in respect to normalized median-centered numeric psychometric scores by the partition around medoids algorithm with the cosine distance measure between observations [@Drost2018; @Schubert2019].
+The clustering procedure was characterized a good explanatory performance (explained clustering variance: ratio of between-cluster sum of squares to total sum of squares) and the superior stability in 10-fold cross-validation [@Lange2004] as compared with hierarchical and KMEANS algorithms.
+The cluster number was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic [@Kassambara2020]. 
+Three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) cluster. 
+Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
+The quality of cluster predictions was assessed by comparing explained clustering variances in the training and test subset, visual inspection of UMAP (uniform manifold approximation and projection) layouts [@McInnes2018; @Konopka2022], of distance heat maps and heat maps of normalized clustering feature levels.
+
+Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters were trained in the training subset of the study cohort by the one-rule algorithm [@VonJouanne-Diedrich2017; @Holte1993].
+The multi-parameter classifier was developed in the training subset with the conditional random forest algorithm (1000 random trees, mtry = 5) [@Hothorn2006; @Strobl2007; @Hothorn2022]. 
+The cluster assignment was subsequently predicted for the test subset observations and accuracy and $\kappa$ statistic for the predictions were computed [@Kuhn2008]. 
+Conditional importance for explanatory variables in the random forest classifiers were expressed as accuracy loss [@Strobl2008].
+
 \newpage
@@
 # Results
+
+# Characteristic of the study cohort
+
+A total of 
+`r nrow(ptsd$cleared)` patients aged $\geq$ 16 years treated at the emergency, surgery, orthopedics and trauma departments at the University Clinic of Innsbruck (Austria) following a non-professional sport accident between 2018 and 2020 were invited to participant in the study survey.
+Out of them, 
+`r length( ptsd$interview_ID)` completed the survey and 
+`r nrow(ptsd$dataset)` individuals with the complete psychometric parameter set were analyzed here amounting to an 
+`r ((nrow(ptsd$dataset)/length(ptsd$interview_ID)) * 100) %>% signif(2)`% overall response rate  (__Figure 1__, __Supplementary Tables S1__ - __S3__).
+
+As compared with the study participants, the individuals excluded from the analysis had lower employment rates, household income and less chronic somatic conditions. 
+The excluded collective tended towards more biking and less ski/snowboard accidents and less professional rescue need. 
+Their injuries were on average less severe as measured by AIS [@Gennarelli2006] with more often upper limb injuries, and lower hospitalization and surgery rates. 
+The excluded individuals tended suffered less often from somatic health consequences of the accident and tended towards lower ratings of post-traumatic syndrome disorder but also diminished post-traumatic growth and less sense of coherence.
+These differences were significant or near-significant with weak effects sizes (V $\leq$ 0.22 or r $\leq$ 0.15). 
+This analysis suffered also from high data missingness in the excluded collective (__Supplementary Table S4__).
+
+In the analyzed cohort, the accidents happened between 
+`r ptsd$dataset$accident_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')` and were managed at the study site between `r ptsd$dataset$ward_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')`.
+The median time between the ward management and survey completion was 
+`r median(ptsd$dataset$obs_time)` days (interquartile range: 
+`r quantile(ptsd$dataset$obs_time, probs = c(0.25, 0.75)) %>% paste(collapse = ' - ')`). 
+The participants were predominantly middle-aged (median `r median(ptsd$dataset$age)` years) and 
+`r signif(table(ptsd$dataset$sex)/sum(table(ptsd$dataset$sex)), 2)[1] * 100`% of them were females. 
+The vast majority had secondary or tertiary education grade and two-third were professionally active. 
+Sport and trauma-risk professions constituted only a minor fraction of the cohort (< 8%). 
+Over 40% of participants had high annual household incomes of $\geq$ 45000 Euro. 
+Less than 10% of participants were smokers or alcohol/drug consumers. 
+Frequencies of somatic and mental conditions were 
+`r signif(table(ptsd$dataset$somatic_comorbidity)/sum(table(ptsd$dataset$somatic_comorbidity)), 2)[2] * 100`% and 
+`r signif(table(ptsd$dataset$psych_comorbidity)/sum(table(ptsd$dataset$psych_comorbidity)), 2)[2] * 100`%, respectively. 
+The rates of smoking (Austria: 20.6%) and somatic (38.3%) illness were lower than in the general population [@StatistikAustria2022a; @StatistikAustria2022]. 
+Four of ten participants were directly affected by or witnessed a traumatic event prior to the accident as gauged by the DIA-X questionnaire [@Maercker2005].
+
+Nearly 40% of participants had sport accidents in the past. 
+Concerning the recent accident, the almost two-third of them were injured during skiing or snowboarding followed by biking and mountain sport. 
+One-third were alone during the accident and, in the vast majority of case, were the sole culprit and victim. 
+Roughly the half of participants could rescue themselves, professional rescue team was involved in 
+`r signif(table(ptsd$dataset$accident_rescue)/sum(table(ptsd$dataset$accident_rescue)), 2)[3] * 100`% of cases. 
+In 
+`r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[2] * 100`% of participants the injury grade was moderate (AIS 2) and in 
+`r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[3] * 100`% severe to life threatening (AIS $\geq$ 3). 
+Upper and lower limb injuries were the most frequent followed by head and face (__Supplementary Figure S1A__). 
+Hospitalization and surgery rates were 
+`r signif(table(ptsd$dataset$hospitalization)/sum(table(ptsd$dataset$hospitalization)), 2)[2] * 100`% and 
+`r signif(table(ptsd$dataset$surgery_done)/sum(table(ptsd$dataset$surgery_done)), 2)[2] * 100`%, respectively. 
+Psychological or psychiatric support was applied to 
+`r signif(table(ptsd$dataset$psych_support_post_accident)/sum(table(ptsd$dataset$psych_support_post_accident)), 2)[2] * 100`% individuals. 
+Among those who had not received it, 
+`r signif(table(ptsd$dataset$psych_support_need)/sum(table(ptsd$dataset$psych_support_need)), 2)[2] * 100`% declared the need for psychological support concerning the accident. 
+Despite somatic health consequences of the accident were reported by 
+`r signif(table(ptsd$dataset$accident_aftermath)/sum(table(ptsd$dataset$accident_aftermath)), 2)[2] * 100`% and unwilling flashbacks during sport in 
+`r signif(table(ptsd$dataset$unwilling_flashback)/sum(table(ptsd$dataset$unwilling_flashback)), 2)[2] * 100`%, the majority of participants returned to the accident sport type (
+`r signif(table(ptsd$dataset$same_sport_type_post_accident)/sum(table(ptsd$dataset$same_sport_type_post_accident)), 2)[2] * 100`%). 
+Yet, two-third described their behavior during sport activity as more cautious (__Table \@ref(tab:tab-cohort-accident)__).
+
+Clinically relevant signs of 
+anxiety (`r signif(table(ptsd$dataset$gad7_total_class)/sum(table(ptsd$dataset$gad7_total_class)), 2)[2] * 100`%), 
+depression (`r signif(table(ptsd$dataset$phq9_total_class)/sum(table(ptsd$dataset$phq9_total_class)), 2)[2] * 100`%) and 
+somatization (`r signif(table(ptsd$dataset$phq_events_total_class)/sum(table(ptsd$dataset$phq_events_total_class)), 2)[2] * 100`%) were rare in the study cohort. 
+The median loss of quality of life was scored with `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS QOL points, which was substantially more than in a generalized Western European population [@Schmidt2006].
+The participants were most satisfied with their overall quality of life and housing (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1B__). 
+Roughly two-third of participants were assigned to the high resilient coping class [@Leppert2008]. 
+The scores of overall post-traumatic growth (median: `r median(ptsd$dataset$ptgi_total) %>% signif(2)`) and its specific domains were lower in the study cohort as compared with the figures originally reported for the PTGI tool [@Tedeschi1996] (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1C__).
+In `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants symptoms of post-traumatic syndrome disorder defined as positivity for at least one PCL-5 DSM-5 tool domain [@Bovin2016] could be observed.
+Among them, domain B symptoms (`r signif(table(ptsd$dataset$dsm5_B_class)/sum(table(ptsd$dataset$dsm5_B_class)), 2)[2] * 100`%) were the most frequent and domain D symptoms were the rarest (`r signif(table(ptsd$dataset$dsm5_D_class)/sum(table(ptsd$dataset$dsm5_D_class)), 2)[2] * 100`%) (__Table \@ref(tab:tab-cohort-mental)__ and __Supplementary Figure S1D__).
+
+For sake of semi-supervised clustering and classification procedure development, the study cohort was split into the training (n = `r nrow(clust_globals$analysis_tbl$training)`) and test subset (n = `r nrow(clust_globals$analysis_tbl$test)`).
+There were no significant differences between the subsets in demographic, socioeconomic, clinical and accident-related parameters. 
+The training subset tended towards more frequent face injuries and low resilient coping class individuals. 
+Yet the effect size of those differences was weak (V < 0.14) (__Supplementary Table S5__).
+
+# Three clusters of mental response in sport accident victims
+
+To comprehend types of mental response to the sport accident, we subjected the study participants to semi-supervised clustering in respect to measures of anxiety, depression, panic, persistent somatic symptoms, resilience, sense of coherence, quality of life, post-traumatic growth and post-traumatic syndrome disorder.
+Among several clustering algorithms compared in the training subset, partition around medoids with cosine distance between observations [@Drost2018; @Schuber2019] demonstrated good explanatory power and superior reproducibility in cross-validation [@Lange2004] (__Supplementary Figure S2A__).
+Three clusters of participants named neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) cluster after the patterns of their characteristic mental features were identified in the training subset.
+Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance in both study cohort subsets (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`) and comparable cluster sizes (__Supplementary Figures S1B__ - __S4__).
+
+Roughly one-third study participants assigned to the neutral clusters were characterized by low scores of anxiety, depression, panic and somatic symptoms along with high rating of coherence, resilient coping and quality of life.
+Scores of post-traumatic growth and post-traumatic syndrome were the lowest in the neutral cluster. 
+The PTG cluster demonstrated similarly low score of major mental health problems as neutral cluster individuals. 
+By contrast, the overall measures of post-traumatic growth and its domains were the highest in the PTG cluster. 
+In addition, ratings of post-traumatic syndrome disorder symptoms were slightly elevated in the PTG as compared with the neutral cluster. 
+The remaining participants subsumed in the PTSD cluster displayed the most intense signs of anxiety, depression, panic and somatic symptoms as well as poor sense of coherence, resilient coping and low quality of life as compared with the neutral and PTG cluster.
+PTSD cluster was hallmarked by the highest scores of post-traumatic syndrome disorder symptoms in every PCL-5 DSM-5 domain. 
+Post-traumatic growth was also present in the PTSD cluster as evident from elevated PTGI scores (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Figure S4__, __Supplementary Table S6__).
+
+Along this line, clinically relevant symptoms of anxiety, depression and somatization were present virtually only in the PTSD clusters of the training and test subsets of the study cohort. 
+Furthermore, frequencies of low and moderate resilient coping classes were by far the highest in the PTSD cluster. 
+Finally, approximately 40% of PTSD cluster individuals were screened positive for at least one post-traumatic syndrome disorder symptom which was more than in the PTG 
+(training: `r signif(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)), 2)[2] * 100`%,
+test: `r signif(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)), 2)[2] * 100`%) or neutral cluster
+(training: `r signif(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)), 2)[2] * 100`%,
+test: `r signif(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)), 2)[2] * 100`%).
+An analogical trend was observed for the remaining post-traumatic syndrome disorder domains (__Figure \@ref(fig:fig-clust-symptoms)__).
+
+Collectively, the PTG cluster may represent individuals with salutary reaction to their sport accident, whereas the PTSD cluster individuals may pose a group of mental health deterioration.
+
+# Demographic, socioeconomic and clinical background of the mental clusters
+
+Next, we sought to explore differences in demography, socioeconomic status, clinical background, accident details and consequences between the mental clusters. 
+Among 
+`r length(clust_bcg$variables)` investigated variables, only 
+`r length(clust_bcg$top_factors)` parameters were found significant or near-significant (p < 0.1) in both the training and test subsets of the study cohort (__Supplementary Table S6__).
+
+A substantial enrichment of females, low-to-middle income and non-tertiary education participants could be observed in the PTSD cluster. 
+The PTG cluster had the highest share of middle-aged individuals. 
+Frequency of somatic conditions were the highest in the PTG and PTSD clusters. 
+Mental illness participants were in turn represented virtually only in the PTSD cluster. 
+Although there were no differences in frequency of previous sport accidents between the clusters, PTG cluster individuals were found to be affected by traumatic events in the past at a higher percentage than neutral cluster individuals.
+These differences were, however, non or borderline significant and of weak effect size (__Figure \@ref(fig:fig-clinic-factors)__, __Supplementary Figure S5__). 
+There were no consistent differences in accident sport type, rescue mode, injury severity or injured body parts between the mental clusters. 
+Yet, participants assigned to the PTG and PTSD clusters were tended towards higher hospitalization and surgery rates (__Figure \@ref(fig:fig-injury)__ and __Supplementary Figure S6__).
+
+PTSD cluster participants reported somatic health consequences of the accident and flashbacks significantly more often than neutral cluster individuals. 
+Additionally, PTG and PTSD cluster individuals tended to describe their behavior during sport activity as more careful as compared with the neutral cluster. 
+Roughly one-sixth PTSD cluster participants declared need for psychological or psychiatric support as compared with none in the neutral clusters of the training or test subset. 
+Only a small fraction the study participants received psychological support following the accident. 
+Although such individuals tended to be enriched in the PTG cluster, their frequency (up to 16%) was low. 
+Hence it is unlikely, that psychological support affected the mental cluster assignment (__Figure \@ref(fig:fig-aftermath)__).
+
+In summary, differences in non-mental features between the clusters were weak. 
+Yet, the PTSD cluster could be characterized by elevated fractions of female, low-to-middle income class individuals, increased rates of somatic and mental conditions, hospitalization and surgery. 
+Furthermore, such individuals were disproportionately often affected by somatic health aftermath and flashbacks.
+
+# Prediction of the mental cluster assignment by demographic, socioeconomic and accident-related factors
+
+Finally, we intended to establish a mental cluster classification procedure based solely on demographic, socioeconomic, clinical and accident-related factors. 
+As investigated by a simple one rule classification algorithm [@VonJouanne-Diedrich2017; @Holte1993], predictive value of single explanatory parameters was low with cluster assignment accuracy $\leq$ `r class_one$fit_stats$global$Accuracy %>% max %>% signif(2)` and $\kappa \leq$ `r class_one$fit_stats$global$Kappa %>% max %>% signif(2)`.
+In such univariable analysis, flashback presence and frequency, somatic health consequences of the accident, age class, annual income and traumatic events in the past were identified as the most predictive factors (__Supplementary Figure S7__).
+
+To account for cumulative effects of multiple weak predictors of the mental cluster assignment, we employed multi-parameter conditional random forest modeling [@Strobl2007; @Hothorn2006; @Hothorn2022].
+The conditional random forest classifier employing the full set of demographic, socioeconomic, clinical and accident-related explanatory factors could correctly assign `r signif(class_rf$fit_stats$global$Accuracy[1], 2) * 100`% training subset observations to their mental clusters
+($\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[1], 2)`). 
+However, the predictive performance in the test subset of the study cohort was only fair with correct mental cluster predictions in `r signif(class_rf$fit_stats$global$Accuracy[2], 2) * 100`% of cases
+($\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[2], 2)`). 
+The predictive performance was the best for PTG cluster individuals of whom >80% and > 62% were classified correctly in the training and test subset, respectively (__Figure \@ref(fig:fig-crf)A__).
+The most important explanatory variables for the mental cluster prediction by conditional random forests were presence and frequency of flashbacks, cautious behavior during sport activity after the accident, annual income, sex and somatic health consequences of the accident.
 
@@
 
+In individuals surveyed on average 3.5 years after the non-professional sport accident, some of its consequences like somatic health problems, unwilling flashbacks as well as phenomena of post-traumatic growth and symptoms of post-traumatic syndrome disorder were still tangible.
+By semi-supervised clustering we could discern three distinct patterns of mental health in this collective with each present in approximately one-third of participants: (1) a neutral mental response reaction, (2) salutatory reaction to the accident characterized by post-traumatic growth and (3) mental health deterioration hallmarked by symptoms of anxiety, depression, somatization, resilience loss and signs of post-traumatic syndrome disorder.
+Notably, separation of the PTG and PTSD clusters was in part blurry. 
+Symptoms of post-traumatic syndrome were observed also in PTG cluster participants albeit less frequently than in the PTSD cluster. 
+Conversely, mental health problems in PTSD cluster participants were paralleled by some signs of post-traumatic growth. 
+In a detailed comparison of `r length(clust_bcg$variables)` factors between the clusters, no distinct demographic, socioeconomic, clinical or accident-related background of the mental clusters could be identified.
+We found a moderate enrichment of females, low-to-moderate income participants, non-academic graduates, chronic somatic and psychiatric conditions as well as individuals requiring hospitalization and surgery in the PTG and, in particular, in the PTSD cluster.
+Injury severity and localization or professional rescue rates indicative of potentially more severe accident were comparable between the clusters. 
+Similarly, the cluster assignment was unlikely affected by psychological support after the accident. 
+In turn, the PTSD and, to a lesser extent, the PTG cluster were characterized by somatic health consequences of the accident, flashbacks, more caution during sport activity and subjective need for psychological support.
+Finally, the lacking non-mental explanatory variables strongly associated with the cluster assignment likely resulted in sub-optimal performance of machine learning classifiers employed for prediction of the cluster assignment and identification of accident victims at risk of mental health problems.
+
+_Hanna and Katharina: your turn now:)_
+
+Our study bears limitations. 
+First, although the key phenomena phenomena in the newly identified mental clusters, post-traumatic growth and post-traumatic syndrome disorder, are clearly associated with trauma [@Tedeschi1996; @Bovin2016], we could not exclude the mental cluster assignment to be affected by other events than the sport accident.
+To account for that, follow-up studies including a control group are needed. 
+Second, our variable set misses potentially important explanatory factors for the mental cluster classification such as length of hospital stay, rehabilitation need or ability to work. 
+Finally, effects of the accident, injury severity, hospitalization or surgery on mental health after an accident may have been obscured in the cross-sectional cohort and needs to be investigated in a more defined collective e.g.
+of high risk sport accident victims or hospitalized individuals.
+
 # Conclusion
+
+_Your part again please_
 
@@
 
-An R data (RDa) file with anonymized patient data will be made available upon request to the corresponding author. 
+An R data (RData) file with anonymized patient data will be made available upon request to the corresponding author. 
 The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.
@@
 
-flextable::flextable(tables$cohort_demo) %>% 
-  flextable::width(1:2, width = 5.5, unit = 'cm') %>% 
+tables$cohort_demo %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
+  flextable::width(1, width = 5.5, unit = 'cm') %>% 
+  flextable::width(2, width = 7.5, unit = 'cm') %>% 
+  footnote(4, 1, 
+           value = as_paragraph('young: 16 - 30, middle: 31 - 65, elderly: \u2265 66 years'), 
+           part = 'body', 
+           ref_symbols = 'a') %>% 
+  footnote(13, 1, 
+           value = as_paragraph('CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale'), 
+           part = 'body', 
+           ref_symbols = 'b') %>% 
   theme_vanilla
@@
   footnote(8, 1, 
-           value = as_paragraph('AIS: acute injury severity score'), 
+           value = as_paragraph('AIS: abbreviated injury scale'), 
            part = 'body', 
@@
 
-flextable::flextable(tables$cohort_mental) %>% 
-  flextable::width(1:2, width = 5.5, unit = 'cm') %>% 
+tables$cohort_mental %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
+  flextable::width(1, width = 5.5, unit = 'cm') %>% 
+  flextable::width(2, width = 7.5, unit = 'cm') %>% 
   footnote(1, 1, 
-           value = as_paragraph('PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; PCL-5 DSM-5: PTSD Checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder'),
            part = 'header', 
@@
 
-```{r fig-consort, fig.width = mm_inch(figures$consort$w), fig.height = mm_inch(figures$consort$h), fig.cap = 'Flow diagram of the analysis inclusion process.'}
+```{r fig-consort, fig.width = mm_inch(figures$consort$w), fig.height = mm_inch(figures$consort$h), fig.cap = 'Flow diagram of the analysis inclusion process and the analysis strategy.'}
 
@@
 
-__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process.__
+__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process and the analysis strategy.__
+
+\newpage
+
+```{r fig-clust-psych, fig.width = mm_inch(figures$clust_psych$w), fig.height = mm_inch(figures$clust_psych$h), fig.cap = 'Scores of psychometry readouts in the mental clusters.'}
+
+figures$clust_psych$plot
+
+```
+
+__Figure \@ref(fig:fig-clust-psych). Scores of psychometry readouts in the mental clusters.__
+
+_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations.
+Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
+Three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder), All clustering parameters differed significantly between the clusters as assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Mean normalized scores of major mental health readouts (A), diminished quality of life (B), post-traumatic growth (C) and post-traumatic symptom disorder (D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines.
+Tinted regions represent 2 $\times$ SEM (standard error of the mean).
+Numbers of observations in the clusters are displayed in the plot captions.
+GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder._
+
+\newpage
+
+```{r fig-clust-symptoms, fig.width = mm_inch(figures$clust_symptoms$w), fig.height = mm_inch(figures$clust_symptoms$h), fig.cap = 'Signs of mental health problems in the mental clusters.'}
+
+figures$clust_symptoms$plot
+
+```
+
+__Figure \@ref(fig:fig-clust-symptoms). Signs of mental health problems in the mental clusters.__
+
+_Frequencies of symptoms of anxiety (A, GAD-7 ≥10), depression (B, PHQ-9 ≥ 11) and somatization (C, PHQ-15 ≥ 11), distribution of resilient coping classes (D, RS13) and symptoms of post-traumatic syndrome disorder (E, PTSD, PCL-5 DSM-5, positivity for PTSD domains and frequency of participants positive for at least one PTSD domain [PTSD+]) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Fractions of symptom-positive and -negative participants in the mental clusters in the training and test subset of the study cohort are presented in stack and bar plots.
+Effect sizes and p-values are shown in the plot captions or Y axes.
+Numbers of observations in the clusters are shown in the X axes or plot captions.
+Significant effects in (E) are highlighted in bold.
+GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTSD: post-traumatic syndrome disorder._
+
+\newpage
+
+```{r fig-clinic-factors, fig.width = mm_inch(figures$clinic_factors$w), fig.height = mm_inch(figures$clinic_factors$h), fig.cap = 'Demographic and clinical factors differing between the mental clusters of accident victims.'}
+
+figures$clinic_factors$plot
+
+```
+
+__Figure \@ref(fig:fig-clinic-factors). Demographic and clinical factors differing between the mental clusters of accident victims.__
+
+_Distribution of gender (A), annual household income (B), and frequencies of self-reported somatic (C) and mental illness (D) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
+Effect sizes and p-values are displayed in the plot captions.
+Numbers of observations in the clusters are presented in the X axes._
+
+\newpage
+
+```{r fig-injury, fig.width = mm_inch(figures$injury$w), fig.height = mm_inch(figures$injury$h), fig.cap = 'Injury severity and location in the mental clusters.'}
+
+figures$injury$plot
+
+```
+
+__Figure \@ref(fig:fig-injury). Injury severity and location in the mental clusters.__
+
+_Distribution of injury severity grades (A, AIS: abbreviate injury scale), hospitalization (B) and surgery rates (C), and injuries of the body parts (D) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack and bar plots.
+Effect sizes and p-values are displayed in the plot captions or Y axes.
+Numbers of observations in the clusters are presented in the X axes or plot captions._
+
+\newpage
+
+```{r fig-aftermath, fig.width = mm_inch(figures$aftermath$w), fig.height = mm_inch(figures$aftermath$h), fig.cap = 'Consequences of the accident in the mental clusters.'}
+
+figures$aftermath$plot
+
+```
+
+__Figure \@ref(fig:fig-aftermath). Consequences of the accident in the mental clusters.__
+
+_Frequencies of self-reported somatic health consequences of the accident (A), flashbacks during sport (B), more or less cautious behavior during sport activity (C), self-reported psychological support need (D) and of a received psychological support (E) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
+Effect sizes and p-values are displayed in the plot captions.
+Numbers of observations in the clusters are presented in the X axes._
+
+\newpage
+
+```{r fig-crf, fig.width = mm_inch(figures$crf$w), fig.height = mm_inch(figures$crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical and accident-related factors.'}
+
+figures$crf$plot
+
+```
+
+__Figure \@ref(fig:fig-crf). Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical and accident-related factors.__
+
+_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident-related factors was trained in the training subset of the study cohort._
+
+_(A) Prediction accuracy of the classifier in the training and test subset of the study cohort presented in a heat map plot of the confusion matrices.
+Color codes for number of cases.
+The overall prediction accuracy, $\kappa$ and observation numbers are displayed in the plot caption.
+Prediction accuracy for particular clusters is shown in the X axis._
+
+_(B) Conditional importance of the explanatory variables for prediction of the mental cluster assignment expressed as accuracy loss ($\Delta$ accuracy)._
 

### 2023-02-24 paper/markdown/supplementary_material.Rmd
Analysis pipeline, Manuscript, Suppelements

--- 2023-02-13 paper/markdown/supplementary_material.Rmd
@@
 ---
-title: "Three distinct patterns of mental health response after a sport accident"
+title: "Three distinct patterns of mental response after a sport accident"
 subtitle: "Supplementary Material"
@@
 
+## Software
+
+Data analysis and transformation was accomplished with R version 4.2.0. 
+The study dataset was imported from an SPSS file with the _foreign_ package [@RCoreTeam]. 
+Tabular data were handled with the _tidyverse_ package bundle [@Wickham2019] and the packages _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
+Test variables were handled with _stringi_ [@Gagolewski2021].
+
+For distribution testing (normality, variance, Gini index), the packages [_ExDA_](https://github.com/PiotrTymoszuk/ExDA), _rstatix_ [@Kassambara2021] and _DescTools_ [@Signorell2022] were utilized. 
+Consistency of psychometric tools was assessed with the packages _psych_ [@Revelle2022], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).
+Clustering tendency was investigated with _factoextra_ [@Kassambara2020] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
+Statistical hypothesis testing and correlation analyses were accomplished with [_ExDA_](https://github.com/PiotrTymoszuk/ExDA), _rstatix_ [@Kassambara2021].
+
+For clustering, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized.
+UMAP (uniform manifold approximation and projection) was done with the _umap_ package [@Konopka2022; @McInnes2018]. 
+Univariable mental cluster assignment classifiers were developed with the _OneR_ [@VonJouanne-Diedrich2017; @Holte1993]. 
+A multi-variable classifier for mental cluster assignment was developed with the conditional random forest algorithm [@Hothorn2006; @Strobl2007] implemented in the _party_ package [@Hothorn2022]. 
+Accuracy of the classifiers was assessed with the packages _caret_ [@Kuhn2008] and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra).
+
+Results were visualized with _ggplot_ (bar plots, box plots and heat maps of confusion matrices) [@Wickham2016], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (violin, stack and ribbon plots) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) (cluster quality control plots, clustering features heat maps, distance heat maps, scatter plots of UMAP layouts).
+Figures were created with the package _cowplot_ [@Wilke2019]. 
+Tables were generated with _flextable_ [@Gohel2022]. 
+The manuscript and Supplementary Material were written in _rmarkdown_ [@Allaire2022] with the package _bookdown_ [@Xie2016]. 
+Figures, tables and R expressions in the markdown documents were managed with the development package [_figur_](https://github.com/PiotrTymoszuk/figur). 
+The markdown documents were rendered with _knitr_ [@Xie2022], _bookdown_ [@Xie2016] with `author-info-blocks.lua` and `scholarly-metadata.lua` scripts by Albert Krewinkel and Jörn Krenzer.
+
+## Data import and transformation
+
+The study data set was imported from an SPSS file with raw study data (function `read.spss()`, package _foreign_). 
+The list of extracted variables with their description is available in __Supplementary Table S\@ref(tab:tab-analysis-vars)__. 
+A total of `r nrow(ptsd$dataset)` participants with the complete set of psychometric battery variables (__Supplementary Table S\@ref(tab:tab-mental-battery)__) were included in the analysis (__Figure 1__).
+
+The psychometric battery (__Supplementary Table S\@ref(tab:tab-mental-battery)__) consisted of German versions of assessment tools for 
+anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
+depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
+panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
+persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
+resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
+loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
+loss of quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
+post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
+post-traumatic syndrome disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016].
+
+Symptoms of anxiety were defined as GAD-7 $geq$ 10, symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], significant somatization/persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002].
+Resilient coping classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].
+
+Items of the EUROHIS QOL 8 tool were scores as with 1 - 5 likert scales (1: no concerns/full satisfaction, 5: extreme concerns, no satisfaction at all) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing).
+The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].
+
+Separate scores were computed for each domain of the PTGI tool (I: relations, II: possibilities, III: personal strength, IV: spiritual strength, V: life appreciation) with each item scores as 0: none, 1: very little, 2: little, 3: moderate, 4: great, 5: extremely great.
+In addition, the total PTGI score was calculated as the sum of all items [@Tedeschi1996].
+
+Separate scores were calculated for domains B, C, D and E of the PCL-5 DSM-5 tool along with the total score being the sum of all items. 
+Each PCL-5 DSM-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
+Participants positive for the B domain or C domain were identified by at least one item scored with 'moderate', participants positive for the D or E domain were identified by at least two item scored with 'moderate'.
+Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 DSM-5 domains [@Bovin2016].
+
+Traumatic events prior to the sport accident were assessed with the DIA-X tool [@Maercker2005]. 
+Direct personal experience or being a witness of a traumatic event specified by the DIA-X questionnaire or by an additional yes/no item ('other traumatic events') was scored as 1. 
+Prior traumatic event was assumed with at least one DIA-X item scored with 1. 
+Prior sport accidents, 
+flashbacks during sport activity, 
+confusion during sport activity, 
+subjective need for psychological support, 
+presence of psychological/psychiatric support/therapy post accident 
+and presence of somatic health consequences after the accident 
+were surveyed as single yes/no items. 
+Flashbacks frequency was assessed in the following categories: none, more than one per month and more than one per year. 
+Smoking was surveyed as a single yes/no question. 
+Substance abuse was investigated with the CAGE tool with $\geq$ 2 points indicative of substantial addiction [@OBrien2008]. 
+Data on hospital treatment, hospitalization and surgery were extracted from electronic patient's records. 
+Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].
+
+Additional information on study variables and their stratification schemes are presented in __Supplementary Table S\@ref(tab:tab-analysis-vars)__.
+
+## Consistency of psychometric tools and power analysis
+
+Consistency of psychometric tools was assessed by factor analysis and McDonald's $\omega$ (function `omega()`, package _psych_) [@McDonald1999; @Revelle2022].
+The number of latent factors for calculation of $\omega$ was identified by inspection of loadings determined by factor analysis (function `reduce_data()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)).
+All psychometric tools used in the study except for the stress PSS-4 scale [@Cohen1983] and the BRCS resilience tool [@Sinclair2004] exhibited good-to-excellent consistency with $\omega$ > 0.8 (__Supplementary Table S\@ref(tab:tab-consistency)__).
+The poorly performing PSS-4 and BRCS scales were excluded from the further analysis.
+
+To find the optimal size of a training subset of the study cohort, random subsamples of the study dataset of varying sizes were investigated (`r paste(pow$sizes, collapse = ', ')` observations, 50 random draws per subsample size).
+For each random subsample, clustering tendency was assessed by Hopkins statistic (function `get_clust_tendency()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
+For n = 200 observations, the median Hopkins statistic was `r pow$hopkins %>% filter(size == 200) %>% .$hopkins_stat %>% median %>% signif(2)` (interquartile range: `r pow$hopkins %>% filter(size == 200) %>% .$hopkins_stat %>% quantile(probs = c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' - ')`) indicative of good clustering tendency.
+The Hopkins statistic value for this subset was comparable with the value for the entire subset (`r signif(pca$clust_tendency$data$hopkins_stat, 2)`). 
+For these reasons, 2/3 of the entire dataset or n = `r count(ptsd$dataset, partition)$n[1]` was considered as the size of the training subset of the study cohort.
+
+## Training/test subset definition
+
+The study participants were assigned to the training and test subsets with the 2:1 size ratio (__Figure 1__). 
+To this end, `r rand$part_obs %>% length` random splits of the datasets were generated. 
+The subset assignment scheme was chosen with possibly small differences in accident year, age, gender, somatic and mental illness rates, frequencies of prior traumatic events (DIA-X) [@Maercker2005] and distribution of injury severity grades (AIS: abbreviated injury scale) [@Gennarelli2006] as investigated by $\chi^2$ and Mann-Whitney test, as appropriate (functions `chisq_test()` and `wilcox_test()`, package _rstatix_).
+
+## Assessment of selection bias
+
+To assess possible selection bias, demographic, socioeconomic, clinical and accident-related parameters were compared between individuals excluded due to missingness of psychometric data or denying survey participation and participants included in he analysis (__Figure 1__).
+Categorical variables were compared by $\chi^2$ test with Cramer V effect size statistic and numeric variables were compared by Mann-Whitney test with r effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+Potential differences between the training and test subset of the study cohort (__Figure 1__) were analyzed in an analogical way. 
+Significant and near-significant (p < 0.1) differences between the included/excluded participants and the training/test subsets are presented in __Supplementary Tables S\@ref(tab:tab-incl-excl)__ and __S\@ref(tab:tab-partition)__, respectively.
+
+## Semi-supervised clustering
+
+Observations of the training subset of the study cohort were subjected to clustering in respect to the numeric psychometric scores (__Supplementary Table S\@ref(tab:tab-mental-battery)__). 
+The score values were normalized and median-centered (function `center_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
+For clustering, the PAM (partition around medoids) algorithm [@Schubert2019] with the cosine distance measure between observations [@Drost2018] was employed. 
+The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
+The choice of the clustering algorithm was motivated by its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares, method `var()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and superior stability in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 7-nearest neighbors classifier, method `cv()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__.
+The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
+Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
+This semi-supervised procedure yielded clustering structures with similar fractions of explained variance in the training (V = `r signif(semi_clust$variance$variance[1], 2)`) and test subset (V = `r signif(semi_clust$variance$variance[2], 2)`).
+In addition, comparable separation between the clusters could be discerned in the training and test subsets by a visual analysis of UMAP layouts, pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__) and heat maps of normalized levels of the clustering variables (__Supplementary Figure S\@ref(fig:fig-feat-hm)__).
+
+Differences in clustering variables, frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical and accident-related factors between the mental clusters were assessed separately in the training and test subset by $\chi^2$ test with Cramer V effect size statistic and Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (__Supplementary Tables S\@ref(tab:tab-clust-fct)__ and __S\@ref(tab:tab-clust-bcg)__).
+
+## Uni- and multi-variable cluster assignment classifiers
+
+Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters were trained in the training subset of the study cohort by the one-rule (oneR) algorithm (function `OneR()`, package _OneR__) [@VonJouanne-Diedrich2017; @Holte1993].
+The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _OneR_). 
+The multi-parameter classifier was developed in the training subset with the conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022] with n = 1000 random trees, the splitting rule corresponding to the maximum test statistic and five variable draws per tree (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 5, mincriterion = qnorm(0.9), ntree = 1000)` control object, package _party_).
+The optimal variable draw values were obtained by optimization (tuning) employing the out-of-bag predictions in the training subsets (not shown). 
+The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _party_).
+
+Accuracy and $\kappa$ statistic for the classifiers' mental cluster assignment predictions in the training and test subsets were computed with the function `multiClassSummary()` from the package _caret_ [@Kuhn2008].
+Conditional importance for explanatory variables expressed as accuracy loss [@Strobl2008] was obtained for the conditional random forest algorithm with the function 'varimp()` from the _party_ package.
+
+## Data and code availability
+
+An R data (RData) file with anonymized patient data will be made available upon request to the corresponding author. 
+The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.
+
 \newpage
@@
 suppl_tables$analysis_vars[1, 1] %>% 
-  mutate(Section = '') %>% 
+  mutate(`Variable name in R` = '') %>% 
   set_names(' ') %>% 
@@
 
-```{r tab-incl-excl, tab.cap = 'Significant and near signifcant (unadjusted p < 0.05) differences between individuals excluded from analysis and analyzed study participants.
+```{r tab-consistency, tab.cap = "Consistency of the psychometric tools used in the study measured by McDonald's omega."}
+
+flextable::flextable(suppl_tables$consistency) %>% 
+  flextable::width(1, width = 3.5, unit = 'cm') %>% 
+  flextable::width(2:3, width = 3, unit = 'cm') %>% 
+  footnote(1, 1, 
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; BRCS: brief resilent coping scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-incl-excl, tab.cap = 'Significant and near-significant (p < 0.1) differences between individuals excluded from analysis and analyzed study participants.
 Numeric variables are presented as medians with interquartile ranges (IQR).
@@
 
-flextable::flextable(suppl_tables$incl_excl) %>% 
+suppl_tables$incl_excl %>% 
+  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
+  flextable %>% 
   flextable::width(1, width = 3.2, unit = 'cm') %>% 
   flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
-  flextable::width(4, width = 2.4, unit = 'cm') %>% 
+  flextable::width(4, width = 2.7, unit = 'cm') %>% 
   flextable::width(5, width = 2.3, unit = 'cm') %>% 
   footnote(1, 4:5, 
-           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03c/\u00B2 test with Cramer V effec size statistic.
-P values were corrected for multiple testing with the false discovery method (FDR)'),
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03c\u00B2 test with Cramer V effec size statistic'), 
            part = 'header', 
            ref_symbols = 'a') %>% 
-  footnote(3:4, 1, 
-           value = as_paragraph('AIS: acute injury severity score'), 
+  footnote(8, 1, 
+           value = as_paragraph('AIS: abbreviated injury scale'), 
            part = 'body', 
            ref_symbols = 'b') %>% 
-  footnote(7:10, 1, 
+  footnote(11, 1, 
+           value = as_paragraph('PCL-5 DSM-5: PTSD checklist for DSM-5'), 
+           part = 'body', 
+           ref_symbols = 'c') %>% 
+  footnote(12:15, 1, 
            value = as_paragraph('PTGI: post-traumatic growth inventory'), 
            part = 'body', 
-           ref_symbols = 'c') %>% 
-  footnote(11, 1, 
+           ref_symbols = 'd') %>% 
+  footnote(16, 1, 
            value = as_paragraph('PSS4: 4-item perceived stress scale'), 
            part = 'body', 
-           ref_symbols = 'd') %>% 
-  footnote(12, 1, 
-           value = as_paragraph('SOC-9L: Leipzig 9-item sense of soherence scale'), 
+           ref_symbols = 'e') %>% 
+  footnote(17, 1, 
+           value = as_paragraph('SOC-9L: Leipzig 9-item sense of coherence scale'), 
            part = 'body', 
-           ref_symbols = 'e') %>% 
+           ref_symbols = 'f') %>% 
   theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-partition, tab.cap = 'Significant and near-significant (p < 0.1) differences between the training and test subsets of the study cohort.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+flextable::flextable(suppl_tables$partition) %>% 
+  flextable::width(1, width = 3.2, unit = 'cm') %>% 
+  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
+  flextable::width(4, width = 2.7, unit = 'cm') %>% 
+  flextable::width(5, width = 2.3, unit = 'cm') %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('\u03c\u00B2 test with Cramer V effec size statistic'), 
+           part = 'body', 
+           ref_symbols = 'a') %>% 
+  footnote(2, 1, 
+           value = as_paragraph('RS13: 13-item resilience scale'), 
+           part = 'body', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-clust-fct, tab.cap = 'Differences in psychometric clustering factors between the mental clusters.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+The table is available in a supplementary Excel file.'}
+
+suppl_tables$clust_fct[1, 1] %>% 
+  mutate(Variable = '') %>% 
+  set_names(' ') %>% 
+  flextable
+
+```
+
+\newpage
+
+```{r tab-clust-bcg, tab.cap = 'Significant and near-significant (p < 0.1) differences in demographic, socioeconomic, clinical and accident-related factors between the mental clusters.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.
+The table is available in a supplementary Excel file.'}
+
+suppl_tables$clust_bcg[1, 1] %>% 
+  mutate(Variable = '') %>% 
+  set_names(' ') %>% 
+  flextable
 
@@
 
+```{r fig-parts-mental-cohort, fig.width = mm_inch(suppl_figures$parts_mental_cohort$w), fig.height = mm_inch(suppl_figures$parts_mental_cohort$h), fig.cap = 'Injured body regions and detailed scoring of quality of life, post-traumatic syndrome disorder and post-traumatic growth in the study cohort.'}
+
+suppl_figures$parts_mental_cohort$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-parts-mental-cohort).
+Injured body regions and detailed scoring of quality of life, post-traumatic syndrome disorder and post-traumatic growth in the study cohort.__
+
+_For categorical variables, percentages of complete observations are presented as bars.
+Numeric variables are presented in violin plots with red diamonds denoting medians, interquartile ranges presented as red whiskers and single observations depicted as points.
+The number of complete cases is displayed in the plot captions._
+_(A) Distribution of injured body regions.
+(B) Scoring of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8).
+(C) Scoring of the post-traumatic syndrome disorder (PTSD) clusters with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD clusters.
+(E) Scoring of the factors of the post-traumatic growth with the post-traumatic growth inventory (PTGI) tool._
+
+\newpage
+
+```{r fig-clust-dev, fig.width = mm_inch(suppl_figures$clust_dev$w), fig.height = mm_inch(suppl_figures$clust_dev$h), fig.cap = 'Definition of the mental clusters in the training subset of the study cohort.'}
+
+suppl_figures$clust_dev$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clust-dev). Definition of the mental clusters in the training subset of the study cohort.__
+
+_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations._
+
+_(A) Comparison of explanatory performance and stability of various clustering algorithms in the training subset.
+The explanatory performance was measured as a fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares).
+The stability was assessed by a rate of correct cluster assignment in 10-fold cross-validation (CV) with cluster assignment in the folds by an inverse distance weighted 7-nearest neighbors classifier.
+Note the superior stability of the PAM/cosine distance algorithm._
+
+_(B) Determination of the cluster number by the bend of the within-cluster sum of squares curve and the peak mean silhouette statistic._
+
+\newpage
+
+```{r fig-semi-clust, fig.width = mm_inch(suppl_figures$semi_clust$w), fig.height = mm_inch(suppl_figures$semi_clust$h), fig.cap = 'Semi-supervised clustering.'}
+
+suppl_figures$semi_clust$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-semi-clust). Semi-supervised clustering.__
+
+_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations.
+Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
+By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder).
+Numbers of observations in the mental clusters are shown in the plot legend in (A)._
+
+_(A) Observations in the training and test subsets of the study cohorts were subjected to two-dimensional UMAP (uniform manifold approximation and projection) in respect to the psychometric scores.
+UMAP layouts are shown in scatter plots.
+Points represent single samples.
+Point color codes for the cluster assignment._
+
+_(B) Pairwise cosine distances between observations in the mental clusters of the training and test subsets presented in heat maps._
+
+\newpage
+
+```{r fig-feat-hm, fig.width = mm_inch(suppl_figures$feat_hm$w), fig.height = mm_inch(suppl_figures$feat_hm$h), fig.cap = 'Levels of psychometric clustering scores in the mental clusters.'}
+
+suppl_figures$feat_hm$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-feat-hm). Levels of psychometric clustering scores in the mental clusters.__
+
+_Levels of psychometric scores used in clustering in the mental clusters.
+Normalized, median-centered score values (Z scores) are presented as heat maps.
+Numbers of observations in the clusters are displayed in the plot captions.
+PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder._
+
+\newpage
+
+```{r fig-demo, fig.width = mm_inch(suppl_figures$demo$w), fig.height = mm_inch(suppl_figures$demo$h), fig.cap = 'Age, education, household income, prior accidents and traumatic events in the mental clusters.'}
+
+suppl_figures$demo$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-demo). Age, education, household income, prior accidents and traumatic events in the mental clusters.__
+
+_Distribution of age classes (A, young: < 30, middle: 30 - 65, elderly > 65 years), education grades (B), frequencies of prior sport accidents (D) and prior traumatic events (E, measured by the DIA-X tool) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
+Effect sizes and p-values are displayed in the plot captions.
+Numbers of observations in the clusters are presented in the X axes._
+
+\newpage
+
+```{r fig-accident, fig.width = mm_inch(suppl_figures$accident$w), fig.height = mm_inch(suppl_figures$accident$h), fig.cap = 'Accident sport type and accident rescue in the mental clusters.'}
+
+suppl_figures$accident$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-accident). Accident sport type and accident rescue in the mental clusters.__
+
+_Distribution of sport types (A) and rescue modes (B, C) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
+Effect sizes and p-values are displayed in the plot captions.
+Numbers of observations in the clusters are presented in the X axes._
+
+\newpage
+
+```{r fig-oneR, fig.width = mm_inch(suppl_figures$oneR$w), fig.height = mm_inch(suppl_figures$oneR$h), fig.cap = 'Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors.'}
+
+suppl_figures$oneR$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-oneR). Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors.__
+
+_Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters were trained in the training subset of the study cohort by the one-rule (oneR) algorithm.
+The cluster assignment was subsequently predicted for the test subset observations.
+Accuracy and $\kappa$ of the classifiers with any predictive value ($\kappa$ > 0 in both subsets) are shown in bar plots._
+
 \newpage

### 2023-02-24 classification scripts/one_rule.R

@@ -1,0 +1,126 @@
+# Analysis of classification capability of single factors
+
+  insert_head()
+  
+# container ------
+  
+  class_one <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## analysis tables for single clusters
+  
+  class_one$clust_tbl <- 
+    list(neutral = class_globals$analysis_tbl, 
+         PTG = class_globals$analysis_tbl, 
+         PTSD = class_globals$analysis_tbl) %>% 
+    map2(., names(.), 
+         function(data, clust) data %>% 
+           map(mutate, 
+               clust_id = ifelse(clust_id == clust, clust, 'rest'), 
+               clust_id = factor(clust_id)))
+
+# univariable modeling ------
+  
+  insert_msg('Univariable modeling')
+  
+  ## global classification
+  
+  class_one$global_models <- class_globals$variables %>% 
+    map(model_one, 
+        train_data = class_globals$analysis_tbl$training, 
+        test_data = class_globals$analysis_tbl$test, 
+        response = 'clust_id') %>% 
+    set_names(class_globals$variables)
+  
+  ## neutral cluster
+  
+  class_one$neutral_models <- class_globals$variables %>% 
+    map(model_one, 
+        train_data = class_one$clust_tbl$neutral$training, 
+        test_data = class_one$clust_tbl$neutral$test, 
+        response = 'clust_id') %>% 
+    set_names(class_globals$variables)
+  
+  ## PTG cluster
+  
+  class_one$PTG_models <- class_globals$variables %>% 
+    map(model_one, 
+        train_data = class_one$clust_tbl$PTG$training, 
+        test_data = class_one$clust_tbl$PTG$test, 
+        response = 'clust_id') %>% 
+    set_names(class_globals$variables)
+  
+  ## PTSD cluster
+  
+  class_one$PTSD_models <- class_globals$variables %>% 
+    map(model_one, 
+        train_data = class_one$clust_tbl$PTSD$training, 
+        test_data = class_one$clust_tbl$PTSD$test, 
+        response = 'clust_id') %>% 
+    set_names(class_globals$variables)
+  
+# model summaries -------
+  
+  insert_msg('Model summaries')
+  
+  class_one$fit_stats[c('global', 
+                        'neutral', 
+                        'PTG', 
+                        'PTSD')] <- class_one[c("global_models", 
[…]

### 2023-02-24 classification scripts/rf.R

@@ -1,0 +1,186 @@
+# Multi-parameter RF classifier with the top factors for cluster classification
+# identified by the OneR procedure
+
+  insert_head()
+  
+# container ------
+  
+  class_rf <- list()
+
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+
+  ## tables of assignments for single clusters
+  
+  class_rf$clust_tbl <- 
+    list(neutral = class_globals$analysis_tbl, 
+         PTG = class_globals$analysis_tbl, 
+         PTSD = class_globals$analysis_tbl) %>% 
+    map2(., names(.), 
+         function(data, clust) data %>% 
+           map(mutate, 
+               clust_id = ifelse(clust_id == clust, clust, 'rest'), 
+               clust_id = factor(clust_id)))
+
+# tuning of the classifier -------
+  
+  insert_msg('Tuning')
+  
+  ## finding the optimal mtry parameter based on OOB prediction accuracy
+  
+  set.seed(1234)
+
+  class_rf$tuning$controls <- 2:12 %>% 
+    map(~cforest_control(teststat = "max",
+                         testtype = "Teststatistic", 
+                         mtry = .x, 
+                         mincriterion = qnorm(0.9), 
+                         ntree = 1000)) %>% 
+    set_names(2:12)
+  
+  class_rf$tuning$models <- class_rf$tuning$controls %>% 
+    map(~cforest(formula = class_globals$formula, 
+                 data = class_globals$analysis_tbl$training, 
+                 controls = .x))
+  
+  ## predictions and fit stats
+  
+  class_rf$tuning$predictions <- class_rf$tuning$models %>% 
+    map(~data.frame(obs = class_globals$analysis_tbl$training$clust_id, 
+                    pred = predict(.x, OOB = TRUE) %>% 
+                      factor(levels = levels(class_globals$analysis_tbl$training$clust_id))))
+  
+  class_rf$tuning$fit_stats <- class_rf$tuning$predictions %>% 
+    map(multiClassSummary, 
+        lev = levels(class_globals$analysis_tbl$training$clust_id)) %>% 
+    map(as.list) %>% 
+    map(as_tibble) %>% 
+    compress(names_to = 'mtry') %>% 
+    mutate(mtry = as.numeric(mtry))
+  
+# Plotting the tuning results ------
+  
+  insert_msg('Plottng the tuning results')
+  
+  ## accuracy, kappa, sensitivity and specificity
+  
+  class_rf$tune_plots <- 
[…]

### 2023-02-24 classification.R

@@ -1,0 +1,142 @@
+# Building a classifier (random forest) to identify the mental health cluster
+# with help of socioeconomic, demographic, clnical and accident-related factors
+
+# tools ------
+
+  library(plyr)
+  library(tidyverse)
+  library(rlang)
+  library(stringi)
+  library(exda)
+  library(soucer)
+  library(furrr)
+  library(rstatix)
+  library(trafo)
+  library(clustTools)
+  library(ggrepel)
+  library(caret)
+  library(caretExtra)
+  library(doParallel)
+  library(OneR)
+  library(party)
+  library(ggwordcloud)
+  
+  
+  insert_head()
+  
+  source_all('./tools/tools.R', 
+             message = TRUE, 
+             crash = TRUE)
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  class_globals <- list()
+
+  ## explanatory variables 
+  
+  class_globals$variables <- ptsd$var_lexicon %>% 
+    filter(type %in% c('characteristic')) %>% 
+    .$variable
+  
+  class_globals$variables <- 
+    class_globals$variables[!class_globals$variables %in% c('accident_injured_persons', 
+                                                            'obs_time', 
+                                                            'accident_year')]
+  
+  ## full model formula
+  
+  class_globals$formula <-  as.formula(paste('clust_id ~ ', 
+                                          paste(class_globals$variables, 
+                                                collapse = '+')))
+  
+  ## cluster assignment schemes
+  
+  class_globals$clust_assignment <- semi_clust$clust_obj %>% 
+    map(~.x$clust_assignment) %>% 
+    map(set_names, c('ID', 'clust_id'))
+  
+  ## analysis table, splitting into the training and test subset
+  ## appending with the cluster assignment scheme
+  
+  class_globals$analysis_tbl <- ptsd$dataset %>% 
+    dlply('partition', 
+          select, 
+          ID, 
+          any_of(class_globals$variables)) %>% 
+    map(as_tibble)
+  
+  class_globals$analysis_tbl <- 
+    map2(class_globals$analysis_tbl, 
+         class_globals$clust_assignment, 
+         left_join, by = 'ID')
+  
+  ## analysis table: recoding the missing injury information
+  
+  class_globals$analysis_tbl <- class_globals$analysis_tbl %>% 
+    map(mutate, 
+        injury_severity_ais = as.character(injury_severity_ais), 
[…]

### 2023-02-24 clustering scripts/background.R

@@ -1,0 +1,167 @@
+# Differences in levels of demographic, socioeconomic and clinical factors 
+# as well as variables related to the accident between the clusters
+
+  insert_head()
+  
+# container -------
+  
+  clust_bcg <- list()
+  
+# parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## explanatory variables 
+  
+  clust_bcg$variables <- ptsd$var_lexicon %>% 
+    filter(type %in% c('characteristic')) %>% 
+    .$variable
+
+  ## cluster assignment schemes
+  
+  clust_bcg$clust_assignment <- semi_clust$clust_obj %>% 
+    map(~.x$clust_assignment) %>% 
+    map(set_names, c('ID', 'clust_id'))
+  
+  ## analysis table, splitting into the training and test subset
+  ## appending with the clust assignment scheme
+  
+  clust_bcg$analysis_tbl <- ptsd$dataset %>% 
+    dlply('partition', 
+          select, 
+          ID, 
+          all_of(clust_bcg$variables)) %>% 
+    map(as_tibble)
+  
+  clust_bcg$analysis_tbl <- 
+    map2(clust_bcg$analysis_tbl, 
+         clust_bcg$clust_assignment, 
+         right_join, by = 'ID')
+  
+  ## test types
+  
+  clust_bcg$test_types <- 
+    clust_bcg$analysis_tbl$training[clust_bcg$variables] %>% 
+    map_lgl(is.numeric) %>% 
+    ifelse('kruskal_eta', 'cramer_v')
+
+# Descriptive stats ------  
+  
+  insert_msg('Descriptive stats')
+  
+  clust_bcg$desc_stats <- clust_bcg$analysis_tbl %>% 
+    future_map(explore, 
+               split_factor = 'clust_id', 
+               variables = clust_bcg$variables, 
+               what = 'table', 
+               pub_styled = TRUE, 
+               .options = furrr_options(seed = TRUE))
+  
+# Testing ------
+  
+  insert_msg('Testing for differences between the clusters')
+  
+  ## all variables are categorical: chi-squared test with Cramer V effect size
+  
+  clust_bcg$test <- clust_bcg$analysis_tbl %>% 
+    future_map(compare_variables, 
+               variables = clust_bcg$variables, 
+               split_factor = 'clust_id', 
+               what = 'eff_size', 
+               types = clust_bcg$test_types, 
[…]

### 2023-02-24 clustering scripts/characteristic.R

@@ -1,0 +1,302 @@
+# Differences in clustering variables between the mental clusters
+
+  insert_head()
+  
+# container -------
+  
+  feat_clust <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  feat_clust$variables <- 
+    c(clust_globals$variables, 
+      ptsd$mental_variables[!ptsd$mental_variables %in% clust_globals$variables], 
+      'dsm5_cluster_class', 
+      'dsm5_B_class', 
+      'dsm5_C_class', 
+      'dsm5_D_class', 
+      'dsm5_E_class', 
+      'rs13_total_class', 
+      'phq9_total_class', 
+      'phq_events_total_class', 
+      'gad7_total_class')
+  
+  ## clustering objects 
+  
+  feat_clust$clust_obj <- semi_clust$clust_obj
+  
+  ## analysis tables: identity and normalized
+  
+  feat_clust$norm_tbl <- feat_clust$clust_obj %>% 
+    map(model.frame) %>% 
+    map(rownames_to_column, 'ID') %>% 
+    map2(., feat_clust$clust_obj, 
+         ~left_join(.x, 
+                    set_names(.y$clust_assignment, c('ID', 'clust_id')), 
+                    by = 'ID')) %>% 
+    map(as_tibble)
+  
+  feat_clust$ident_tbl <- ptsd$dataset %>% 
+    dlply('partition', 
+          select, 
+          ID, 
+          all_of(feat_clust$variables)) %>% 
+    map2(., feat_clust$clust_obj, 
+         ~left_join(.x, 
+                    set_names(.y$clust_assignment, c('ID', 'clust_id')), 
+                    by = 'ID')) %>% 
+    map(as_tibble)
+  
+  ## test types
+  
+  feat_clust$test_types <- feat_clust$ident_tbl$training[feat_clust$variables] %>% 
+    map_lgl(is.numeric)
+  
+  feat_clust$test_types <- ifelse(feat_clust$test_types, 
+                                  'kruskal_eta', 
+                                  'cramer_v')
+  
+# Descriptive stats -------
+  
+  insert_msg('Descriptive stats')
+  
+  feat_clust$desc_stats <- feat_clust$ident_tbl %>% 
+    map(explore, 
+        variables = feat_clust$variables, 
+        split_factor = 'clust_id', 
+        what = 'table', 
+        pub_styled = TRUE)
+  
+# Testing for differences -----
+  
[…]

### 2023-02-24 clustering scripts/clustering.R

@@ -1,0 +1,206 @@
+# Semi-supervised clustering.
+#
+# The most stable clustering structure (PAM/cosine) is trained in the training
+# subset and the cluster assignment predicted for the test subset 
+# (7-NN, inverse distance weighted). 
+# Goodness of prediction: clustering variance
+#
+# Diagnostic plots for the training subset and variable importance measures
+
+  insert_head()
+  
+# container ------
+  
+  semi_clust <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  semi_clust$analysis_tbl <- clust_globals$analysis_tbl
+  
+# Training and predictions -----
+  
+  insert_msg('Devlopment of the clustering structures')
+  
+  ## training, renaming the clusters
+  
+  semi_clust$clust_obj$training <- 
+    kcluster(semi_clust$analysis_tbl$training, 
+             distance_method = 'cosine', 
+             clust_fun = 'pam', 
+             k = 3, 
+             seed = 1234)
+  
+  semi_clust$clust_obj$training$clust_assignment <- 
+    semi_clust$clust_obj$training$clust_assignment %>% 
+    mutate(clust_id = car::recode(as.character(clust_id), 
+                                  "'1' = 'PTSD'; 
+                                  '2' = 'neutral'; 
+                                  '3' = 'PTG'"), 
+           clust_id = factor(clust_id, c('neutral', 'PTG', 'PTSD'))) 
+  
+  ## test
+  
+  semi_clust$clust_obj$test <- 
+    predict(semi_clust$clust_obj$training, 
+            newdata = semi_clust$analysis_tbl$test, 
+            type = 'propagation', 
+            kNN = 7, 
+            resolve_ties = TRUE, 
+            simple_vote = FALSE)
+  
+# Clustering variances ------
+  
+  insert_msg('Clustering variances')
+  
+  semi_clust$variance <- semi_clust$clust_obj %>% 
+    map(var) %>% 
+    map(~.x$frac_var) %>% 
+    compress(names_to = 'partition', 
+             values_to = 'variance') %>% 
+    map_dfc(unlist)
+  
+  ## plotting the variance values
+  
+  semi_clust$variance_plot <- semi_clust$variance %>% 
+    ggplot(aes(x = variance, 
+               y = reorder(partition, variance), 
+               fill = partition)) + 
+    geom_bar(stat = 'identity', 
+             color = 'gray20') + 
[…]

### 2023-02-24 clustering scripts/development.R

@@ -1,0 +1,131 @@
+# Development of the optimal clustering procedure in the training dataset
+
+  insert_head()
+  
+# container list ------
+  
+  clust_devel <- list()
+  
+# globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## analysis table
+  
+  clust_devel$analysis_tbl <- clust_globals$analysis_tbl$training
+
+  ## distances
+  
+  clust_devel$distances <- 
+    c('euclidean', 'manhattan', 'cosine')
+
+# Clustering algorithms ------
+  
+  insert_msg('Generating clustering objects')
+  
+  ## HCL, Ward D2
+
+  clust_devel$algos[paste0('hcl_', clust_devel$distances)] <- 
+    clust_devel$distances %>% 
+    map(hcluster, 
+        data = clust_devel$analysis_tbl, 
+        k = 3,  
+        hc_method = 'ward.D2')
+  
+  ## K-means
+  
+  clust_devel$algos[paste0('kmeans_', clust_devel$distances)] <-  
+    clust_devel$distances %>% 
+    map(kcluster, 
+        data = clust_devel$analysis_tbl, 
+        k = 3, 
+        clust_fun = 'kmeans') 
+  
+  ## PAM
+  
+  clust_devel$algos[paste0('pam_', clust_devel$distances)] <-  
+    clust_devel$distances %>% 
+    map(kcluster, 
+        data = clust_devel$analysis_tbl, 
+        k = 3, 
+        clust_fun = 'pam') 
+
+# Clustering variances ------
+  
+  insert_msg('Clustering variances')
+  
+  clust_devel$variance <- clust_devel$algos %>% 
+    map(var) %>% 
+    map(~.x$frac_var) %>% 
+    compress(names_to = 'method', 
+             values_to = 'variance') %>% 
+    mutate(variance = unlist(variance))
+  
+# Cross-validation ----
+  
+  insert_msg('Cross-validation')
+  
+  plan('multisession')
+  
+  clust_devel$cv <- clust_devel$algos %>% 
+    future_map(cv, 
+               nfolds = 10, 
+               kNN = 7, 
+               resolve_ties = TRUE, 
+               simple_vote = FALSE, 
+               .parallel = FALSE, 
+               .options = furrr_options(seed = TRUE, 
+                                        packages = c('clustTools', 
+                                                     'somKernels'))) 
+  
+  plan('sequential')
+  
+  clust_devel$cv <- clust_devel$cv %>% 
[…]

### 2023-02-24 clustering.R

@@ -1,0 +1,61 @@
+# Semi-supervised clustering. The cluster structure is tuned and trained 
+# in the training subset of the cohort and subsequently validated in the 
+# test subset (cluster assignment by inverse distance weighted kNN)
+#
+# Levels of psychometric variables are compared between the clusters
+# and finally, demographic, socioeconomic, clinical and accident-related 
+# measures are compared between the clusters
+
+
+# tools ------
+
+  library(plyr)
+  library(tidyverse)
+  library(rlang)
+  library(stringi)
+  library(exda)
+  library(soucer)
+  library(furrr)
+  library(rstatix)
+  library(trafo)
+  library(psych)
+  library(clustTools)
+  library(ggrepel)
+  
+  insert_head()
+  
+  source_all('./tools/tools.R', 
+             message = TRUE, 
+             crash = TRUE)
+  
+# analysis globals -------
+  
+  insert_msg('Clustering globals')
+  
+  clust_globals <- list()
+  
+  ## clustering variables
+  
+  clust_globals$variables <- ptsd$mental_variables
+  
+  ## normalized, median-centered psychometric data
+
+  clust_globals$analysis_tbl <- ptsd$dataset %>% 
+    dlply('partition', as_tibble) %>% 
+    map(select, ID, all_of(clust_globals$variable)) %>% 
+    map(column_to_rownames, 'ID') %>% 
+    map(center_data, type = 'median')
+  
+# clustering analysis scripts -----
+  
+  insert_msg('Clustering analysis scripts')
+  
+  c('./clustering scripts/development.R', 
+    './clustering scripts/clustering.R', 
+    './clustering scripts/characteristic.R', 
+    './clustering scripts/background.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# END ------
+  
+  insert_tail()

### 2023-02-24 exec.R

@@ -5,4 +5,6 @@
   print(source_all(c('import.R', 
                      'exploration.R', 
                      'factor.R', 
+                     'clustering.R', 
+                     'classification.R', 
                      'paper.R')))

### 2023-02-24 exploration scripts/bias.R

@@ -29,7 +29,12 @@
         all_of(eda_globals$variables$variable)) %>% 
     compress(names_to = 'analysis_status') %>% 
     mutate(analysis_status = factor(analysis_status, 
-                                    c('included', 'excluded')))
+                                    c('included', 'excluded')), 
+           accident_injured_persons = ifelse(accident_injured_persons == 'no information', 
+                                             NA, as.character(accident_injured_persons)), 
+           accident_injured_persons = factor(accident_injured_persons, 
+                                             levels(ptsd$dataset$accident_injured_persons)), 
+           accident_injured_persons = droplevels(accident_injured_persons))
   
 # Descriptive stats -------
   
@@ -50,8 +55,6 @@
   ##
   ## working with safely(): some variables are so missing in the excluded 
   ## set that no testing is possible
-  ##
-  ## FDR correction
   
   excl$test <- 
     list(variables = eda_globals$variables$variable, 
@@ -65,24 +68,14 @@
                 pub_styled = TRUE, 
                 .options = furrr_options(seed = TRUE)) %>% 
     map_dfr(~.x$result) %>% 
-    re_adjust %>% 
-    mutate(plot_cap = paste(eff_size, significance, sep = ', '), 
-           y_lab = ifelse(test == 'Chi-squared test', 
-                          '% of strata', 
-                          exchange(variable, 
-                                   dict = ptsd$var_lexicon, 
-                                   key = 'variable', 
-                                   label = 'axis_lab')), 
-           plot_type = ifelse(test == 'Chi-squared test', 
-                              'stack', 
-                              'violin'))
+    format_fct_test
   
 # Significant and near significant differences -----
   
   insert_msg('Significant and near significant differences')
   
   excl$top_factors <- excl$test %>% 
-    filter(p_value < 0.05) %>% 
+    filter(p_value < 0.1) %>% 
     .$variable
   
 # A ready to use table with significant and near significant differences -------
@@ -98,33 +91,18 @@
     reduce(left_join, by = 'variable') %>% 
     format_summ_tbl %>% 
     set_names(c('Variable', 'Included', 'Excluded', 
-                'Significance (FDR)', 'Effect size'))
+                'Significance', 'Effect size'))
   
 # Plots ------
   
   insert_msg('Plots')
   
   excl$plots <- 
-    list(variable = excl$test$variable, 
[…]

### 2023-02-24 exploration scripts/cohort.R

@@ -25,6 +25,7 @@
   cohort$var_class <- 
     list(demo = c('obs_time', 
                   'age', 
+                  'age_class', 
                   'sex', 
                   'residence_alpine_region', 
                   'education', 
@@ -33,9 +34,8 @@
                   'trauma_risk_profession', 
                   'household_income_class', 
                   'smoking_status', 
-                  'drug_status', 
+                  'cage_total_class', 
                   'somatic_comorbidity', 
-                  #'somatic_comorbidity_type', 
                   'psych_comorbidity', 
                   'traumatic_event', 
                   'prime_trauma_event', 
@@ -47,12 +47,15 @@
                       'accident_injured_persons', 
                       'accident_rescue', 
                       'injury_sev_strata', 
+                      'hospitalization', 
+                      'surgery_done', 
                       'psych_support_post_accident', 
+                      'psych_support_need', 
                       'accident_aftermath', 
                       'same_sport_type_post_accident', 
-                      'caution_post_accident'), 
+                      'caution_post_accident', 
+                      'flashback_frequency'), 
          mental = c(## gneral measures of mental health and QoL
-                    'pss4_total', 
                     'gad7_total', 
                     'gad7_total_class', 
                     'phq9_total', 
@@ -63,13 +66,10 @@
                     'soc9l_total', 
                     'rs13_total', 
                     'rs13_total_class', 
-                    'cage_total_class', 
                     ## accident-specific measures
-                    #'unwilling_flashback', 
-                    'flashback_frequency', 
+                    'ptgi_total', 
                     'dsm5_total', 
-                    'dsm5_cluster_class', 
-                    'ptgi_total'))
+                    'dsm5_cluster_class'))
   
   ## splitting the main table
   
@@ -232,11 +232,7 @@
                    plot_subtitle = paste('n =', 
                                          cohort$ptsd_cluster$frequency$n_total[[1]])) + 
     scale_y_discrete(limits = rev(cohort$ptsd_cluster$variables), 
-                     labels = exchange(rev(cohort$ptsd_cluster$variables), 
-                                       dict = ptsd$var_lexicon, 
-                                       key = 'variable', 
-                                       value = 'label') %>% 
-                       stri_extract(regex = 'B|C|D|E'))
+                     labels = function(x) stri_extract(x, regex = 'B|C|D|E'))
   
 # END -----
   

### 2023-02-24 exploration scripts/consistency.R

@@ -29,6 +29,7 @@
                                        'ptgi_fctIV', 
                                        'ptgi_fctV'), 
                         rs13_total = paste0('rs13_q', 1:13), 
+                        brcs_total = paste0('brcs_q', 1:4), 
                         cage_total = paste0('cage_q', 1:4), 
                         pss4_total = paste0('pss4_q', 1:4), 
                         phq9_total = paste0('phqd_psych_satisf_q', 1:9), 
@@ -199,6 +200,7 @@
     c(dsm5_total = 1, 
       ptgi_total = 2, 
       rs13_total = 2, 
+      brcs_total = 1, 
       cage_total = 1, 
       pss4_total = 1, 
       phq9_total = 4, 

### 2023-02-24 exploration scripts/distribution.R

@@ -80,6 +80,37 @@
          subtitle = 'Mean to variance ratio. Poisson: mean = variance', 
          x = 'mean:variance ratio')
   
+# Gini indexes -------  
+  
+  insert_msg('Gini indexes')
+  
+  ## numeric values
+  
+  distr$gini_tbl <- 
+    ptsd$dataset[eda_globals$variables$variable] %>% 
+    map(as.numeric) %>% 
+    map(~.x[!is.na(.x)]) %>% 
+    map_dbl(Gini) %>% 
+    compress(names_to = 'variable', 
+             values_to = 'Gini')
+  
+  ## plot
+  
+  distr$gini_plot <- distr$gini_tbl %>% 
+    ggplot(aes(x = Gini,
+               y = reorder(variable, Gini))) + 
+    geom_bar(stat = 'identity', 
+             color = 'gray20', 
+             fill = 'cornsilk') + 
+    scale_y_discrete(labels = eda_globals$variables$variable %>% 
+                       exchange(ptsd$var_lexicon, 
+                                key = 'variable', 
+                                value = 'label')) + 
+    globals$common_theme + 
+    theme(axis.title.y = element_blank()) + 
+    labs(title = 'Variable variability', 
+         x = 'Gini index')
+  
 # END ------
   
   insert_tail()

### 2023-02-24 exploration scripts/partition.R

@@ -1,0 +1,82 @@
+# comparison of the training and test datasets
+
+  insert_head()
+
+# container ------
+
+  partition <- list()
+
+# parallel backend ------
+
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# descriptive stats -------
+  
+  insert_msg('Descriptive stats')
+
+  partition$desc_stats <- ptsd$dataset %>% 
+    explore(variables = eda_globals$variables$variable, 
+            split_factor = 'partition', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# testing -------
+  
+  insert_msg('Testing for differences by data partition')
+  
+  ## Mann-Whitney or Chi-squared test
+  ## multiple variables are not normally distributed
+  ##
+  ## FDR correction
+  
+  partition$test <- 
+    compare_variables(ptsd$dataset, 
+                      variables = eda_globals$variables$variable, 
+                      split_factor = 'partition', 
+                      what = 'eff_size', 
+                      types = eda_globals$variables$types, 
+                      ci = FALSE, 
+                      exact = FALSE, 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('ptsd', 'eda_globals'))) %>% 
+    format_fct_test
+  
+# Significant and near significant differences -----
+  
+  insert_msg('Significant and near significant differences')
+  
+  partition$top_factors <- partition$test %>% 
+    filter(p_value < 0.1) %>% 
+    .$variable
+  
+# A ready to use table with significant and near significant differences -------
+  
+  insert_msg('Near- and significant difference table')
+  
+  partition$result_tbl <- 
+    list(partition$desc_stats %>% 
+           reduce(left_join, by = 'variable') %>% 
+           set_names(c('variable', names(partition$desc_stats))), 
+         partition$test[c('variable', 'significance', 'eff_size')]) %>% 
+    map(filter, variable %in% partition$top_factors) %>% 
+    reduce(left_join, by = 'variable') %>% 
+    format_summ_tbl %>% 
+    set_names(c('Variable', 'Training', 'Test', 
+                'Significance', 'Effect size'))
+  
+# Plots ------
+  
+  insert_msg('Plots')
+  
+  excl$plots <- plot_fct(ptsd$dataset, 
+                         test_data = partition$test, 
+                         factor = 'partition', 
[…]

### 2023-02-24 exploration scripts/pca.R

@@ -1,0 +1,200 @@
+# Principal component analysis and testing of the clustering tendency
+# normalization and median centering
+
+  insert_head()
+  
+# container ------
+  
+  pca <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables: total scores for PTSD, PTG and EUROHIS QoL are skipped
+  ## cause the scales are per definition multi-factorial
+  
+  pca$variables <- ptsd$mental_variables
+  
+  pca$variables <- 
+    pca$variables[!pca$variables %in% c('dsm5_total', 
+                                        'ptgi_total', 
+                                        'eurohis_total')]
+  
+  pca$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, all_of(pca$variables)) %>% 
+    column_to_rownames('ID') %>% 
+    center_data(type = 'median')
+  
+  ## colors
+  
+  pca$comp_colors <- c('firebrick4', 
+                       'darkolivegreen4', 
+                       'steelblue3', 
+                       'plum4', 
+                       'cornflowerblue', 
+                       'bisque2', 
+                       'coral4', 
+                       'orangered2', 
+                       'firebrick2', 
+                       'cornsilk3')
+  
+# PCA -------
+  
+  insert_msg('PCA')
+  
+  pca$pca_obj <- pca$analysis_tbl %>% 
+    reduce_data(kdim = 10, red_fun = 'pca')
+  
+  ## variable renaming
+  
+  pca$pca_obj$loadings <- pca$pca_obj$loadings %>% 
+    mutate(var_r = variable) %>% 
+    exchange(variable = 'variable', 
+             dict = ptsd$var_lexicon, 
+             value = 'label') %>% 
+    mutate(variable = stri_replace(variable, 
+                                   regex = '\\s{1}score$', 
+                                   replacement = ''))
+  
+# PCA plots: scree and score, the first two dimensions -----
+  
+  insert_msg('PCA plots')
+  
+  pca$pca_plots[c('scree', 
+                  'scores')] <- c('scree', 
+                                  'scores') %>% 
+    map(~plot(pca$pca_obj, 
+              type = .x, 
+              cust_theme = globals$common_theme, 
+              label_points = FALSE)) %>%
+    map(~.x + labs(subtitle = '10-d PCA, median-centered psych scores'))
+  
+# Eigenvectors per component and in total -------
+  
[…]

### 2023-02-24 exploration scripts/power.R

@@ -1,0 +1,151 @@
+# Subset size for clustering with PAM/cosine
+
+  insert_head()
+  
+# container -------
+  
+  pow <- list()
+  
+# parallel backend ------
+  
+  plan('multisession')
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ## clustering variables
+  
+  pow$variables <- pow$mental_variables
+  
+  ## normalized, median-centered psychometric data
+  
+  pow$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, all_of(clust_globals$variable)) %>% 
+    column_to_rownames('ID') %>% 
+    center_data(type = 'median')
+
+  ## samples of the dataset of varying sizes
+  
+  pow$sizes <- c(50, 100, 150, 200, 250, 300)
+  
+  set.seed(1234)
+  
+  for(i in paste0('run_', 1:50)) {
+    
+    pow$samples[[i]] <- pow$sizes %>% 
+      map(~pow$analysis_tbl[sample(1: nrow(pow$analysis_tbl), 
+                                   size = .x, 
+                                   replace = FALSE), ]) %>% 
+      set_names(pow$sizes)
+    
+  }
+  
+  pow$samples <- pow$samples %>% 
+    unlist(recursive = FALSE)
+  
+# Clustering tendency -------
+  
+  insert_msg('Clustering tendencies')
+  
+  pow$clust_tendency <- pow$samples %>% 
+    future_map(~get_clust_tendency(.x, n = 0.5 * nrow(.x)), 
+               .options = furrr_options(seed = TRUE))
+  
+  ## extraction of the Hopkins stats
+  
+  pow$hopkins <- pow$clust_tendency %>% 
+    map_dbl(~.x$hopkins_stat) %>% 
+    compress(names_to = 'run', 
+             values_to = 'hopkins_stat') %>% 
+    mutate(size = stri_split_fixed(run, pattern = '.', simplify = TRUE)[, 2], 
+           size = as.numeric(size), 
+           run = stri_split_fixed(run, pattern = '.', simplify = TRUE)[, 1])
+  
+  ## plotting the Hopkins stats as a function of the dataset size
+  
+  pow$hopkins_plot <- pow$hopkins %>% 
+    ggplot(aes(x = factor(size), 
+               y = hopkins_stat)) + 
+    geom_boxplot(outlier.color = NA, 
+                 fill = 'coral3', 
+                 alpha = 0.3) + 
[…]

### 2023-02-24 exploration.R

@@ -1,6 +1,5 @@
 # Exploratory data analysis: 
 # cohort characteristic, numeric variable distribution, distribution plots
-# Export of the formatted data tables
 
 # tools ------
 
@@ -15,6 +14,8 @@
   library(trafo)
   library(psych)
   library(clustTools)
+  library(ggrepel)
+  library(DescTools)
 
   insert_head()
   
@@ -31,20 +32,8 @@
   eda_globals$variables <- ptsd$var_lexicon %>% 
     filter(type %in% c('characteristic', 'response')) %>% 
     mutate(type = ifelse(format == 'numeric', 'violin', 'stack'), 
-           y_lab = ifelse(format == 'numeric', 
-                          exchange(variable, 
-                                   dict = ptsd$var_lexicon, 
-                                   value = 'axis_lab'), 
-                          '% of strata'), 
-           plot_title = stri_capitalize(exchange(variable, 
-                                                 dict = ptsd$var_lexicon, 
-                                                 value = 'label')), 
-           plot_subtitle = exchange(variable, 
-                                    dict = ptsd$var_lexicon, 
-                                    value = 'description'), 
-           types = ifelse(type == 'violin', 'wilcoxon_r', 'cramer_v'), 
-           types2 = ifelse(type == 'violin', 'kruskal_etasq', 'cramer_v')) %>% 
-    select(variable, type, y_lab, plot_title, plot_subtitle, types, types2)
+           types = ifelse(type == 'violin', 'wilcoxon_r', 'cramer_v')) %>% 
+    select(variable, type, types)
   
   eda_globals$numeric_vars <- eda_globals$variables %>% 
     filter(type == 'violin') %>% 
@@ -54,7 +43,7 @@
     filter(type != 'violin') %>% 
     .$variable
   
-# launching the exploration and export scripts -----
+# launching the exploration scripts -----
   
   insert_msg('Launching the exploration scripts')
   
@@ -62,9 +51,26 @@
     './exploration scripts/cohort.R', 
     './exploration scripts/distribution.R', 
     './exploration scripts/consistency.R', 
-    './exploration scripts/bias.R') %>% 
+    './exploration scripts/bias.R', 
+    './exploration scripts/partition.R', 
+    './exploration scripts/pca.R') %>% 
     source_all(message = TRUE, 
                crash = TRUE)
+  
+  ## cached result of the power estimation analysis
+  
+  if(file.exists('./cache/power.RData')) {
+    
[…]

### 2023-02-24 factor scripts/afermath.R

@@ -1,0 +1,151 @@
+# Effects of somatic accident consequences
+
+  insert_head()
+
+# container -----
+
+  conseq <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  conseq$variables <- fct_globals$variables$variable
+  
+  conseq$variables <- conseq$variables[conseq$variables != 'accident_aftermath']
+  
+  ## analysis tables
+  
+  conseq$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, 
+           accident_aftermath, 
+           all_of(conseq$variables)) %>% 
+    filter(!is.na(accident_aftermath)) %>% 
+    mutate(accident_aftermath = car::recode(accident_aftermath, 
+                                            "'no' = 'no conseqences'; 
+                                            'yes' = 'conseqences'"), 
+           accident_aftermath = factor(accident_aftermath, 
+                                       c('no conseqences', 
+                                         'conseqences')))
+  
+  ## statistical test type
+  
+  conseq$test_type <- 
+    ptsd$dataset[conseq$variables] %>% 
+    map_lgl(is.numeric)
+  
+  conseq$test_type <- ifelse(conseq$test_type, 
+                             'wilcoxon_r', 'cramer_v')
+  
+# Descriptive statistic ------
+  
+  insert_msg('Descriptive statistics')
+  
+  conseq$desc_stats <- conseq$analysis_tbl %>% 
+    explore(variables = conseq$variables, 
+            split_factor = 'accident_aftermath', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for differences between the conseq modes ------
+  
+  insert_msg('Testing for differences between the conseq strata')
+  
+  conseq$test <- conseq$analysis_tbl %>%
+    compare_variables(variables = conseq$variables, 
+                      split_factor = 'accident_aftermath', 
+                      what = 'eff_size', 
+                      types = conseq$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('conseq'))) %>% 
+    format_fct_test
+  
+# Significant and near significant differences -------
+  
+  insert_msg('Significant differences')
+  
+  conseq$top_factors <- conseq$test %>% 
[…]

### 2023-02-24 factor scripts/age.R

@@ -55,19 +55,18 @@
                       types = age$test_type, 
                       exact = FALSE, 
                       ci = FALSE, 
-                      adj_method = 'BH', 
                       pub_styled = TRUE, 
                       .parallel = TRUE, 
                       .paropts = furrr_options(seed = TRUE, 
                                                globals = c('age'))) %>% 
     format_fct_test
   
-# Significant and near significant differences -------
+# Significant differences -------
   
-  insert_msg('Significant and near significant differences')
+  insert_msg('Significant differences')
   
   age$top_factors <- age$test %>% 
-    filter(p_adjusted < 0.1) %>% 
+    filter(p_value < 0.05) %>% 
     .$variable
 
 # Single variable plots ------
@@ -83,8 +82,6 @@
 # Specific plots: PTSD, PTG and QoL scores ------
   
   insert_msg('Plot panels')
-  
-  ## variables
   
   age$panels <- 
     plot_fct_panels(data = age$analysis_tbl, 
@@ -105,7 +102,7 @@
     reduce(left_join, by = 'variable') %>% 
     format_summ_tbl %>% 
     set_names(c('Variable', 'Young adult', 'Middle-aged', 
-                'Elderly', 'significance (FDR)', 'Effect size'))
+                'Elderly', 'significance', 'Effect size'))
   
 # END ------
   

### 2023-02-24 factor scripts/gender.R

@@ -46,7 +46,6 @@
                       types = gender$test_type, 
                       exact = FALSE, 
                       ci = FALSE, 
-                      adj_method = 'BH', 
                       pub_styled = TRUE, 
                       .parallel = TRUE, 
                       .paropts = furrr_options(seed = TRUE, 
@@ -58,7 +57,7 @@
   insert_msg('Significant and near significant differences')
   
   gender$top_factors <- gender$test %>% 
-    filter(p_adjusted < 0.1) %>% 
+    filter(p_value < 0.05) %>% 
     .$variable
   
 # Single variable plots ------
@@ -94,7 +93,7 @@
     reduce(left_join, by = 'variable') %>% 
     format_summ_tbl %>% 
     set_names(c('Variable', 'Female', 'Male', 
-                'Significance (FDR)', 'Effect size'))
+                'Significance', 'Effect size'))
   
 # END -----
   

### 2023-02-24 factor scripts/mental_illness.R

@@ -17,10 +17,22 @@
   mental$variables <- 
     mental$variables[mental$variables != 'psych_comorbidity']
   
+  ## analysis table
+  
+  mental$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, psych_comorbidity, 
+           all_of(mental$variables)) %>% 
+    mutate(psych_comorbidity = car::recode(as.character(psych_comorbidity), 
+                                           "'no' = 'no mental illness'; 
+                                           'yes' = 'mental illness'"), 
+           psych_comorbidity = factor(psych_comorbidity, 
+                                      c('no mental illness', 
+                                        'mental illness')))
+  
   ## statistical test type
   
   mental$test_type <- 
-    ptsd$dataset[mental$variables] %>% 
+    mental$analysis_tbl[mental$variables] %>% 
     map_lgl(is.numeric)
   
   mental$test_type <- ifelse(mental$test_type, 
@@ -30,7 +42,7 @@
   
   insert_msg('Descriptive statistics')
   
-  mental$desc_stats <- ptsd$dataset %>% 
+  mental$desc_stats <-  mental$analysis_tbl %>% 
     explore(variables = mental$variables, 
             split_factor = 'psych_comorbidity', 
             what = 'table', 
@@ -40,18 +52,17 @@
   
   insert_msg('Testing for differences between the mental illness strata')
   
-  mental$test <- ptsd$dataset %>%
+  mental$test <-  mental$analysis_tbl %>%
     compare_variables(variables = mental$variables, 
                       split_factor = 'psych_comorbidity', 
                       what = 'eff_size', 
                       types = mental$test_type, 
                       exact = FALSE, 
                       ci = FALSE, 
-                      adj_method = 'BH', 
                       pub_styled = TRUE, 
                       .parallel = TRUE, 
                       .paropts = furrr_options(seed = TRUE, 
-                                               globals = c('ptsd'))) %>% 
+                                               globals = c('mental'))) %>% 
     format_fct_test
   
 # Significant and near significant differences -------
@@ -59,7 +70,7 @@
   insert_msg('Significant and near significant differences')
   
   mental$top_factors <- mental$test %>% 
-    filter(p_adjusted < 0.1) %>% 
+    filter(p_value < 0.05) %>% 
     .$variable
   
 # Single variable plots ------
@@ -67,7 +78,7 @@
   insert_msg('Single variable plots')
[…]

### 2023-02-24 factor scripts/modeling.R

@@ -1,0 +1,280 @@
+# Modeling effects of the  interaction between the injury severity 
+# and psychological support after the accident on flashbacks 
+# (logistic regression), PTSD and PTG scoring (negative binomial GLM)
+
+  insert_head()
+  
+# container ------
+  
+  mod <- list()
+  
+# analysis globals ------
+  
+  ## analysis responses
+  
+  mod$responses <- 
+    c('dsm5_total', 
+      'dsm5_B', 
+      'dsm5_C', 
+      'dsm5_D', 
+      'dsm5_E', 
+      'ptgi_total', 
+      'ptgi_fctI', 
+      'ptgi_fctII', 
+      'ptgi_fctIII', 
+      'ptgi_fctIV', 
+      'ptgi_fctV', 
+      'unwilling_flashback')
+  
+  ## explanatory variables
+  
+  mod$variables <- 
+    c('injury_sev_strata', 
+      'psych_support_post_accident', 
+      'accident_rescue', 
+      'age', 
+      'sex', 
+      'psych_comorbidity', 
+      'accident_aftermath', 
+      'traumatic_event')
+  
+  ## formulas
+  
+  mod$formulas <- 
+    mod$responses %>% 
+    paste('~', paste(mod$variables, collapse = ' + ')) %>% 
+    map(as.formula) %>% 
+    set_names(mod$responses)
+  
+  ## modeling table
+  
+  mod$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, 
+           all_of(mod$responses), 
+           all_of(mod$variables)) %>% 
+    mutate(age = cut(age, 
+                     c(-Inf, 30, 65, Inf), 
+                     c('young', 'middle', 'elderly')), 
+           accident_rescue = car::recode(as.character(accident_rescue), 
+                                         "'self' = 'self/partner'; 
+                                         'tour partner' = 'self/partner'; 
+                                         'third party' = 'self/partner'; 
+                                         'rescue team' = 'professional'"), 
+           accident_rescue = factor(accident_rescue, 
+                                    c('self/partner', 'professional'))) %>% 
+    filter(complete.cases(.))
+  
+# Construction of the full models -------
+  
+  insert_msg('Construction of the full models')
+  
+  mod$full_models <- 
+    list(formula = mod$formulas, 
+         mod_fun = list(glm.nb, 
+                        glm.nb, 
+                        glm.nb, 
+                        glm.nb, 
+                        glm.nb, 
+                        glm.nb, 
[…]

### 2023-02-24 factor scripts/prior_trauma.R

@@ -1,0 +1,133 @@
+# Effects of prior traumatic events
+
+  insert_head()
+
+# container -----
+
+  trauma <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  trauma$variables <- fct_globals$variables$variable
+  
+  trauma$variables <- trauma$variables[trauma$variables != 'traumatic_event']
+  
+  ## analysis tables
+  
+  trauma$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, 
+           traumatic_event , 
+           all_of(trauma$variables)) %>% 
+    filter(!is.na(traumatic_event)) %>% 
+    mutate(traumatic_event  = car::recode(traumatic_event , 
+                                            "'no' = 'no trauma'; 
+                                            'yes' = 'trauma'"), 
+           traumatic_event  = factor(traumatic_event , 
+                                       c('no trauma', 
+                                         'trauma')))
+  
+  ## statistical test type
+  
+  trauma$test_type <- 
+    ptsd$dataset[trauma$variables] %>% 
+    map_lgl(is.numeric)
+  
+  trauma$test_type <- ifelse(trauma$test_type, 
+                             'wilcoxon_r', 'cramer_v')
+  
+# Descriptive statistic ------
+  
+  insert_msg('Descriptive statistics')
+  
+  trauma$desc_stats <- trauma$analysis_tbl %>% 
+    explore(variables = trauma$variables, 
+            split_factor = 'traumatic_event', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for differences between the trauma modes ------
+  
+  insert_msg('Testing for differences between the trauma strata')
+  
+  trauma$test <- trauma$analysis_tbl %>%
+    compare_variables(variables = trauma$variables, 
+                      split_factor = 'traumatic_event', 
+                      what = 'eff_size', 
+                      types = trauma$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('trauma'))) %>% 
+    format_fct_test
+  
+# Significant differences -------
+  
+  insert_msg('Significant differences')
+  
+  trauma$top_factors <- trauma$test %>% 
+    filter(p_value < 0.05) %>% 
+    .$variable
+  
+# Single variable plots ------
+  
[…]

### 2023-02-24 factor scripts/rescue.R

@@ -1,0 +1,149 @@
+# Effects of rescue (self/partner vs professional)
+
+  insert_head()
+
+# container -----
+
+  rescue <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  rescue$variables <- fct_globals$variables$variable
+  
+  rescue$variables <- rescue$variables[rescue$variables != 'accident_rescue']
+  
+  ## analysis tables
+  
+  rescue$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, 
+           accident_rescue, 
+           all_of(rescue$variables)) %>% 
+    filter(!is.na(accident_rescue))
+  
+  ## statistical test type
+  
+  rescue$test_type <- 
+    ptsd$dataset[rescue$variables] %>% 
+    map_lgl(is.numeric)
+  
+  rescue$test_type <- ifelse(rescue$test_type, 
+                             'wilcoxon_r', 'cramer_v')
+  
+# Descriptive statistic ------
+  
+  insert_msg('Descriptive statistics')
+  
+  rescue$desc_stats <- rescue$analysis_tbl %>% 
+    explore(variables = rescue$variables, 
+            split_factor = 'accident_rescue', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for differences between the rescue modes ------
+  
+  insert_msg('Testing for differences between the rescue strata')
+  
+  rescue$test <- rescue$analysis_tbl %>%
+    compare_variables(variables = rescue$variables, 
+                      split_factor = 'accident_rescue', 
+                      what = 'eff_size', 
+                      types = rescue$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('rescue'))) %>% 
+    format_fct_test
+  
+# Significant differences -------
+  
+  insert_msg('Significant differences')
+  
+  rescue$top_factors <- rescue$test %>% 
+    filter(p_adjusted < 0.05) %>% 
+    .$variable
+  
+# Single variable plots ------
+  
+  insert_msg('Single variable plots')
+  
+  rescue$plots <- 
+    plot_fct(data = rescue$analysis_tbl, 
+             test_data = rescue$test, 
+             factor = 'accident_rescue', 
[…]

### 2023-02-24 factor scripts/severity.R

@@ -1,0 +1,149 @@
+# Effects of injury severity (AIS strata)
+
+  insert_head()
+  
+# container ------
+  
+  sever <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  sever$variables <- fct_globals$variables$variable
+  
+  sever$variables <- sever$variables[sever$variables != 'injury_sev_strata']
+  
+  ## analysis table
+  
+  sever$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, injury_sev_strata, 
+           all_of(sever$variables)) %>% 
+    filter(!is.na(injury_sev_strata)) %>% 
+    mutate(injury_sev_strata = paste('AIS', as.character(injury_sev_strata)), 
+           injury_sev_strata = factor(injury_sev_strata))
+
+  ## statistical test type
+  
+  sever$test_type <- 
+    sever$analysis_tbl[sever$variables] %>% 
+    map_lgl(is.numeric)
+  
+  sever$test_type <- ifelse(sever$test_type, 
+                            'kruskal_eta', 'cramer_v')
+
+# Descriptive stats ------
+
+  insert_msg('Descriptive statistics')
+  
+  sever$desc_stats <- sever$analysis_tbl %>% 
+    explore(variables = sever$variables, 
+            split_factor = 'injury_sev_strata', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Serial testing for differences between the severity strata ------
+  
+  insert_msg('Testing for differences between the injury severity strata')
+  
+  sever$test <- sever$analysis_tbl %>%
+    compare_variables(variables = sever$variables, 
+                      split_factor = 'injury_sev_strata', 
+                      what = 'eff_size', 
+                      types = sever$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('sever'))) %>% 
+    format_fct_test
+  
+# Significant differences -------
+  
+  insert_msg('Significant differences')
+  
+  sever$top_factors <- sever$test %>% 
+    filter(p_value < 0.05) %>% 
+    .$variable
+  
+# Plots for single variables -------
+  
+  insert_msg('Plots for single variables')
+  
+  sever$plots <- 
[…]

### 2023-02-24 factor scripts/support.R

@@ -1,0 +1,148 @@
+# Effects of psychological support after the accident
+
+  insert_head()
+  
+# container -----
+  
+  support <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  support$variables <- fct_globals$variables$variable
+  
+  support$variables <- 
+    support$variables[support$variables != 'psych_support_post_accident']
+  
+  ## analysis table
+  
+  support$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, psych_support_post_accident, 
+           all_of(support$variables)) %>% 
+    filter(!is.na(psych_support_post_accident)) %>% 
+    mutate(psych_support_post_accident = car::recode(psych_support_post_accident, 
+                                                     "'no' = 'no support'; 
+                                                     'yes' = 'support'"))
+
+  ## statistical test type
+  
+  support$test_type <- 
+    ptsd$dataset[support$variables] %>% 
+    map_lgl(is.numeric)
+  
+  support$test_type <- ifelse(support$test_type, 
+                              'wilcoxon_r', 'cramer_v')
+  
+# Descriptive stats ------
+  
+  insert_msg('Descriptive statistics')
+  
+  support$desc_stats <- support$analysis_tbl %>% 
+    explore(variables = support$variables, 
+            split_factor = 'psych_support_post_accident', 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Serial testing for differences between the support strata ------
+  
+  insert_msg('Testing for differences between the psych. support strata')
+  
+  support$test <- support$analysis_tbl %>%
+    compare_variables(variables = support$variables, 
+                      split_factor = 'psych_support_post_accident', 
+                      what = 'eff_size', 
+                      types = support$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      pub_styled = TRUE, 
+                      .parallel = TRUE, 
+                      .paropts = furrr_options(seed = TRUE, 
+                                               globals = c('support'))) %>% 
+    format_fct_test
+  
+# Significant differences -------
+  
+  insert_msg('Significant differences')
+  
+  support$top_factors <- support$test %>% 
+    filter(p_value < 0.05) %>% 
+    .$variable
+  
+# Plots for single variables -------
+  
+  insert_msg('Plots for single variables')
+  
+  support$plots <- 
+    plot_fct(data = support$analysis_tbl, 
[…]

### 2023-02-24 factor.R

@@ -1,5 +1,8 @@
-# Association of age, gender, mental illness, injury severity 
+# Association of age, gender, mental illness, rescue mode, injury severity 
 # and psychological support with accident circumstances and mental health
+#
+# This factors were described to be associated with e.g. PTSD in literature
+# Yet, none of them is a strong marker
 
 # tools -----
 
@@ -13,8 +16,16 @@
   library(rstatix)
   library(trafo)
   library(clustTools)
+  library(ggtext)
+  library(MASS)
+  library(lmqc)
+  library(doParallel)
+  library(caret)
+  library(caretExtra)
   
   insert_head()
+  
+  select <- dplyr::select
   
   source_all('./tools/tools.R', 
              message = TRUE, 
@@ -35,8 +46,13 @@
   
   c('./factor scripts/age.R', 
     './factor scripts/gender.R', 
-    './factor scripts/mental_illness.R') %>% 
-    source_all(message = TRUE, encoding = TRUE)
+    './factor scripts/mental_illness.R', 
+    './factor scripts/prior_trauma.R', 
+    './factor scripts/rescue.R',
+    './factor scripts/severity.R', 
+    './factor scripts/support.R', 
+    './factor scripts/modeling.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
   
 # END ------
   

### 2023-02-24 import scripts/randomization.R

@@ -1,0 +1,111 @@
+# Finding the optimal split of the cohort's dataset 
+# into a training and a test subset used in modeling 
+# and multi-parameter classification
+
+  insert_head()
+  
+# container -----
+  
+  rand <- list()
+  
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables to be kept possibly indifferent between the partitions
+  
+  rand$variables <- c('accident_year', 
+                      'age', 
+                      'sex', 
+                      'somatic_comorbidity', 
+                      'psych_comorbidity', 
+                      'traumatic_event', 
+                      'injury_sev_strata')
+  
+  ## analysis table
+  
+  rand$analysis_tbl <- ptsd$dataset %>% 
+    select(ID, all_of(rand$variables)) %>% 
+    mutate(obs = 1:nrow(.))
+  
+# creating data partitions ------
+  
+  insert_msg('Creating data partitions')
+  
+  set.seed(1234)
+  
+  rand$part_obs <- 
+    createDataPartition(y = rand$analysis_tbl$accident_year, 
+                        p = 1/3, 
+                        times = 1000)
+  
+  rand$part_ids <- set_names(names(rand$part_obs), 
+                             names(rand$part_obs))
+  
+  for(i in rand$part_ids) {
+    
+    rand$analysis_tbl <- rand$analysis_tbl %>% 
+      mutate(!!i := ifelse(obs %in% rand$part_obs[[i]], 
+                           'test', 'training'), 
+             !!i := factor(.data[[i]], c('training', 'test')))
+    
+  }
+  
+# comparing the partitions, categorical variables -------
+  
+  insert_msg('Comparing the partitions, categorical vars')
+  
+  rand$comparison[rand$variables[rand$variables != 'age']] <- 
+    rand$variables[rand$variables != 'age'] %>% 
+    map(freq_tester, 
+        data = rand$analysis_tbl, 
+        part_variables = rand$part_ids)
+
+# age ------
+  
+  insert_msg('Comparing the partitions: age')
+  
+  ## descriptive stats: medians
+  
+  rand$comparison$age$median <- rand$part_ids %>% 
+    future_map(~group_by(rand$analysis_tbl, .data[[.x]]) %>% 
[…]

### 2023-02-24 import.R

@@ -10,6 +10,9 @@
   library(tidyverse)
   library(rlang)
   library(trafo)
+  library(caret)
+  library(rstatix)
+  library(furrr)
 
   source_all('./tools/tools.R', 
              message = TRUE, crash = TRUE)
@@ -83,6 +86,11 @@
                            'Lower limbs', 
                            'Other external') %>% 
     set_names(globals$injury_vars)
+  
+  ## data partition colors
+  
+  globals$part_colors <- c(training = 'steelblue', 
+                           test = 'coral3')
 
 # reading the recoding scheme -----
   
@@ -140,16 +148,46 @@
   insert_msg('Removal of the zero-age and child age records')
   
   ptsd$cleared <- ptsd$cleared %>% 
-    filter(age >= 16)
+    filter(is.na(age) | age >= 16)
   
 # manual polishing -----
   
   insert_msg('Manual polishing')
   
   ptsd$cleared <- ptsd$cleared %>% 
-    mutate(hosp_date = pss2date(hosp_date), 
+    mutate(age_class = cut(age, 
+                           c(-Inf, 30, 65, Inf), 
+                           c('young', 'middle', 'elderly')), 
+           ward_date = pss2date(ward_date), 
+           hospitalization = ifelse(hospitalization == 'X', 'yes', 'no'), 
+           hospitalization = factor(hospitalization, c('no', 'yes')), 
            survey_date = pss2date(survey_date), 
-           obs_time = floor(as.numeric(survey_date - hosp_date)), 
+           obs_time = floor(as.numeric(survey_date - ward_date)), 
+           accident_month = stri_split_fixed(accident_date, 
+                                             pattern = '-', 
+                                             simplify = TRUE)[, 2], 
+           accident_month = as.numeric(accident_month), 
+           accident_season = ifelse(accident_month %in% c(12, 1, 2), 
+                                    'winter', 
+                                    ifelse(accident_month %in% c(3, 4, 5), 
+                                           'spring', 
+                                           ifelse(accident_month %in% c(6, 7, 8), 
+                                                  'summer', 
+                                                  'fall'))),
+           accident_season = ifelse(is.na(accident_season), 
+                                    'no information', accident_season), 
+           accident_season = factor(accident_season, 
+                                    c('spring', 'summer', 'fall', 
+                                      'winter', 'no information')), 
+           accident_hour = stri_extract(accident_time, 
+                                        regex = '^\\d{2}'), 
+           accident_hour = as.numeric(accident_hour), 
[…]

### 2023-02-24 paper scripts/figures.R

@@ -17,8 +17,430 @@
               ref_name = 'consort', 
               caption = paste('Flow diagram of the analysis', 
                               'inclusion process.'), 
-              w = 90, 
-              h = 2166/2106 * 90)
+              w = 120, 
+              h = 3846/2916 * 120)
+  
+# Figure 2: psychometry scores in the clusters -------
+  
+  insert_msg('Figure 2: psychometry scoring in the clusters')
+  
+  ## upper panel: stress, anxiety, depression, somatization, panic
+  ## resilience and sense-of coherence
+  
+  figures$clust_psych$upper <- feat_clust$ribbon_plots
+  
+  for(i in names(figures$clust_psych$upper)) {
+    
+    figures$clust_psych$upper[[i]]$data <- 
+      figures$clust_psych$upper[[i]]$data %>% 
+      filter(variable %in% c('gad7_total', 
+                             'phq9_total', 
+                             'phq_events_total', 
+                             'phqd_panic_total', 
+                             'rs13_total', 
+                             'soc9l_total'))
+    
+  }
+  
+  figures$clust_psych$upper <- figures$clust_psych$upper %>% 
+    map2(., 
+         paste0('Major mental health readouts, ', 
+                c('training', 'test')), 
+         ~.x + 
+           scale_y_discrete(limits = rev(c('gad7_total', 
+                                           'phq9_total', 
+                                           'phq_events_total', 
+                                           'phqd_panic_total', 
+                                           'soc9l_total', 
+                                           'rs13_total')), 
+                            labels = psych_labeller) + 
+           labs(title = .y))
+  
+  ## second panel
+  
+  figures$clust_psych$qol <- feat_clust$ribbon_plots
+  
+  for(i in names(figures$clust_psych$qol)) {
+    
+    figures$clust_psych$qol[[i]]$data <- 
+      figures$clust_psych$qol[[i]]$data %>% 
+      filter(stri_detect(variable, fixed = 'eurohis'))
+    
+  }
+  
+  figures$clust_psych$qol <- figures$clust_psych$qol %>% 
+    map2(., 
+         paste0('Dimished QoL, EUROHIS-QOL 8, ', 
+                c('training', 'test')), 
+         ~.x + 
+           scale_y_discrete(limits = rev(c('eurohis_total', 
+                                           'eurohis_qol', 
+                                           'eurohis_selfesteem', 
+                                           'eurohis_energy', 
+                                           'eurohis_activity', 
+                                           'eurohis_health', 
+                                           'eurohis_relationship', 
[…]

### 2023-02-24 paper scripts/render.R

@@ -14,6 +14,19 @@
 # paper -----
   
   insert_msg('Rendering the paper')
+  
+  my_word <- function(...) {
+    
+    form <- word_document2(number_sections = FALSE, 
+                           reference_docx = 'ms_template.docx')
+    
+    form$pandoc$lua_filters <- c(form$pandoc$lua_filters, 
+                                 'scholarly-metadata.lua', 
+                                 'author-info-blocks.lua')
+    
+    form
+    
+  }
 
   render('./paper/markdown/manuscript.Rmd', 
          output_format = my_word(), 

### 2023-02-24 paper scripts/supplementary.R

@@ -11,18 +11,21 @@
   insert_msg('Figure S1: injured parts and mental health, cohort')
   
   suppl_figures$parts_mental_cohort <- 
-    list(cohort$injury$plot, 
+    list(cohort$injury$plot + 
+           expand_limits(x = 50), 
          cohort$mental_details$plots$eurohis, 
+         cohort$mental_details$plots$ptgi, 
+         ggdraw(), 
          cohort$mental_details$plots$dsm, 
-         cohort$ptsd_cluster$plot, 
-         cohort$mental_details$plots$ptgi) %>% 
+         cohort$ptsd_cluster$plot + 
+           expand_limits(x = 13)) %>% 
     plot_grid(plotlist = ., 
               ncol = 2, 
               align = 'hv', 
               axis = 'tblr', 
               labels = c('A', 'B', 'C', '', 'D'), 
               label_size = 10, 
-              rel_heights = c(1.4, 1, 1.1)) %>% 
+              rel_heights = c(1.4, 1.1, 1)) %>% 
     as_figure(label = 'figure_s1_parts_mental_cohort', 
               ref_name = 'parts_mental_cohort', 
               caption = paste('Injured body regions and detailed scoring', 
@@ -32,164 +35,231 @@
               w = 180, 
               h = 210)
   
-# Figure S2: mental features associated with participant's age ------
+# Figure S2: cluster development ------
   
-  insert_msg('Figure S2: age and mental health')
+  insert_msg('Figure S2: cluster development')
   
-  ## upper panel: panic, resilience and sense of coherence
+  ## upper panel: variance and cross-validation
   
-  suppl_figures$age_mental$upper <- age$plots[c("phqd_panic_total", 
-                                                "rs13_total", 
-                                                "soc9l_total")] %>% 
-    map(~.x + theme(legend.position = 'none')) %>% 
+  suppl_figures$clust_dev$upper <- 
+    plot_grid(clust_devel$result_plot, 
+              ncol = 2, 
+              rel_widths = c(0.8, 0.2))
+  
+  ## bottom panel: WSS and silhouette
+  
+  suppl_figures$clust_dev$bottom <- semi_clust$diagn_plots %>% 
+    map(~.x + 
+          labs(subtitle = paste0(.x$labels$subtitle, 
+                                ', n = ', 
+                                nrow(clust_globals$analysis_tbl$training))) + 
+          theme(plot.tag = element_blank())) %>% 
[…]

### 2023-02-24 paper scripts/tables.R

@@ -51,13 +51,36 @@
   insert_msg('Table S1: variables used in the analysis pipeline')
   
   suppl_tables$analysis_vars <- ptsd$var_lexicon %>% 
-    filter(variable %in% eda_globals$variables$variable) %>% 
-    transmute(Section = section, 
+    filter(type %in% c('characteristic', 'response')) %>% 
+    transmute(`Variable name in R` = variable, 
               Variable = label, 
-              Description = description) %>% 
+              Description = description,
+              Format = format,  
+              Unit = unit)
+  
+  ## finding the possible factor levels
+  
+  fct_levs <- suppl_tables$analysis_vars %>% 
+    filter(Format == 'factor') %>% 
+    .$`Variable name in R`
+  
+  fct_levs <- ptsd$dataset[fct_levs] %>% 
+    map(levels) %>% 
+    compress(names_to = 'Variable name in R', 
+             values_to = 'Categories') %>% 
+    mutate(Categories = map_chr(Categories, paste, collapse = ', '))
+    
+  ## the entire table
+  
+  suppl_tables$analysis_vars <- 
+    left_join(suppl_tables$analysis_vars, 
+              fct_levs, 
+              by = 'Variable name in R') %>% 
     mdtable(label = 'table_s1_analysis_variables', 
             ref_name = 'analysis_vars', 
             caption = paste('Variables used in the analysis pipeline.'))
+  
+  rm(fct_levs)
   
 # Supplementary Table S2: mental health assessment battery -----
   
@@ -72,15 +95,42 @@
             ref_name = 'mental_battery', 
             caption = paste('Mental health assessment battery.'))
   
-# Supplementary Table S3: differences between included and excluded participants -------
+# Supplementary Table S3: mental scale consistency -------
   
-  insert_msg('Table S3: included versus excluded')
+  insert_msg('Table S3: mental scale consistency')
+  
+  suppl_tables$consistency <- 
+    left_join(compress(cons$pca_factor_n, 
+                       names_to = 'scale', 
+                       values_to = 'n'), 
+              cons$omega_tbl, 
+              by = 'scale') %>% 
+    filter(scale != 'cage_total') %>% 
+    arrange(-omega) %>% 
+    mutate(scale = ifelse(scale != 'brcs_total', 
+                          exchange(scale, 
+                                   dict = ptsd$var_lexicon, 
+                                   key = 'variable', 
+                                   value = 'label'), 
+                          'BRCS'), 
[…]

### 2023-02-24 scratch pad.R

@@ -1,0 +1,52 @@
+# parallel backend ------
+
+insert_msg('Parallel backend')
+
+registerDoParallel(cores = 7)
+
+# tuning of the classifier ------
+
+insert_msg('Tuning')
+
+set.seed(1234)
+
+tune_model <- 
+  train(form = as.formula(paste('clust_id ~ ', 
+                                paste(class_globals$variables, 
+                                      collapse = '+'))), 
+        data = class_globals$analysis_tbl$training, 
+        method = 'C5.0', 
+        metric = 'Kappa', 
+        trControl = trainControl(method = 'cv', 
+                                 number = 10, 
+                                 savePredictions = 'final', 
+                                 returnData = TRUE, 
+                                 returnResamp = 'final', 
+                                 classProbs = TRUE))
+
+
+
+tune_model %>% 
+  as_caretx %>% 
+  predict(newdata = class_globals$analysis_tbl$test) %>% 
+  map(plot, 'confusion')
+
+
+varImp(tune_model)
+
+library(caretEnsemble)
+
+test_ensemble <- 
+  caretList(form = as.formula(paste('clust_id ~ ', 
+                                    paste(class_globals$variables, 
+                                          collapse = '+'))), 
+            data = class_globals$analysis_tbl$training, 
+            methodList=c("ranger", "C5.0"), 
+            trControl = trainControl(method = 'cv', 
+                                     number = 10, 
+                                     savePredictions = 'final', 
+                                     returnData = TRUE, 
+                                     returnResamp = 'final', 
+                                     classProbs = TRUE))
+
+stopImplicitCluster()

### 2023-02-24 tools/tools.R

@@ -172,6 +172,43 @@
   
   pss2date <- function(x) as.Date(x/86400, origin = "1582-10-14")
   
+# randomization ------
+  
+  freq_tester <- function(data, 
+                          variable, 
+                          part_variables) {
+    
+    ## computes fractions and tests for differences in distribution
+    ## between data partitions
+    
+    ## data splits
+    
+    splits <- part_variables %>% 
+      future_map(~split(data, data[[.x]]) %>% 
+                   map(~table(.x[[variable]])))
+    
+    ## fractions
+    
+    fractions <- splits %>% 
+      map(~map(.x, 
+               ~.x/sum(.x)) %>% 
+            reduce(rbind) %>% 
+            `row.names<-`(., c('training', 'test')))
+    
+    ## chi-squared tests
+    
+    tests <- splits %>% 
+      map(reduce, rbind) %>% 
+      map(`row.names<-`, c('training', 'test')) %>% 
+      map(chisq_test) %>% 
+      compress(names_to = 'partition')
+    
+    list(fractions = fractions, 
+         tests = tests)
+    
+  }
+  
+  
 # plot axis labels with n numbers ------
   
   label_n <- function(data, split_factor, sep = '\nn = ') {
@@ -226,6 +263,7 @@
                              x_lab = '% of cohort', 
                              bar_color = 'steelblue', 
                              fill_scale = scale_fill_brewer(), 
+                             color_scale = scale_color_brewer(), 
                              show_freqs = TRUE, ...) {
     
     ## plots a simple bar plot of frequencies
@@ -261,12 +299,154 @@
     
     if(show_freqs) {
       
-      freq_plot <- freq_plot + 
-        geom_text(aes(label = signif(.data[[freq_var]], 2)), ...)
+      if(!is.null(split_factor)) {
+        
+        freq_plot <- freq_plot + 
+          geom_text(aes(label = signif(.data[[freq_var]], 2), 
+                        color = .data[[split_factor]]), ...) + 
+          color_scale
+        
+      } else {
+        
+        freq_plot <- freq_plot + 
+          geom_text(aes(label = signif(.data[[freq_var]], 2)), ...)
+        
+      }
       
     }
     
     return(freq_plot)
+    
+  }
+  
+  plot_ptsd_freq <- function(data, 
+                             test_data, 
+                             split_factor = 'psych_comorbidity', 
+                             plot_title = 'PCL-5 DSM-5 clusters', 
+                             plot_subtitle = NULL, 
[…]

### 2023-02-27 paper/markdown/manuscript.Rmd
Feedback changes
Changes in text as proposed by Katharina

@@
       institute: [psych]
-      email: Katharina.Huefner@i-med.ac.at
+      email: Katharina.Huefner@tirol-kliniken.at
       correspondence: true
@@
 
-keywords: "sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth"
+keywords: "sport accident, mental health, post-traumatic stress disorder, post-traumatic growth"
 
@@
 
-__Methods:__ A cross-sectional survey study was conducted with 307 individuals treated at the Innsbruck University Hospital (Austria) following a non-professional sport accident. 
-Measures of anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic syndrome (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data.
-Patterns of mental readouts were defined by semi-supervised clustering and factors affecting the cluster assignment were identified by statistical testing and conditional random forest machine learning.
+__Methods:__ A cross-sectional survey study was conducted with `r nrow(ptsd$dataset)` individuals treated at the Innsbruck University Hospital (Austria) following a non-professional sport accident. 
+Measures of anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data.
+Patterns of mental readouts were defined by semi-supervised clustering and factors affecting the cluster assignment were identified by conditional random forest machine learning.
 
-__Results:__ Participants were assigned to three mental clusters characterized by a neutral mental response, post-traumatic growth and symptoms of post-traumatic syndrome disorder, respectively. 
-Overall differences in demographic, socioeconomic, injury- and accident-related variables between the clusters were low-to-moderate. 
-The post-traumatic growth and post-traumatic syndrome clusters were characterized by relatively frequent hospitalization, surgery, somatic illness and female gender. 
-The most predictive factors for the post-traumatic syndrome cluster were more cautious behavior and flashbacks during sport activity, low income, mental disorders, somatic consequences of the accident and subjective need for psychological support.
+__Results:__ Participants were assigned to three mental clusters characterized by a neutral mental response, post-traumatic growth and post-traumatic stress symptoms, respectively. 
+Overall differences in demographic, socioeconomic, injury- and accident-related variables between the clusters were weak-to-moderate. 
+The post-traumatic growth and post-traumatic stress clusters were characterized by relatively frequent hospitalization, surgery, pre-existing somatic illness and female gender. 
+The key characteristics the post-traumatic stress cluster were more caution and flashbacks during sport activity, low income, pre-existing mental disorders, somatic consequences of the accident and subjective need for psychological support.
 
-__Conclusion:__ A subset of non-professional sport accident victims develops signs of mental health problems including symptoms of post-traumatic syndrome disorder. 
-Patients with low socioeconomic status, somatic and mental conditions, and hospitalization represent a vulnerable group which may benefit most from a psychological intervention.
+__Conclusion:__ A subset of non-professional sport accident victims develops signs of mental health problems including symptoms of post-traumatic stress disorder. 
+Patients with low socioeconomic status, pre-existing somatic and mental conditions, and hospitalization represent a vulnerable group which may benefit most from a psychological intervention.
 
@@
 
-sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth
+sport accident, mental health, post-traumatic stress disorder, post-traumatic growth
 
@@
 post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
-post-traumatic syndrome disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
+post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
 The tools displayed good-to-excellent consistency as measured by McDonald's $\omega$ > 0.8 [@McDonald1999] (__Supplementary Table S3__).
 
-Symptoms of anxiety were defined as GAD-7 $geq$ 10, symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], significant somatization/persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002].
+Symptoms of anxiety were defined as GAD-7 $\geq$ 10, symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], significant somatization/persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002].
 Resilient coping classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].
@@
 The cluster number was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic [@Kassambara2020]. 
-Three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) cluster. 
+Three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder) cluster. 
 Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
@@
 Their injuries were on average less severe as measured by AIS [@Gennarelli2006] with more often upper limb injuries, and lower hospitalization and surgery rates. 
-The excluded individuals tended suffered less often from somatic health consequences of the accident and tended towards lower ratings of post-traumatic syndrome disorder but also diminished post-traumatic growth and less sense of coherence.
+The excluded individuals tended suffered less often from somatic health consequences of the accident and tended towards lower ratings of post-traumatic stress disorder but also diminished post-traumatic growth and less sense of coherence.
 These differences were significant or near-significant with weak effects sizes (V $\leq$ 0.22 or r $\leq$ 0.15). 
@@
 `r ptsd$dataset$accident_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')` and were managed at the study site between `r ptsd$dataset$ward_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')`.
-The median time between the ward management and survey completion was 
+The median time between the emergency unit or ward management and survey completion was 
 `r median(ptsd$dataset$obs_time)` days (interquartile range: 
@@
 somatization (`r signif(table(ptsd$dataset$phq_events_total_class)/sum(table(ptsd$dataset$phq_events_total_class)), 2)[2] * 100`%) were rare in the study cohort. 
-The median loss of quality of life was scored with `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS QOL points, which was substantially more than in a generalized Western European population [@Schmidt2006].
+The median loss of quality of life was scored with `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS QOL points. 
+The overall quality of life was hence substantially worse than in a generalized Western European population [@Schmidt2006]. 
 The participants were most satisfied with their overall quality of life and housing (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1B__). 
@@
 The scores of overall post-traumatic growth (median: `r median(ptsd$dataset$ptgi_total) %>% signif(2)`) and its specific domains were lower in the study cohort as compared with the figures originally reported for the PTGI tool [@Tedeschi1996] (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1C__).
-In `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants symptoms of post-traumatic syndrome disorder defined as positivity for at least one PCL-5 DSM-5 tool domain [@Bovin2016] could be observed.
+In `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants symptoms of post-traumatic stress disorder defined as positivity for at least one PCL-5 DSM-5 tool domain [@Bovin2016] could be observed.
 Among them, domain B symptoms (`r signif(table(ptsd$dataset$dsm5_B_class)/sum(table(ptsd$dataset$dsm5_B_class)), 2)[2] * 100`%) were the most frequent and domain D symptoms were the rarest (`r signif(table(ptsd$dataset$dsm5_D_class)/sum(table(ptsd$dataset$dsm5_D_class)), 2)[2] * 100`%) (__Table \@ref(tab:tab-cohort-mental)__ and __Supplementary Figure S1D__).
@@
 
-To comprehend types of mental response to the sport accident, we subjected the study participants to semi-supervised clustering in respect to measures of anxiety, depression, panic, persistent somatic symptoms, resilience, sense of coherence, quality of life, post-traumatic growth and post-traumatic syndrome disorder.
-Among several clustering algorithms compared in the training subset, partition around medoids with cosine distance between observations [@Drost2018; @Schuber2019] demonstrated good explanatory power and superior reproducibility in cross-validation [@Lange2004] (__Supplementary Figure S2A__).
-Three clusters of participants named neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) cluster after the patterns of their characteristic mental features were identified in the training subset.
+To comprehend types of mental response to the sport accident, we subjected the study participants to semi-supervised clustering in respect to measures of anxiety, depression, panic, persistent somatic symptoms, resilience, sense of coherence, quality of life, post-traumatic growth and post-traumatic stress disorder.
+Among several clustering algorithms compared in the training subset, partition around medoids with cosine distance between observations [@Drost2018; @Schubert2019] demonstrated good explanatory power and superior reproducibility in cross-validation [@Lange2004] (__Supplementary Figure S2A__).
+Three clusters of participants named neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder) cluster after the patterns of their characteristic mental features were identified in the training subset.
 Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance in both study cohort subsets (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`) and comparable cluster sizes (__Supplementary Figures S1B__ - __S4__).
@@
 Roughly one-third study participants assigned to the neutral clusters were characterized by low scores of anxiety, depression, panic and somatic symptoms along with high rating of coherence, resilient coping and quality of life.
-Scores of post-traumatic growth and post-traumatic syndrome were the lowest in the neutral cluster. 
+Scores of post-traumatic growth and post-traumatic stresswere the lowest in the neutral cluster. 
 The PTG cluster demonstrated similarly low score of major mental health problems as neutral cluster individuals. 
 By contrast, the overall measures of post-traumatic growth and its domains were the highest in the PTG cluster. 
-In addition, ratings of post-traumatic syndrome disorder symptoms were slightly elevated in the PTG as compared with the neutral cluster. 
+In addition, ratings of post-traumatic stress disorder symptoms were slightly elevated in the PTG as compared with the neutral cluster. 
 The remaining participants subsumed in the PTSD cluster displayed the most intense signs of anxiety, depression, panic and somatic symptoms as well as poor sense of coherence, resilient coping and low quality of life as compared with the neutral and PTG cluster.
-PTSD cluster was hallmarked by the highest scores of post-traumatic syndrome disorder symptoms in every PCL-5 DSM-5 domain. 
+PTSD cluster was hallmarked by the highest scores of post-traumatic stress disorder symptoms in every PCL-5 DSM-5 domain. 
 Post-traumatic growth was also present in the PTSD cluster as evident from elevated PTGI scores (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Figure S4__, __Supplementary Table S6__).
@@
 Furthermore, frequencies of low and moderate resilient coping classes were by far the highest in the PTSD cluster. 
-Finally, approximately 40% of PTSD cluster individuals were screened positive for at least one post-traumatic syndrome disorder symptom which was more than in the PTG 
+Finally, approximately 40% of PTSD cluster individuals were screened positive for at least one post-traumatic stress disorder symptom which was more than in the PTG 
 (training: `r signif(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)), 2)[2] * 100`%,
@@
 test: `r signif(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)), 2)[2] * 100`%).
-An analogical trend was observed for the remaining post-traumatic syndrome disorder domains (__Figure \@ref(fig:fig-clust-symptoms)__).
+An analogical trend was observed for the remaining post-traumatic stress disorder domains (__Figure \@ref(fig:fig-clust-symptoms)__).
 
@@
 
-In individuals surveyed on average 3.5 years after the non-professional sport accident, some of its consequences like somatic health problems, unwilling flashbacks as well as phenomena of post-traumatic growth and symptoms of post-traumatic syndrome disorder were still tangible.
-By semi-supervised clustering we could discern three distinct patterns of mental health in this collective with each present in approximately one-third of participants: (1) a neutral mental response reaction, (2) salutatory reaction to the accident characterized by post-traumatic growth and (3) mental health deterioration hallmarked by symptoms of anxiety, depression, somatization, resilience loss and signs of post-traumatic syndrome disorder.
+In individuals surveyed on average 3.5 years after the non-professional sport accident, some of its consequences like somatic health problems, unwilling flashbacks as well as phenomena of post-traumatic growth and symptoms of post-traumatic stress disorder were still tangible.
+By semi-supervised clustering we could discern three distinct patterns of mental health in this collective with each present in approximately one-third of participants: (1) a neutral mental response reaction, (2) salutatory reaction to the accident characterized by post-traumatic growth and (3) mental health deterioration hallmarked by symptoms of anxiety, depression, somatization, resilience loss and signs of post-traumatic stress disorder.
 Notably, separation of the PTG and PTSD clusters was in part blurry. 
-Symptoms of post-traumatic syndrome were observed also in PTG cluster participants albeit less frequently than in the PTSD cluster. 
+Symptoms of post-traumatic stresswere observed also in PTG cluster participants albeit less frequently than in the PTSD cluster. 
 Conversely, mental health problems in PTSD cluster participants were paralleled by some signs of post-traumatic growth. 
@@
 Our study bears limitations. 
-First, although the key phenomena phenomena in the newly identified mental clusters, post-traumatic growth and post-traumatic syndrome disorder, are clearly associated with trauma [@Tedeschi1996; @Bovin2016], we could not exclude the mental cluster assignment to be affected by other events than the sport accident.
+First, although the key phenomena phenomena in the newly identified mental clusters, post-traumatic growth and post-traumatic stress disorder, are clearly associated with trauma [@Tedeschi1996; @Bovin2016], we could not exclude the mental cluster assignment to be affected by other events than the sport accident.
 To account for that, follow-up studies including a control group are needed. 
@@
   footnote(1, 1, 
-           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder'),
            part = 'header', 
@@
 Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
-Three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder), All clustering parameters differed significantly between the clusters as assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder), All clustering parameters differed significantly between the clusters as assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
 Mean normalized scores of major mental health readouts (A), diminished quality of life (B), post-traumatic growth (C) and post-traumatic symptom disorder (D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines.
@@
 Numbers of observations in the clusters are displayed in the plot captions.
-GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder._
+GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
 
@@
 
-_Frequencies of symptoms of anxiety (A, GAD-7 ≥10), depression (B, PHQ-9 ≥ 11) and somatization (C, PHQ-15 ≥ 11), distribution of resilient coping classes (D, RS13) and symptoms of post-traumatic syndrome disorder (E, PTSD, PCL-5 DSM-5, positivity for PTSD domains and frequency of participants positive for at least one PTSD domain [PTSD+]) in the mental clusters.
+_Frequencies of symptoms of anxiety (A, GAD-7 ≥10), depression (B, PHQ-9 ≥ 11) and somatization (C, PHQ-15 ≥ 11), distribution of resilient coping classes (D, RS13) and symptoms of post-traumatic stress disorder (E, PTSD, PCL-5 DSM-5, positivity for PTSD domains and frequency of participants positive for at least one PTSD domain [PTSD+]) in the mental clusters.
 Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
@@
 Significant effects in (E) are highlighted in bold.
-GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTSD: post-traumatic syndrome disorder._
+GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder._
 
@@
 
-_Distribution of gender (A), annual household income (B), and frequencies of self-reported somatic (C) and mental illness (D) in the mental clusters.
+_Distribution of gender (A), annual household income (B), and frequencies of self-reported pre-existing somatic conditions (C) and self-reported pre-existing mental illness (D) in the mental clusters.
 Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
@@
 
-_Distribution of injury severity grades (A, AIS: abbreviate injury scale), hospitalization (B) and surgery rates (C), and injuries of the body parts (D) in the mental clusters.
+_Distribution of injury severity grades (A, AIS: abbreviated injury scale), hospitalization (B) and surgery rates (C), and injuries of the body parts (D) in the mental clusters.
 Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.

### 2023-02-27 paper/markdown/supplementary_material.Rmd
Feedback changes
Changes in text as proposed by Katharina

@@
 post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
-post-traumatic syndrome disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016].
+post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016].
 
@@
 To this end, `r rand$part_obs %>% length` random splits of the datasets were generated. 
-The subset assignment scheme was chosen with possibly small differences in accident year, age, gender, somatic and mental illness rates, frequencies of prior traumatic events (DIA-X) [@Maercker2005] and distribution of injury severity grades (AIS: abbreviated injury scale) [@Gennarelli2006] as investigated by $\chi^2$ and Mann-Whitney test, as appropriate (functions `chisq_test()` and `wilcox_test()`, package _rstatix_).
+The subset assignment scheme was chosen with possibly small differences in accident year, age, gender, pre-existing somatic and pre-existing mental illness rates, frequencies of prior traumatic events (DIA-X) [@Maercker2005] and distribution of injury severity grades (AIS: abbreviated injury scale) [@Gennarelli2006] as investigated by $\chi^2$ and Mann-Whitney test, as appropriate (functions `chisq_test()` and `wilcox_test()`, package _rstatix_).
 
@@
 The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
-By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
+By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
 Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
@@
   footnote(1, 4:5, 
-           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03c\u00B2 test with Cramer V effec size statistic'), 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic'), 
            part = 'header', 
@@
   footnote(1, 4:5, 
-           value = as_paragraph('\u03c\u00B2 test with Cramer V effec size statistic'), 
+           value = as_paragraph('\u03C7\u00B2 test with Cramer V effect size statistic'), 
            part = 'body', 
@@
 
-```{r fig-parts-mental-cohort, fig.width = mm_inch(suppl_figures$parts_mental_cohort$w), fig.height = mm_inch(suppl_figures$parts_mental_cohort$h), fig.cap = 'Injured body regions and detailed scoring of quality of life, post-traumatic syndrome disorder and post-traumatic growth in the study cohort.'}
+```{r fig-parts-mental-cohort, fig.width = mm_inch(suppl_figures$parts_mental_cohort$w), fig.height = mm_inch(suppl_figures$parts_mental_cohort$h), fig.cap = 'Injured body regions and detailed scoring of quality of life, post-traumatic stress disorder and post-traumatic growth in the study cohort.'}
 
@@
 __Supplementary Figure S\@ref(fig:fig-parts-mental-cohort).
-Injured body regions and detailed scoring of quality of life, post-traumatic syndrome disorder and post-traumatic growth in the study cohort.__
+Injured body regions and detailed scoring of quality of life, post-traumatic stress disorder and post-traumatic growth in the study cohort.__
 
@@
 (B) Scoring of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8).
-(C) Scoring of the post-traumatic syndrome disorder (PTSD) clusters with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD clusters.
+(C) Scoring of the post-traumatic stress disorder (PTSD) clusters with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD clusters.
 (E) Scoring of the factors of the post-traumatic growth with the post-traumatic growth inventory (PTGI) tool._
@@
 Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
-By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder).
+By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder).
 Numbers of observations in the mental clusters are shown in the plot legend in (A)._
@@
 Numbers of observations in the clusters are displayed in the plot captions.
-PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder._
+PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
 
@@
 
+```{r fig-early-crf, fig.width = mm_inch(suppl_figures$early_crf$w), fig.height = mm_inch(suppl_figures$early_crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.'}
+
+suppl_figures$early_crf$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-early-crf). Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.__
+
+_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident-related factors available during early acute medical management of the accident was trained in the training subset of the study cohort._
+
+_(A) Prediction accuracy of the classifier in the training and test subset of the study cohort presented in a heat map plot of the confusion matrices.
+Color codes for number of cases.
+The overall prediction accuracy, $\kappa$ and observation numbers are displayed in the plot caption.
+Prediction accuracy for particular clusters is shown in the X axis._
+
+_(B) Conditional importance of the explanatory variables for prediction of the mental cluster assignment expressed as accuracy loss ($\Delta$ accuracy)._
+
+\newpage
+
 # References

### 2023-02-27 classification scripts/early_rf.R

@@ -1,19 +1,20 @@
-# Multi-parameter RF classifier with the top factors for cluster classification
-# identified by the OneR procedure
+# Development of a random forest classifier employing solely 'early' 
+# explanatory variables, i.e. features available during an accident 
+# and acute medical management
 
   insert_head()
   
 # container ------
   
-  class_rf <- list()
-
-# analysis globals -------
+  class_early <- list()
+  
+# analysis globals ------
   
   insert_msg('Analysis globals')
-
+  
   ## tables of assignments for single clusters
   
-  class_rf$clust_tbl <- 
+  class_early$clust_tbl <- 
     list(neutral = class_globals$analysis_tbl, 
          PTG = class_globals$analysis_tbl, 
          PTS = class_globals$analysis_tbl) %>% 
@@ -22,16 +23,16 @@
            map(mutate, 
                clust_id = ifelse(clust_id == clust, clust, 'rest'), 
                clust_id = factor(clust_id)))
-
-# tuning of the classifier -------
+  
+# tuning ------
   
   insert_msg('Tuning')
   
   ## finding the optimal mtry parameter based on OOB prediction accuracy
   
   set.seed(1234)
-
-  class_rf$tuning$controls <- 2:12 %>% 
+  
+  class_early$tuning$controls <- 2:12 %>% 
     map(~cforest_control(teststat = "max",
                          testtype = "Teststatistic", 
                          mtry = .x, 
@@ -39,19 +40,19 @@
                          ntree = 1000)) %>% 
     set_names(2:12)
   
-  class_rf$tuning$models <- class_rf$tuning$controls %>% 
-    map(~cforest(formula = class_globals$formula, 
+  class_early$tuning$models <- class_early$tuning$controls %>% 
+    map(~cforest(formula = class_globals$early_formula, 
                  data = class_globals$analysis_tbl$training, 
                  controls = .x))
   
   ## predictions and fit stats
   
-  class_rf$tuning$predictions <- class_rf$tuning$models %>% 
+  class_early$tuning$predictions <- class_early$tuning$models %>% 
     map(~data.frame(obs = class_globals$analysis_tbl$training$clust_id, 
                     pred = predict(.x, OOB = TRUE) %>% 
                       factor(levels = levels(class_globals$analysis_tbl$training$clust_id))))
   
[…]

### 2023-02-27 classification scripts/one_rule.R

@@ -15,7 +15,7 @@
   class_one$clust_tbl <- 
     list(neutral = class_globals$analysis_tbl, 
          PTG = class_globals$analysis_tbl, 
-         PTSD = class_globals$analysis_tbl) %>% 
+         PTS = class_globals$analysis_tbl) %>% 
     map2(., names(.), 
          function(data, clust) data %>% 
            map(mutate, 
@@ -53,12 +53,12 @@
         response = 'clust_id') %>% 
     set_names(class_globals$variables)
   
-  ## PTSD cluster
+  ## PTS cluster
   
-  class_one$PTSD_models <- class_globals$variables %>% 
+  class_one$PTS_models <- class_globals$variables %>% 
     map(model_one, 
-        train_data = class_one$clust_tbl$PTSD$training, 
-        test_data = class_one$clust_tbl$PTSD$test, 
+        train_data = class_one$clust_tbl$PTS$training, 
+        test_data = class_one$clust_tbl$PTS$test, 
         response = 'clust_id') %>% 
     set_names(class_globals$variables)
   
@@ -69,10 +69,10 @@
   class_one$fit_stats[c('global', 
                         'neutral', 
                         'PTG', 
-                        'PTSD')] <- class_one[c("global_models", 
+                        'PTS')] <- class_one[c("global_models", 
                                                 "neutral_models", 
                                                 "PTG_models", 
-                                                "PTSD_models")] %>% 
+                                                "PTS_models")] %>% 
     map(~map(.x, ~.x$stats)) %>% 
     map(compress, names_to = 'variable')
   

### 2023-02-27 classification.R

@@ -47,9 +47,30 @@
   
   ## full model formula
   
-  class_globals$formula <-  as.formula(paste('clust_id ~ ', 
-                                          paste(class_globals$variables, 
-                                                collapse = '+')))
+  class_globals$formula <-  
+    as.formula(paste('clust_id ~ ', 
+                     paste(class_globals$variables, 
+                           collapse = '+')))
+  
+  
+  ## explanatory variables for early diagnosis
+  
+  class_globals$early_variables <- 
+    class_globals$variables[!class_globals$variables %in% c('psych_support_post_accident', 
+                                                            'psych_support_need', 
+                                                            'accident_aftermath', 
+                                                            'same_sport_type_post_accident', 
+                                                            'caution_post_accident', 
+                                                            'unwilling_flashback', 
+                                                            'flashback_frequency', 
+                                                            'confusion_during_sport')]
+  
+  ## early model formula
+  
+  class_globals$early_formula <- 
+    as.formula(paste('clust_id ~ ', 
+                     paste(class_globals$early_variables, 
+                           collapse = '+')))
   
   ## cluster assignment schemes
   

### 2023-02-27 clustering scripts/background.R

@@ -157,7 +157,7 @@
          right_join, by = 'variable') %>% 
     map(format_summ_tbl) %>% 
     map(set_names, 
-        c('Variable', 'Neutral cluster', 'PTG cluster', 'PTGI cluster', 
+        c('Variable', 'Neutral cluster', 'PTG cluster', 'PTS cluster', 
           'Significance', 'Effect size'))
   
 # END ------

### 2023-02-27 clustering scripts/characteristic.R

@@ -282,7 +282,7 @@
                  regex = '\\nn\\s{1}=\\s{1}\\d+$', 
                  replacement = '')) %>% 
     map(set_names, 
-        c('Variable', 'Neutral cluster', 'PTG cluster', 'PTSD cluster', 
+        c('Variable', 'Neutral cluster', 'PTG cluster', 'PTS cluster', 
           'Significance', 
           'Effect size'))
   

### 2023-02-27 clustering scripts/clustering.R

@@ -35,10 +35,10 @@
   semi_clust$clust_obj$training$clust_assignment <- 
     semi_clust$clust_obj$training$clust_assignment %>% 
     mutate(clust_id = car::recode(as.character(clust_id), 
-                                  "'1' = 'PTSD'; 
+                                  "'1' = 'PTS'; 
                                   '2' = 'neutral'; 
                                   '3' = 'PTG'"), 
-           clust_id = factor(clust_id, c('neutral', 'PTG', 'PTSD'))) 
+           clust_id = factor(clust_id, c('neutral', 'PTG', 'PTS'))) 
   
   ## test
   

### 2023-02-27 import.R

@@ -62,7 +62,7 @@
   
   globals$clust_colors <- c('neutral' = 'cadetblue3',
                             'PTG' = 'darkolivegreen3', 
-                            'PTSD' = 'coral4')
+                            'PTS' = 'coral4')
   
   ## injury regions
   

### 2023-02-27 paper scripts/figures.R

@@ -266,29 +266,38 @@
               'somatic_comorbidity', 
               'psych_comorbidity')]) %>% 
     transpose %>% 
-    map(function(x) plot_grid(plotlist = map(x, 
-                                             ~.x + 
-                                               theme(legend.position = 'none', 
-                                                     plot.margin = ggplot2::margin(l = 2, 
-                                                                                   r = 2, 
-                                                                                   t = 3, 
-                                                                                   b = 3, 
-                                                                                   unit = 'mm'))), 
-                              ncol = 2, 
-                              align = 'hv') %>% 
-          plot_grid(get_legend(x[[1]] + 
-                                 theme(legend.position = 'bottom') + 
-                                 scale_fill_brewer(palette = 'Reds', 
-                                                   labels = function(x) x %>% 
-                                                     stri_replace_all(regex = '000\\s{1}', 
-                                                                      replacement = 'K') %>% 
-                                                     stri_replace(fixed = 'no income', 
-                                                                  replacement = 'none') %>% 
-                                                     stri_replace(fixed = 'EUR', 
-                                                                  replacement = '')) + 
-                                 theme(legend.position = 'bottom')), 
-                    nrow = 2, 
-                    rel_heights = c(0.85, 0.15))) %>% 
+    map2(., 
+         c('Sex', 
+           'Income/year', 
+           'Somatic illness', 
+           'Mental illness'), 
+         function(x, y) plot_grid(plotlist = map2(x, 
+                                                  paste(y, 
+                                                        c('training', 'test'), 
+                                                        sep = ', '),  
+                                                 ~.x + 
+                                                   labs(title = .y) + 
+                                                   theme(legend.position = 'none', 
+                                                         plot.margin = ggplot2::margin(l = 2, 
+                                                                                       r = 2, 
+                                                                                       t = 3, 
+                                                                                       b = 3, 
+                                                                                       unit = 'mm'))), 
+                                  ncol = 2, 
+                                  align = 'hv') %>% 
+           plot_grid(get_legend(x[[1]] + 
+                                  theme(legend.position = 'bottom') + 
+                                  scale_fill_brewer(palette = 'Reds', 
+                                                    labels = function(x) x %>% 
+                                                      stri_replace_all(regex = '000\\s{1}', 
+                                                                       replacement = 'K') %>% 
+                                                      stri_replace(fixed = 'no income', 
[…]

### 2023-02-27 paper scripts/supplementary.R

@@ -203,7 +203,7 @@
               w = 180, 
               h = 140)
   
-# Figure S6: additinal accident details -------
+# Figure S6: additional accident details -------
   
   insert_msg('Figure S6: additional accident details')
 
@@ -260,6 +260,37 @@
               w = 180, 
               h = 120)
   
+  
+# Figure S8: mental cluster classification, early predictors -------
+  
+  insert_msg('Figure S8: Mental cluster classification, cRF, early preds')
+  
+  suppl_figures$early_crf <- class_early$confusion_plots$global %>% 
+    map2(., 
+         paste('Cluster classification,', 
+               c('training', 'test')), 
+         ~.x + 
+           theme(legend.position = 'none') + 
+           labs(title = .y)) %>% 
+    plot_grid(plotlist = ., 
+              nrow = 2, 
+              align = 'hv') %>% 
+    plot_grid(class_early$importance_plots$global + 
+                theme(plot.title.position = 'plot', 
+                      plot.title = element_text(hjust = 0.3)), 
+              ncol = 2, 
+              rel_widths = c(0.9, 1.1), 
+              labels = LETTERS, 
+              label_size = 10) %>% 
+    as_figure(label = 'figure_s8_rf_early_classifier', 
+              ref_name = 'early_crf', 
+              caption = paste('Assignment of accident victims to', 
+                              'the mental clusters based on', 
+                              'explanatory factors available', 
+                              'during acute medical management', 
+                              'of the accident.'), 
+              w = 180, 
+              h = 150)  
   
 # Saving the figures on the disc -------
   

### 2023-03-30 paper/markdown/manuscript.Rmd
Changes in figures, tables and paper text
Changes in the table layout, shortening of the text, focus on early predictors of mental problems post accident

--- 2023-02-14 paper/markdown/manuscript.Rmd
@@
 ---
-title: "Three distinct patterns of mental health response after a sport accident"
+title: "Three distinct patterns of mental response after mountain sport accidents"
 subtitle: "Manuscript"
@@
       institute: [psych]
-      email: Katharina.Huefner@i-med.ac.at
+      email: katharina.huefner@tirol-kliniken.at
       correspondence: true
@@
   - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria
+  - daas: Data Analysis as a Service Tirol, Innsbruck, Austria
 
-keywords: "sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth"
+keywords: "sport accident, mental health, post-traumatic stress disorder, post-traumatic growth"
 
@@
 
+__Background:__ Psychological consequences of mountain sport accidents are poorly characterized. 
+We investigated patterns of mental response to mountain sport accidents and their victims at risk of mental deterioration.
+
+__Methods:__ A cross-sectional survey was conducted with `r nrow(ptsd$dataset)` adult victims of mountain sport accidents treated at the Innsbruck University Hospital (Austria). 
+Anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data.
+Patterns of mental readouts were defined by semi-supervised medoid clustering and models of cluster assignment based on non-mental measures were developed by the random forest algorithm.
+
+__Results:__ Three mental clusters of accident victims were hallmarked by a neutral mental response, post-traumatic growth and post-traumatic burden of mental disorders, respectively. 
+Despite relatively frequent hospitalization, surgery, pre-existing mental or somatic illness, low income, and female gender in the post-traumatic growth and post-traumatic burden clusters, overall differences in non-mental variables between the clusters were weak.
+Consequently, no reliable model of the cluster assignment employing explanatory factors available during acute medical accident management could be established.
+
+__Conclusion:__ A subset of mountain sport accident victims develops symptoms of anxiety, depression, post-traumatic stress disorder and other mental health problems. 
+Since identification of these vulnerable patients with demographic, socioeconomic, accident- and injury-related predictors remains challenging, early and low-threshold psychological support is key to prevention of mental disorders following mountain sport accidents.
+
 \newpage
@@
 
-sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth
+mountain sport accident, mental health, post-traumatic stress disorder, post-traumatic growth, clustering, random forest, machine learning
 
@@
 
+## Ethics
+
+The study was conducted in accordance with the Declaration of Helsinki and European data policies. 
+All participants gave electronically signed written informed consent to participate. 
+Participants' data were processed in anonymized form. 
+The study protocol was approved by the ethics committee of the Medical University of Innsbruck (approval number: _please fill in!_).
+
+## Participants
+
+In- and outpatients treated at the orthopedics and trauma departments of the Medical University of Innsbruck (Austria) between 2018 and 2020 were screened for participation. 
+Patients fulfilling the inclusion criteria: age $\geq$ 16 years, proficiency in German and treatment due to a mountain sport accident (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey.
+Out of them, `r length(ptsd$interview_ID)` completed the survey and `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S2__).
+
+## Procedures
+
+Details on study procedures and variables are provided in __Supplementary Methods__ and __Supplementary Table S1__.
+
+The psychometric battery consisted of German versions of assessment tools for 
+anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
+depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
+panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
+persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
+resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
+loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
+quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
+post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
+post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
+The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__). 
+Signs of anxiety (GAD-7 $\geq$ 10), depression (PHQ-9 $\geq$ 11) [@Manea2012], persistent somatic symptoms (PHQ-15 $\geq$ 11) [@Kroenke2002] and resilient coping classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73)[@Leppert2008] were defined as described before.
+The total PTGI and PCL-5 DSM-5 scores were calculated as the sum of all items of the respective tools [@Tedeschi1996; @Bovin2016]. 
+Significant PTSD symptoms were defined as positivity in at least one PCL-5 DSM-5 domain [@Bovin2016].
+
+Traumatic events before the accident were assessed with the DIA-X tool [@Maercker2005]. 
+Flashbacks frequency was surveyed as none, > 1/month and > 1/year. 
+Smoking was surveyed as a single yes/no question. 
+Alcohol abuse was investigated with the CAGE tool [@OBrien2008]. 
+Data on injury (AIS: abbreviated injury scale) [@Gennarelli2006] and medical treatment were extracted from electronic patient's records.
+
+The training and test participant subsets (2:1 size ratio) were obtained by random splitting which minimized differences in accident year, age, gender, pre-existing traumatic events, somatic and mental illness, and injury severity (__Figure \@ref(fig:fig-consort)__).
+The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as assessed by Hopkins metrics = `r pow$hopkins %>% filter(size == 200) %>% .$hopkins_stat %>% median %>% signif(2)` [@Kassambara2020].
+
+## Analysis endpoints
+
+The primary analysis endpoint was identification of subsets of mountain sport accident victims by clustering in respect to scores of anxiety, depression, panic, persistent somatic symptoms, quality of life, post-traumatic growth and post-traumatic stress disorder.
+The secondary analysis endpoints were demographic, socioeconomic and clinical characteristic of the mental clusters and development of a mental cluster assignment model to discern accident victims at risk of mental disorders.
+
+## Statistical analysis
+
+Details on statistical analysis are provided in __Supplementary Methods__
+
+Statistical analysis was done with R version 4.2.3. 
+Numeric variables are displayed as medians with interquartile ranges. 
+Categorical variables are presented as percentages and counts. 
+Differences in numeric variables were analyzed by Mann-Whitney or Kruskal-Wallis test with r or  $\eta^2$ effect size statistics. 
+Differences in categorical variable distribution were assessed by $\chi^2$ test with Cramer V effect size statistic [@Kassambara2021].
+
+The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
+This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares) and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical and KMEANS algorithms.
+The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
+The training subset observations were assigned to the clusters with an inverse distance weighted 7-nearest neighbor classifier.
+
+The uni- and multi-parameter classifiers were trained with the one-rule [@VonJouanne-Diedrich2017; @Holte1993] and conditional random forest algorithms [@Hothorn2006; @Strobl2007; @Hothorn2022], respectively.
+The cluster assignment was predicted for the test subset and assessed by the accuracy and $\kappa$ statistics [@Kuhn2008]. 
+Conditional variable importance in the random forest classifiers was expressed as accuracy loss [@Strobl2008].
+
 \newpage
@@
 # Results
+
+# Characteristic of the study cohort
+
+In- and outpatient adult victims of mountain sport accidents treated at the Medical University of Innsbruck (Austria) between 2018 and 2020 were invited to participate (n = `r nrow(ptsd$cleared)`). 
+Out of them, 
+`r length( ptsd$interview_ID)` completed the study survey and 
+`r nrow(ptsd$dataset)` individuals with the complete psychometric parameter set were analyzed (overall response rate: `r ((nrow(ptsd$dataset)/nrow(ptsd$cleared)) * 100) %>% signif(2)`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).
+
+As compared with the participants, the individuals excluded from the analysis had lower income, less frequent ski and more frequent biking accidents, less severe injuries, and lower hospitalization and surgery rates.
+The excluded individuals had also lower ratings of post-traumatic growth ratings and lower sense of coherence. 
+Effects of these differences were weak (V $\leq$ 0.22 or r $\leq$ 0.15, __Supplementary Table S4__).
+
+In the analyzed cohort, the median time between the medical accident management and survey completion was 
+`r median(ptsd$dataset$obs_time)` days. 
+The participants were predominantly middle-aged (median `r median(ptsd$dataset$age)` years) and 
+`r signif(table(ptsd$dataset$sex)/sum(table(ptsd$dataset$sex)), 2)[1] * 100`% of them were females. 
+The vast majority had secondary or tertiary education grade and were professionally active. 
+Sport and trauma-risk professions constituted < 8% of the cohort. 
+High annual household incomes of $\geq$ 45000 Euro was reported by > 40% of participants. 
+Less than 10% of participants were smokers or at risk of alcohol addiction. 
+Chronic somatic conditions prior to the accident were reported by `r signif(table(ptsd$dataset$somatic_comorbidity)/sum(table(ptsd$dataset$somatic_comorbidity)), 2)[2] * 100`% of participants with cardiovascular, neurological and metabolic illness being the most frequent.
+The rates of smoking (Austria: 20.6%) and somatic illness (38.3%) were lower than in the general Austrian population [@StatistikAustria2022a; @StatistikAustria2022].
+Psychiatric disorders before the accident affected `r signif(table(ptsd$dataset$psych_comorbidity)/sum(table(ptsd$dataset$psych_comorbidity)), 2)[2] * 100`% of the cohort. 
+Four of ten participants were affected by or witnessed a traumatic event prior to the accident (__Table \@ref(tab:tab-cohort-demo)__).
+
+Nearly 40% of participants had sport accidents in the past. 
+Almost two-thirds of the investigated accidents happened during skiing or snowboarding followed by biking and classical mountain sports such as climbing, hiking or mountaineering. 
+One-third participants were alone during the accident and, in most cases, were the sole culprit and victim. 
+Roughly half of participants could rescue themselves, `r signif(table(ptsd$dataset$accident_rescue)/sum(table(ptsd$dataset$accident_rescue)), 2)[3] * 100`% required a professional rescue. 
+In `r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[2] * 100`% of participants the injury was moderate (AIS 2) and in 
+`r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[3] * 100`% severe-to-critical (AIS $\geq$ 3). 
+Limb injuries were the most common followed by head and face (__Supplementary Figure S1A__). 
+Hospitalization and surgery rates were 
+`r signif(table(ptsd$dataset$hospitalization)/sum(table(ptsd$dataset$hospitalization)), 2)[2] * 100`% and 
+`r signif(table(ptsd$dataset$surgery_done)/sum(table(ptsd$dataset$surgery_done)), 2)[2] * 100`%, respectively. 
+Psychological or psychiatric support after the accident was applied to 
+`r signif(table(ptsd$dataset$psych_support_post_accident)/sum(table(ptsd$dataset$psych_support_post_accident)), 2)[2] * 100`% individuals. 
+Among those who had not received it, 
+`r signif(table(ptsd$dataset$psych_support_need)/sum(table(ptsd$dataset$psych_support_need)), 2)[2] * 100`% declared psychological support need. 
+Despite fairly common somatic health consequences of the accident 
+(`r signif(table(ptsd$dataset$accident_aftermath)/sum(table(ptsd$dataset$accident_aftermath)), 2)[2] * 100`%) and flashbacks during sport (`r signif(table(ptsd$dataset$unwilling_flashback)/sum(table(ptsd$dataset$unwilling_flashback)), 2)[2] * 100`%), most participants returned to the accident sport type (
+`r signif(table(ptsd$dataset$same_sport_type_post_accident)/sum(table(ptsd$dataset$same_sport_type_post_accident)), 2)[2] * 100`%). 
+Yet, two-third described their behavior during sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).
+
+Signs of 
+anxiety (`r signif(table(ptsd$dataset$gad7_total_class)/sum(table(ptsd$dataset$gad7_total_class)), 2)[2] * 100`%), 
+depression (`r signif(table(ptsd$dataset$phq9_total_class)/sum(table(ptsd$dataset$phq9_total_class)), 2)[2] * 100`%) and 
+persistent somatic symptoms (`r signif(table(ptsd$dataset$phq_events_total_class)/sum(table(ptsd$dataset$phq_events_total_class)), 2)[2] * 100`%) were rare in the study cohort (__Table \@ref(tab:tab-cohort-mental)__).
+Quality of life was rated with median `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS points, which was comparable to the generalized Western European population [@Schmidt2006]; overall quality of life and housing domains received the highest ratings.
+Roughly two-thirds of participants were assigned to the high resilient coping class [@Leppert2008]. 
+Overall post-traumatic growth (median: `r median(ptsd$dataset$ptgi_total) %>% signif(2)`) and its specific domains were rated lower than originally reported for the PTGI tool [@Tedeschi1996]. 
+Symptoms of post-traumatic stress disorder were observed in `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants with domain B symptoms (`r signif(table(ptsd$dataset$dsm5_B_class)/sum(table(ptsd$dataset$dsm5_B_class)), 2)[2] * 100`%) being most common (__Table \@ref(tab:tab-cohort-mental)__ and __Supplementary Figure S1__).
+
+For clustering and modeling, the cohort was split into the training (n = `r nrow(clust_globals$analysis_tbl$training)`) and test subset (n = `r nrow(clust_globals$analysis_tbl$test)`). 
+There were no significant differences between the subsets in demographic, socioeconomic, clinical and accident-related parameters (not shown).
+
+# Three clusters of mental response in sport accident victims
+
+To explore mental responses to mountain sport accidents, we subjected the participants to clustering in respect to a broad range of numeric psychometry variables (__Supplementary Table S2__). 
+Among clustering algorithms compared in the training subset, partition around medoids with cosine distance [@Drost2018; @Schubert2019] demonstrated good explanatory power and superior reproducibility in cross-validation [@Lange2004] (__Supplementary Figure S2A__).
+Three mental clusters: 'neutral', 'PTG' (post-traumatic growth) and 'PTB' (post-traumatic burden), named after their key mental characteristic were identified in the training subset. 
+Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`), comparable cluster sizes and good visual clusetr separation in both study cohort subsets (__Supplementary Figures S2B__ - __S4__).
+
+The neutral cluster encompassing roughly one-third participants was characterized by low levels of anxiety, depression, panic, somatic symptoms, post-traumatic stress and post-traumatic growth along with high rating of coherence, resilient coping and quality of life.
+The PTG cluster demonstrated similarly low scores of major mental health disorders. 
+By contrast, levels of post-traumatic growth were the highest in the PTG cluster. 
+Furthermore, ratings of post-traumatic stress were slightly higher in the PTG than in the neutral cluster. 
+The remaining PTB cluster displayed the highest scores of anxiety, depression, panic, somatic symptoms and post-traumatic stress as well as poor sense of coherence, resilient coping and low quality of life.
+Post-traumatic growth scores in the PTB cluster were higher than in neutral but lower than in PTB cluster participants (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Table S5__).
+Consequently, clinically relevant symptoms of anxiety, depression and persistent somatic symptoms were present virtually only in the PTB cluster. 
+Furthermore, frequencies of low and moderate resilient classes peaked in the PTB cluster. 
+Finally, symptoms of post-traumatic stress disorder were way more frequent in the PTB cluster (any symptoms: > 36%) than in the remaining clusters (__Figure \@ref(fig:fig-clust-symptoms)__).
+
+Collectively, the PTG cluster may represent individuals with salutary reaction to mountain sport accident, whereas the PTB cluster may pose a group at risk of mental health deterioration.
+
+# Demographic, socioeconomic and clinical background of the mental clusters
+
+Among 
+`r length(clust_bcg$variables)` investigated demographic, socioeconomic, clinical, accident- and accident consequence-related variables, only 
+`r length(clust_bcg$top_factors)` parameters were found to differ significantly between the mental clusters in both the training and test subsets. 
+The effect size of such differences was weak (__Supplementary Table S6__).
+
+A substantial enrichment of females, low-to-middle income and non-tertiary education participants was observed in the PTB cluster. 
+The PTG cluster had the highest share of middle-aged individuals. 
+Frequency of somatic conditions before the accident were the highest in the PTG and PTB clusters. 
+Participants with pre-existing mental illness were in turn represented virtually only in the PTB cluster. 
+Although frequencies of previous sport accidents were comparable between the clusters, PTG cluster individuals were affected by prior traumatic events more often than neutral cluster individuals. 
+These differences were, however, non or borderline significant and of weak effect size (__Figure \@ref(fig:fig-clinic-factors)__, __Supplementary Figure S5__). 
+There were no consistent differences in accident sport type, rescue, injury severity and location between the mental clusters. 
+Yet, PTG and PTB cluster individuals tended towards higher hospitalization and surgery rates (__Figure \@ref(fig:fig-injury)__ and __Supplementary Figure S6__).
+
+Somatic health consequences of the accident and flashbacks were significantly more common in the PTB than the neutral cluster. 
+Additionally, PTG and PTB cluster individuals tended towards more caution during sport as compared with the neutral cluster. 
+Roughly one-sixth PTB cluster individuals reported need for psychological support following the accident as compared with none in the neutral cluster. 
+Rates psychological support after the accident tended to be higher in the PTG (12 - 16%) than in the remaining clusters, yet this effect was significant only in the training subset (__Figure \@ref(fig:fig-aftermath)__).
+Given the low frequency of psychological support after the accident in the study cohort and particular clusters, it is unlikely, that psychological support affected the mental cluster assignment .
+
+# Prediction of the mental cluster assignment by demographic, socioeconomic and accident-related factors
+
+Finally, we intended to model the mental cluster assignment with `r length(class_globals$early_variables)` demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident (__Supplementary Table S7__).
+Such models would enable early identification of mountain sport accident victims at risk of mental health problems.
+
+Predictive value [@VonJouanne-Diedrich2017; @Holte1993] of single early explanatory parameters was poor (maximum accuracy: `r class_one$fit_stats$global$Accuracy %>% max %>% signif(2)`, maximum $\kappa$: `r class_one$fit_stats$global$Kappa %>% max %>% signif(2)`) and inconsistent between the subsets of the study cohort.
+In such univariable analysis, age class, annual income and traumatic events in the past were the most predictive factors (__Supplementary Figure S7__).
+
+To account for cumulative effects and interactions of multiple weak predictors of the mental cluster assignment, we resorted to conditional random forest modeling [@Strobl2007; @Hothorn2006; @Hothorn2022].
+This multi-parameter model employing early candidate cluster predictors could correctly assign `r signif(class_early$fit_stats$global$Accuracy[1], 2) * 100`% training subset observations to their mental clusters
+($\kappa$ = `r signif(class_early$fit_stats$global$Kappa[1], 2)`). 
+However, its predictive performance in the test subset was poor (accuracy: `r signif(class_early$fit_stats$global$Accuracy[2], 2) * 100`, $\kappa$ = `r signif(class_early$fit_stats$global$Kappa[2], 2)`).
+The prediction quality was the best for the PTG cluster (accuracy, training: 89.5%, test: 59.4%, __Figure \@ref(fig:fig-early-crf)A__).
+The most important early explanatory variables for the cluster prediction by conditional random forests were annual income, sex, mode of rescue, leg injury and pre-existing mental illness (__Figure \@ref(fig:fig-early-crf)B__).
+Inclusion of late accident consequences such as cautious behavior during sport, flashbacks or somatic health burden of the accident in the random forest model could only marginally improve its performance (training: accuracy = `r signif(class_rf$fit_stats$global$Accuracy[1], 2)`, $\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[1], 2)`; test subset: accuracy = `r signif(class_rf$fit_stats$global$Accuracy[2], 2)`, $\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[2], 2)`).
+The most important late predictors of the cluster assignment were caution during sport and flashbacks (__Supplementary Figure S8__).
+
+Collectively, we could not establish a reliable model for identification of patients at risk of mental disorders represented by the PTB cluster with a broad ensemble of early non-mental predictors.
 
@@
 
+In individuals surveyed on average 3.5 years after the non-professional sport accident, some of its consequences like somatic health problems, unwilling flashbacks as well as phenomena of post-traumatic growth and symptoms of post-traumatic stress disorder were still tangible.
+By semi-supervised clustering we could discern three distinct patterns of mental health in this collective with each present in approximately one-third of participants: (1) a neutral mental response reaction, (2) salutatory reaction to the accident characterized by post-traumatic growth and (3) mental health deterioration hallmarked by mental accident burden such as symptoms of anxiety, depression, somatization, resilience loss and signs of post-traumatic stress disorder.
+Notably, separation of the PTG and PTB clusters was in part blurry. 
+Symptoms of post-traumatic stress were observed also in PTG cluster participants albeit less frequently than in the PTB cluster. 
+Conversely, mental health problems in PTB cluster participants were paralleled by some signs of post-traumatic growth. 
+In a detailed comparison of `r length(clust_bcg$variables)` factors between the clusters, no distinct demographic, socioeconomic, clinical or accident-related background of the mental clusters could be identified.
+We found a moderate enrichment of females, low-to-moderate income participants, non-academic graduates, chronic somatic and psychiatric conditions as well as individuals requiring hospitalization and surgery in the PTG and, in particular, in the PTB cluster.
+Injury severity and localization or professional rescue rates indicative of potentially more severe accident were comparable between the clusters. 
+Similarly, the cluster assignment was unlikely affected by psychological support after the accident. 
+In turn, the PTB and, to a lesser extent, the PTG cluster were characterized by somatic health consequences of the accident, flashbacks, more caution during sport activity and subjective need for psychological support.
+Finally, the lacking non-mental explanatory variables strongly associated with the cluster assignment likely resulted in sub-optimal performance of machine learning classifiers employed for prediction of the cluster assignment and identification of accident victims at risk of mental health problems.
+
+_Hanna and Katharina: your turn now:)_
+
+Our study bears limitations. 
+First, although the key phenomena phenomena in the newly identified mental clusters, post-traumatic growth and post-traumatic stress disorder, are clearly associated with trauma [@Tedeschi1996; @Bovin2016], we could not exclude the mental cluster assignment to be affected by other events than the sport accident.
+To account for that, follow-up studies including a control group are needed. 
+Second, our variable set misses potentially important explanatory factors for the mental cluster classification such as length of hospital stay, rehabilitation need or ability to work. 
+Finally, effects of the accident, injury severity, hospitalization or surgery on mental health after an accident may have been obscured in the cross-sectional cohort and needs to be investigated in a more defined collective e.g.
+of high risk sport accident victims or hospitalized individuals.
+
 # Conclusion
+
+_Your part again please_
 
@@
   flextable::width(2, width = 7.5, unit = 'cm') %>% 
+  footnote(1, 1, 
+           value = as_paragraph('CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; DIA-X: Diagnostic Expert System, traumatic event score'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
   theme_vanilla
@@
   footnote(2, 1, 
-           value = as_paragraph('ski/snowdoard: alpine skiing, snowboarding and cross-country skiing\nsledding: sledding or bobsled\nmountain: hiking, climbing and skitouring\nbiking: mountainbike, tour and road cycling'),
+           value = as_paragraph('ski/snowdoard: alpine skiing, snowboarding and cross-country skiing\nsledding: sledding or bobsled\nclimbing/hiking/mountaineering: hiking, rock and ice climbing, mountaineering, skitouring\nbiking: mountainbike, tour and road cycling'),
            part = 'body', 
@@
   footnote(1, 1, 
-           value = as_paragraph('PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder'),
            part = 'header', 
@@
 
-__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process.__
+__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process and the analysis strategy.__
+
+\newpage
+
+```{r fig-clust-psych, fig.width = mm_inch(figures$clust_psych$w), fig.height = mm_inch(figures$clust_psych$h), fig.cap = 'Scores of psychometry readouts in the mental clusters.'}
+
+figures$clust_psych$plot
+
+```
+
+__Figure \@ref(fig:fig-clust-psych). Scores of psychometry readouts in the mental clusters.__
+
+_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations.
+Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
+Three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden), All clustering parameters differed significantly between the clusters as assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Mean normalized scores of major mental health readouts (A), diminished quality of life (B), post-traumatic growth (C) and post-traumatic symptom disorder (D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines.
+Tinted regions represent 2 $\times$ SEM (standard error of the mean).
+Numbers of observations in the clusters are displayed in the plot captions.
+GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
+
+\newpage
+
+```{r fig-clust-symptoms, fig.width = mm_inch(figures$clust_symptoms$w), fig.height = mm_inch(figures$clust_symptoms$h), fig.cap = 'Signs of mental health problems in the mental clusters.'}
+
+figures$clust_symptoms$plot
+
+```
+
+__Figure \@ref(fig:fig-clust-symptoms). Signs of mental health problems in the mental clusters.__
+
+_Frequencies of symptoms of anxiety (A, GAD-7 ≥10), depression (B, PHQ-9 ≥ 11) and somatization (C, PHQ-15 ≥ 11), distribution of resilient coping classes (D, RS13) and symptoms of post-traumatic stress disorder (E, PTSD, PCL-5 DSM-5, positivity for PTSD domains and frequency of participants positive for at least one PTSD domain [PTSD+]) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Fractions of symptom-positive and -negative participants in the mental clusters in the training and test subset of the study cohort are presented in stack and bar plots.
+Effect sizes and p-values are shown in the plot captions or Y axes.
+Numbers of observations in the clusters are shown in the X axes or plot captions.
+Significant effects in (E) are highlighted in bold.
+GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder._
+
+\newpage
+
+```{r fig-clinic-factors, fig.width = mm_inch(figures$clinic_factors$w), fig.height = mm_inch(figures$clinic_factors$h), fig.cap = 'Demographic and clinical factors differing between the mental clusters of accident victims.'}
+
+figures$clinic_factors$plot
+
+```
+
+__Figure \@ref(fig:fig-clinic-factors). Demographic and clinical factors differing between the mental clusters of accident victims.__
+
+_Distribution of gender (A), annual household income (B), and frequencies of self-reported pre-existing somatic conditions (C) and self-reported pre-existing mental illness (D) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
+Effect sizes and p-values are displayed in the plot captions.
+Numbers of observations in the clusters are presented in the X axes._
+
+\newpage
+
+```{r fig-injury, fig.width = mm_inch(figures$injury$w), fig.height = mm_inch(figures$injury$h), fig.cap = 'Injury severity and location in the mental clusters.'}
+
+figures$injury$plot
+
+```
+
+__Figure \@ref(fig:fig-injury). Injury severity and location in the mental clusters.__
+
+_Distribution of injury severity grades (A, AIS: abbreviated injury scale), hospitalization (B) and surgery rates (C), and injuries of the body parts (D) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack and bar plots.
+Effect sizes and p-values are displayed in the plot captions or Y axes.
+Numbers of observations in the clusters are presented in the X axes or plot captions._
+
+\newpage
+
+```{r fig-aftermath, fig.width = mm_inch(figures$aftermath$w), fig.height = mm_inch(figures$aftermath$h), fig.cap = 'Consequences of the accident in the mental clusters.'}
+
+figures$aftermath$plot
+
+```
+
+__Figure \@ref(fig:fig-aftermath). Consequences of the accident in the mental clusters.__
+
+_Frequencies of self-reported somatic health consequences of the accident (A), flashbacks during sport (B), more or less cautious behavior during sport activity (C), self-reported psychological support need (D) and of a received psychological support (E) in the mental clusters.
+Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
+Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
+Effect sizes and p-values are displayed in the plot captions.
+Numbers of observations in the clusters are presented in the X axes._
+
+\newpage
+
+```{r fig-early-crf, fig.width = mm_inch(figures$early_crf$w), fig.height = mm_inch(figures$early_crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.'}
+
+figures$early_crf$plot
+
+```
+
+__Figure \@ref(fig:fig-early-crf). Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.__
+
+_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident was trained in the training subset of the study cohort._
+
+_(A) Prediction accuracy of the classifier in the training and test subset of the study cohort presented in a heat map plot of the confusion matrices.
+Color codes for number of cases.
+The overall prediction accuracy, $\kappa$ and observation numbers are displayed in the plot caption.
+Prediction accuracy for particular clusters is shown in the X axis._
+
+_(B) Conditional importance of the explanatory variables for prediction of the mental cluster assignment expressed as accuracy loss ($\Delta$ accuracy)._
 

### 2023-03-30 paper/markdown/supplementary_material.Rmd
Changes in figures, tables and paper text
Changes in the table layout, shortening of the text, focus on early predictors of mental problems post accident

@@
 ---
-title: "Three distinct patterns of mental response after a sport accident"
+title: "Three distinct patterns of mental response after mountain sport accidents"
 subtitle: "Supplementary Material"
@@
 
-Items of the EUROHIS QOL 8 tool were scores as with 1 - 5 likert scales (1: no concerns/full satisfaction, 5: extreme concerns, no satisfaction at all) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing).
+Items of the EUROHIS QOL 8 tool were scores as with 1 - 5 likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction, ) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing).
 The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].
@@
 
-Traumatic events prior to the sport accident were assessed with the DIA-X tool [@Maercker2005]. 
+Traumatic events prior to the sport accident were assessed with the DIA-X tool (Diagnostic Expert System) [@Maercker2005]. 
 Direct personal experience or being a witness of a traumatic event specified by the DIA-X questionnaire or by an additional yes/no item ('other traumatic events') was scored as 1. 
@@
 Smoking was surveyed as a single yes/no question. 
-Substance abuse was investigated with the CAGE tool with $\geq$ 2 points indicative of substantial addiction [@OBrien2008]. 
-Data on hospital treatment, hospitalization and surgery were extracted from electronic patient's records. 
+Alcohol abuse was investigated with the CAGE tool with $\geq$ 2 points indicative of substantial addiction [@OBrien2008]. 
+Data on hospital treatment and surgery were extracted from electronic patient's records. 
 Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].
@@
 Potential differences between the training and test subset of the study cohort (__Figure 1__) were analyzed in an analogical way. 
-Significant and near-significant (p < 0.1) differences between the included/excluded participants and the training/test subsets are presented in __Supplementary Tables S\@ref(tab:tab-incl-excl)__ and __S\@ref(tab:tab-partition)__, respectively.
+Significant (p < 0.05) differences between the included/excluded participants are presented in __Supplementary Tables S\@ref(tab:tab-incl-excl)__.
 
@@
 The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
-By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
+By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
 Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
@@
 
-Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters were trained in the training subset of the study cohort by the one-rule (oneR) algorithm (function `OneR()`, package _OneR__) [@VonJouanne-Diedrich2017; @Holte1993].
+Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters (__Supplementary Table S\@ref(tab:tab-mod-variables)__) were trained in the training subset of the study cohort by the one-rule (oneR) algorithm (function `OneR()`, package _OneR__) [@VonJouanne-Diedrich2017; @Holte1993].
 The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _OneR_).
-The multi-parameter classifier was developed in the training subset with the conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022] with n = 1000 random trees, the splitting rule corresponding to the maximum test statistic and five variable draws per tree (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 5, mincriterion = qnorm(0.9), ntree = 1000)` control object, package _party_).
-The optimal variable draw values were obtained by optimization (tuning) employing the out-of-bag predictions in the training subsets (not shown). 
+
+Two multi-parameter classifiers of the mental cluster assignment were developed in the training subset:
+
+1. a model employing candidate early predictors of cluster assignment, i.e. demographic, socioeconomic, medical history and accident related explanatory factors
+
+2. a model including additionally late consequences of the accident such as somatic health aftermath, flashbacks or cautious behavior during sport
+
+The multi-parameter classifiers (__Supplementary Table S\@ref(tab:tab-mod-variables)__) were trained in the training subset with the conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022] with n = 1000 random trees and the splitting rule corresponding to the maximum test statistic.
+In the model 1, the `mtry` parameter was set to 7, `minsplit` to 10 and `mincriterion` to 1.28 (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 7, mincriterion = qnorm(0.9), minsplit = 10, ntree = 1000)` control object, package _party_).
+In the model 2, the `mtry` parameter was set to 6, `minsplit` to 20 and `mincriterion` to 0 (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 6, mincriterion = 0, minsplit = 20, ntree = 1000)` control object, package _party_).
+The optimal parameter sets were obtained by optimization (tuning) employing the out-of-bag predictions in the training subsets (not shown). 
 The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _party_).
@@
 
-```{r tab-mental-battery, tab.cap = 'Mental health assessment battery. The table is available in a supplementary Excel file.'}
+\newpage
 
-suppl_tables$mental_battery[1, 1] %>% 
-  mutate(Section = '') %>% 
-  set_names(' ') %>% 
-  flextable
+```{r tab-mental-battery, tab.cap = 'Mental health assessment battery.'}
+
+suppl_tables$mental_battery[c(2, 3)] %>% 
+  flextable %>% 
+  width(1, width = 6.5, unit = 'cm') %>% 
+  width(2, width = 10.5, unit = 'cm') %>% 
+  theme_vanilla
 
@@
 
-```{r tab-incl-excl, tab.cap = 'Significant and near-significant (p < 0.1) differences between individuals excluded from analysis and analyzed study participants.
+```{r tab-incl-excl, tab.cap = 'Significant (p < 0.05) differences between individuals excluded from analysis and analyzed study participants.
 Numeric variables are presented as medians with interquartile ranges (IQR).
@@
   flextable::width(5, width = 2.3, unit = 'cm') %>% 
-  footnote(1, 4:5, 
-           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic'), 
+  footnote(1, 1, 
+           value = as_paragraph('AIS: abbreviated injury scale; PTGI: post-traumatic growth inventory; SOC-9L: Leipzig 9-item sense of coherence scale; K: 1000 Euro; EUR: Euro.'), 
            part = 'header', 
            ref_symbols = 'a') %>% 
-  footnote(8, 1, 
-           value = as_paragraph('AIS: abbreviated injury scale'), 
-           part = 'body', 
-           ref_symbols = 'b') %>% 
-  footnote(11, 1, 
-           value = as_paragraph('PCL-5 DSM-5: PTSD checklist for DSM-5'), 
-           part = 'body', 
-           ref_symbols = 'c') %>% 
-  footnote(12:15, 1, 
-           value = as_paragraph('PTGI: post-traumatic growth inventory'), 
-           part = 'body', 
-           ref_symbols = 'd') %>% 
-  footnote(16, 1, 
-           value = as_paragraph('PSS4: 4-item perceived stress scale'), 
-           part = 'body', 
-           ref_symbols = 'e') %>% 
-  footnote(17, 1, 
-           value = as_paragraph('SOC-9L: Leipzig 9-item sense of coherence scale'), 
-           part = 'body', 
-           ref_symbols = 'f') %>% 
-  theme_vanilla
-
-```
-
-\newpage
-
-```{r tab-partition, tab.cap = 'Significant and near-significant (p < 0.1) differences between the training and test subsets of the study cohort.
-Numeric variables are presented as medians with interquartile ranges (IQR).
-Categorical variables are presented as percentages and counts within the complete observation set.'}
-
-flextable::flextable(suppl_tables$partition) %>% 
-  flextable::width(1, width = 3.2, unit = 'cm') %>% 
-  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
-  flextable::width(4, width = 2.7, unit = 'cm') %>% 
-  flextable::width(5, width = 2.3, unit = 'cm') %>% 
   footnote(1, 4:5, 
-           value = as_paragraph('\u03C7\u00B2 test with Cramer V effect size statistic'), 
-           part = 'body', 
-           ref_symbols = 'a') %>% 
-  footnote(2, 1, 
-           value = as_paragraph('RS13: 13-item resilience scale'), 
-           part = 'body', 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic, '), 
+           part = 'header', 
            ref_symbols = 'b') %>% 
@@
 
-```{r tab-clust-bcg, tab.cap = 'Significant and near-significant (p < 0.1) differences in demographic, socioeconomic, clinical and accident-related factors between the mental clusters.
+```{r tab-clust-bcg, tab.cap = 'Significant (p < 0.05) differences in demographic, socioeconomic, clinical and accident-related factors between the mental clusters.
 Numeric variables are presented as medians with interquartile ranges (IQR).
-Categorical variables are presented as percentages and counts within the complete observation set.
-The table is available in a supplementary Excel file.'}
+Categorical variables are presented as percentages and counts within the complete observation set.'}
 
-suppl_tables$clust_bcg[1, 1] %>% 
-  mutate(Variable = '') %>% 
-  set_names(' ') %>% 
-  flextable
+suppl_tables$clust_bcg %>% 
+  flextable %>% 
+  width(1, width = 1.7, unit = 'cm') %>% 
+  width(2, width = 3.7, unit = 'cm') %>% 
+  width(3:5, width = 3.5, unit = 'cm') %>% 
+  width(6, width = 2.4, unit = 'cm') %>% 
+  width(7, width = 2.2, unit = 'cm') %>% 
+  merge_v(1) %>% 
+  footnote(1, 2, 
+           value = as_paragraph('CAGE: Cut, Annoyed, Guilty, and Eye scale; AIS: abbreviated injury score.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 6:7, 
+           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors (early: available during acute medical management of the accident) used for modeling of the mental cluster assignment.'}
+
+flextable::flextable(suppl_tables$mod_variables) %>% 
+  width(1, width = 3, unit = 'cm') %>% 
+  width(2, width = 12, unit = 'cm') %>% 
+  theme_vanilla
 
@@
 Numeric variables are presented in violin plots with red diamonds denoting medians, interquartile ranges presented as red whiskers and single observations depicted as points.
+Quality of life scores are expressed as categorical variables and presented as percentages of participants in stack plots.
 The number of complete cases is displayed in the plot captions._
@@
 (B) Scoring of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8).
-(C) Scoring of the post-traumatic stress disorder (PTSD) clusters with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD clusters.
+(C) Scoring of the post-traumatic stress disorder (PTSD) domains with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD domains.
 (E) Scoring of the factors of the post-traumatic growth with the post-traumatic growth inventory (PTGI) tool._
@@
 Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
-By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder).
+By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden).
 Numbers of observations in the mental clusters are shown in the plot legend in (A)._
@@
 
-```{r fig-oneR, fig.width = mm_inch(suppl_figures$oneR$w), fig.height = mm_inch(suppl_figures$oneR$h), fig.cap = 'Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors.'}
+```{r fig-oneR, fig.width = mm_inch(suppl_figures$oneR$w), fig.height = mm_inch(suppl_figures$oneR$h), fig.cap = 'Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-oneR). Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors.__
+__Supplementary Figure S\@ref(fig:fig-oneR).
+Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident.__
 
-_Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters were trained in the training subset of the study cohort by the one-rule (oneR) algorithm.
+_Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters available during acute medical management of the mountain sport accident were trained in the training subset of the study cohort by the one-rule (oneR) algorithm.
 The cluster assignment was subsequently predicted for the test subset observations.
@@
 
-```{r fig-early-crf, fig.width = mm_inch(suppl_figures$early_crf$w), fig.height = mm_inch(suppl_figures$early_crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.'}
+```{r fig-crf, fig.width = mm_inch(suppl_figures$crf$w), fig.height = mm_inch(suppl_figures$crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical, accident- and accident consequence-related factors by the random forest algorithm.'}
 
-suppl_figures$early_crf$plot
+suppl_figures$crf$plot
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-early-crf). Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.__
+__Supplementary Figure S\@ref(fig:fig-crf).
+Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical, accident- and accident consequence-related factors by the random forest algorithm.__
 
-_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident-related factors available during early acute medical management of the accident was trained in the training subset of the study cohort._
+_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident- and accident consequence-related factors was trained in the training subset of the study cohort._
 

### 2023-03-31 paper/markdown/supplementary_material.Rmd
Footnotes, Table S2

@@
   width(2, width = 10.5, unit = 'cm') %>% 
+  footnote(1, 1:2, 
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; BRCS: brief resilent coping scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
   theme_vanilla

### 2023-04-18 classification scripts/early_mix.R
Cluster definition by mixture modelling
Trial of mental cluster definition with a non-parametric EM mixture model: PTB and PTG clusters are similar, hence not included in the manuscript

@@ -1,0 +1,158 @@
+# Development of a random forest classifier employing solely 'early' 
+# explanatory variables, i.e. features available during an accident 
+# and acute medical management
+
+  insert_head()
+  
+# container ------
+  
+  mix_early <- list()
+  
+# parallel backend ------
+  
+  plan('multisession')
+
+# tuning of the classifier -------
+  
+  insert_msg('Tuning')
+  
+  ## finding the optimal mtry parameter based on OOB prediction accuracy
+  
+  set.seed(1234)
+  
+  mix_early$tune_grid <- 
+    expand.grid(mtry = 2:10, 
+                teststat = c('quad', 'max'), 
+                testtype = c('Teststatistic'), 
+                mincriterion = c(0, 0.95, qnorm(0.9)), 
+                minsplit = c(10, 20), 
+                stringsAsFactors = FALSE)
+  
+  mix_early$tuning$controls <- mix_early$tune_grid %>% 
+    pmap(cforest_control, 
+         ntree = 1000)
+  
+  mix_early$tuning$models <- mix_early$tuning$controls %>% 
+    future_map(~cforest(formula = class_globals$early_formula$mix, 
+                        data = class_globals$analysis_tbl$training, 
+                        controls = .x), 
+               .options = furrr_options(seed = TRUE))
+  
+  ## predictions and fit stats
+  
+  mix_early$tuning$predictions <- mix_early$tuning$models %>% 
+    map(~data.frame(obs = class_globals$analysis_tbl$training$clust_mix, 
+                    pred = predict(.x, OOB = TRUE) %>% 
+                      factor(levels = levels(class_globals$analysis_tbl$training$clust_mix))))
+  
+  mix_early$tuning$fit_stats <- mix_early$tuning$predictions %>% 
+    map(multiClassSummary, 
+        lev = levels(class_globals$analysis_tbl$training$clust_mix)) %>% 
+    map(as.list) %>% 
+    map_dfr(as_tibble)
+  
+  mix_early$tuning$fit_stats <- 
+    cbind(mix_early$tuning$fit_stats, 
+          mix_early$tune_grid) %>% 
+    as_tibble
+  
+  mix_early$tuning$models <- NULL
+  
+  mix_early <- compact(mix_early)
+  
+  plan('sequential')
+  
+  gc()
+  
+# Plotting the tuning results ------
+  
+  insert_msg('Plottng the tuning results')
+  
+  ## accuracy, kappa, sensitivity and specificity
+  
+  mix_early$tune_plots <- 
[…]

### 2023-04-18 classification scripts/early_pam.R

@@ -1,0 +1,158 @@
+# Development of a random forest classifier employing solely 'early' 
+# explanatory variables, i.e. features available during an accident 
+# and acute medical management
+
+  insert_head()
+  
+# container ------
+  
+  pam_early <- list()
+  
+# parallel backend ------
+  
+  plan('multisession')
+
+# tuning of the classifier -------
+  
+  insert_msg('Tuning')
+  
+  ## finding the optimal mtry parameter based on OOB prediction accuracy
+  
+  set.seed(1234)
+  
+  pam_early$tune_grid <- 
+    expand.grid(mtry = 2:10, 
+                teststat = c('quad', 'max'), 
+                testtype = c('Teststatistic'), 
+                mincriterion = c(0, 0.95, qnorm(0.9)), 
+                minsplit = c(10, 20), 
+                stringsAsFactors = FALSE)
+  
+  pam_early$tuning$controls <- pam_early$tune_grid %>% 
+    pmap(cforest_control, 
+         ntree = 1000)
+  
+  pam_early$tuning$models <- pam_early$tuning$controls %>% 
+    future_map(~cforest(formula = class_globals$early_formula$pam, 
+                        data = class_globals$analysis_tbl$training, 
+                        controls = .x), 
+               .options = furrr_options(seed = TRUE))
+  
+  ## predictions and fit stats
+  
+  pam_early$tuning$predictions <- pam_early$tuning$models %>% 
+    map(~data.frame(obs = class_globals$analysis_tbl$training$clust_pam, 
+                    pred = predict(.x, OOB = TRUE) %>% 
+                      factor(levels = levels(class_globals$analysis_tbl$training$clust_pam))))
+  
+  pam_early$tuning$fit_stats <- pam_early$tuning$predictions %>% 
+    map(multiClassSummary, 
+        lev = levels(class_globals$analysis_tbl$training$clust_pam)) %>% 
+    map(as.list) %>% 
+    map_dfr(as_tibble)
+  
+  pam_early$tuning$fit_stats <- 
+    cbind(pam_early$tuning$fit_stats, 
+          pam_early$tune_grid) %>% 
+    as_tibble
+  
+  pam_early$tuning$models <- NULL
+  
+  pam_early <- compact(pam_early)
+  
+  plan('sequential')
+  
+  gc()
+  
+# Plotting the tuning results ------
+  
+  insert_msg('Plottng the tuning results')
+  
+  ## accuracy, kappa, sensitivity and specificity
+  
+  pam_early$tune_plots <- 
[…]

### 2023-04-18 classification scripts/full_mix.R

@@ -1,0 +1,158 @@
+# Development of a random forest classifier employing solely 'early' 
+# explanatory variables, i.e. features available during an accident 
+# and acute medical management
+
+  insert_head()
+  
+# container ------
+  
+  mix_full <- list()
+  
+# parallel backend ------
+  
+  plan('multisession')
+
+# tuning of the classifier -------
+  
+  insert_msg('Tuning')
+  
+  ## finding the optimal mtry parameter based on OOB prediction accuracy
+  
+  set.seed(1234)
+  
+  mix_full$tune_grid <- 
+    expand.grid(mtry = 2:10, 
+                teststat = c('quad', 'max'), 
+                testtype = c('Teststatistic'), 
+                mincriterion = c(0, 0.95, qnorm(0.9)), 
+                minsplit = c(10, 20), 
+                stringsAsFactors = FALSE)
+  
+  mix_full$tuning$controls <- mix_full$tune_grid %>% 
+    pmap(cforest_control, 
+         ntree = 1000)
+  
+  mix_full$tuning$models <- mix_full$tuning$controls %>% 
+    future_map(~cforest(formula = class_globals$formula$mix, 
+                        data = class_globals$analysis_tbl$training, 
+                        controls = .x), 
+               .options = furrr_options(seed = TRUE))
+  
+  ## predictions and fit stats
+  
+  mix_full$tuning$predictions <- mix_full$tuning$models %>% 
+    map(~data.frame(obs = class_globals$analysis_tbl$training$clust_mix, 
+                    pred = predict(.x, OOB = TRUE) %>% 
+                      factor(levels = levels(class_globals$analysis_tbl$training$clust_mix))))
+  
+  mix_full$tuning$fit_stats <- mix_full$tuning$predictions %>% 
+    map(multiClassSummary, 
+        lev = levels(class_globals$analysis_tbl$training$clust_mix)) %>% 
+    map(as.list) %>% 
+    map_dfr(as_tibble)
+  
+  mix_full$tuning$fit_stats <- 
+    cbind(mix_full$tuning$fit_stats, 
+          mix_full$tune_grid) %>% 
+    as_tibble
+  
+  mix_full$tuning$models <- NULL
+  
+  mix_full <- compact(mix_full)
+  
+  plan('sequential')
+  
+  gc()
+  
+# Plotting the tuning results ------
+  
+  insert_msg('Plottng the tuning results')
+  
+  ## accuracy, kappa, sensitivity and specificity
+  
+  mix_full$tune_plots <- 
[…]

### 2023-04-18 classification scripts/full_pam.R

@@ -1,0 +1,158 @@
+# Development of a random forest classifier employing all 
+# explanatory variables, i.e. features available during an accident 
+# and acute medical management and during the recovery
+
+  insert_head()
+  
+# container ------
+  
+  pam_full <- list()
+  
+# parallel backend ------
+  
+  plan('multisession')
+
+# tuning of the classifier -------
+  
+  insert_msg('Tuning')
+  
+  ## finding the optimal mtry parameter based on OOB prediction accuracy
+  
+  set.seed(1234)
+  
+  pam_full$tune_grid <- 
+    expand.grid(mtry = 2:10, 
+                teststat = c('quad', 'max'), 
+                testtype = c('Teststatistic'), 
+                mincriterion = c(0, 0.95, qnorm(0.9)), 
+                minsplit = c(10, 20), 
+                stringsAsFactors = FALSE)
+  
+  pam_full$tuning$controls <- pam_full$tune_grid %>% 
+    pmap(cforest_control, 
+         ntree = 1000)
+  
+  pam_full$tuning$models <- pam_full$tuning$controls %>% 
+    future_map(~cforest(formula = class_globals$formula$pam, 
+                        data = class_globals$analysis_tbl$training, 
+                        controls = .x), 
+               .options = furrr_options(seed = TRUE))
+  
+  ## predictions and fit stats
+  
+  pam_full$tuning$predictions <- pam_full$tuning$models %>% 
+    map(~data.frame(obs = class_globals$analysis_tbl$training$clust_pam, 
+                    pred = predict(.x, OOB = TRUE) %>% 
+                      factor(levels = levels(class_globals$analysis_tbl$training$clust_pam))))
+  
+  pam_full$tuning$fit_stats <- pam_full$tuning$predictions %>% 
+    map(multiClassSummary, 
+        lev = levels(class_globals$analysis_tbl$training$clust_pam)) %>% 
+    map(as.list) %>% 
+    map_dfr(as_tibble)
+  
+  pam_full$tuning$fit_stats <- 
+    cbind(pam_full$tuning$fit_stats, 
+          pam_full$tune_grid) %>% 
+    as_tibble
+  
+  pam_full$tuning$models <- NULL
+  
+  pam_full <- compact(pam_full)
+  
+  plan('sequential')
+  
+  gc()
+  
+# Plotting the tuning results ------
+  
+  insert_msg('Plottng the tuning results')
+  
+  ## accuracy, kappa, sensitivity and specificity
+  
+  pam_full$tune_plots <- 
[…]

### 2023-04-18 classification.R

@@ -1,5 +1,5 @@
 # Building a classifier (random forest) to identify the mental health cluster
-# with help of socioeconomic, demographic, clnical and accident-related factors
+# with help of socioeconomic, demographic, clinical and accident-related factors
 
 # tools ------
 
@@ -48,8 +48,13 @@
   
   ## full model formula
   
-  class_globals$formula <-  
-    as.formula(paste('clust_id ~ ', 
+  class_globals$formula$pam <-  
+    as.formula(paste('clust_pam ~ ', 
+                     paste(class_globals$variables, 
+                           collapse = '+')))
+  
+  class_globals$formula$mix <-  
+    as.formula(paste('clust_mix ~ ', 
                      paste(class_globals$variables, 
                            collapse = '+')))
   
@@ -66,32 +71,42 @@
                                                             'flashback_frequency', 
                                                             'confusion_during_sport')]
   
-  ## early model formula
+  ## early model formulas
   
-  class_globals$early_formula <- 
-    as.formula(paste('clust_id ~ ', 
+  class_globals$early_formula$pam <- 
+    as.formula(paste('clust_pam ~ ', 
+                     paste(class_globals$early_variables, 
+                           collapse = '+')))
+  
+  class_globals$early_formula$mix <- 
+    as.formula(paste('clust_mix ~ ', 
                      paste(class_globals$early_variables, 
                            collapse = '+')))
   
   ## cluster assignment schemes
   
-  class_globals$clust_assignment <- semi_clust$clust_obj %>% 
+  class_globals$pam_assignment <- semi_clust$clust_obj %>% 
     map(~.x$clust_assignment) %>% 
-    map(set_names, c('ID', 'clust_id'))
+    map(set_names, c('ID', 'clust_pam'))
+  
+  class_globals$mix_assignment <- semi_mix$clust_obj %>% 
+    map(~.x$clust_assignment) %>% 
+    map(set_names, c('ID', 'clust_mix'))
   
   ## analysis table, splitting into the training and test subset
   ## appending with the cluster assignment scheme
   
   class_globals$analysis_tbl <- ptsd$dataset %>% 
-    dlply('partition', 
-          select, 
-          ID, 
-          any_of(class_globals$variables)) %>% 
-    map(as_tibble)
-  
+    blast(partition) %>% 
+    map(select, 
+        ID, 
+        any_of(class_globals$variables))
+
[…]

### 2023-04-18 clustering scripts/bnp_development.R

@@ -1,0 +1,57 @@
+# Clustering with the BNP (Bayesian non-parametric) EM model
+# likely a bug in teh code (package is not at CRAN anymore)
+
+  insert_head()
+  
+# tools -----
+  
+  library(BNPMIXcluster)
+  
+# container -----
+  
+  clust_bnp <- list()
+  
+# analysis variables and data -----
+  
+  insert_msg('Analysis globals')
+  
+  clust_bnp$variables <- clust_globals$variables
+  
+  clust_bnp$var_types <- 
+    c(dsm5_total = 'c', 
+      dsm5_B = 'c', 
+      dsm5_C = 'c', 
+      dsm5_D = 'c', 
+      dsm5_E = 'c', 
+      ptgi_total = 'c', 
+      ptgi_fctI = 'c', 
+      ptgi_fctII = 'c', 
+      ptgi_fctIII = 'c', 
+      ptgi_fctIV = 'c', 
+      ptgi_fctV = 'c', 
+      rs13_total = 'c', 
+      soc9l_total = 'c', 
+      phq9_total = 'c', 
+      gad7_total = 'c', 
+      phqd_panic_total = 'c', 
+      phq_events_total = 'c', 
+      eurohis_total = 'c', 
+      eurohis_qol = 'o', 
+      eurohis_health = 'o', 
+      eurohis_energy = 'o', 
+      eurohis_finances = 'o', 
+      eurohis_activity = 'o', 
+      eurohis_selfesteem = 'o', 
+      eurohis_relationship = 'o', 
+      eurohis_housing = 'o')
+  
+  clust_bnp$analysis_tbl <- clust_globals$analysis_tbl$training
+  
+# Modeling -------
+  
+  insert_msg('Modeling')
+  
+  clust_bnp$models <- MIXclustering(Y = as.matrix(clust_bnp$analysis_tbl), 
+                                    var_type = clust_bnp$var_types)
+  
+# END ------

### 2023-04-18 clustering scripts/mix_background.R

@@ -1,0 +1,179 @@
+# Differences in levels of demographic, socioeconomic and clinical factors 
+# as well as variables related to the accident between the clusters
+
+  insert_head()
+  
+# container -------
+  
+  mix_bcg <- list()
+  
+# parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## explanatory variables 
+  
+  mix_bcg$variables <- ptsd$var_lexicon %>% 
+    filter(type %in% c('characteristic')) %>% 
+    .$variable
+
+  ## cluster assignment schemes
+  
+  mix_bcg$clust_assignment <- semi_mix$clust_obj %>% 
+    map(~.x$clust_assignment) %>% 
+    map(set_names, c('ID', 'clust_id'))
+  
+  ## analysis table, splitting into the training and test subset
+  ## appending with the clust assignment scheme
+  
+  mix_bcg$analysis_tbl <- ptsd$dataset %>% 
+    blast(partition) %>% 
+    map(select, 
+        ID, 
+        all_of(mix_bcg$variables))
+  
+  mix_bcg$analysis_tbl <- 
+    map2(mix_bcg$analysis_tbl, 
+         mix_bcg$clust_assignment, 
+         right_join, by = 'ID') %>% 
+    map(~map_dfc(.x, unname))
+  
+  ## test types
+  
+  mix_bcg$test_types <- 
+    mix_bcg$analysis_tbl$training[mix_bcg$variables] %>% 
+    map_lgl(is.numeric) %>% 
+    ifelse('kruskal_eta', 'cramer_v')
+
+# Descriptive stats ------  
+  
+  insert_msg('Descriptive stats')
+  
+  mix_bcg$desc_stats <- mix_bcg$analysis_tbl %>% 
+    future_map(explore, 
+               split_factor = 'clust_id', 
+               variables = mix_bcg$variables, 
+               what = 'table', 
+               pub_styled = TRUE, 
+               .options = furrr_options(seed = TRUE))
+  
+# Testing ------
+  
+  insert_msg('Testing for differences between the clusters')
+  
+  ## all variables are categorical: chi-squared test with Cramer V effect size
+  
+  mix_bcg$test <- mix_bcg$analysis_tbl %>% 
+    future_map(compare_variables, 
+               variables = mix_bcg$variables, 
+               split_factor = 'clust_id', 
+               what = 'eff_size', 
[…]

### 2023-04-18 clustering scripts/mix_characteristic.R

@@ -1,0 +1,304 @@
+# Differences in clustering variables between the mental clusters
+
+  insert_head()
+  
+# container -------
+  
+  feat_mix <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  feat_mix$variables <- 
+    c(clust_globals$variables, 
+      ptsd$mental_variables[!ptsd$mental_variables %in% clust_globals$variables], 
+      'dsm5_cluster_class', 
+      'dsm5_B_class', 
+      'dsm5_C_class', 
+      'dsm5_D_class', 
+      'dsm5_E_class', 
+      'rs13_total_class', 
+      'phq9_total_class', 
+      'phq_events_total_class', 
+      'gad7_total_class')
+  
+  ## clustering objects 
+  
+  feat_mix$clust_obj <- semi_mix$clust_obj
+  
+  ## analysis tables: identity and normalized
+  
+  feat_mix$norm_tbl <- feat_mix$clust_obj %>% 
+    map(model.frame) %>% 
+    map(rownames_to_column, 'ID') %>% 
+    map2(., feat_mix$clust_obj, 
+         ~left_join(.x, 
+                    set_names(.y$clust_assignment, c('ID', 'clust_id')), 
+                    by = 'ID')) %>% 
+    map(as_tibble) %>% 
+    map(~map_dfc(.x, unname))
+  
+  feat_mix$ident_tbl <- ptsd$dataset %>% 
+    dlply('partition', 
+          select, 
+          ID, 
+          all_of(feat_mix$variables)) %>% 
+    map2(., feat_mix$clust_obj, 
+         ~left_join(.x, 
+                    set_names(.y$clust_assignment, c('ID', 'clust_id')), 
+                    by = 'ID')) %>% 
+    map(as_tibble) %>% 
+    map(~map_dfc(.x, unname))  
+  
+  ## test types
+  
+  feat_mix$test_types <- feat_mix$ident_tbl$training[feat_mix$variables] %>% 
+    map_lgl(is.numeric)
+  
+  feat_mix$test_types <- ifelse(feat_mix$test_types, 
+                                  'kruskal_eta', 
+                                  'cramer_v')
+  
+# Descriptive stats -------
+  
+  insert_msg('Descriptive stats')
+  
+  feat_mix$desc_stats <- feat_mix$ident_tbl %>% 
+    map(explore, 
+        variables = feat_mix$variables, 
+        split_factor = 'clust_id', 
[…]

### 2023-04-18 clustering scripts/mix_clustering.R

@@ -1,0 +1,184 @@
+# Assignment of the test cohort samples to clusters developed with 
+# mixture modeling by 7-NN inverse distance weighted classifier
+# Characteristic of the clustering structures in the training and test subset
+
+  insert_head()
+  
+# container -----
+  
+  semi_mix <- list()
+  
+# Analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  semi_mix$analysis_tbl <- clust_globals$analysis_tbl
+  
+# Training and predictions ------
+  
+  insert_msg('Training and testing')
+  
+  semi_mix$clust_obj$training <- clust_em$clust_obj
+  
+  semi_mix$clust_obj$test <- 
+    predict(semi_mix$clust_obj$training, 
+            newdata = semi_mix$analysis_tbl$test, 
+            type = 'propagation', 
+            kNN = 7, 
+            resolve_ties = TRUE, 
+            simple_vote = FALSE)
+  
+  semi_mix$clust_obj$test$clust_assignment <- 
+    semi_mix$clust_obj$test$clust_assignment %>% 
+    mutate(clust_id = factor(clust_id, 
+                             levels(semi_mix$clust_obj$training$clust_assignment$clust_id)))
+  
+# Clustering variances ------
+  
+  insert_msg('Clustering variances')
+  
+  semi_mix$variance <- semi_mix$clust_obj %>% 
+    map(var) %>% 
+    map(~.x$frac_var) %>% 
+    compress(names_to = 'partition', 
+             values_to = 'variance') %>% 
+    map_dfc(unlist)
+  
+  ## plotting the variance values
+  
+  semi_mix$variance_plot <- semi_mix$variance %>% 
+    ggplot(aes(x = variance, 
+               y = reorder(partition, variance), 
+               fill = partition)) + 
+    geom_bar(stat = 'identity', 
+             color = 'gray20') + 
+    geom_text(aes(label = signif(variance, 2)), 
+              size = 2.5, 
+              hjust = 1.4, 
+              color = 'white') + 
+    scale_fill_manual(values = c(training = 'steelblue4', 
+                                 test = 'coral4'), 
+                      name = 'Subset') +
+    guides(fill = 'none') + 
+    scale_x_continuous(limits = c(0, 1)) + 
+    globals$common_theme + 
+    theme(axis.title.y = element_blank()) + 
+    labs(title = 'Semi-supervised clustering', 
+         x = 'Fraction explained clustering variance')
+  
+# Cluster percentages --------
[…]

### 2023-04-18 clustering scripts/mix_development.R

@@ -1,0 +1,248 @@
+# A model-based approach employing non-parametric EM finite mixture modeling
+
+  insert_head()
+
+# container ------
+  
+  clust_em <- list()
+  
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## analysis table
+  
+  clust_em$analysis_tbl <- clust_globals$analysis_tbl$training
+  
+  ## number of variables and observations
+  
+  clust_em$n_numbers[c('observations', 'variables')] <- 
+    list(nrow, ncol) %>% 
+    map(~.x(clust_em$analysis_tbl))
+  
+  ## range of class numbers
+  
+  clust_em$class_no <- 1:10
+  
+# tuning: optimal class number ------
+  
+  insert_msg('Optimal class number')
+  
+  set.seed(1234)
+  
+  ## models
+  
+  clust_em$tuning$models <- clust_em$class_no %>% 
+    future_map(~npEM(x = as.matrix(clust_em$analysis_tbl), 
+                     mu0 = .x, 
+                     #samebw = FALSE, 
+                     verb = FALSE), 
+               .options = furrr_options(seed = TRUE)) %>% 
+    set_names(paste0('class_', clust_em$class_no))
+  
+  ## log-likelihood (final) and BIC
+
+  clust_em$tuning$stats <- clust_em$tuning$models %>% 
+    map(~.x$loglik * -1) %>% 
+    map_dbl(~.x[length(.x)]) %>% 
+    compress(names_to = 'k', 
+             values_to = 'loglik') %>% 
+    mutate(k = stri_extract(k, regex = '\\d+'), 
+           k = as.numeric(k), 
+           bic = 2 * loglik + log(clust_em$n_numbers$observations) * clust_em$n_numbers$variables)
+  
+  ## plotting the BIC against the cluster number
+  
+  clust_em$tuning$plot <- clust_em$tuning$stats %>% 
+    ggplot(aes(x = k, 
+               y = bic)) + 
+    geom_vline(xintercept = 3, 
+               linetype = 'dashed') + 
+    geom_path(color ='steelblue') + 
+    scale_x_continuous(breaks = clust_em$class_no) + 
+    globals$common_theme + 
+    labs(title = 'EM model fit', 
+         x = 'Number of clusters, k', 
+         y = 'BIC', 
[…]

### 2023-04-18 clustering.R

@@ -21,8 +21,11 @@
   library(psych)
   library(clustTools)
   library(ggrepel)
-  
+  library(mixtools)
+
   insert_head()
+  
+  set_rownames <- trafo::set_rownames
   
   source_all('./tools/tools.R', 
              message = TRUE, 
@@ -38,14 +41,17 @@
   
   clust_globals$variables <- ptsd$mental_variables
   
-  ## normalized, median-centered psychometric data
+  ## date: no transformation, 
+  ## analysis table: normalized, median-centered psychometric data
 
-  clust_globals$analysis_tbl <- ptsd$dataset %>% 
-    dlply('partition', as_tibble) %>% 
+  clust_globals$data <- ptsd$dataset %>% 
+    blast(partition) %>% 
     map(select, ID, all_of(clust_globals$variable)) %>% 
-    map(column_to_rownames, 'ID') %>% 
-    map(center_data, type = 'median')
+    map(column_to_rownames, 'ID')
   
+  clust_globals$analysis_tbl <- clust_globals$data %>% 
+    map(center_data, type = 'median')
+
 # clustering analysis scripts -----
   
   insert_msg('Clustering analysis scripts')
@@ -53,7 +59,9 @@
   c('./clustering scripts/development.R', 
     './clustering scripts/clustering.R', 
     './clustering scripts/characteristic.R', 
-    './clustering scripts/background.R') %>% 
+    './clustering scripts/background.R', 
+    './clustering scripts/mix_development.R', 
+    './clustering scripts/mix_clustering.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
 # END ------

### 2023-04-18 paper scripts/figures.R

@@ -428,7 +428,7 @@
   
   insert_msg('Figure 7: Mental cluster classification, cRF, early preds')
   
-  figures$early_crf <- class_early$confusion_plots$global %>% 
+  figures$early_crf <- pam_early$confusion_plots %>% 
     map2(., 
          paste('Cluster classification,', 
                c('training', 'test')), 
@@ -438,7 +438,7 @@
     plot_grid(plotlist = ., 
               nrow = 2, 
               align = 'hv') %>% 
-    plot_grid(class_early$importance_plots$global + 
+    plot_grid(pam_early$importance_plot + 
                 theme(plot.title.position = 'plot', 
                       plot.title = element_text(hjust = 0.3)), 
               ncol = 2, 

### 2023-04-18 paper scripts/render.R

@@ -2,6 +2,12 @@
 
   insert_head()
   
+# reading the bibliography -------
+  
+  insert_msg('Reading the bibliography')
+  
+  alpine_bib <- read_bib('./paper/markdown/ptsd_biblio.bib')
+  
 # supplementary material ------
   
   insert_msg('Rendering the supplements')

### 2023-04-18 paper scripts/supplementary.R

@@ -236,43 +236,12 @@
                               'in the mental clusters.'), 
               w = 180, 
               h = 220)
+
+# Figure S7: mental cluster classification, cRF -------
   
-# Figure S7: OneR, accuracy and kappa -------
-  
-  insert_msg('Figure S7: oneR')
-  
-  suppl_figures$oneR <- 
-    class_one$plots$global[c("Accuracy", "Kappa")] %>% 
-    map2(., 
-         c('Cluster classification, single factors, accuracy', 
-           'Cluster classification, single factors, kappa'), 
-         ~.x + 
-           labs(title = .y) +
-           theme(legend.position = 'none', 
-                 plot.title.position = 'plot', 
-                 plot.title = element_text(hjust = 0.3))) %>% 
-    plot_grid(plotlist = ., 
-              ncol = 2, 
-              align = 'hv', 
-              axis = 'tblr') %>% 
-    plot_grid(get_legend(class_one$plots$global[[1]] + 
-                           theme(legend.position = 'bottom')), 
-              nrow = 2, 
-              rel_heights = c(0.9, 0.1)) %>% 
-    as_figure(label = 'figure_s7_cluster_classification_single_factors', 
-              ref_name = 'oneR', 
-              caption = paste('Prediction of the mental cluster assignment', 
-                              'by single demographic, socioeconomic,', 
-                              'clinical and accident-related factors.'), 
-              w = 180, 
-              h = 120)
-  
-  
-# Figure S8: mental cluster classification, cRF -------
-  
-  insert_msg('Figure S8: Mental cluster classification, cRF')
-  
-  suppl_figures$crf <- class_rf$confusion_plots$global %>% 
+  insert_msg('Figure S7: Mental cluster classification, cRF')
+
+  suppl_figures$crf <- pam_full$confusion_plots %>% 
     map2(., 
          paste('Cluster classification,', 
                c('training', 'test')), 
@@ -282,19 +251,20 @@
     plot_grid(plotlist = ., 
               nrow = 2, 
               align = 'hv') %>% 
-    plot_grid(class_rf$importance_plots$global + 
+    plot_grid(pam_full$importance_plot + 
                 theme(plot.title.position = 'plot', 
                       plot.title = element_text(hjust = 0.3)), 
               ncol = 2, 
               rel_widths = c(0.9, 1.1), 
               labels = LETTERS, 
               label_size = 10) %>% 
-    as_figure(label = 'figure_s8_rf_classifier', 
[…]

### 2023-04-18 tools/tools.R

@@ -843,6 +843,34 @@
       theme_void()
     
   }
+  
+# Variable importance ------
+  
+  plot_crf_importance <- function(data, 
+                                  plot_title = NULL, 
+                                  plot_subtitle = NULL, 
+                                  x_lab = '\u0394 accuracy', 
+                                  fill_color = 'cornsilk'){
+    
+    ## plots conditional random forest importance measures
+    
+    data %>% 
+      ggplot(aes(x = importance, 
+                 y = reorder(variable, importance))) + 
+      geom_bar(stat = 'identity', 
+               color = 'gray20', 
+               fill = fill_color) + 
+      scale_y_discrete(labels = exchange(class_globals$variables, 
+                                         dict = ptsd$var_lexicon, 
+                                         key = 'variable', 
+                                         value = 'label')) +
+      globals$common_theme + 
+      theme(axis.title.y = element_blank()) + 
+      labs(title = plot_title,
+           subtitle = plot_subtitle, 
+           x = x_lab)
+    
+  }
 
 # Labellers --------
   

### 2023-04-18 paper/markdown/manuscript.Rmd
Introduction, figures re-arranged
Results of OneR modeling excluded from the manuscript

@@
 
+Stress-related manifestations following an experience or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence amount to post-traumatic stress disorder (PTSD).
+PTSD symptoms include hyperarousal, panic, somatic symptoms, sleep disorders and unwilling recalls of the trauma (flashbacks) [@AmericanPsychiatricAssociation2013; @Bryant2019; @Howie2019] and are classified as chronic when lasting over three months [@Seedat2005].
+Of note, PTSD may overlap with signs of depression, anxiety and substance abuse [@Brady2000; @Gagnon-Sanschagrin2022]. 
+Although most individuals have experienced at least one traumatic event during the lifetime [@Benjet2016; @Breslau2004], the prevalence of PTSD in the general population was estimated for 6 - 9% [@VanAmeringen2008; @Kessler2005].
+Factors pre-disposing for PTSD include repeated trauma exposition e.g.
+in a family [@Cloitre2009] or professional setting [@Berger2012], female gender [@Christiansen2020] or young age [@Kongshoej2022].
+Yet, effects of early demographic or clinical predictors available during acute medical or psychiatric management of traumatic events on PTSD risk are weak [@Papini2018]. 
+Hence identification of PTSD-vulnerable population is challenging and hampers effective prevention measures [@Papini2018; @Galatzer-Levy2014]. 
+Post-traumatic growth describes a salutary adjustment psychological reaction to a traumatic experience including boosted attitude to life, self and relationship [@Tedeschi1996]. 
+It is disputable, if resilience, i.e.
+ability to adapt to stress or trauma [@AmericanPsychiatricAssociation2013], represents a pre-existing personality trait or a feature that can be modified by a potentially traumatic event [@Agaibi2005].
+The interplay between PTSD, post-traumatic growth and resilience is still unclear. 
+While resilience is considered to protect from PTSD development, symptoms of both PTSD and post-traumatic growth may co-occur in a subset of trauma affected individuals [@Levine2009; @Bensimon2012].
+
+While outdoor physical activity has been shown to foster mental health [@Niedermeier2017a; @Niedermeier2017], mountain sports are also associated with an elevated risk of injury [@Faulhaber2020].
+Physical injury frequently perceived as a life-threatening induces a response of fear [@Segal2010], and up to 26% of acute trauma injury patients were reported to develop PTSD [@Joseph2020]. 
+PTSD frequency in accident victims was found to be associated with head injuries, hospitalization length and pain [@Galatzer-Levy2014]. 
+Literature on mental distress following traumatic events related to mountain sports is scarce and discrepant. 
+In studies with mountain guides  and mountain rescue personnel, 71 - 78% of participants had experiences traumatic events [@Sommer2004; @Mikutta2022]. 
+Intriguingly, the frequency of PTSD symptoms ranged between 1 - 2.7% which was less than in the general population and, particularly, much lower than in other vulnerable groups such as emergency workers [@Sommer2004; @Mikutta2022; @Berger2012].
+Moreover, significantly higher resilience levels than in the general population were reported [@Mikutta2022]. 
+By contrast, PTSD symptom prevalence reported by Traber et al. in a similar collective of mountain professionals was 22% [@Traber2020]. 
+Concerning alpine accidents, survivors of avalanches were affected by acute and long-term PTSD symptoms such as hyperarousal or sleep disorders at rates between 11 - 16% [@Thordardottir2015; @Leonard2021].
+Still, a more general psychological reaction to a mountain accident, which may trigger PTSD-like distress, post-traumatic growth or resilience boost, has not been yet characterized.
+
+Herein, we sought to explore patterns of mental health in a collective of `r nrow(ptsd$dataset)` mountain accident victims in Tyrol, Austria, screened for signs of PTSG, post-traumatic growth, deteriorating quality of life and major psychiatric problems.
+Additionally, we searched for early demographic, socioeconomic and clinical predictors of mental health disorders following a mountain accidents which may help to tailor early preventive psychiatric interventions.
+
 \newpage
@@
 
-The uni- and multi-parameter classifiers were trained with the one-rule [@VonJouanne-Diedrich2017; @Holte1993] and conditional random forest algorithms [@Hothorn2006; @Strobl2007; @Hothorn2022], respectively.
+Multi-parameter classifiers were trained with the conditional random forest algorithms [@Hothorn2006; @Strobl2007; @Hothorn2022]. 
 The cluster assignment was predicted for the test subset and assessed by the accuracy and $\kappa$ statistics [@Kuhn2008]. 
-Conditional variable importance in the random forest classifiers was expressed as accuracy loss [@Strobl2008].
+Permutation variable importance in the random forest classifiers was expressed as accuracy loss [@Strobl2008].
 
@@
 Three mental clusters: 'neutral', 'PTG' (post-traumatic growth) and 'PTB' (post-traumatic burden), named after their key mental characteristic were identified in the training subset. 
-Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`), comparable cluster sizes and good visual clusetr separation in both study cohort subsets (__Supplementary Figures S2B__ - __S4__).
+Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`), comparable cluster sizes and good visual cluster separation in both study cohort subsets (__Supplementary Figures S2B__ - __S4__).
 
@@
 
-Predictive value [@VonJouanne-Diedrich2017; @Holte1993] of single early explanatory parameters was poor (maximum accuracy: `r class_one$fit_stats$global$Accuracy %>% max %>% signif(2)`, maximum $\kappa$: `r class_one$fit_stats$global$Kappa %>% max %>% signif(2)`) and inconsistent between the subsets of the study cohort.
-In such univariable analysis, age class, annual income and traumatic events in the past were the most predictive factors (__Supplementary Figure S7__).
-
-To account for cumulative effects and interactions of multiple weak predictors of the mental cluster assignment, we resorted to conditional random forest modeling [@Strobl2007; @Hothorn2006; @Hothorn2022].
-This multi-parameter model employing early candidate cluster predictors could correctly assign `r signif(class_early$fit_stats$global$Accuracy[1], 2) * 100`% training subset observations to their mental clusters
-($\kappa$ = `r signif(class_early$fit_stats$global$Kappa[1], 2)`). 
-However, its predictive performance in the test subset was poor (accuracy: `r signif(class_early$fit_stats$global$Accuracy[2], 2) * 100`, $\kappa$ = `r signif(class_early$fit_stats$global$Kappa[2], 2)`).
+This multi-parameter conditional random forest model [@Strobl2007; @Hothorn2006; @Hothorn2022; @Gagnon-Sanschagrin2022] employing early candidate cluster predictors could correctly assign `r signif(pam_early$fit_stats$Accuracy[1], 2) * 100`% training subset observations to their mental clusters
+($\kappa$ = `r signif(pam_early$fit_stats$Kappa[1], 2)`). 
+However, its predictive performance in the test subset was poor (accuracy: 
+`r signif(pam_early$fit_stats$Accuracy[2], 2) * 100`, 
+$\kappa$ = `r signif(pam_early$fit_stats$Kappa[2], 2)`). 
 The prediction quality was the best for the PTG cluster (accuracy, training: 89.5%, test: 59.4%, __Figure \@ref(fig:fig-early-crf)A__).
 The most important early explanatory variables for the cluster prediction by conditional random forests were annual income, sex, mode of rescue, leg injury and pre-existing mental illness (__Figure \@ref(fig:fig-early-crf)B__).
-Inclusion of late accident consequences such as cautious behavior during sport, flashbacks or somatic health burden of the accident in the random forest model could only marginally improve its performance (training: accuracy = `r signif(class_rf$fit_stats$global$Accuracy[1], 2)`, $\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[1], 2)`; test subset: accuracy = `r signif(class_rf$fit_stats$global$Accuracy[2], 2)`, $\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[2], 2)`).
-The most important late predictors of the cluster assignment were caution during sport and flashbacks (__Supplementary Figure S8__).
+Inclusion of late accident consequences such as cautious behavior during sport, flashbacks or somatic health burden of the accident in the random forest model could only marginally improve its performance (training: accuracy =
+`r signif(pam_full$fit_stats$Accuracy[1], 2)`, 
+$\kappa$ = `r signif(pam_full$fit_stats$Kappa[1], 2)`; test subset: accuracy = `r signif(pam_full$fit_stats$Accuracy[2], 2)`, 
+$\kappa$ = `r signif(pam_full$fit_stats$Kappa[2], 2)`). 
+The most important late predictors of the cluster assignment were caution during sport and flashbacks (__Supplementary Figure S7__).
 

### 2023-04-18 paper/markdown/supplementary_material.Rmd
Introduction, figures re-arranged
Results of OneR modeling excluded from the manuscript

@@
 
-## Uni- and multi-variable cluster assignment classifiers
-
-Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters (__Supplementary Table S\@ref(tab:tab-mod-variables)__) were trained in the training subset of the study cohort by the one-rule (oneR) algorithm (function `OneR()`, package _OneR__) [@VonJouanne-Diedrich2017; @Holte1993].
-The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _OneR_).
+## Cluster assignment classifiers
 
@@
 Accuracy and $\kappa$ statistic for the classifiers' mental cluster assignment predictions in the training and test subsets were computed with the function `multiClassSummary()` from the package _caret_ [@Kuhn2008].
-Conditional importance for explanatory variables expressed as accuracy loss [@Strobl2008] was obtained for the conditional random forest algorithm with the function 'varimp()` from the _party_ package.
+Permutation importance for explanatory variables expressed as accuracy loss [@Strobl2008] was obtained for the conditional random forest algorithm with the function 'varimp()` from the _party_ package.
 
@@
 
-```{r fig-oneR, fig.width = mm_inch(suppl_figures$oneR$w), fig.height = mm_inch(suppl_figures$oneR$h), fig.cap = 'Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident.'}
-
-suppl_figures$oneR$plot
-
-```
-
-__Supplementary Figure S\@ref(fig:fig-oneR).
-Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident.__
-
-_Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters available during acute medical management of the mountain sport accident were trained in the training subset of the study cohort by the one-rule (oneR) algorithm.
-The cluster assignment was subsequently predicted for the test subset observations.
-Accuracy and $\kappa$ of the classifiers with any predictive value ($\kappa$ > 0 in both subsets) are shown in bar plots._
-
-\newpage
-
 ```{r fig-crf, fig.width = mm_inch(suppl_figures$crf$w), fig.height = mm_inch(suppl_figures$crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical, accident- and accident consequence-related factors by the random forest algorithm.'}

### 2023-08-31 paper/markdown/manuscript.Rmd
Author's feedback
Incorporation of the most recent author's feedback into the manuscript text and Supplementary Material

--- 2023-03-30 paper/markdown/manuscript.Rmd
@@
 ---
-title: "Three distinct patterns of mental response after mountain sport accidents"
+title: "Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center"
 subtitle: "Manuscript"
@@
 author: 
+  - Hanna Salvotti, MD:
+      institute: [psych, regensburg]
+  - Piotr Tymoszuk, PhD:
+      institute: [daas]
+  - Mathias Ströhle, MD:
+      institute: [kufstein]
+  - Peter Paal, MD: 
+      institute: [salzburg, kurasi, himed, icar]
+  - Hermann Brugger, MD: 
+      institute: [eurac, ismm]
+  - Martin Faulhaber, PhD: 
+      institute: [sport]
+  - Nicola Kruijer, MD: 
+      institute: [trauma]
+  - Barbara Sperner-Unterweger, MD: 
+      institute: [psych]
   - Katharina Hüfner, MD:
@@
       correspondence: true
-  - Hanna Salvotti, MD:
-      institute: [psych]
-  - Piotr Tymoszuk, PhD:
-      institute: [daas]
 
 institute:
-  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria
+  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria.
+  - regensburg: Department of Neurosurgery, University Hospital of Regensburg, Regensburg, Germany
   - daas: Data Analysis as a Service Tirol, Innsbruck, Austria
+  - kufstein: Department of Anesthesiology and Critical Care Medicine, Bezirkskrankenhaus Kufstein, Kufstein, Austria.
+  - salzburg: Department of Anesthesiology and Intensive Care Medicine, Paracelsus Medical University, Salzburg, Austria.
+  - kurasi:  Austrian Board of Mountain Safety (Österreichisches Kuratorium fur Alpine Sicherheit), Innsbruck, Austria.
+  - himed: Austrian Society of Mountain and High-Altitude Medicine, Mieming, Austria. 
+  - icar: International Commission for Mountain Emergency Medicine (ICAR), Kloten, Switzerland.
+  - eurac: EURAC Research, Institute for Mountain Emergency Medicine, Bolzano/Bozen, Italy.
+  - ismm: International Society of Mountain Medicine (ISMM).
+  - sport: Department of Sport Science, University of Innsbruck, Innsbruck, Austria.
+  - trauma: Department of Orthopedics and Traumatology, Medical University of Innsbruck, Innsbruck, Austria.
 
@@
 
-__Background:__ Psychological consequences of mountain sport accidents are poorly characterized. 
-We investigated patterns of mental response to mountain sport accidents and their victims at risk of mental deterioration.
+__Background:__ Mountain sports are inherently linked to physical injury risk. 
+We aimed to characterize mental health following mountain sport accidents requiring professional medical management.
 
-__Methods:__ A cross-sectional survey was conducted with `r nrow(ptsd$dataset)` adult victims of mountain sport accidents treated at the Innsbruck University Hospital (Austria). 
-Anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data.
-Patterns of mental readouts were defined by semi-supervised medoid clustering and models of cluster assignment based on non-mental measures were developed by the random forest algorithm.
+__Methods:__ Adult victims of mountain sport accidents treated at the Innsbruck University Hospital (Austria) between 2008 and 2020 (n = `r nrow(ptsd$dataset)`) participated in a cross-sectional survey study at median `r round(median(ptsd$dataset$obs_time)/30.436)` months after the trauma center admission.
+Symptoms of anxiety, depression, somatization and panic disorder (PHQ), resilience (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS-QOL), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD, PCL-5), sociodemographic and clinical information were obtained from an on-line survey and electronic health records.
+Mental health patterns were investigated by semi-supervised medoid clustering and modeled by machine learning.
 
-__Results:__ Three mental clusters of accident victims were hallmarked by a neutral mental response, post-traumatic growth and post-traumatic burden of mental disorders, respectively. 
-Despite relatively frequent hospitalization, surgery, pre-existing mental or somatic illness, low income, and female gender in the post-traumatic growth and post-traumatic burden clusters, overall differences in non-mental variables between the clusters were weak.
-Consequently, no reliable model of the cluster assignment employing explanatory factors available during acute medical accident management could be established.
+__Results:__ At least one diagnostic criterion of PTSD was met by `r signif(table(ptsd$dataset$dsm5_cluster_class)[2]/nrow(ptsd$dataset), 2) * 100`% of participants. 
+Three comparably sized subsets of patients were identified: (1) neutral, (2) post-traumatic growth and (3) post-traumatic stress cluster.
+The post-traumatic stress cluster was characterized by frequent mental disorder symptoms, and low resilience, sense of coherence and quality of life as well as by low age, the highest frequency of pre-existing mental disorders and common physical health consequences of the accident.
+Reliability of machine learning-based prediction of the cluster assignment based on `r length(class_globals$variables$early)` variables available during acute treatment was limited.
 
-__Conclusion:__ A subset of mountain sport accident victims develops symptoms of anxiety, depression, post-traumatic stress disorder and other mental health problems. 
-Since identification of these vulnerable patients with demographic, socioeconomic, accident- and injury-related predictors remains challenging, early and low-threshold psychological support is key to prevention of mental disorders following mountain sport accidents.
+__Conclusion:__ A subset of mountain sport accident victims shows symptoms of mental health disorders including PTSD. 
+Since early identification of these vulnerable patients remains challenging, low-threshold psychological support is key for a successful interdisciplinary management of mountain sport accidents.
 
@@
 
-mountain sport accident, mental health, post-traumatic stress disorder, post-traumatic growth, clustering, random forest, machine learning
+mountain sport accident, mental health, post-traumatic stress disorder, post-traumatic growth, machine learning
 
@@
 # Introduction
+
+Physical activity in mountains, such as hiking, climbing, alpine skiing or ski touring, gain popularity, a trend that has been accelerated by the COVID-19 pandemic [@Hedenborg2022]. 
+Outdoor physical activity in general and mountain sport in particular exerts beneficial effects on mental health [@Ower2019; @Niedermeier2017a; @Niedermeier2017]. 
+However, mountain sports are also associated with the risk of injury [@Faulhaber2017; @Faulhaber2020], which could foster development of mental disorders including post-traumatic stress disorder (PTSD).
+
+PTSD symptoms include hyperarousal, panic, somatic symptoms, sleep disorders and unwilling recalls of the trauma (flashbacks) [@Bryant2019; @AmericanPsychiatricAssociation2013; @Howie2019], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence.
+PTSD may co-occur with symptoms of depression and anxiety and substance abuse disorders [@Gagnon-Sanschagrin2022; @Brady2000]. 
+Although most individuals have experienced at least one traumatic event during the lifetime [@Benjet2016; @Breslau2004], the prevalence of PTSD in the general European population has been estimated for 0.38 to 6.67% [@Trautmann2018; @Burri2014].
+Yet, up to 26% of acute physical trauma patients have been reported to develop PTSD [@Joseph2020].
+
+So far it has proven difficult to identify early predictors of subsequent PTSD development in physical trauma patients. 
+While some recent studies on specific populations have identified risk factors such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020].
+These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019]. 
+Many trauma patients at risk of PTSD undergo intensive acute medical management at emergency departments, which poses a challenge for implementation of early diagnostic and psychological preventive measures [@Linares2017].
+Furthermore, efficacy of such acute psychological interventions at preventing PTSD in trauma patients is uncertain, probably due to large heterogeneity of study cohorts [@Qi2016].
+
+Post-traumatic growth describes a salutary adjustment psychological reaction to a traumatic experience with strengthened perceptions of self, others, and meaning of the event [@Tedeschi1996]. 
+Resilience, i.e. the ability to master adversities has been proposed to bridge traumatic events and post-traumatic growth. 
+It was also suggested to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012]. 
+Overall, mountain sport was associated with high resilience values [@Ower2019].
+
+Reports on mental health following accidents during mountain sports are scarce. 
+The available data concern mountain guides, rescuers or avalanche accident survivors [@Leonard2021; @Sommer2004; @Mikutta2022; @Thordardottir2015; @Traber2020]. 
+Considering the large number of mountain accident victims, and heterogeneity of accident and injuries, accurate characteristic of mental health of the affected patients is urgently needed. 
+In particular, tools allowing for identification of patients at risk of later mental health problems during acute medical management of the mountain accident would facilitate the early targeted psychological support and help to preserve the positive mental health effects of physical activity.
+
+Herein, we analyzed a broad palette of mental health readouts such as anxiety, depression, resilience, sense of coherence, quality of life, post-traumatic growth and signs of PTSD in a collective of `r nrow(ptsd$dataset)` patients treated for injuries during mountain sport accidents at the Medical University of Innsbruck, Austria by means of semi-supervised clustering.
+Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain accident with a machine learning approach.
 
@@
 Participants' data were processed in anonymized form. 
-The study protocol was approved by the ethics committee of the Medical University of Innsbruck (approval number: _please fill in!_).
+The study protocol was approved by the ethics committee of the Medical University of Innsbruck (approval number: 1472/2020).
 
@@
 
-In- and outpatients treated at the orthopedics and trauma departments of the Medical University of Innsbruck (Austria) between 2018 and 2020 were screened for participation. 
-Patients fulfilling the inclusion criteria: age $\geq$ 16 years, proficiency in German and treatment due to a mountain sport accident (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey.
-Out of them, `r length(ptsd$interview_ID)` completed the survey and `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S2__).
+Patients treated for a mountain sport accident at the Department for Orthopedics and Traumatology at the Medical University of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 (i.e.
+at least 6 months prior to the start of the study) were screened for participation.
+Patients fulfilling the inclusion criteria: age $\geq$ 18 years, residence in a German-speaking country and proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail.
+Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey and `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S2__).
 
@@
 
-The psychometric battery consisted of German versions of assessment tools for 
+Sociodemographic (e.g.
+age, sex, education, profession, income, prior sport accidents and traumatic events), accident- and recovery-related (e.g.
+rescue, psychological support, persistent physical health consequences) and psychometric variables were recorded with a cross-sectional on-line survey.
+
+The psychometric battery consisted of German versions of assessment tools for symptoms of 
 anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
-depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
+depression (PHQ: patient health questionnaire, PHQ-9, depression module) [@Lowe2002; @Grafe2004], 
 panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
-persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
-resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
-loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
+persistent somatic symptoms (PHQ-15, somatization module) [@Kroenke2002], 
+resilience (RS13: 13-item resilience scale) [@Leppert2008], 
+lack of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
 quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
@@
 The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).
-Signs of anxiety (GAD-7 $\geq$ 10), depression (PHQ-9 $\geq$ 11) [@Manea2012], persistent somatic symptoms (PHQ-15 $\geq$ 11) [@Kroenke2002] and resilient coping classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73)[@Leppert2008] were defined as described before.
-The total PTGI and PCL-5 DSM-5 scores were calculated as the sum of all items of the respective tools [@Tedeschi1996; @Bovin2016]. 
-Significant PTSD symptoms were defined as positivity in at least one PCL-5 DSM-5 domain [@Bovin2016].
+
+Clinically significant symptoms of anxiety (GAD-7 $\geq$ 10), 
+depression (PHQ-9 $\geq$ 11) [@Manea2012], 
+persistent somatic symptoms (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
+resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73)[@Leppert2008] 
+were defined with published cutoffs.
+
+Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
+Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
+Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
+Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
+Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].
 
 Traumatic events before the accident were assessed with the DIA-X tool [@Maercker2005]. 
-Flashbacks frequency was surveyed as none, > 1/month and > 1/year. 
+Flashbacks frequency was surveyed as none, > 1/year and > 1/month. 
 Smoking was surveyed as a single yes/no question. 
-Alcohol abuse was investigated with the CAGE tool [@OBrien2008]. 
-Data on injury (AIS: abbreviated injury scale) [@Gennarelli2006] and medical treatment were extracted from electronic patient's records.
+Problematic alcohol use was investigated with the CAGE tool [@OBrien2008]. 
+Data on the type of the accident mountain sport, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient's records.
 
-The training and test participant subsets (2:1 size ratio) were obtained by random splitting which minimized differences in accident year, age, gender, pre-existing traumatic events, somatic and mental illness, and injury severity (__Figure \@ref(fig:fig-consort)__).
-The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as assessed by Hopkins metrics = `r pow$hopkins %>% filter(size == 200) %>% .$hopkins_stat %>% median %>% signif(2)` [@Kassambara2020].
+The training and test participant subsets (3:1 size ratio) were obtained by random splitting which minimized differences in in sociodemographic, medical history, clinical and accident- and injury-related variables assessed by Gower distance (__Figure \@ref(fig:fig-consort)__).
+The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as assessed by Hopkins metrics of `r signif(pca$clust_tendency$data$hopkins_stat, 2)` indicative of substantial spontaneous clustering tendency [@Kassambara2020].
 
@@
 
-The primary analysis endpoint was identification of subsets of mountain sport accident victims by clustering in respect to scores of anxiety, depression, panic, persistent somatic symptoms, quality of life, post-traumatic growth and post-traumatic stress disorder.
-The secondary analysis endpoints were demographic, socioeconomic and clinical characteristic of the mental clusters and development of a mental cluster assignment model to discern accident victims at risk of mental disorders.
+The primary analysis endpoint was assessment of mental health of mountain accident victims by clustering using standardized psychometric tools for symptoms of anxiety, depression, panic, persistent somatic symptoms, quality of life, resilience, sense of coherence, post-traumatic growth and PTSD.
+The secondary analysis endpoints were 
+(1) to compare demographic, socioeconomic and clinical characteristics between the mental health clusters of mountain accident victims and 
+(2) to identify patients at risk of post-traumatic mental disorders based on explanatory variables available during early medical treatment of the mountain sport accident.
 
@@
 The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
-This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares) and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical and KMEANS algorithms.
+This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares) and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical clustering and KMEANS algorithms.
 The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
-The training subset observations were assigned to the clusters with an inverse distance weighted 7-nearest neighbor classifier.
+The training subset observations were assigned to the clusters with an inverse distance-weighted 27-nearest neighbor classifier.
 
-The uni- and multi-parameter classifiers were trained with the one-rule [@VonJouanne-Diedrich2017; @Holte1993] and conditional random forest algorithms [@Hothorn2006; @Strobl2007; @Hothorn2022], respectively.
-The cluster assignment was predicted for the test subset and assessed by the accuracy and $\kappa$ statistics [@Kuhn2008]. 
-Conditional variable importance in the random forest classifiers was expressed as accuracy loss [@Strobl2008].
+Multi-parameter machine learning classifiers of the mental health cluster assignment were trained with 
+the canonical random forest algorithm [@Wright2017; @Breiman2001], 
+regularized neural network with a single hidden layer [@Ripley2014], 
+support vector machine algorithm with radial kernel [@Karatzoglou2004; @Weston1998], 
+recursive partitioning [@Therneau2022; @Breiman2017], 
+multinomial elastic net regression [@Zou2005; @Friedman2010], 
+and conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022]. 
+The cluster assignment was predicted for the test subset and the classifiers' performance at predicting the cluster assignment was assessed by the accuracy and $\kappa$ statistics, Brier score as well as sensitivity and specificity of predicted assignment to the post-traumatic stress cluster [@Cohen1960; @Brier1950; @Kuhn2008].
 
@@
 
-# Characteristic of the study cohort
+## Characteristic of the study cohort
 
-In- and outpatient adult victims of mountain sport accidents treated at the Medical University of Innsbruck (Austria) between 2018 and 2020 were invited to participate (n = `r nrow(ptsd$cleared)`). 
-Out of them, 
-`r length( ptsd$interview_ID)` completed the study survey and 
-`r nrow(ptsd$dataset)` individuals with the complete psychometric parameter set were analyzed (overall response rate: `r ((nrow(ptsd$dataset)/nrow(ptsd$cleared)) * 100) %>% signif(2)`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).
+The study collective consisted of victims of mountain sport accidents which happened in Tyrol, Austria, a renowned and popular alpine touristic region. 
+In- and outpatient adult survivors of mountain sport accidents and treated at the Department of Orthopedics and Traumatology at Medical University of Innsbruck (Tyrol, Austria) between 1^st^ January 2018 and 31^st^ December 2020 were invited to participate (n = `r nrow(ptsd$cleared)`).
+Among the invited patients, 
+`r length(ptsd$interview_ID)` completed the study survey and 
+`r nrow(ptsd$dataset)` individuals with the complete psychometric data set were analyzed (overall analysis inclusion rate: 
+`r ((nrow(ptsd$dataset)/nrow(ptsd$cleared)) * 100) %>% signif(2)`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__). 
+The median time between the trauma center admission and survey completion was 
+`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
+As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and less surgery diagnoses (__Supplementary Table S4__).
+The survey participants excluded from the analysis due to missingness of psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__).
+Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).
 
-As compared with the participants, the individuals excluded from the analysis had lower income, less frequent ski and more frequent biking accidents, less severe injuries, and lower hospitalization and surgery rates.
-The excluded individuals had also lower ratings of post-traumatic growth ratings and lower sense of coherence. 
-Effects of these differences were weak (V $\leq$ 0.22 or r $\leq$ 0.15, __Supplementary Table S4__).
+The analyzed participants were predominantly middle-aged 
+(median: `r get_num_stats(ptsd$dataset, 'age', collapse = ', interquartile range: ')` years) and 
+`r get_percent(ptsd$dataset, 'sex')$percent['female']`% of them were females.
+The vast majority had secondary or tertiary education grade 
+(`r sum(get_percent(ptsd$dataset, 'education')$percent[c('secondary', 'tertiary')])`%) and were professionally active 
+(`r get_percent(ptsd$dataset, 'employment_status')$percent['employed']`%). 
+Less than 8% of participants worked in a trauma-exposed or mountain sport profession. 
+High annual household incomes of $\geq$ 45000 Euro was reported by 
+`r get_percent(ptsd$dataset, 'high_income')$percent['yes']`% of participants. 
+Less than 10% of participants were smokers or at risk of problematic alcohol use. 
+Pre-existing physical disorders were reported by 
+`r get_percent(ptsd$dataset, 'somatic_comorbidity')$percent['yes']`% of participants with cardiovascular, neurological and metabolic illness being the most frequent. 
+Four of ten participants had experiences by or witnessed a traumatic event prior to the accident, 
+`r sum(get_percent(ptsd$dataset, 'traumatic_number')$percent[c('2', '3+')])`% participants experienced two or more traumatic events. 
+Mental disorders prior the accident affected 
+`r get_percent(ptsd$dataset, 'psych_comorbidity')$percent['yes']`% of the cohort with affective 
+(`r get_percent(mental$analysis_tbl, 'affective')$percent['yes']`%) and somatoform disorders 
+(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental conditions (__Table \@ref(tab:tab-cohort-demo)__).
 
-In the analyzed cohort, the median time between the medical accident management and survey completion was 
-`r median(ptsd$dataset$obs_time)` days. 
-The participants were predominantly middle-aged (median `r median(ptsd$dataset$age)` years) and 
-`r signif(table(ptsd$dataset$sex)/sum(table(ptsd$dataset$sex)), 2)[1] * 100`% of them were females. 
-The vast majority had secondary or tertiary education grade and were professionally active. 
-Sport and trauma-risk professions constituted < 8% of the cohort. 
-High annual household incomes of $\geq$ 45000 Euro was reported by > 40% of participants. 
-Less than 10% of participants were smokers or at risk of alcohol addiction. 
-Chronic somatic conditions prior to the accident were reported by `r signif(table(ptsd$dataset$somatic_comorbidity)/sum(table(ptsd$dataset$somatic_comorbidity)), 2)[2] * 100`% of participants with cardiovascular, neurological and metabolic illness being the most frequent.
-The rates of smoking (Austria: 20.6%) and somatic illness (38.3%) were lower than in the general Austrian population [@StatistikAustria2022a; @StatistikAustria2022].
-Psychiatric disorders before the accident affected `r signif(table(ptsd$dataset$psych_comorbidity)/sum(table(ptsd$dataset$psych_comorbidity)), 2)[2] * 100`% of the cohort. 
-Four of ten participants were affected by or witnessed a traumatic event prior to the accident (__Table \@ref(tab:tab-cohort-demo)__).
+Mountain sport accidents in the past were reported by `r get_percent(ptsd$dataset, 'prior_accident')$percent['yes']`% of participants. 
+Most of the investigated accidents occurred during alpine skiing, snowboarding or cross-country skiing 
+(`r get_percent(ptsd$dataset, 'sport_type')$percent['ski/snowboard/cross-country']`%) followed by biking 
+(`r get_percent(ptsd$dataset, 'sport_type')$percent['biking']`%) and classical mountain sports such as climbing, hiking, mountaineering or back-country skiing 
+(`r get_percent(ptsd$dataset, 'climbing/hiking/mountaineering/skitour')$percent['climbing/hiking/mountaineering/skitour']`%). 
+One-third of participants were alone during the accident 
+(`r get_percent(ptsd$dataset, 'accident_alone')$percent['yes']`%) and, in most cases, were the sole culprit 
+(`r get_percent(ptsd$dataset, 'accident_culprit')$percent['self']`). 
+Professional rescue service was involved in 
+`r get_percent(ptsd$dataset, 'accident_rescue')$percent['rescue team']`% of the accidents. 
+In `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['2']`% of participants the injury was moderate (AIS 2) and in 
+`r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['3+']`% severe-to-critical (AIS $\geq$ 3). 
+Limb injuries were the most common followed by head and face (__Supplementary Figure S1A__). 
+Hospitalization and surgery rates following the accident were 
+`r get_percent(ptsd$dataset, 'hospitalization')$percent['yes']`% and 
+`r get_percent(ptsd$dataset, 'surgery_done')$percent['yes']`%, respectively. 
+Psychological or psychiatric support after the accident was provided to 
+`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`% individuals. 
+A subset of participants (`get_percent(ptsd$dataset, 'psych_support_need')$percent['yes']`%), who had not received it, declared need for psychological support following the accident. 
+Despite persisting physical consequences of the accident, which affected 
+`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks during mountain sport at the time of the questionnaire completion in 
+`r get_percent(ptsd$dataset, 'unwilling_flashback')$percent['yes']`% of the cohort, 
+most participants returned to the same mountain sport following the accident 
+(`r get_percent(ptsd$dataset, 'same_sport_type_post_accident')$percent['yes']`%). 
+Yet, `r get_percent(ptsd$dataset, 'caution_post_accident')$percent['more cautious']`% described their behavior during mountain sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).
 
-Nearly 40% of participants had sport accidents in the past. 
-Almost two-thirds of the investigated accidents happened during skiing or snowboarding followed by biking and classical mountain sports such as climbing, hiking or mountaineering. 
-One-third participants were alone during the accident and, in most cases, were the sole culprit and victim. 
-Roughly half of participants could rescue themselves, `r signif(table(ptsd$dataset$accident_rescue)/sum(table(ptsd$dataset$accident_rescue)), 2)[3] * 100`% required a professional rescue. 
-In `r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[2] * 100`% of participants the injury was moderate (AIS 2) and in 
-`r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[3] * 100`% severe-to-critical (AIS $\geq$ 3). 
-Limb injuries were the most common followed by head and face (__Supplementary Figure S1A__). 
-Hospitalization and surgery rates were 
-`r signif(table(ptsd$dataset$hospitalization)/sum(table(ptsd$dataset$hospitalization)), 2)[2] * 100`% and 
-`r signif(table(ptsd$dataset$surgery_done)/sum(table(ptsd$dataset$surgery_done)), 2)[2] * 100`%, respectively. 
-Psychological or psychiatric support after the accident was applied to 
-`r signif(table(ptsd$dataset$psych_support_post_accident)/sum(table(ptsd$dataset$psych_support_post_accident)), 2)[2] * 100`% individuals. 
-Among those who had not received it, 
-`r signif(table(ptsd$dataset$psych_support_need)/sum(table(ptsd$dataset$psych_support_need)), 2)[2] * 100`% declared psychological support need. 
-Despite fairly common somatic health consequences of the accident 
-(`r signif(table(ptsd$dataset$accident_aftermath)/sum(table(ptsd$dataset$accident_aftermath)), 2)[2] * 100`%) and flashbacks during sport (`r signif(table(ptsd$dataset$unwilling_flashback)/sum(table(ptsd$dataset$unwilling_flashback)), 2)[2] * 100`%), most participants returned to the accident sport type (
-`r signif(table(ptsd$dataset$same_sport_type_post_accident)/sum(table(ptsd$dataset$same_sport_type_post_accident)), 2)[2] * 100`%). 
-Yet, two-third described their behavior during sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).
+At least one diagnostic criterion of PTSD defined by the PCL-5 tool (B, C, D or E) were met by 
+`r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% of participants. 
+The domain B PTSD symptoms were the most common (`r get_percent(ptsd$dataset, 'dsm5_B_class')$percent['positive']`%) and the domain D symptoms were the least frequent (`r get_percent(ptsd$dataset, 'dsm5_D_class')$percent['positive']`%).
+Solely `r sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)` patients 
+(`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) were screened positive for all four diagnostic criteria of PTSD (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1BC__).
+Clinically relevant symptoms of 
+anxiety (`r get_percent(ptsd$dataset, 'gad7_total_class')$percent['positive']`), 
+depression (`r get_percent(ptsd$dataset, 'phq9_total_class')$percent['positive']`%) 
+and somatic symptoms (`r get_percent(ptsd$dataset, 'phq_events_total_class')$percent['positive']`%) were rare in this cohort consisting of `r get_percent(ptsd$dataset, 'rs13_total_class')$percent['high']`% highly resilient individuals (__Table \@ref(tab:tab-cohort-mental)__).
 
-Signs of 
-anxiety (`r signif(table(ptsd$dataset$gad7_total_class)/sum(table(ptsd$dataset$gad7_total_class)), 2)[2] * 100`%), 
-depression (`r signif(table(ptsd$dataset$phq9_total_class)/sum(table(ptsd$dataset$phq9_total_class)), 2)[2] * 100`%) and 
-persistent somatic symptoms (`r signif(table(ptsd$dataset$phq_events_total_class)/sum(table(ptsd$dataset$phq_events_total_class)), 2)[2] * 100`%) were rare in the study cohort (__Table \@ref(tab:tab-cohort-mental)__).
-Quality of life was rated with median `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS points, which was comparable to the generalized Western European population [@Schmidt2006]; overall quality of life and housing domains received the highest ratings.
-Roughly two-thirds of participants were assigned to the high resilient coping class [@Leppert2008]. 
-Overall post-traumatic growth (median: `r median(ptsd$dataset$ptgi_total) %>% signif(2)`) and its specific domains were rated lower than originally reported for the PTGI tool [@Tedeschi1996]. 
-Symptoms of post-traumatic stress disorder were observed in `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants with domain B symptoms (`r signif(table(ptsd$dataset$dsm5_B_class)/sum(table(ptsd$dataset$dsm5_B_class)), 2)[2] * 100`%) being most common (__Table \@ref(tab:tab-cohort-mental)__ and __Supplementary Figure S1__).
+For clustering and modeling, the cohort was split into the 
+training (n = `r table(ptsd$dataset$partition)['training']`) and 
+test subset (n = `r table(ptsd$dataset$partition)['test']`). 
+Significant differences of weak effect size between these subsets concerned resilience class distribution and scores of quality of life (__Supplementary Table S6__).
 
-For clustering and modeling, the cohort was split into the training (n = `r nrow(clust_globals$analysis_tbl$training)`) and test subset (n = `r nrow(clust_globals$analysis_tbl$test)`). 
-There were no significant differences between the subsets in demographic, socioeconomic, clinical and accident-related parameters (not shown).
+## Three clusters of mental response in sport accident victims
 
-# Three clusters of mental response in sport accident victims
+To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to partition around medoids clustering in respect to a broad range of psychometry scores [@Drost2018; @Schubert2019] (__Supplementary Table S2__).
+Three mental clusters: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress), named after their key mental characteristic were identified in the training subset (__Supplementary Figure S3__).
+Subsequently, the mental cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
+(training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`), 
+comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).
 
-To explore mental responses to mountain sport accidents, we subjected the participants to clustering in respect to a broad range of numeric psychometry variables (__Supplementary Table S2__). 
-Among clustering algorithms compared in the training subset, partition around medoids with cosine distance [@Drost2018; @Schubert2019] demonstrated good explanatory power and superior reproducibility in cross-validation [@Lange2004] (__Supplementary Figure S2A__).
-Three mental clusters: 'neutral', 'PTG' (post-traumatic growth) and 'PTB' (post-traumatic burden), named after their key mental characteristic were identified in the training subset. 
-Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`), comparable cluster sizes and good visual clusetr separation in both study cohort subsets (__Supplementary Figures S2B__ - __S4__).
+The mental clusters were of approximately equal size. 
+The neutral cluster was characterized by low levels of anxiety, depression, panic, somatic symptoms, post-traumatic stress and post-traumatic growth along with high rating of sense of coherence, resilience and quality of life.
+The PTG cluster demonstrated similarly low scores of major mental health disorders, high resilience and sense of coherence. 
+Its key characteristic were the highest levels of post-traumatic growth. 
+The remaining PTS cluster displayed the highest scores of for anxiety, depression, panic disorder, somatic symptoms and post-traumatic stress as well as poor sense of coherence, and low resilience and low quality of life.
+Post-traumatic growth scores in the PTS cluster were higher than in neutral but lower than in PTG cluster participants (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Table S7__). 
+Consequently, clinically relevant symptoms of 
+anxiety (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'gad7_total_class')$percent['positive']`%), 
+depression (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq9_total_class')$percent['positive']`%) and somatic symptoms (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq_events_total_class')$percent['positive']`%)
+were present virtually only in the PTS cluster. 
+Furthermore, frequencies of 
+low (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'rs13_total_class')$percent['low']`%) and 
+moderate resilience classes (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'rs13_total_class')$percent['moderate']`%) peaked in the PTS cluster. 
+Finally, the frequency of flashbacks, PTSD symptoms specified by the B, C, D and E PCL-5 domains and the frequency of patients meeting at least one PCL-5 PTSD diagnostic criterion 
+(PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'dsm5_cluster_class')$percent['positive']`%) were the highest in PTS cluster individuals (__Figure \@ref(fig:fig-clust-symptoms)__, __Supplementary Table S7__).
 
-The neutral cluster encompassing roughly one-third participants was characterized by low levels of anxiety, depression, panic, somatic symptoms, post-traumatic stress and post-traumatic growth along with high rating of coherence, resilient coping and quality of life.
-The PTG cluster demonstrated similarly low scores of major mental health disorders. 
-By contrast, levels of post-traumatic growth were the highest in the PTG cluster. 
-Furthermore, ratings of post-traumatic stress were slightly higher in the PTG than in the neutral cluster. 
-The remaining PTB cluster displayed the highest scores of anxiety, depression, panic, somatic symptoms and post-traumatic stress as well as poor sense of coherence, resilient coping and low quality of life.
-Post-traumatic growth scores in the PTB cluster were higher than in neutral but lower than in PTB cluster participants (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Table S5__).
-Consequently, clinically relevant symptoms of anxiety, depression and persistent somatic symptoms were present virtually only in the PTB cluster. 
-Furthermore, frequencies of low and moderate resilient classes peaked in the PTB cluster. 
-Finally, symptoms of post-traumatic stress disorder were way more frequent in the PTB cluster (any symptoms: > 36%) than in the remaining clusters (__Figure \@ref(fig:fig-clust-symptoms)__).
-
-Collectively, the PTG cluster may represent individuals with salutary reaction to mountain sport accident, whereas the PTB cluster may pose a group at risk of mental health deterioration.
-
-# Demographic, socioeconomic and clinical background of the mental clusters
+## Demographic, socioeconomic and clinical background of the mental health clusters
 
 Among 
-`r length(clust_bcg$variables)` investigated demographic, socioeconomic, clinical, accident- and accident consequence-related variables, only 
-`r length(clust_bcg$top_factors)` parameters were found to differ significantly between the mental clusters in both the training and test subsets. 
-The effect size of such differences was weak (__Supplementary Table S6__).
+`r length(clust_bcg$variables[!stri_detect(clust_bcg$variables, regex = '^(dsm|rs13|phq|gad|flash|unwilling)')])` investigated demographic, socioeconomic, clinical, accident- and recovery-related variables, only
+`r length(clust_bcg$top_factors[!stri_detect(clust_bcg$top_factors, regex = '^(dsm|rs13|phq|gad|flash|unwilling)')])` features (age, pre-existing physical illness, pre-existing mental disorder, psychological support need, persistent physical health consequences of the accident, and caution during sport activity) were found to differ significantly between the mental clusters.
+The effect size of these differences was weak (__Supplementary Table S8__).
 
-A substantial enrichment of females, low-to-middle income and non-tertiary education participants was observed in the PTB cluster. 
-The PTG cluster had the highest share of middle-aged individuals. 
-Frequency of somatic conditions before the accident were the highest in the PTG and PTB clusters. 
-Participants with pre-existing mental illness were in turn represented virtually only in the PTB cluster. 
-Although frequencies of previous sport accidents were comparable between the clusters, PTG cluster individuals were affected by prior traumatic events more often than neutral cluster individuals. 
-These differences were, however, non or borderline significant and of weak effect size (__Figure \@ref(fig:fig-clinic-factors)__, __Supplementary Figure S5__). 
-There were no consistent differences in accident sport type, rescue, injury severity and location between the mental clusters. 
-Yet, PTG and PTB cluster individuals tended towards higher hospitalization and surgery rates (__Figure \@ref(fig:fig-injury)__ and __Supplementary Figure S6__).
+In more detail, PTS cluster patients were the youngest of all participants and had the highest frequency of pre-existing mental conditions. 
+The neutral cluster was in turn substantially enriched in males, and tertiary education and high income individuals; these effects were, however not significant. 
+Participants suffering from pre-existing somatic conditions were significantly enriched in the PTG and PTS clusters. 
+A similar, yet not significant tendency was observed for frequency of prior traumatic events, which was higher in the PTG and PTS clusters as compared with neutral cluster patients. 
+There were no substantial differences in frequency of participants with trauma-risk professions and employment structure between the clusters (__Figure \@ref(fig:fig-clinic-factors)__, __Supplementary Figure S7__).
+Participants reporting being alone during the accident, the only culprit of the accident, and requiring comrade or professional rescue were enriched in the PTG cluster. 
+In turn, frequency of self-rescuers tended to be the highest in the neutral cluster. 
+These differences in accident and rescue circumstances were not significant. 
+We could not observe any substantial differences in accident season, mountain sport type, and numbers of injured persons between the clusters (__Figure \@ref(fig:fig-injury)A__, __Supplementary Figure S7__).
+There were no significant differences in severity and injury and its location between the clusters. 
+However, moderate-to-severe injuries (AIS $\geq$ 2), hospitalization and surgery rates were higher in the PTG and PTS clusters as compared with neutral cluster patients (__Figure \@ref(fig:fig-injury)__).
 
-Somatic health consequences of the accident and flashbacks were significantly more common in the PTB than the neutral cluster. 
-Additionally, PTG and PTB cluster individuals tended towards more caution during sport as compared with the neutral cluster. 
-Roughly one-sixth PTB cluster individuals reported need for psychological support following the accident as compared with none in the neutral cluster. 
-Rates psychological support after the accident tended to be higher in the PTG (12 - 16%) than in the remaining clusters, yet this effect was significant only in the training subset (__Figure \@ref(fig:fig-aftermath)__).
-Given the low frequency of psychological support after the accident in the study cohort and particular clusters, it is unlikely, that psychological support affected the mental cluster assignment .
+There were no significant differences in rates of psychological or psychiatric support after the accident between the clusters. 
+Among PTG and PTS cluster individuals, `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'psych_support_need')$percent['yes']`% and `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'psych_support_need')$percent['yes']`%, respectively, reported need for psychological support following the accident as compared with none in the neutral cluster.
+Frequency of participants suffering from persistent physical health consequences of the accident was the highest in the PTS cluster (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'accident_aftermath')$percent['yes']`%,
+PTG: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'accident_aftermath')$percent['yes']`%, 
+neutral: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'neutral'), 'accident_aftermath')$percent['yes']`%). 
+Analogically, PTS cluster patients reported the highest rates of cautious behavior during mountain sport following the accident 
+(PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'caution_post_accident')$percent['more cautious']`%, 
+PTG: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'caution_post_accident')$percent['more cautious']`%, 
+neutral: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'neutral'), 'caution_post_accident')$percent['more cautious']`%). 
+By contrast, percentages of participants who returned to the same mountain sport after the accident were comparable between the mental clusters (__Figure \@ref(fig:fig-aftermath)__).
 
-# Prediction of the mental cluster assignment by demographic, socioeconomic and accident-related factors
+## Prediction of the mental cluster assignment by demographic, socioeconomic and accident-related factors
 
-Finally, we intended to model the mental cluster assignment with `r length(class_globals$early_variables)` demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident (__Supplementary Table S7__).
+Finally, we intended to model the mental cluster assignment with  demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident (__Supplementary Table S9__) with seven popular machine learning algorithms: the random forests [@Wright2017; @Breiman2001], neural networks [@Ripley2014], support vector machines [@Weston1998], recursive partitioning [@Therneau2022], discriminant analysis [@Ahdesmaki2010; @Ahdesmaki2022], elastic net regression [@Zou2005; @Friedman2010], and conditional random forests [@Hothorn2006; @Strobl2007; @Hothorn2022] (__Supplementary Table S10__).
 Such models would enable early identification of mountain sport accident victims at risk of mental health problems.
 
-Predictive value [@VonJouanne-Diedrich2017; @Holte1993] of single early explanatory parameters was poor (maximum accuracy: `r class_one$fit_stats$global$Accuracy %>% max %>% signif(2)`, maximum $\kappa$: `r class_one$fit_stats$global$Kappa %>% max %>% signif(2)`) and inconsistent between the subsets of the study cohort.
-In such univariable analysis, age class, annual income and traumatic events in the past were the most predictive factors (__Supplementary Figure S7__).
-
-To account for cumulative effects and interactions of multiple weak predictors of the mental cluster assignment, we resorted to conditional random forest modeling [@Strobl2007; @Hothorn2006; @Hothorn2022].
-This multi-parameter model employing early candidate cluster predictors could correctly assign `r signif(class_early$fit_stats$global$Accuracy[1], 2) * 100`% training subset observations to their mental clusters
-($\kappa$ = `r signif(class_early$fit_stats$global$Kappa[1], 2)`). 
-However, its predictive performance in the test subset was poor (accuracy: `r signif(class_early$fit_stats$global$Accuracy[2], 2) * 100`, $\kappa$ = `r signif(class_early$fit_stats$global$Kappa[2], 2)`).
-The prediction quality was the best for the PTG cluster (accuracy, training: 89.5%, test: 59.4%, __Figure \@ref(fig:fig-early-crf)A__).
-The most important early explanatory variables for the cluster prediction by conditional random forests were annual income, sex, mode of rescue, leg injury and pre-existing mental illness (__Figure \@ref(fig:fig-early-crf)B__).
-Inclusion of late accident consequences such as cautious behavior during sport, flashbacks or somatic health burden of the accident in the random forest model could only marginally improve its performance (training: accuracy = `r signif(class_rf$fit_stats$global$Accuracy[1], 2)`, $\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[1], 2)`; test subset: accuracy = `r signif(class_rf$fit_stats$global$Accuracy[2], 2)`, $\kappa$ = `r signif(class_rf$fit_stats$global$Kappa[2], 2)`).
-The most important late predictors of the cluster assignment were caution during sport and flashbacks (__Supplementary Figure S8__).
-
-Collectively, we could not establish a reliable model for identification of patients at risk of mental disorders represented by the PTB cluster with a broad ensemble of early non-mental predictors.
-
-\newpage
+The cluster assignment models employing early predictors demonstrated a moderate-to-excellent prediction performance in the training subset of the study cohort 
+(`r paste0((filter(early_class$overall_stats, dataset == 'train')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`% of correctly assigned observations, 
+Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$kappa), 2) %>% paste(collapse = ' - ')`, Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$bs), 2) %>% paste(collapse = ' - ')`).
+However, their accuracy in cross-validation and in the test subset was poor 
+(test: `r paste0((filter(early_class$overall_stats, dataset == 'test')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`% of correctly assigned observations, 
+Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$kappa), 2) %>% paste(collapse = ' - ')`, 
+Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$bs), 2) %>% paste(collapse = ' - ')`). 
+Concerning the early identification of the vulnerable PTS cluster patients, the conditional random forest, random forest and support vector machine models were able to differentiate PTS cluster patients from the pooled neutral + PTG cluster subset with the best, at least
+`r min(signif(early_class$clust_stats %>% filter(dataset == 'test', clust_id == 'PTS') %>% top_n(n = 3, sens) %>% .$sens, 2) * 100)`% sensitivity in the test subset of the study cohort (__Figure \@ref(fig:fig-early-class)__, __Supplementary Table S12__).
+Pre-existing mental disorder or physical illness, hospitalization and surgery, injury severity, age, rescue mode, education, annual income, and traumatic events in the past were among the most important explanatory factors employed by these models (__Supplementary Figure S8__).
+Of note, inclusion of follow-up-related variables such as psychological support, physical health consequences or caution during sport activity in the machine learning models of the mental cluster assignment did not improve their accuracy in cross-validation or the test subset of the study cohort
+(test: `r paste0((filter(full_class$overall_stats, dataset == 'test')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`% of correctly assigned observations, 
+Cohen's $\kappa$: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$kappa), 2) %>% paste(collapse = ' - ')`, 
+Brier score: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$bs), 2) %>% paste(collapse = ' - ')`).
 
@@
 
-In individuals surveyed on average 3.5 years after the non-professional sport accident, some of its consequences like somatic health problems, unwilling flashbacks as well as phenomena of post-traumatic growth and symptoms of post-traumatic stress disorder were still tangible.
-By semi-supervised clustering we could discern three distinct patterns of mental health in this collective with each present in approximately one-third of participants: (1) a neutral mental response reaction, (2) salutatory reaction to the accident characterized by post-traumatic growth and (3) mental health deterioration hallmarked by mental accident burden such as symptoms of anxiety, depression, somatization, resilience loss and signs of post-traumatic stress disorder.
-Notably, separation of the PTG and PTB clusters was in part blurry. 
-Symptoms of post-traumatic stress were observed also in PTG cluster participants albeit less frequently than in the PTB cluster. 
-Conversely, mental health problems in PTB cluster participants were paralleled by some signs of post-traumatic growth. 
-In a detailed comparison of `r length(clust_bcg$variables)` factors between the clusters, no distinct demographic, socioeconomic, clinical or accident-related background of the mental clusters could be identified.
-We found a moderate enrichment of females, low-to-moderate income participants, non-academic graduates, chronic somatic and psychiatric conditions as well as individuals requiring hospitalization and surgery in the PTG and, in particular, in the PTB cluster.
-Injury severity and localization or professional rescue rates indicative of potentially more severe accident were comparable between the clusters. 
-Similarly, the cluster assignment was unlikely affected by psychological support after the accident. 
-In turn, the PTB and, to a lesser extent, the PTG cluster were characterized by somatic health consequences of the accident, flashbacks, more caution during sport activity and subjective need for psychological support.
-Finally, the lacking non-mental explanatory variables strongly associated with the cluster assignment likely resulted in sub-optimal performance of machine learning classifiers employed for prediction of the cluster assignment and identification of accident victims at risk of mental health problems.
+Herein, we cross-sectionally explored mental health in victims of mountain sport accidents treated at a tertiary trauma center at median `r round(median(ptsd$dataset$obs_time)/30.436)` months after the trauma center admission.
+We identified three approximately equally sized subsets of participants with distinct mental health characteristic: 
+(1) the neutral mental health cluster, 
+(2) the PTG cluster characterized by predominantly salutatory reaction to the accident manifested by post-traumatic growth, and 
+(3) the PTS cluster hallmarked by symptoms of PTSD, anxiety, depression, panic disorder, persistent somatic symptoms as well as low resilience, quality of life and sense of coherence.
+PTS cluster patients were characterized by lower age, more frequent pre-existing physical and mental conditions, self-reported need for psychological support, long-term physical health consequences of the accident and cautious behavior during sport.
+The PTS cluster was also substantially enriched in hospitalized and surgery patients. 
+However, even with robust machine learning algorithms we could not establish a reliable model to identify PTS cluster patients at risk of long-term mental health problems with data available during acute medical management of the mountain sport accident.
 
-_Hanna and Katharina: your turn now:)_
+To the best of our knowledge this is the first study evaluating a broad palette of mental features including symptoms of anxiety and depression, somatic symptoms, resilience, post-traumatic growth and symptoms of PTSD in survivors of mountain sport accidents.
+This population is of special interest because, on one hand, mountain sport is generally considered beneficial for mental health [@Ower2019; @Niedermeier2017a]. 
+On the other hand, mountain sport is associated with specific individual physical and mental traits, and bears potential of danger [@Faulhaber2017; @Faulhaber2020].
+Studies conducted on mountain guides and mountain rescue personnel, showed that 71% - 78% of them experienced a traumatic event [@Mikutta2022; @Sommer2004]. 
+Intriguingly, the frequency of PTSD symptoms in those populations was estimated for 1% - 5.9% [@Traber2020; @Mikutta2022; @Sommer2004], which was comparable to the general population [@Frans2005; @PTSD2023; @Trautmann2018; @Burri2014] and, particularly, lower than in other potentially vulnerable groups such as emergency workers [@Berger2012].
+By contrast, a recent Swiss study found a prevalence of PTSD in up to 22% of mountain rescuers once differentiating between clinical interviews and self-reporting of symptoms for the assessment of PTSD [@Traber2020].
+Concerning mountain accidents, survivors of avalanches were affected by acute and long-term PTSD symptoms such as hyperarousal or sleep disorders at rates between 11 - 16% [@Thordardottir2015; @Leonard2021].
+This figure is comparable with `r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% participants of our study screened positive for at least one of of the PTSD symptom domains defined by the PCL-5 tool [@Bovin2016].
+Of note, the frequency of at least one PCL-5 PTSD symptom in our cohort recruited among patients of a trauma department is similar to the 11.8% - 26.6% prevalence reported in literature for emergency department or acute trauma patients [@Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Joseph2020].
+However, in our cohort, solely 
+`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`% of participants demonstrated concomitant PCL-5 domain B, c, D and E symptoms suggestive of fully manifest PTSD.
+Resilience has been proposed as a protective trait against mental health deterioration following trauma [@Ahmed2007; @Bensimon2012] and high resilience levels in mountain rescuers may be explain low prevalence of PTSD in this population despite trauma exposure [@Mikutta2022].
+Two-thirds of our study participants were classified as highly resilient [@Leppert2008]. 
+This observation may hence explain the low rate of manifest PTSD as well as lower frequencies of clinically relevant symptoms of anxiety and depression in our cohort than estimated for the general Austrian population [@Humer2022].
+The high resilience in the study cohort may also pertain to the high rating of quality of life 
+(EUROHIS, median: `r get_num_stats(ptsd$dataset, 'eurohis_total', collapse = ', interquartile range: ')`) comparable to the general western European population [@Schmidt2006]. 
+Post-traumatic growth represents a salutogenic reaction to a traumatic event with positive changes in personality, self-assurance, experience and inter-personal relationships [@Tedeschi1996], which may overlap with resilience and PTSD symptoms [@Liu2017; @Bensimon2012].
+This phenomenon was evident also in our study, where PTG cluster participants were characterized by a pronounced post-traumatic growth and good resilience scoring, which was however paralleled by flashbacks and single PTSD symptoms.
+
+By semi-supervised clustering, we identified mountain sport accident victims with co-existing symptoms of PTSD, depression, anxiety, somatic symptoms, as well as low sense of coherence and quality of life.
+Those patients assigned to the PTS subset may represent a potentially clinically relevant risk group of post-traumatic mental disorders. 
+The psychometric characteristic of the PTS cluster corresponds to the multi-faceted mental symptoms described for PTSD individuals [@Gagnon-Sanschagrin2022; @Brady2000; @Bryant2019; @Howie2019]. 
+Regarding early predictors of mental health problems, the PTS cluster was characterized by a significantly younger age as well as high rates of pre-existing physical and mental disorders. 
+It was also substantially enriched in females, low-to-moderate income patients, individuals requiring comrade or professional rescue, hospitalization and surgery as compared with neutral cluster patients.
+Of note, 
+female gender [@Shalev2019; @Christiansen2020; @Seedat2005; @Baecher2018], 
+young age at traumatization [@Schultebraucks2020; @Kongshoej2022], 
+low income [@Carbone2019] and 
+pre-existing mental disorders [@Papini2018; @Sandweiss2011; @Baecher2018] has been proposed as risk factors of PTSD. 
+The recovery features of the PTS cluster were the highest rates of self-reported need for psychological support, persistent physical health consequences of the accident and the highest percentage of more cautious behavior during sport activity as compared with the neutral or PTG cluster.
+However, differences in the early or follow-up variables between the mental clusters in the study cohort were only of weak effect size, which underlines the lack of strong predictive markers and moderate explanatory power of published models predicting post-traumatic mental problems [@Shalev2019; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018; @Cohen2013; @Rice2005].
+PTSD risk has been previously associated with injury severity, head injuries, hospitalization length and pain [@Baecher2018; @Blanchard1995; @Galatzer-Levy2014]. 
+We could not establish any link between the clusters and, in particular, the PTS cluster with objective injury severity measured with the AIS scale [@Gennarelli2006] or injury location. 
+Intriguingly, subjective injury severity, which was not surveyed in our cohort, was shown to be a stronger predictor of PTSD symptoms than objective injury severity [@Gabert-Quillen2011]. 
+This phenomenon together with the longer survey - treatment time interval than in most other studies and differences in composition of the study sample, may explain the lacking effect of injury on later mental disorders in our collective.
+Importantly, the percentage of participants having received psychological or psychiatric support after the accident was generally low (`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`%) in the study cohort and did not differ between the mental clusters.
+Hence, it is unlikely that psychological or psychiatric intervention had any impact on classification of the participants.
+
+Early identification of trauma patients at risk of development of PTSD or follow-up mental health disorders has previously been tackled by diverse machine learning algorithms [@Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
+In our hands, none of seven machine learning algorithms employing different mathematical principles, random forest, neural network, support vector machines, recursive partitioning, conditional random forest, discriminant analysis or elastic net regression, was able to reliably predict the mental cluster assignment based on `r length(class_globals$variables$early)` demographic, socioeconomic, clinical and accident-related variables available during acute medical treatment of the mountain sport accident.
+This phenomenon is likely attributed to the lacking strong non-mental explanatory variables differentiating between the mental clusters and, of clinical importance, highly specific markers of post-traumatic stress patients.
+Interestingly, inclusion of peri-traumatic mental symptoms like PTSD features, flashbacks, hyperarousal, or subjective need for support was shown to be crucial for optimal prediction of long-term mental disorders or PTSD during recovery [@Gabert-Quillen2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
+Unfortunately, such variables were not available in our cohort investigated at least six months after the accident.
+
+# Limitations
 
 Our study bears limitations. 
-First, although the key phenomena phenomena in the newly identified mental clusters, post-traumatic growth and post-traumatic stress disorder, are clearly associated with trauma [@Tedeschi1996; @Bovin2016], we could not exclude the mental cluster assignment to be affected by other events than the sport accident.
-To account for that, follow-up studies including a control group are needed. 
-Second, our variable set misses potentially important explanatory factors for the mental cluster classification such as length of hospital stay, rehabilitation need or ability to work. 
-Finally, effects of the accident, injury severity, hospitalization or surgery on mental health after an accident may have been obscured in the cross-sectional cohort and needs to be investigated in a more defined collective e.g.
-of high risk sport accident victims or hospitalized individuals.
+The overall analysis inclusion rate was low and, as indicated by comparison of the excluded and analyzed participants, the results were potentially confounded by a selection bias towards better patients with higher social status and more severe injuries.
+Although the key of the mental clusters, post-traumatic growth and PTSD, were clearly associated with trauma [@AmericanPsychiatricAssociation2013; @Tedeschi1996; @Bovin2016], the mental cluster assignment could have been affected by other events than the sport accident.
+Furthermore, our data set misses important explanatory factors which may facilitate distinction between the mental clusters such as rehabilitation need or ability to work. 
+Analogically, due to the cross-sectional design of the study, we were not able to assess peri-traumatic mental disorder symptoms, which has been proven crucial for identification of vulnerable patients  [@Gabert-Quillen2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
+Finally, a longitudinal survey may have helped to assess individual trajectories of mental health reaction to the mountain sport accident.
 
-# Conclusion
+# Conclusions and outlook
 
-_Your part again please_
+Our study represents the first attempt to characterize post-traumatic stress and further facets of mental health following mountain sport accidents. 
+We identified symptoms of mental health impairment in roughly one third of the victims. 
+However, we could not identify robust predictors for post-traumatic stress among variables available during early medical treatment.
+Hence, considering the increasing popularity of mountain sports, low-threshold information and psychological support for trauma department patients treated seems the most appropriate strategy for successful multi-disciplinary medial management of mountain sport accidents.
+Such information campaign can include 
+informative leaflets, 
+additional information in the discharge letter, 
+as well as websites or mobile applications explaining frequent warning signs and symptoms of post-traumatic stress disorder and providing contact data for points of psychological support. 
+We are currently implementing such measures at our hospital.
 
 \newpage
-
-# Acknowledgments
-
-# Author's contribution
-
-# Conflict of interest
 
@@
 
-tables$cohort_demo %>% 
-  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
-  flextable %>% 
-  flextable::width(1, width = 5.5, unit = 'cm') %>% 
-  flextable::width(2, width = 7.5, unit = 'cm') %>% 
+flextable::flextable(tables$cohort_demo) %>% 
+  set_widths(c(5.5, 7.5)) %>% 
   footnote(1, 1, 
-           value = as_paragraph('CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; DIA-X: Diagnostic Expert System, traumatic event score'), 
+           value = as_paragraph('CAGE: Cut/Annoyed/Guilty/Eye substance abuse scale; DIA-X: Diagnostic Expert System, traumatic event score.'), 
            part = 'header', 
@@
 
-tables$cohort_accident %>% 
-  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
-  flextable %>% 
-  flextable::width(1, width = 5.5, unit = 'cm') %>% 
-  flextable::width(2, width = 7.5, unit = 'cm') %>% 
+flextable::flextable(tables$cohort_accident) %>% 
+  set_widths(c(7.5, 7.5)) %>% 
   footnote(2, 1, 
-           value = as_paragraph('ski/snowdoard: alpine skiing, snowboarding and cross-country skiing\nsledding: sledding or bobsled\nclimbing/hiking/mountaineering: hiking, rock and ice climbing, mountaineering, skitouring\nbiking: mountainbike, tour and road cycling'),
+           value = as_paragraph('ski/snowdoard/cross-country: alpine skiing, snowboarding and cross-country skiing\nsledding: sledding or bobsled\nclimbing/hiking/mountaineering: hiking, rock and ice climbing, mountaineering, skitouring\nbiking: mountainbike, tour and road cycling.'),
            part = 'body', 
@@
   footnote(8, 1, 
-           value = as_paragraph('AIS: abbreviated injury scale'), 
+           value = as_paragraph('AIS: abbreviated injury scale.'), 
            part = 'body', 
@@
   flextable %>% 
-  flextable::width(1, width = 5.5, unit = 'cm') %>% 
-  flextable::width(2, width = 7.5, unit = 'cm') %>% 
+  set_widths(c(5.5, 7.5)) %>% 
   footnote(1, 1, 
-           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder.'),
            part = 'header', 
@@
 
-```{r fig-consort, fig.width = mm_inch(figures$consort$w), fig.height = mm_inch(figures$consort$h), fig.cap = 'Flow diagram of the analysis inclusion process and the analysis strategy.'}
+```{r fig-consort, fig.width = figur::convert(figures$consort, to = 'in')$w, fig.height = figur::convert(figures$consort, to = 'in')$h, fig.cap = 'Flow diagram of the analysis inclusion process.'}
 
@@
 
-__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process and the analysis strategy.__
+__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process.__
 
@@
 
-```{r fig-clust-psych, fig.width = mm_inch(figures$clust_psych$w), fig.height = mm_inch(figures$clust_psych$h), fig.cap = 'Scores of psychometry readouts in the mental clusters.'}
+```{r fig-clust-psych, fig.width = figur::convert(figures$clust_psych, to = 'in')$w, fig.height = figur::convert(figures$clust_psych, to = 'in')$h, fig.cap = 'Scores of mental disorder symptoms, sense of coherence, resilience, quality of life, post-traumatic growth, and post-traumatic stress disorder in the mental clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-clust-psych). Scores of psychometry readouts in the mental clusters.__
+__Figure \@ref(fig:fig-clust-psych).
+Scores of mental disorder symptoms, sense of coherence, resilience, quality of life, post-traumatic growth, and post-traumatic stress disorder in the mental clusters.__
 
-_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations.
-Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
-Three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden), All clustering parameters differed significantly between the clusters as assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
-Mean normalized scores of major mental health readouts (A), diminished quality of life (B), post-traumatic growth (C) and post-traumatic symptom disorder (D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines.
+_Study participants in the training subset of the cohort were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) mental clusters defined in respect to psychometric scores by the partition around medoids (PAM) algorithm.
+Study participants in the test subset of the cohort were assigned to the mental clusters by the inverse distance weighted 7-nearest neighbors classifier.
+Differences in the cluster-defining psychometric scores between the clusters were assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+P values were adjusted for multiple testing with the false discovery rate method.
+Mean normalized scores of major mental health readouts, sense of coherence and resilience (A), quality of life (B), post-traumatic growth (C) and post-traumatic stress disorder (PTSD, D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines.
 Tinted regions represent 2 $\times$ SEM (standard error of the mean).
 Numbers of observations in the clusters are displayed in the plot captions.
-GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
+Significant effects are highlighted in bold._
+_GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; som.
+sympt.: somatic symptoms; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
 
@@
 
-```{r fig-clust-symptoms, fig.width = mm_inch(figures$clust_symptoms$w), fig.height = mm_inch(figures$clust_symptoms$h), fig.cap = 'Signs of mental health problems in the mental clusters.'}
+```{r fig-clust-symptoms, fig.width = figur::convert(figures$clust_symptoms, to = 'in')$w, fig.height = figur::convert(figures$clust_symptoms, to = 'in')$h, fig.cap = 'Symptoms of mental disorders and resilience classes in the mental clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-clust-symptoms). Signs of mental health problems in the mental clusters.__
+__Figure \@ref(fig:fig-clust-symptoms). Symptoms of mental disorders and resilience classes in the mental clusters.__
 
-_Frequencies of symptoms of anxiety (A, GAD-7 ≥10), depression (B, PHQ-9 ≥ 11) and somatization (C, PHQ-15 ≥ 11), distribution of resilient coping classes (D, RS13) and symptoms of post-traumatic stress disorder (E, PTSD, PCL-5 DSM-5, positivity for PTSD domains and frequency of participants positive for at least one PTSD domain [PTSD+]) in the mental clusters.
-Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
-Fractions of symptom-positive and -negative participants in the mental clusters in the training and test subset of the study cohort are presented in stack and bar plots.
-Effect sizes and p-values are shown in the plot captions or Y axes.
-Numbers of observations in the clusters are shown in the X axes or plot captions.
-Significant effects in (E) are highlighted in bold.
-GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder._
+_Differences in frequency of mental disorder symptoms and flashbacks, and distribution of resilience classes between the mental clusters (neutral, post-traumatic growth [PTG] and post-traumatic stress [PTS]) were investigated in the entire study cohort by $\chi^2$ test with Cramer's V effect size statistic.
+P values were adjusted for multiple testing with the false discovery rate method._
+
+_(A) Symptoms of mental disorders, presence of flashbacks during sport activity and symptoms of post-traumatic stress disorder (PTSD).
+Percentages of affected individuals in the mental clusters are presented in bar plots.
+Effect sizes and p values are indicated in the Y axes.
+Significant effects are highlighted in bold.
+Numbers of observations in the clusters are indicated in the plot captions._
+
+_(B) Resilience classes and frequency of flashbacks during sport activity in the mental clusters presented in stack plots.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations in the clusters are indicated in the X axes._
+
+_GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder._
 
@@
 
-```{r fig-clinic-factors, fig.width = mm_inch(figures$clinic_factors$w), fig.height = mm_inch(figures$clinic_factors$h), fig.cap = 'Demographic and clinical factors differing between the mental clusters of accident victims.'}
+```{r fig-clinic-factors, fig.width = figur::convert(figures$clinic_factors, to = 'in')$w, fig.height = figur::convert(figures$clinic_factors, to = 'in')$h, fig.cap = 'Sociodemographic and medical history characteristic of the mental clusters of accident victims.'}
 
@@
 
-__Figure \@ref(fig:fig-clinic-factors). Demographic and clinical factors differing between the mental clusters of accident victims.__
+__Figure \@ref(fig:fig-clinic-factors). Sociodemographic and medical history characteristic of the mental clusters of accident victims.__
 
-_Distribution of gender (A), annual household income (B), and frequencies of self-reported pre-existing somatic conditions (C) and self-reported pre-existing mental illness (D) in the mental clusters.
-Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
-Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
-Effect sizes and p-values are displayed in the plot captions.
-Numbers of observations in the clusters are presented in the X axes._
+_Differences in age during the accident between the mental clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were assessed in the entire cohort by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Differences in genders, education, annual income classes (K: 1000 Euro), percentages of participants with trauma risk profession, traumatic events in the past (DIA-X: Diagnostic Expert System, traumatic event score), as well as pre-existing physical illness and pre-existing professionally diagnosed mental disorders between the mental clusters were investigated in the entire cohort by $\chi^2$ test with Cramer's V effect size statistic.
+P values were adjusted for multiple testing  with the false discovery rate method.
+Age is presented in violin plots with single observations visualized as points and medians with interquartile ranges depicted by red diamonds and whiskers.
+The remaining variables are presented as percentages of the clusters in stack plots.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations in the clusters are indicated in the X axes._
 
@@
 
-```{r fig-injury, fig.width = mm_inch(figures$injury$w), fig.height = mm_inch(figures$injury$h), fig.cap = 'Injury severity and location in the mental clusters.'}
+```{r fig-injury, fig.width = figur::convert(figures$injury, to = 'in')$w, fig.height = figur::convert(figures$injury, to = 'in')$h, fig.cap = 'Accident culprits, accident rescue, injury severity and injured body parts in the mental clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-injury). Injury severity and location in the mental clusters.__
+__Figure \@ref(fig:fig-injury). Accident culprits, accident rescue, injury severity and injured body parts in the mental clusters.__
 
-_Distribution of injury severity grades (A, AIS: abbreviated injury scale), hospitalization (B) and surgery rates (C), and injuries of the body parts (D) in the mental clusters.
-Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
-Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack and bar plots.
-Effect sizes and p-values are displayed in the plot captions or Y axes.
-Numbers of observations in the clusters are presented in the X axes or plot captions._
+_Differences in accident culprit, rescue mode, injury severity, hospitalization and surgery rates, and injured body regions between the mental clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were investigated in the entire cohort by $\chi^2$ test with Cramer's V effect size statistic.
+P values were adjusted for multiple testing with the false discovery rate method._
+
+_(A, B) Percentages of self and non-self accident culprit, distribution of the rescue modes and injury severity classes (AIS: abbreviated injury scale), rates of hospitalization and surgery in the mental clusters were presented in stack plots.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations in the clusters are indicated in the X axes._
+
+_(C) Percentages of injured body regions were presented in a bar plot.
+Effect sizes and p values are indicated in the Y axis.
+Numbers of observations in the clusters are displayed in the plot captions._
 
@@
 
-```{r fig-aftermath, fig.width = mm_inch(figures$aftermath$w), fig.height = mm_inch(figures$aftermath$h), fig.cap = 'Consequences of the accident in the mental clusters.'}
+```{r fig-aftermath, fig.width = figur::convert(figures$aftermath, to = 'in')$w, fig.height = figur::convert(figures$aftermath, to = 'in')$h, fig.cap = 'Psychological support and consequences of the accident in the mental clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-aftermath). Consequences of the accident in the mental clusters.__
+__Figure \@ref(fig:fig-aftermath). Psychological support and consequences of the accident in the mental clusters.__
 
-_Frequencies of self-reported somatic health consequences of the accident (A), flashbacks during sport (B), more or less cautious behavior during sport activity (C), self-reported psychological support need (D) and of a received psychological support (E) in the mental clusters.
-Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
-Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
-Effect sizes and p-values are displayed in the plot captions.
-Numbers of observations in the clusters are presented in the X axes._
+_Differences in rates of received psychological or psychiatric support, self-reported need for psychological support, persistent physical health consequences of the accident, cautious behavior during mountain sport activity and percentages of participants having returned to the same mountain sport activity between the mental clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic.
+P values were adjusted for multiple testing with the false discovery rate method.
+Percentages within the clusters are presented in stack plots.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations in the clusters are indicated in the X axes._
 
@@
 
-```{r fig-early-crf, fig.width = mm_inch(figures$early_crf$w), fig.height = mm_inch(figures$early_crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.'}
+```{r fig-early-class, fig.width = figur::convert(figures$early_class, to = 'in')$w, fig.height = figur::convert(figures$early_class, to = 'in')$h, fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.'}
 
-figures$early_crf$plot
+figures$early_class$plot
 
@@
 
-__Figure \@ref(fig:fig-early-crf). Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.__
+__Figure \@ref(fig:fig-early-class). Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.__
 
-_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident was trained in the training subset of the study cohort._
+_The mental cluster assignment was modeled with demographic, medical history and accident-related explanatory factors available during acute medical management of the accident.
+Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set._
 
-_(A) Prediction accuracy of the classifier in the training and test subset of the study cohort presented in a heat map plot of the confusion matrices.
-Color codes for number of cases.
-The overall prediction accuracy, $\kappa$ and observation numbers are displayed in the plot caption.
-Prediction accuracy for particular clusters is shown in the X axis._
+_(A) Accuracy of the predicted Cluster assignment and predictive performance of the modeling algorithms was assessed by overall cluster assignment accuracy, Cohen's $\kappa$ inter-rater accuracy metric (high values indicate good accuracy) and Brier score (low values indicate good performance) in the training and test subsets of the study cohort.
+Performance metrics are presented in scatter plots.
+Point size codes for the overall cluster assignment accuracy.
+Point color codes for the modeling algorithm.
+Numbers of complete observations are displayed in the plot captions._
 
-_(B) Conditional importance of the explanatory variables for prediction of the mental cluster assignment expressed as accuracy loss ($\Delta$ accuracy)._
+_(B) Sensitivity (Se) and specificity (Sp) of detection of participants assigned to the PTS cluster (post-traumatic stress) investigated by receiver-operating characteristic in the training and test subset of the study cohort.
+Sensitivity and specificity values are indicted in the plots.
+Line color codes for the modeling algorithm.
+Numbers of complete observations and observations in the PTS cluster are indicated in the plot captions._
+
+_RF: random forest; NNet: neural network with a single hidden layer; SVM/radial: support vector machines with radial kernel; RPart: recursive partitioning; SDA: shrinkage discriminant analysis; cForest: conditional random forest; Elastic Net: elastic net multinomial regression._
 

### 2023-08-31 paper/markdown/supplementary_material.Rmd
Author's feedback
Incorporation of the most recent author's feedback into the manuscript text and Supplementary Material

@@
 ---
-title: "Three distinct patterns of mental response after mountain sport accidents"
+title: "Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center"
 subtitle: "Supplementary Material"
@@
 
-Data analysis and transformation was accomplished with R version 4.2.0. 
+Data analysis and transformation was accomplished with R version 4.2.3. 
 The study dataset was imported from an SPSS file with the _foreign_ package [@RCoreTeam]. 
-Tabular data were handled with the _tidyverse_ package bundle [@Wickham2019] and the packages _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
-Test variables were handled with _stringi_ [@Gagolewski2021].
+Tabular data were handled with the _tidyverse_ package bundle [@Wickham2019], and the packages _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
+Text variables were handled with _stringi_ [@Gagolewski2021].
 
@@
 Clustering tendency was investigated with _factoextra_ [@Kassambara2020] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
-Statistical hypothesis testing and correlation analyses were accomplished with [_ExDA_](https://github.com/PiotrTymoszuk/ExDA), _rstatix_ [@Kassambara2021].
+Statistical hypothesis testing and correlation analyses were accomplished with [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and _rstatix_ [@Kassambara2021].
 
-For clustering, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized.
-UMAP (uniform manifold approximation and projection) was done with the _umap_ package [@Konopka2022; @McInnes2018]. 
-Univariable mental cluster assignment classifiers were developed with the _OneR_ [@VonJouanne-Diedrich2017; @Holte1993]. 
-A multi-variable classifier for mental cluster assignment was developed with the conditional random forest algorithm [@Hothorn2006; @Strobl2007] implemented in the _party_ package [@Hothorn2022]. 
-Accuracy of the classifiers was assessed with the packages _caret_ [@Kuhn2008] and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra).
+For semi-supervised clustering, diagnostic and performance of the clustering analysis, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized.
+UMAP (uniform manifold approximation and projection) was done with the _umap_ and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) packages [@Konopka2022; @McInnes2018]. 
+Cross-distances between the training and test subsets of the study cohort and cross-distances between clusters in the training and test subsets of the study cohort were computed with _proxy_ [@Meyer2022].
 
-Results were visualized with _ggplot_ (bar plots, box plots and heat maps of confusion matrices) [@Wickham2016], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (violin, stack and ribbon plots) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) (cluster quality control plots, clustering features heat maps, distance heat maps, scatter plots of UMAP layouts).
-Figures were created with the package _cowplot_ [@Wilke2019]. 
+Multi-variable classifiers of cluster assignment were developed with the following algorithms: 
+random forests [@Breiman2001; @Wright2017], 
+neural network [@Ripley2014], 
+support vector machines with radial kernel [@Weston1998; @Karatzoglou2004], 
+recursive partitioning [@Therneau2022; @Breiman2017], 
+conditional random forests [@Hothorn2006; @Strobl2007; @Hothorn2022], 
+shrinkage discriminant analysis [@Ahdesmaki2010; @Ahdesmaki2022], 
+and elastic net multinomial regression [@Zou2005; @Friedman2010]. 
+For tuning, training, prediction and assessment of performance of the classifiers, the packages _caret_ [@Kuhn2008] and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) were used.
+
+Results were visualized with _ggplot_ (bar plots, box plots, heat maps of cross-distances, scatter plots) [@Wickham2016], _plotroc_ [@Sachs2017], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (violin, stack and ribbon plots) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) (cluster quality control plots, heat maps of clustering features, distance heat maps, scatter plots of UMAP layouts) and _ComplexUpset_ (visualization of overlap with upset plots) [@Krassowski2021].
+Figures were created with the packages _cowplot_ [@Wilke2019] and _patchwork_ [@Pedersen2023]. 
 Tables were generated with _flextable_ [@Gohel2022]. 
-The manuscript and Supplementary Material were written in _rmarkdown_ [@Allaire2022] with the package _bookdown_ [@Xie2016]. 
+The manuscript and Supplementary Material were written in the _rmarkdown_ environment [@Allaire2022] with the package _bookdown_ [@Xie2016]. 
 Figures, tables and R expressions in the markdown documents were managed with the development package [_figur_](https://github.com/PiotrTymoszuk/figur). 
-The markdown documents were rendered with _knitr_ [@Xie2022], _bookdown_ [@Xie2016] with `author-info-blocks.lua` and `scholarly-metadata.lua` scripts by Albert Krewinkel and Jörn Krenzer.
+The markdown documents were rendered with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages and with [`author-info-blocks.lua`](https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/author-info-blocks.lua) and [`scholarly-metadata.lua`](https://github.com/pandoc/lua-filters/blob/master/scholarly-metadata/scholarly-metadata.lua) scripts by Albert Krewinkel and Jörn Krenzer.
 
@@
 persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
-resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
+resilience (RS13: 13-item resilience scale) [@Leppert2008], 
 loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
-loss of quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
+quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
 post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
@@
 
-Symptoms of anxiety were defined as GAD-7 $geq$ 10, symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], significant somatization/persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002].
-Resilient coping classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].
+Clinically relevant symptoms of anxiety were defined as GAD-7 $geq$ 10, 
+clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], 
+significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
+Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].
 
-Items of the EUROHIS QOL 8 tool were scores as with 1 - 5 likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction, ) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing).
+Items of the EUROHIS QOL 8 tool were scored as with 1 - 5 likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing).
 The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].
@@
 Each PCL-5 DSM-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
-Participants positive for the B domain or C domain were identified by at least one item scored with 'moderate', participants positive for the D or E domain were identified by at least two item scored with 'moderate'.
+Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
+Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
 Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 DSM-5 domains [@Bovin2016].
@@
 Prior sport accidents, 
-flashbacks during sport activity, 
-confusion during sport activity, 
-subjective need for psychological support, 
-presence of psychological/psychiatric support/therapy post accident 
-and presence of somatic health consequences after the accident 
+flashbacks during mountain sport activity, 
+confusion during mountain sport activity, 
+self-reported need for psychological support, 
+psychological/psychiatric support/therapy after the accident 
+and presence of persistent physical health consequences after the accident 
 were surveyed as single yes/no items. 
-Flashbacks frequency was assessed in the following categories: none, more than one per month and more than one per year. 
+Flashbacks frequency during mountain sport activity was assessed in the following categories: none, more than one per year and more than one per month. 
 Smoking was surveyed as a single yes/no question. 
-Alcohol abuse was investigated with the CAGE tool with $\geq$ 2 points indicative of substantial addiction [@OBrien2008]. 
-Data on hospital treatment and surgery were extracted from electronic patient's records. 
+Alcohol use was investigated with the CAGE tool with $\geq$ 2 points indicative of problematic alcohol consumption [@OBrien2008]. 
+Data on the type of the accident date and daytime, accident mountain sport, injury diagnosis, injured body regions, injury severity, hospital treatment, surgery and number of ICD-10 surgical diagnoses  were extracted from electronic patient's records.
 Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].
@@
 Consistency of psychometric tools was assessed by factor analysis and McDonald's $\omega$ (function `omega()`, package _psych_) [@McDonald1999; @Revelle2022]. 
-The number of latent factors for calculation of $\omega$ was identified by inspection of loadings determined by factor analysis (function `reduce_data()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)).
+The number of latent factors for calculation of $\omega$ was identified by inspection of loadings determined by factor analysis (function `reduce_data()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Bartlett1937].
 All psychometric tools used in the study except for the stress PSS-4 scale [@Cohen1983] and the BRCS resilience tool [@Sinclair2004] exhibited good-to-excellent consistency with $\omega$ > 0.8 (__Supplementary Table S\@ref(tab:tab-consistency)__).
@@
 For each random subsample, clustering tendency was assessed by Hopkins statistic (function `get_clust_tendency()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
-For n = 200 observations, the median Hopkins statistic was `r pow$hopkins %>% filter(size == 200) %>% .$hopkins_stat %>% median %>% signif(2)` (interquartile range: `r pow$hopkins %>% filter(size == 200) %>% .$hopkins_stat %>% quantile(probs = c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' - ')`) indicative of good clustering tendency.
-The Hopkins statistic value for this subset was comparable with the value for the entire subset (`r signif(pca$clust_tendency$data$hopkins_stat, 2)`). 
-For these reasons, 2/3 of the entire dataset or n = `r count(ptsd$dataset, partition)$n[1]` was considered as the size of the training subset of the study cohort.
+For n = 250 observations, the median Hopkins statistic was 
+`r pow$hopkins %>% filter(size == 250) %>% .$hopkins_stat %>% median %>% signif(2)` (interquartile range: 
+`r pow$hopkins %>% filter(size == 250) %>% .$hopkins_stat %>% quantile(probs = c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' - ')`) indicative of good clustering tendency. 
+The Hopkins statistic value for this subset was comparable with the value for the entire subset 
+(`r signif(pca$clust_tendency$data$hopkins_stat, 2)`). 
+For these reasons, 3/4 of the entire dataset or n = 
+`r count(ptsd$dataset, partition)$n[1]` was considered as the adequate size of the training subset of the study cohort for a reproducible clustering analysis.
 
@@
 
-The study participants were assigned to the training and test subsets with the 2:1 size ratio (__Figure 1__). 
-To this end, `r rand$part_obs %>% length` random splits of the datasets were generated. 
-The subset assignment scheme was chosen with possibly small differences in accident year, age, gender, pre-existing somatic and pre-existing mental illness rates, frequencies of prior traumatic events (DIA-X) [@Maercker2005] and distribution of injury severity grades (AIS: abbreviated injury scale) [@Gennarelli2006] as investigated by $\chi^2$ and Mann-Whitney test, as appropriate (functions `chisq_test()` and `wilcox_test()`, package _rstatix_).
+The study participants were assigned to the training and test subsets with the 3:1 size ratio (__Figure 1__). 
+To this end, `r nrow(part$dists)` random splits of the datasets were generated. 
+The subset assignment scheme was chosen with the possibly smallest differences in sociodemographic, medical history, clinical and accident- and injury-related variables assessed by Gower distance between the training and test subsets (function `dist()`, package _proxy_) [@Meyer2022].
+
+## Statistical hypothesis testing, effect size and multiple testing correction
+
+Differences in numeric variables were assessed with Mann-Whitney test with r effect size statistic or Kruskal-Wallis test with $\eta^2$ effect size statistic for two and more than two analysis groups, respectively.
+Differences in frequency of categories of qualitative variables between analysis groups were investigated with $\chi^2$ test with Cramer's V effect size statistic. 
+P values were corrected for multiple testing with the false discovery rate method separately for each analysis task (e.g. comparison of clusters) [@Benjamini1995]. 
+Effects with p < 0.05 following the false discovery rate adjustment were considered significant. 
+Effect size of accuracy of predicted cluster assignment by machine learning classifiers was assessed by Cohen's $\kappa$ inter-rater reliability statistic [@Cohen1960; @McHugh2012]. 
+Intervals of effect sizes were defined as follows [@McHugh2012; @Field2013; @Cohen2013]:
+
+* r statistic: weak: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+* Cramer's V statistic: weak: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+* $\eta^2$ statistic: weak: < 0.06, moderate: 0.06 - 0.14, large: $\geq$ 0.14
+
+* Cohen's $\kappa$ reliability: none: < 0.2, minimal: 0.2 - 0.4, weak: 0.4 - 0.6, moderate: 0.6 - 0.8, strong: $\geq$ 0.8
 
@@
 
-To assess possible selection bias, demographic, socioeconomic, clinical and accident-related parameters were compared between individuals excluded due to missingness of psychometric data or denying survey participation and participants included in he analysis (__Figure 1__).
-Categorical variables were compared by $\chi^2$ test with Cramer V effect size statistic and numeric variables were compared by Mann-Whitney test with r effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+To assess the possible selection bias, demographic, socioeconomic, clinical, accident- and recovery-related parameters were compared between individuals excluded due to missingness of psychometric data or denying survey participation and participants included in the analysis (__Figure 1__).
+Categorical variables were compared by $\chi^2$ test with Cramer V effect size statistic and numeric variables were compared by Mann-Whitney test with r effect size statistic 
+(function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
 Potential differences between the training and test subset of the study cohort (__Figure 1__) were analyzed in an analogical way. 
-Significant (p < 0.05) differences between the included/excluded participants are presented in __Supplementary Tables S\@ref(tab:tab-incl-excl)__.
+Significant (p < 0.05) differences between the included/excluded participants are presented in __Supplementary Table S\@ref(tab:tab-no-response)__ and __S\@ref(tab:tab-missing)__.
 
@@
 The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
-The choice of the clustering algorithm was motivated by its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares, method `var()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and superior stability in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 7-nearest neighbors classifier, method `cv()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__.
-The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
-By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
-Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
+The choice of the clustering algorithm was motivated by its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares, method `var()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier, method `cv()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__.
+The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020].
+By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).
+
+Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
 This semi-supervised procedure yielded clustering structures with similar fractions of explained variance in the training (V = `r signif(semi_clust$variance$variance[1], 2)`) and test subset (V = `r signif(semi_clust$variance$variance[2], 2)`).
-In addition, comparable separation between the clusters could be discerned in the training and test subsets by a visual analysis of UMAP layouts, pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__) and heat maps of normalized levels of the clustering variables (__Supplementary Figure S\@ref(fig:fig-feat-hm)__).
+Comparably good separation of the clusters could be discerned in the training and test subsets by 
+a visual analysis of UMAP layouts and 
+pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
+Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
+An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental clusters (i.e.
+neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g.
+neutral vs PTG) (distances computed with the function `dist()`, package _proxy_; __Supplementary Figure S\@ref(fig:fig-cosine)B__).
+Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).
 
-Differences in clustering variables, frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical and accident-related factors between the mental clusters were assessed separately in the training and test subset by $\chi^2$ test with Cramer V effect size statistic and Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (__Supplementary Tables S\@ref(tab:tab-clust-fct)__ and __S\@ref(tab:tab-clust-bcg)__).
+Differences in frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical, accident- and recovery-related factors between the mental clusters were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic and by Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Supplementary Table S\@ref(tab:tab-clust-bcg)__).
 
@@
 
-Two multi-parameter classifiers of the mental cluster assignment were developed in the training subset:
+Two types of multi-parameter machine learning classifiers of the mental cluster assignment were developed in the training subset:
 
-1. a model employing candidate early predictors of cluster assignment, i.e. demographic, socioeconomic, medical history and accident related explanatory factors
+1.
+models employing candidate early predictors of cluster assignment, i.e.
+demographic, socioeconomic, medical history and accident-related explanatory factors available during acute medical management of the accident
 
-2. a model including additionally late consequences of the accident such as somatic health aftermath, flashbacks or cautious behavior during sport
+2. models including additionally recovery-related predictors such as persistent physical health consequences, flashbacks or cautious behavior during sport
 
-The multi-parameter classifiers (__Supplementary Table S\@ref(tab:tab-mod-variables)__) were trained in the training subset with the conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022] with n = 1000 random trees and the splitting rule corresponding to the maximum test statistic.
-In the model 1, the `mtry` parameter was set to 7, `minsplit` to 10 and `mincriterion` to 1.28 (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 7, mincriterion = qnorm(0.9), minsplit = 10, ntree = 1000)` control object, package _party_).
-In the model 2, the `mtry` parameter was set to 6, `minsplit` to 20 and `mincriterion` to 0 (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 6, mincriterion = 0, minsplit = 20, ntree = 1000)` control object, package _party_).
-The optimal parameter sets were obtained by optimization (tuning) employing the out-of-bag predictions in the training subsets (not shown). 
-The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _party_).
+Of note, psychometric variables used for definition of the mental clusters as well as mental disorder symptoms, presence and frequency of flashbacks were excluded from the explanatory variable sets. 
+The explanatory variables are listed in __Supplementary Table S\@ref(tab:tab-mod-variables)__.
 
-Accuracy and $\kappa$ statistic for the classifiers' mental cluster assignment predictions in the training and test subsets were computed with the function `multiClassSummary()` from the package _caret_ [@Kuhn2008].
-Permutation importance for explanatory variables expressed as accuracy loss [@Strobl2008] was obtained for the conditional random forest algorithm with the function 'varimp()` from the _party_ package.
+The models employed the following algorithms: 
+canonical random forests [@Breiman2001; @Wright2017], 
+regularized neural networks with a single hidden layer [@Ripley2014], 
+support vector machines with radial kernel [@Weston1998; @Karatzoglou2004], 
+recursive partitioning [@Therneau2022; @Breiman2017], 
+shrinkage discriminant analysis [@Ahdesmaki2010; @Ahdesmaki2022], 
+conditional random forest [@Hothorn2006; @Strobl2007; @Hothorn2022], 
+and elastic net multinomial regression [@Zou2005; @Friedman2010]. 
+The optimal values of the algorithms' parameters were found by 10-fold cross-validation-based tuning with the maximal value of Cohen's $\kappa$ [@Cohen1960] as the tuning criterion. 
+The tuning and fitting ('training') in the training subset of the study cohort was done with the wrapper function `train()` provided by the package _caret_ [@Kuhn2008]. 
+For the random forest and conditional random forest algorithms, 1000 random trees each were constructed. 
+Test statistic, p values, number of splits and other parameters of the conditional forest models were controlled with the `ForestControl` object returned by the convenience wrapper `cforest_unbiased()` provided by the _party_ package [@Hothorn2022].
+The optimal algorithm parameter sets are listed in __Supplementary Table S\@ref(tab:tab-tuning)__. 
+Predictions of the cluster assignment in the test subset of the study cohort were obtained with the `predict()` method from the [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) package.
+
+Overall accuracy and Cohen's $\kappa$ statistics [@Cohen1960] were computed with the `summary()` method from the [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) package. 
+Brier scores [@Brier1950] were computed with the following formula:
+
+$$BS = \frac{1}{N}\times\Sigma_{i = 1}^{N}\Sigma_{c = 1}^R(f_{ic} - o_{ic})^2$$
+where $f_{tc}$ is the output probability of assignment of the i-th observation to the c-th cluster, 
+$o_{ic}$ is the numeric-coded actual assignment of the i-th observation to the c-th cluster, 
+$R$ is the total cluster number and 
+$N$ is the total observation number. 
+Brier skill scores [@Brier1950] were computed with the following formula:
+
+$$BSS = 1 - \frac{BS_{model}}{BS_{random}}$$
+
+where $BS_{model}$ is the Brier score of the model and 
+$BS_{random}$ is the Brier score obtained for a purely random cluster assignment. 
+Brier scores, Brier skill scores, sensitivity and specificity of the PTS cluster assignment were computed with in-house developed R functions. 
+Performance statistics for the training subset, 10-fold cross-validation and test subset of the study cohort are listed in __Supplementary Table S\@ref(tab:tab-early-classifiers)__ and __Supplementary Table S\@ref(tab:tab-full-classifiers)__ for the classifiers employing the early and full predictor set, respectively.
+
+Variable importance statistic specific for the machine learning algorithm were extracted from the _caret_ models with the `varImp()` function from package _caret_ [@Kuhn2008]. 
+They were: permutation importance for the random forest and conditional forest algorithm [@Hothorn2022; @Breiman2001], 
+connection weight importance statistic for the neural network [@Garson1991], 
+sum reduction in classification error attributed to each variable at each split for recursive partitioning [@Breiman2017], 
+and linear model coefficient $\beta$ for elastic net regression [@Zou2005]. 
+For support vector machines and discriminant analysis, the variable importance was computed based on area under the ROC curve for single predictors for discrimination between the cluster pairs [@Kuhn2008].
 
@@
 
-An R data (RData) file with anonymized patient data will be made available upon request to the corresponding author. 
+An RData file with anonymized patient data will be made available upon request to the corresponding author. 
 The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.
@@
 
-```{r tab-analysis-vars, tab.cap = 'Variables used in the analysis pipeline. The table is available in a supplementary Excel file.'}
+```{r tab-analysis-vars, tab.cap = 'Variables used in the analysis pipeline.'}
 
@@
   set_names(' ') %>% 
-  flextable
+  flextable %>% 
+  set_widths(16)
 
@@
 
-suppl_tables$mental_battery[c(2, 3)] %>% 
-  flextable %>% 
-  width(1, width = 6.5, unit = 'cm') %>% 
-  width(2, width = 10.5, unit = 'cm') %>% 
-  footnote(1, 1:2, 
-           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; BRCS: brief resilent coping scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+flextable::flextable(suppl_tables$mental_battery) %>% 
+  set_widths(c(2.3, 7, 10)) %>% 
+  merge_v(1) %>% 
+  footnote(1, 2:3, 
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
            part = 'header', 
@@
 flextable::flextable(suppl_tables$consistency) %>% 
-  flextable::width(1, width = 3.5, unit = 'cm') %>% 
-  flextable::width(2:3, width = 3, unit = 'cm') %>% 
+  set_widths(c(3.5, 3, 3)) %>% 
    footnote(1, 1, 
-           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; BRCS: brief resilent coping scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
            part = 'header', 
@@
 
-```{r tab-incl-excl, tab.cap = 'Significant (p < 0.05) differences between individuals excluded from analysis and analyzed study participants.
+```{r tab-no-response, tab.cap = 'Significant differences between patients who did not respond to the study invitation and the analyzed study participants.
 Numeric variables are presented as medians with interquartile ranges (IQR).
@@
 
-suppl_tables$incl_excl %>% 
-  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
-  flextable %>% 
-  flextable::width(1, width = 3.2, unit = 'cm') %>% 
-  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
-  flextable::width(4, width = 2.7, unit = 'cm') %>% 
-  flextable::width(5, width = 2.3, unit = 'cm') %>% 
+flextable::flextable(suppl_tables$no_response) %>% 
+  set_widths(c(3.2, 5.8, 5.8, 2.5, 2.2)) %>% 
    footnote(1, 1, 
-           value = as_paragraph('AIS: abbreviated injury scale; PTGI: post-traumatic growth inventory; SOC-9L: Leipzig 9-item sense of coherence scale; K: 1000 Euro; EUR: Euro.'), 
+           value = as_paragraph('AIS: abbreviated injury scale'), 
            part = 'header', 
@@
   footnote(1, 4:5, 
-           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic, '), 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.
+P values corrected for multiple testing with the false discovery rate method.'),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-missing, tab.cap = 'Significant differences between the study survey responders excluded from analysis due to missingness of psychometric data and the analyzed study participants.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+flextable::flextable(suppl_tables$missing) %>% 
+  set_widths(c(3.2, 5.8, 5.8, 2.5, 2.2)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('AIS: abbreviated injury scale; K: 1000 Euro; EUR: Euro.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.
+P values corrected for multiple testing with the false discovery rate method.'),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-train-test, tab.cap = 'Significant differences between the training and test subset of the study cohort.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+flextable::flextable(suppl_tables$train_test) %>% 
+  set_widths(c(4, 4, 4, 2.5, 2.2)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('RS13: 13-item resilience scale; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.
+P values corrected for multiple testing with the false discovery rate method.'),
            part = 'header', 
@@
 Numeric variables are presented as medians with interquartile ranges (IQR).
+Statistical significance was determined by false discovery rate-corrected Kruskal-Wallis test with eta-square effect size statistic.
 The table is available in a supplementary Excel file.'}
@@
 
-```{r tab-clust-bcg, tab.cap = 'Significant (p < 0.05) differences in demographic, socioeconomic, clinical and accident-related factors between the mental clusters.
+```{r tab-clust-bcg, tab.cap = 'Significant differences in demographic, socioeconomic, clinical, accident- and recovery-related factors, and mental disorder symptoms between the mental clusters in the entire cohort.
 Numeric variables are presented as medians with interquartile ranges (IQR).
@@
 
-suppl_tables$clust_bcg %>% 
-  flextable %>% 
-  width(1, width = 1.7, unit = 'cm') %>% 
-  width(2, width = 3.7, unit = 'cm') %>% 
-  width(3:5, width = 3.5, unit = 'cm') %>% 
-  width(6, width = 2.4, unit = 'cm') %>% 
-  width(7, width = 2.2, unit = 'cm') %>% 
-  merge_v(1) %>% 
-  footnote(1, 2, 
-           value = as_paragraph('CAGE: Cut, Annoyed, Guilty, and Eye scale; AIS: abbreviated injury score.'), 
+flextable::flextable(suppl_tables$clust_bcg) %>% 
+  set_widths(c(3, rep(4.3, 3), 2.5, 2.3)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('AIS: abbreviated injury score; PTSD: post-traumatic stress disorder; RS13: 13-item resilience scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire.'),
            part = 'header', 
            ref_symbols = 'a') %>% 
-  footnote(1, 6:7, 
-           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
+  footnote(1, 5:6, 
+           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic.
+Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.'),
            part = 'header', 
@@
 
-```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors (early: available during acute medical management of the accident) used for modeling of the mental cluster assignment.'}
+```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors used for modeling of the mental cluster assignment.'}
 
 flextable::flextable(suppl_tables$mod_variables) %>% 
-  width(1, width = 3, unit = 'cm') %>% 
-  width(2, width = 12, unit = 'cm') %>% 
+  set_widths(c(2.5, 15.5)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('early predictors: variables available during acute medical management of the accident; full predictor set: variables available during acute medical management of the accident and during follow-up.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-tuning, tab.cap = 'The optimal combinations of machine learning algorithm parameters found in 10-fold cross-validation of the training subset of the study cohort.'}
+
+flextable::flextable(suppl_tables$tuning) %>% 
+  set_widths(c(2, 2.5, 8.5)) %>% 
+  merge_v(1) %>%
+  footnote(1, 1, 
+           value = as_paragraph('early predictors: variables available during acute medical management of the accident; full predictor set: variables available during acute medical management of the accident and during follow-up.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('RF: random forest; NNet: neural network with a single hidden layer; SVM/radial: support vector machines with radial kernel; RPart: recursive partitioning; SDA: shrinkage discriminant analysis; cForest: conditional random forest; Elastic Net: elastic net multinomial regression.'),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-early-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental cluster assignment.
+Models employing early predictors available during acute medical management of the accident.'}
+
+flextable::flextable(suppl_tables$early_classifiers) %>% 
+  set_widths(c(2.3, 2.5, rep(2.3, 6))) %>% 
+  merge_v(1) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('RF: random forest; NNet: neural network with a single hidden layer; SVM/radial: support vector machines with radial kernel; RPart: recursive partitioning; SDA: shrinkage discriminant analysis; cForest: conditional random forest; Elastic Net: elastic net multinomial regression.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('CV: cross-validation'), 
+           part = 'header',  
+           ref_symbols = 'b') %>% 
+  footnote(1, 6, 
+           value = as_paragraph('Brier Skill Score comparing the given classifier with the purely random cluster assignment.'), 
+           part = 'header', 
+           ref_symbols = 'c') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-full-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental cluster assignment.
+Models employing the full predictor set available during acute medical management of the accident and follow-up.'}
+
+flextable::flextable(suppl_tables$full_classifiers) %>% 
+  set_widths(c(2.3, 2.5, rep(2.3, 6))) %>% 
+  merge_v(1) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('RF: random forest; NNet: neural network with a single hidden layer; SVM/radial: support vector machines with radial kernel; RPart: recursive partitioning; SDA: shrinkage discriminant analysis; cForest: conditional random forest; Elastic Net: elastic net multinomial regression.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('CV: cross-validation'), 
+           part = 'header',  
+           ref_symbols = 'b') %>% 
+  footnote(1, 6, 
+           value = as_paragraph('Brier Skill Score comparing the given classifier with the purely random cluster assignment.'), 
+           part = 'header', 
+           ref_symbols = 'c') %>% 
   theme_vanilla
@@
 
-```{r fig-parts-mental-cohort, fig.width = mm_inch(suppl_figures$parts_mental_cohort$w), fig.height = mm_inch(suppl_figures$parts_mental_cohort$h), fig.cap = 'Injured body regions and detailed scoring of quality of life, post-traumatic stress disorder and post-traumatic growth in the study cohort.'}
+```{r fig-parts-ptsd, fig.width = figur::convert(suppl_figures$parts_ptsd, to = 'in')$w, fig.height = figur::convert(suppl_figures$parts_ptsd, to = 'in')$h, fig.cap = 'Injured body regions and symptoms of PTSD in the study cohort.'}
 
-suppl_figures$parts_mental_cohort$plot
+suppl_figures$parts_ptsd$plot
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-parts-mental-cohort).
-Injured body regions and detailed scoring of quality of life, post-traumatic stress disorder and post-traumatic growth in the study cohort.__
+__Supplementary Figure S\@ref(fig:fig-parts-ptsd). Injured body regions and symptoms of PTSD in the study cohort.__
 
-_For categorical variables, percentages of complete observations are presented as bars.
-Numeric variables are presented in violin plots with red diamonds denoting medians, interquartile ranges presented as red whiskers and single observations depicted as points.
-Quality of life scores are expressed as categorical variables and presented as percentages of participants in stack plots.
-The number of complete cases is displayed in the plot captions._
-_(A) Distribution of injured body regions.
-(B) Scoring of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8).
-(C) Scoring of the post-traumatic stress disorder (PTSD) domains with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD domains.
-(E) Scoring of the factors of the post-traumatic growth with the post-traumatic growth inventory (PTGI) tool._
+_(A) Distribution of injured body regions presented in a bar plot. The number of complete observations is indicated in the plot caption._
+
+_(B) Scores of particular domains of the PCL-5 tool evaluating symptoms of post-traumatic stress disorder (PTSD) presented in violin plots.
+Single observations are visualized as points.
+Medians and interquartile ranges are represented by red diamonds and whiskers.
+The number of complete observations is indicated in the plot caption._
+
+_(C) Frequency and overlap of clinically relevant PTSD symptoms captured by particular domains of the PCL-5 scale presented in an upset plot.
+Percentages of participants screened positive for the domain B, C, D or E PTSD symptoms are displayed in as red bars.
+Percentages of participants with single and overlapping domain B, C, D or E PTSD symptoms are presented as green bars.
+The number of complete observations is indicated in the plot caption._
 
@@
 
-```{r fig-clust-dev, fig.width = mm_inch(suppl_figures$clust_dev$w), fig.height = mm_inch(suppl_figures$clust_dev$h), fig.cap = 'Definition of the mental clusters in the training subset of the study cohort.'}
+```{r fig-qol-ptg, fig.width = figur::convert(suppl_figures$qol_ptg, to = 'in')$w, fig.height = figur::convert(suppl_figures$qol_ptg, to = 'in')$h, fig.cap = 'Scores of quality of life and post-traumatic growth in the study cohort.'}
+
+suppl_figures$qol_ptg$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-qol-ptg). Scores of quality of life and post-traumatic growth in the study cohort.__
+
+_(A) Scores of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8) presented in a bar plot. The number of complete observations is displayed in the plot caption._
+
+_(B) Scores of the domains of the post-traumatic growth inventory (PTGI) scale presented in a violin plots.
+Single observations are visualized as points.
+Red diamonds with whiskers represent medians with interquartile ranges.
+The number of complete observations is indicated in the plot caption._
+
+\newpage
+
+```{r fig-clust-dev, fig.width = figur::convert(suppl_figures$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_figures$clust_dev, to = 'in')$h, fig.cap = 'Definition of the mental clusters in the training subset of the study cohort.'}
 
@@
 
-_(A) Comparison of explanatory performance and stability of various clustering algorithms in the training subset.
+_(A) Comparison of explanatory performance and reproducibility of various clustering algorithms in the training subset.
 The explanatory performance was measured as a fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares).
-The stability was assessed by a rate of correct cluster assignment in 10-fold cross-validation (CV) with cluster assignment in the folds by an inverse distance weighted 7-nearest neighbors classifier.
+The reproducibility was assessed by the rate of correct cluster assignment in 10-fold cross-validation (CV) with cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier.
 Note the superior stability of the PAM/cosine distance algorithm._
@@
 
-```{r fig-semi-clust, fig.width = mm_inch(suppl_figures$semi_clust$w), fig.height = mm_inch(suppl_figures$semi_clust$h), fig.cap = 'Semi-supervised clustering.'}
+```{r fig-semi-clust, fig.width = figur::convert(suppl_figures$semi_clust, to = 'in')$w, fig.height = figur::convert(suppl_figures$semi_clust, to = 'in')$h, fig.cap = 'Semi-supervised clustering.'}
 
@@
 _The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations.
-Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier.
-By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden).
+Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 27-nearest neighbors classifier.
+By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress).
 Numbers of observations in the mental clusters are shown in the plot legend in (A)._
@@
 
-```{r fig-feat-hm, fig.width = mm_inch(suppl_figures$feat_hm$w), fig.height = mm_inch(suppl_figures$feat_hm$h), fig.cap = 'Levels of psychometric clustering scores in the mental clusters.'}
+```{r fig-cosine, fig.width = figur::convert(suppl_figures$cosine, to = 'in')$w, fig.height = figur::convert(suppl_figures$cosine, to = 'in')$h, fig.cap = 'Distribution of the mental clusters and cosine distances between the mental clusters in the training and test subset of the study cohort.'}
+
+suppl_figures$cosine$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-cosine). Distribution of the mental clusters and cosine distances between the mental clusters in the training and test subset of the study cohort.__
+
+_Study cohorts observations were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) clusters by semi-supervised clustering as presented in Supplementary Figure S\@ref(fig:fig-semi-clust)._
+
+_(A) Percentages of observations in the training and test subsets of the study cohort in the mental clusters. Numbers of complete observations are indicated in the Y axis._
+
+_(B) Cosine cross-distances between the mental clusters in the training and test subset of the study dataset visualized as a heat map. Cross-distances are indicated in the tiles._
+
+\newpage
+
+```{r fig-feat-hm, fig.width = figur::convert(suppl_figures$feat_hm, to = 'in')$w, fig.height = figur::convert(suppl_figures$feat_hm, to = 'in')$h, fig.cap = 'Levels of psychometric scores used for the cluster definition in the mental clusters.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-feat-hm). Levels of psychometric clustering scores in the mental clusters.__
+__Supplementary Figure S\@ref(fig:fig-feat-hm). Levels of psychometric scores used for the cluster definition in the mental clusters.__
 
@@
 
-```{r fig-demo, fig.width = mm_inch(suppl_figures$demo$w), fig.height = mm_inch(suppl_figures$demo$h), fig.cap = 'Age, education, household income, prior accidents and traumatic events in the mental clusters.'}
+```{r fig-demo, fig.width = figur::convert(suppl_figures$demo, to = 'in')$w, fig.height = figur::convert(suppl_figures$demo, to = 'in')$h, fig.cap = 'Employment status, prior sport accidents, and accident details in the mental clusters.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-demo). Age, education, household income, prior accidents and traumatic events in the mental clusters.__
+__Supplementary Figure S\@ref(fig:fig-demo). Employment status, prior sport accidents, and accident details in the mental clusters.__
 
-_Distribution of age classes (A, young: < 30, middle: 30 - 65, elderly > 65 years), education grades (B), frequencies of prior sport accidents (D) and prior traumatic events (E, measured by the DIA-X tool) in the mental clusters.
+_Distribution of employment status, frequency of sport accidents in the past, distribution of the accidents in the seasons, accident sport types, frequency of being alone during the accident and number of injured persons in the mental clusters.
 Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
-Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
+P values were corrected for multiple testing with the false discovery rate method.
+Percentages of variable's categories in the entire study cohort are presented in stack plots.
 Effect sizes and p-values are displayed in the plot captions.
@@
 
-```{r fig-accident, fig.width = mm_inch(suppl_figures$accident$w), fig.height = mm_inch(suppl_figures$accident$h), fig.cap = 'Accident sport type and accident rescue in the mental clusters.'}
+```{r fig-importance-early, fig.width = figur::convert(suppl_figures$importance_early, to = 'in')$w, fig.height = figur::convert(suppl_figures$importance_early, to = 'in')$h, fig.cap = 'Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
+Predictors available during acute medical management of the accident.'}
 
-suppl_figures$accident$plot
+suppl_figures$importance_early$plot
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-accident). Accident sport type and accident rescue in the mental clusters.__
+__Supplementary Figure S\@ref(fig:fig-importance-early).
+Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
+Predictors available during acute medical management of the accident.__
 
-_Distribution of sport types (A) and rescue modes (B, C) in the mental clusters.
-Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
-Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots.
-Effect sizes and p-values are displayed in the plot captions.
-Numbers of observations in the clusters are presented in the X axes._
+_The cluster assignment was modeled with demographic, medical history and accident-related explanatory factors available during acute medical management of the accident.
+Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set.
+Variable importance metrics of machine learning algorithms with the best performance at predicting the mental cluster assignment in the test subset of the study cohorts were computed.
+Importance metrics for the top 20 most important variables are presented as bar plots._
+_(A) Random forest, (B) support vector machines with radial kernel, (C) shrinkage discriminant analysis, (D) conditional random forest._
 
@@
 
-```{r fig-crf, fig.width = mm_inch(suppl_figures$crf$w), fig.height = mm_inch(suppl_figures$crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical, accident- and accident consequence-related factors by the random forest algorithm.'}
+```{r fig-full-class, fig.width = figur::convert(suppl_figures$full_class, to = 'in')$w, fig.height = figur::convert(suppl_figures$full_class, to = 'in')$h, fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident and long-term follow-up.'}
 
-suppl_figures$crf$plot
+suppl_figures$full_class$plot
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-crf).
-Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical, accident- and accident consequence-related factors by the random forest algorithm.__
+__Supplementary Figure S\@ref(fig:fig-full-class).
+Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident and long-term follow-up.__
 
-_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident- and accident consequence-related factors was trained in the training subset of the study cohort._
+_The cluster assignment was modeled with demographic, medical history, accident- and recovery-related explanatory factors available during acute medical management of the accident and follow-up.
+Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set._
 
-_(A) Prediction accuracy of the classifier in the training and test subset of the study cohort presented in a heat map plot of the confusion matrices.
-Color codes for number of cases.
-The overall prediction accuracy, $\kappa$ and observation numbers are displayed in the plot caption.
-Prediction accuracy for particular clusters is shown in the X axis._
+_(A) Accuracy of the predicted Cluster assignment and predictive performance of the modeling algorithms was assessed by overall cluster assignment accuracy, Cohen's $\kappa$ inter-rater accuracy metric (high values indicate good accuracy) and Brier score (low values indicate good performance) in the training and test subsets of the study cohort.
+Performance metrics are presented in scatter plots.
+Point size codes for the overall cluster assignment accuracy.
+Point color codes for the modeling algorithm.
+Numbers of complete observations are displayed in the plot captions._
 
-_(B) Conditional importance of the explanatory variables for prediction of the mental cluster assignment expressed as accuracy loss ($\Delta$ accuracy)._
+_(B) Sensitivity (Se) and specificity (Sp) of detection of participants assigned to the PTS cluster (post-traumatic stress) investigated by receiver-operating characteristic in the training and test subset of the study cohort.
+Sensitivity and specificity values are indicated in the plots.
+Line color codes for the modeling algorithm.
+Numbers of complete observations and observations in the PTS cluster are indicated in the plot captions._
+
+_RF: random forest; NNet: neural network with a single hidden layer; SVM/radial: support vector machines with radial kernel; RPart: recursive partitioning; SDA: shrinkage discriminant analysis; cForest: conditional random forest; Elastic Net: elastic net multinomial regression._
+
+\newpage
+
+```{r fig-importance-full, fig.width = figur::convert(suppl_figures$importance_full, to = 'in')$w, fig.height = figur::convert(suppl_figures$importance_full, to = 'in')$h, fig.cap = 'Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
+Predictors available during acute medical management of the accident and during the follow-up.'}
+
+suppl_figures$importance_full$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-importance-full).
+Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
+Predictors available during acute medical management of the accident and during the follow-up.__
+
+_The cluster assignment was modeled with demographic, medical history, accident- and recovery-related explanatory factors available during acute medical management of the accident and follow-up.
+Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set.
+Variable importance metrics of machine learning algorithms with the best performance at predicting the mental cluster assignment in the test subset of the study cohorts were computed.
+Importance metrics for the top 20 most important variables are presented as bar plots._
+_(A) Random forest, (B) support vector machines with radial kernel, (C) shrinkage discriminant analysis, (D) conditional random forest._
 

### 2023-09-04 paper/markdown/manuscript.Rmd
Update of the manuscript
Update of the manuscript, figures and supplementary figures - using improved functionalities of the 'clustTools' package

@@
 The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
-This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares) and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical clustering and KMEANS algorithms.
+This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares), separation between clusters (average silhouette width) [@Rousseeuw1987] and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical clustering and KMEANS algorithms.
 The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
@@
 Subsequently, the mental cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
-(training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`), 
+(training: `r signif(semi_clust$stats$variance[1], 2)`, test: `r signif(semi_clust$stats$variance[2], 2)`), 
+comparable average silhouette statistics 
+(training: `r signif(semi_clust$stats$mean_sil[1], 2)`, test: `r signif(semi_clust$stats$mean_sil[2], 2)`) [@Rousseeuw1987], 
 comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).

### 2023-09-04 paper/markdown/supplementary_material.Rmd
Update of the manuscript
Update of the manuscript, figures and supplementary figures - using improved functionalities of the 'clustTools' package

@@
 UMAP (uniform manifold approximation and projection) was done with the _umap_ and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) packages [@Konopka2022; @McInnes2018]. 
-Cross-distances between the training and test subsets of the study cohort and cross-distances between clusters in the training and test subsets of the study cohort were computed with _proxy_ [@Meyer2022].
+Cross-distances between the training and test subsets of the study cohort and cross-distances between clusters in the training and test subsets of the study cohort were computed with _philentropy_ [@Drost2018] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).
 
@@
 To find the optimal size of a training subset of the study cohort, random subsamples of the study dataset of varying sizes were investigated (`r paste(pow$sizes, collapse = ', ')` observations, 50 random draws per subsample size).
-For each random subsample, clustering tendency was assessed by Hopkins statistic (function `get_clust_tendency()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
+For each random subsample, clustering tendency was assessed by Hopkins statistic (function `get_clust_tendency()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Hopkins1954]. 
 For n = 250 observations, the median Hopkins statistic was 
@@
 Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
-This semi-supervised procedure yielded clustering structures with similar fractions of explained variance in the training (V = `r signif(semi_clust$variance$variance[1], 2)`) and test subset (V = `r signif(semi_clust$variance$variance[2], 2)`).
+This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
+training (variance = `r signif(semi_clust$stats$variance[1], 2)`, 
+average silhouette = `r signif(semi_clust$stats$mean_sil[1], 2)`) and 
+test subset (variance = `r signif(semi_clust$stats$variance[2], 2)`, 
+average silhouette = `r signif(semi_clust$stats$mean_sil[2], 2)`). 
 Comparably good separation of the clusters could be discerned in the training and test subsets by 
@@
 neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g.
-neutral vs PTG) (distances computed with the function `dist()`, package _proxy_; __Supplementary Figure S\@ref(fig:fig-cosine)B__).
+neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__).
 Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).
@@
 
-_(A) Comparison of explanatory performance and reproducibility of various clustering algorithms in the training subset.
+_(A) Comparison of explanatory performance, cluster separation and reproducibility of various clustering algorithms in the training subset.
 The explanatory performance was measured as a fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares).
+The cluster separation was assessed with average silhouette width.
 The reproducibility was assessed by the rate of correct cluster assignment in 10-fold cross-validation (CV) with cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier.

### 2023-10-11 paper/markdown/manuscript.Rmd
clustTools and caretExtra compatibility
Compatibility with the newest package versions and new tools

@@
 (`r paste0((filter(early_class$overall_stats, dataset == 'train')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`% of correctly assigned observations, 
-Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$kappa), 2) %>% paste(collapse = ' - ')`, Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$bs), 2) %>% paste(collapse = ' - ')`).
+Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$kappa), 2) %>% paste(collapse = ' - ')`, Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$brier_score), 2) %>% paste(collapse = ' - ')`).
 However, their accuracy in cross-validation and in the test subset was poor 
@@
 Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$kappa), 2) %>% paste(collapse = ' - ')`, 
-Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$bs), 2) %>% paste(collapse = ' - ')`). 
+Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`). 
 Concerning the early identification of the vulnerable PTS cluster patients, the conditional random forest, random forest and support vector machine models were able to differentiate PTS cluster patients from the pooled neutral + PTG cluster subset with the best, at least
@@
 Cohen's $\kappa$: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$kappa), 2) %>% paste(collapse = ' - ')`, 
-Brier score: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$bs), 2) %>% paste(collapse = ' - ')`).
+Brier score: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`).
 

### 2023-10-11 paper/markdown/supplementary_material.Rmd
clustTools and caretExtra compatibility
Compatibility with the newest package versions and new tools

@@
 $N$ is the total observation number.
-Brier skill scores [@Brier1950] were computed with the following formula:
 
-$$BSS = 1 - \frac{BS_{model}}{BS_{random}}$$
-
-where $BS_{model}$ is the Brier score of the model and 
-$BS_{random}$ is the Brier score obtained for a purely random cluster assignment. 
-Brier scores, Brier skill scores, sensitivity and specificity of the PTS cluster assignment were computed with in-house developed R functions. 
 Performance statistics for the training subset, 10-fold cross-validation and test subset of the study cohort are listed in __Supplementary Table S\@ref(tab:tab-early-classifiers)__ and __Supplementary Table S\@ref(tab:tab-full-classifiers)__ for the classifiers employing the early and full predictor set, respectively.
@@
 flextable::flextable(suppl_tables$early_classifiers) %>% 
-  set_widths(c(2.3, 2.5, rep(2.3, 6))) %>% 
+  set_widths(c(2.3, 2.5, rep(2.3, 5))) %>% 
   merge_v(1) %>% 
@@
            ref_symbols = 'b') %>% 
-  footnote(1, 6, 
-           value = as_paragraph('Brier Skill Score comparing the given classifier with the purely random cluster assignment.'), 
-           part = 'header', 
-           ref_symbols = 'c') %>% 
   theme_vanilla
@@
 flextable::flextable(suppl_tables$full_classifiers) %>% 
-  set_widths(c(2.3, 2.5, rep(2.3, 6))) %>% 
+  set_widths(c(2.3, 2.5, rep(2.3, 5))) %>% 
   merge_v(1) %>% 

### 2023-10-11 classification scripts/early_models.R

@@ -35,14 +35,6 @@
   
   insert_msg('Overall model performance stats')
 
-  ## square errors and Brier scores, the reference 
-  ## is a purely random classifier
-  
-  set.seed(12345)
-  
-  early_class$brier <- early_class$predictions %>% 
-    map(class_brier, return_ref = TRUE)
-  
   ## overall performance stats
   
   early_class$overall_stats <- early_class$predictions %>% 
@@ -51,17 +43,10 @@
     map(map, column_to_rownames, 'statistic') %>% 
     map(map, t) %>% 
     map(map, as.data.frame) %>% 
-    map(compress, names_to = 'dataset')
-
-  ## a common table with the performance stats
-  
-  early_class$overall_stats <- 
-    map2(early_class$overall_stats, 
-         map(early_class$brier, ~.x$total_bs), 
-         left_join, by = 'dataset') %>% 
-    map(as_tibble) %>% 
+    map(compress, names_to = 'dataset') %>% 
     compress(names_to = 'method') %>% 
-    mutate(dataset = factor(dataset, c('train', 'cv', 'test')))
+    mutate(dataset = factor(dataset, c('train', 'cv', 'test'))) %>% 
+    as_tibble
   
 # Plot of the model performance stats ------- 
   
@@ -89,11 +74,10 @@
   
   insert_msg('Plots of Brier square errors')
   
-  early_class$square_error_plots <- 
-    list(sq_table = map(early_class$brier, 
-                        ~.x$squares), 
-         plot_title = class_globals$algo_labs[names(early_class$brier)]) %>% 
-    pmap(plot_clust_squares)
+  early_class$square_error_plots <- early_class$predictions %>% 
+    map(map, 
+        plot, 
+        type = 'class_p')
 
 # Cluster prediction stats ------
   

### 2023-10-11 classification scripts/ensemble.R

@@ -165,34 +165,16 @@
 # Fit stat summary ------
   
   insert_msg('Overall fit stat summary')
-  
-  ## square errors and Brier scores, the reference 
-  ## is a purely random classifier
-  
-  set.seed(12345)
-  
-  ens_class$brier <- ens_class$predictions %>% 
-    map(class_brier, return_ref = TRUE)
-  
-  ## overall performance stats
-  
+
   ens_class$overall_stats <- ens_class$predictions %>% 
     map(map, summary) %>% 
     map(map, select, statistic, estimate) %>% 
     map(map, column_to_rownames, 'statistic') %>% 
     map(map, t) %>% 
     map(map, as.data.frame) %>% 
-    map(compress, names_to = 'dataset')
-  
-  ## a common table with the performance stats
-  
-  ens_class$overall_stats <- 
-    map2(ens_class$overall_stats, 
-         map(ens_class$brier, ~.x$total_bs), 
-         left_join, by = 'dataset') %>% 
-    map(as_tibble) %>% 
+    map(compress, names_to = 'dataset') %>% 
     compress(names_to = 'method') %>% 
-    mutate(dataset = factor(dataset, c('train', 'cv', 'test')))
+    as_tibble
 
 # Plot of the model performance stats ------- 
   
@@ -222,11 +204,10 @@
   
   insert_msg('Plots of Brier square errors')
   
-  ens_class$square_error_plots <- 
-    list(sq_table = map(ens_class$brier, 
-                        ~.x$squares), 
-         plot_title = class_globals$predictor_labs[names(ens_class$brier)]) %>% 
-    pmap(plot_clust_squares)  
+  ens_class$square_error_plots <- ens_class$predictions %>% 
+    map(map, 
+        plot, 
+        type = 'class_p')
   
 # Cluster-specific stats ------
   

### 2023-10-11 classification scripts/full_models.R

@@ -35,33 +35,16 @@
   
   insert_msg('Overall model performance stats')
 
-  ## square errors and Brier scores, the reference 
-  ## is a purely random classifier
-  
-  set.seed(12345)
-  
-  full_class$brier <- full_class$predictions %>% 
-    map(class_brier, return_ref = TRUE)
-  
-  ## overall performance stats
-  
   full_class$overall_stats <- full_class$predictions %>% 
     map(map, summary) %>% 
     map(map, select, statistic, estimate) %>% 
     map(map, column_to_rownames, 'statistic') %>% 
     map(map, t) %>% 
     map(map, as.data.frame) %>% 
-    map(compress, names_to = 'dataset')
-
-  ## a common table with the performance stats
-  
-  full_class$overall_stats <- 
-    map2(full_class$overall_stats, 
-         map(full_class$brier, ~.x$total_bs), 
-         left_join, by = 'dataset') %>% 
-    map(as_tibble) %>% 
+    map(compress, names_to = 'dataset') %>% 
     compress(names_to = 'method') %>% 
-    mutate(dataset = factor(dataset, c('train', 'cv', 'test')))
+    mutate(dataset = factor(dataset, c('train', 'cv', 'test'))) %>% 
+    as_tibble
   
 # Plot of the model performance stats ------- 
   
@@ -89,11 +72,10 @@
   
   insert_msg('Plots of Brier square errors')
   
-  full_class$square_error_plots <- 
-    list(sq_table = map(full_class$brier, 
-                        ~.x$squares), 
-         plot_title = class_globals$algo_labs[names(full_class$brier)]) %>% 
-    pmap(plot_clust_squares)
+  full_class$square_error_plots <- full_class$predictions %>% 
+    map(map, 
+        plot,
+        type = 'class_p')
 
 # Cluster prediction stats ------
   

### 2023-10-11 clustering scripts/clustering.R

@@ -270,7 +270,7 @@
                 x = paste0('Mentral cluster, ', 
                            tolower(.y), 
                            ' subset'), 
-                y = paste0('Mentral cluster, ', 
+                y = paste0('Mental cluster, ', 
                            tolower(.y), 
                            ' subset')))
 

### 2023-10-11 clustering scripts/development.R

@@ -93,8 +93,9 @@
   plan('sequential')
   
   clust_devel$cv <- clust_devel$cv %>% 
-    map(~.x$summary) %>% 
-    compress(names_to = 'method')
+    map(summary) %>% 
+    compress(names_to = 'method') %>% 
+    select(method, accuracy_mean)
     
 # Common results table and result visualization ------
   
@@ -111,15 +112,14 @@
                               stri_split_fixed(method, 
                                                pattern = '_', 
                                                simplify = TRUE)[, 2], 
-                              sep = ', '), 
-           cv_accuracy = 1 - mean_error)
+                              sep = ', '))
   
 # Plot ------
   
   insert_msg('Plot')
   
   clust_devel$result_plot <- clust_devel$result_tbl %>% 
-    pivot_longer(cols = c(variance, mean_sil, cv_accuracy), 
+    pivot_longer(cols = c(variance, mean_sil, accuracy_mean), 
                  names_to = 'statistic', 
                  values_to = 'value') %>% 
     ggplot(aes(x = value, 
@@ -130,15 +130,15 @@
              position = position_dodge(0.9)) + 
     scale_fill_manual(values = c(variance = 'steelblue2', 
                                  mean_sil = 'indianred3', 
-                                 cv_accuracy = 'darkolivegreen4'), 
+                                 accuracy_mean = 'darkolivegreen4'), 
                       labels = c(variance = 'Explained variance', 
                                  mean_sil = 'Mean silhouette', 
-                                 cv_accuracy = 'CV accuracy'), 
+                                 accuracy_mean = 'CV accuracy'), 
                       name = '') + 
     facet_grid(. ~ statistic, 
                labeller = as_labeller(c(variance = 'Explained variance', 
                                         mean_sil = 'Mean silhouette', 
-                                        cv_accuracy = 'CV accuracy')), 
+                                        accuracy_mean = 'CV accuracy')), 
                scales = 'free') + 
     globals$common_theme + 
     theme(axis.title.y = element_blank(), 

### 2023-10-11 paper scripts/tables.R

@@ -314,11 +314,11 @@
     map(map_dfc, function(x) if(is.numeric(x)) signif(x, 2) else x) %>% 
     map(select, 
         method, dataset, 
-        correct_rate, kappa, bs, bss, sens, spec) %>% 
+        correct_rate, kappa, brier_score, sens, spec) %>% 
     map(set_names, 
         c('Algorithm', 'Data subset', 
           'Accuracy', "Cohen's \u03BA", 
-          'Brier score', 'Brier skill score', 
+          'Brier score', 
           'Sensitivity, PTS cluster', 'Specificity, PTS cluster'))
 
   suppl_tables[c('early_classifiers', 'full_classifiers')] <- 

### 2023-10-11 tools/tools.R

@@ -746,10 +746,10 @@
                                  title_prefix = 'Mental cluster prediction,', 
                                  plot_subtitle = NULL) {
     
-    list(x = c('kappa', 'correct_rate', 'bs', 'bss'), 
+    list(x = c('kappa', 'correct_rate', 'brier_score'), 
          y = paste(title_prefix, 
-                   c("\u03BA", 'accuracy', 'Brier score', 'Brier skill score')), 
-         z = c("Cohen's \u03BA", 'Accuracy', 'Brier score', 'Brier skill score')) %>% 
+                   c("\u03BA", 'accuracy', 'Brier score')), 
+         z = c("Cohen's \u03BA", 'Accuracy', 'Brier score')) %>% 
       pmap(function(x, y, z) data %>% 
              ggplot(aes(x = .data[[x]], 
                         y = reorder(.data[[y_var]], .data[[x]]), 
@@ -765,7 +765,7 @@
              theme(axis.title.y = element_blank()) + 
              labs(title = y, 
                   x = z)) %>% 
-      set_names(c('kappa', 'correct_rate', 'bs', 'bss'))
+      set_names(c('kappa', 'correct_rate', 'bs'))
     
   }
   
@@ -829,7 +829,7 @@
              dataset = factor(dataset, c('train', 'cv', 'test')), 
              plot_cap = paste0('Accuracy = ', signif(correct_rate, 2), 
                                ', \u03BA = ', signif(kappa, 2), 
-                               ', BS = ', signif(bs, 2))) %>% 
+                               ', BS = ', signif(brier_score, 2))) %>% 
       blast(all_of(split_factor)) %>% 
       map(blast, dataset) %>% 
       map(map, ~.x$plot_cap)
@@ -990,79 +990,7 @@
     
     
   }
-  
-  plot_clust_squares <- function(sq_table, 
-                                 palette = globals$part_colors, 
-                                 labels = globals$part_labels, 
-                                 facet = TRUE, 
-                                 trend = TRUE, 
-                                 point_alpha = 0.75, 
-                                 plot_title = NULL, ...) {
-    
-    # plotting data and accessories
-    
-    sq_table <- sq_table %>% 
-      mutate(dataset = factor(dataset, 
-                              c('train', 'cv', 'test')))
-    
-    ref_squares <- sq_table$ref_error %>% 
-      mean
-    
-    n_numbers <- sq_table %>% 
-      count(dataset) %>% 
-      mutate(dataset = labels[dataset])
[…]

### 2023-11-07 clustering scripts/characteristic.R
Feedback, October 2023, code
And analysis of psychometric factors as a function of observation time

@@ -175,15 +175,11 @@
         label = stri_replace(label, 
                              fixed = ' score', 
                              replacement = ''), 
-        label = stri_replace(label, 
-                             fixed = 'sum', 
-                             replacement = 'sum score'), 
-        label = stri_replace(label, 
-                             regex = '(EUROHIS-QOL\\s{1}8\\s{1})|(PTGI\\s{1})|(PCL-5\\s{1}DSM-5\\s{1})', 
-                             replacement = ''), 
-        label = stri_replace(label, 
-                             fixed = 'somatic symptoms', 
-                             replacement = 'som. sympt'), 
+        label = ifelse(stri_detect(label, fixed = 'sum'), 
+                       label, 
+                       stri_replace(label, 
+                                    regex = '(EUROHIS-QOL\\s{1}8\\s{1})|(PTGI\\s{1})|(PCL-5\\s{1})', 
+                                    replacement = '')), 
         label = ifelse(p_adjusted < 0.05, 
                        paste0('<b>', label, '</b>'), 
                        label)) %>% 

### 2023-11-07 paper/markdown/manuscript.Rmd
Feedback, October 2023, markdown and figures

--- 2023-09-04 paper/markdown/manuscript.Rmd
@@
   - Mathias Ströhle, MD:
-      institute: [kufstein]
+      institute: [kufstein, himed]
   - Peter Paal, MD: 
-      institute: [salzburg, kurasi, himed, icar]
+      institute: [himed, salzburg, kurasi, icar]
   - Hermann Brugger, MD: 
-      institute: [eurac, ismm]
+      institute: [icar. eurac, ismm]
   - Martin Faulhaber, PhD: 
-      institute: [sport]
+      institute: [himed, sport]
   - Nicola Kruijer, MD: 
@@
   - Katharina Hüfner, MD:
-      institute: [psych]
+      institute: [psych, himed]
       email: katharina.huefner@tirol-kliniken.at
@@
   - regensburg: Department of Neurosurgery, University Hospital of Regensburg, Regensburg, Germany
-  - daas: Data Analysis as a Service Tirol, Innsbruck, Austria
+  - daas: Data Analysis as a Service Tirol, Wörgl, Austria
   - kufstein: Department of Anesthesiology and Critical Care Medicine, Bezirkskrankenhaus Kufstein, Kufstein, Austria.
-  - salzburg: Department of Anesthesiology and Intensive Care Medicine, Paracelsus Medical University, Salzburg, Austria.
+  - himed: Austrian Society of Mountain and High-Altitude Medicine, Mieming, Austria. 
+  - salzburg: Department of Anesthesiology and Critial Care Medicine, Paracelsus Medical University, Salzburg, Austria.
   - kurasi:  Austrian Board of Mountain Safety (Österreichisches Kuratorium fur Alpine Sicherheit), Innsbruck, Austria.
-  - himed: Austrian Society of Mountain and High-Altitude Medicine, Mieming, Austria. 
   - icar: International Commission for Mountain Emergency Medicine (ICAR), Kloten, Switzerland.
@@
 
-__Background:__ Mountain sports are inherently linked to physical injury risk. 
-We aimed to characterize mental health following mountain sport accidents requiring professional medical management.
+__Background:__ The restorative effect of physical activity in natural alpine environments on mental and physical health is increasingly recognized. 
+However, mountain sports are also linked to a risk of accidents. 
+We aimed to characterize mental health in individuals following mountain sport accidents requiring professional medical management.
 
-__Methods:__ Adult victims of mountain sport accidents treated at the Innsbruck University Hospital (Austria) between 2008 and 2020 (n = `r nrow(ptsd$dataset)`) participated in a cross-sectional survey study at median `r round(median(ptsd$dataset$obs_time)/30.436)` months after the trauma center admission.
-Symptoms of anxiety, depression, somatization and panic disorder (PHQ), resilience (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS-QOL), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD, PCL-5), sociodemographic and clinical information were obtained from an on-line survey and electronic health records.
-Mental health patterns were investigated by semi-supervised medoid clustering and modeled by machine learning.
+__Methods:__ Adult victims of mountain sport accidents treated at the hospital of the Medical University of Innsbruck (Austria) between 2018 and 2020 completed a cross-sectional survey at a median `r round(median(ptsd$dataset$obs_time)/30.436)` months following the admission (n = `r nrow(ptsd$dataset)`).
+Symptoms of post-traumatic stress disorder (PTSD, PCL-5), anxiety, depression, and somatization (PHQ), resilience (RS-13), sense of coherence (SOC-9L), post-traumatic growth (PTGI) and quality of life (EUROHIS-QOL), as well as sociodemographic and clinical information were obtained from an online survey and extracted from electronic health records.
+Mental health outcome patterns were investigated by semi-supervised medoid clustering and modeled by machine learning.
 
-__Results:__ At least one diagnostic criterion of PTSD was met by `r signif(table(ptsd$dataset$dsm5_cluster_class)[2]/nrow(ptsd$dataset), 2) * 100`% of participants. 
-Three comparably sized subsets of patients were identified: (1) neutral, (2) post-traumatic growth and (3) post-traumatic stress cluster.
-The post-traumatic stress cluster was characterized by frequent mental disorder symptoms, and low resilience, sense of coherence and quality of life as well as by low age, the highest frequency of pre-existing mental disorders and common physical health consequences of the accident.
-Reliability of machine learning-based prediction of the cluster assignment based on `r length(class_globals$variables$early)` variables available during acute treatment was limited.
+__Results:__ Symptoms of PTSD were observed in `r signif(table(ptsd$dataset$dsm5_cluster_class)[2]/nrow(ptsd$dataset), 2) * 100`% of participants. 
+Three comparably sized subsets of participants were identified: a (1) neutral, (2) post-traumatic growth and (3) post-traumatic stress cluster. 
+The post-traumatic stress cluster was characterized by high prevalence of symptoms of mental disorder, low resilience, low sense of coherence and low quality of life as well as by younger age, the highest frequency of pre-existing mental disorders and persisting physical health consequences of the accident.
+Individuals in this cluster self-reported a  need for psychological support following the accident and more cautious behavior during mountain sports since the accident.
+Reliability of machine learning-based prediction of the cluster assignment based on `r length(class_globals$variables$early)` variables available during acute medical treatment of accident victims was limited.
 
-__Conclusion:__ A subset of mountain sport accident victims shows symptoms of mental health disorders including PTSD. 
-Since early identification of these vulnerable patients remains challenging, low-threshold psychological support is key for a successful interdisciplinary management of mountain sport accidents.
+__Conclusion:__ A subset of individuals develop symptoms of mental health disorders including symptoms of PTSD when assessed at least 6 months after the mountain sport accident. 
+Since early identification of these vulnerable patients remains challenging, psychoeducational measures for all patients and low-threshold access to psychological support is key for a successful interdisciplinary management of victims of mountain sport accidents.
 
@@
 
-mountain sport accident, mental health, post-traumatic stress disorder, post-traumatic growth, machine learning
+mountain sport accident, machine learning, mental health, post-traumatic growth, post-traumatic stress disorder
 
@@
 
-Physical activity in mountains, such as hiking, climbing, alpine skiing or ski touring, gain popularity, a trend that has been accelerated by the COVID-19 pandemic [@Hedenborg2022]. 
-Outdoor physical activity in general and mountain sport in particular exerts beneficial effects on mental health [@Ower2019; @Niedermeier2017a; @Niedermeier2017]. 
-However, mountain sports are also associated with the risk of injury [@Faulhaber2017; @Faulhaber2020], which could foster development of mental disorders including post-traumatic stress disorder (PTSD).
+Physical activity in mountains, such as hiking, climbing, alpine skiing or ski touring, are gaining increasing popularity, a trend that has been accelerated by the COVID-19 pandemic [@Hedenborg2022]. 
+Outdoor physical activity in general and mountain sports in particular exert beneficial effects on physical and mental health [@Ower2019; @Niedermeier2017a; @Niedermeier2017]. 
+However, mountain sports are also associated with the risk of accidents and injury [@Faulhaber2017; @Faulhaber2020], which could consequently lead to development of post-traumatic stress disorder (PTSD) or other stress-related mental disorders.
 
-PTSD symptoms include hyperarousal, panic, somatic symptoms, sleep disorders and unwilling recalls of the trauma (flashbacks) [@Bryant2019; @AmericanPsychiatricAssociation2013; @Howie2019], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence.
+Diagnostic symptoms of PTSD  include re-experiencing (criterion B), avoidance (criterion C), negative alterations in cognition and mood (criterion D) and hyper-arousal (criterion E) [@AmericanPsychiatricAssociation2013], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence.
 PTSD may co-occur with symptoms of depression and anxiety and substance abuse disorders [@Gagnon-Sanschagrin2022; @Brady2000]. 
-Although most individuals have experienced at least one traumatic event during the lifetime [@Benjet2016; @Breslau2004], the prevalence of PTSD in the general European population has been estimated for 0.38 to 6.67% [@Trautmann2018; @Burri2014].
+Although most individuals have experienced at least one traumatic event during their lifetime [@Benjet2016], the prevalence of PTSD in the general European population has been estimated between 0.38 to 6.67% [@Burri2014].
 Yet, up to 26% of acute physical trauma patients have been reported to develop PTSD [@Joseph2020].
 
-So far it has proven difficult to identify early predictors of subsequent PTSD development in physical trauma patients. 
-While some recent studies on specific populations have identified risk factors such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020].
+So far it has proven difficult to reliably identify early predictors of PTSD development in accident victims. 
+While some recent studies on specific populations have identified risk factors, such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020].
 These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019].
-Many trauma patients at risk of PTSD undergo intensive acute medical management at emergency departments, which poses a challenge for implementation of early diagnostic and psychological preventive measures [@Linares2017].
-Furthermore, efficacy of such acute psychological interventions at preventing PTSD in trauma patients is uncertain, probably due to large heterogeneity of study cohorts [@Qi2016].
 
-Post-traumatic growth describes a salutary adjustment psychological reaction to a traumatic experience with strengthened perceptions of self, others, and meaning of the event [@Tedeschi1996]. 
-Resilience, i.e. the ability to master adversities has been proposed to bridge traumatic events and post-traumatic growth. 
-It was also suggested to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012]. 
-Overall, mountain sport was associated with high resilience values [@Ower2019].
+Resilience, i.e.
+the ability to recover from adversities can protect an individual from developing PTSD and can help to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012].
+Generally individuals who perform mountain sports display high resilience values [@Ower2019]. 
+Post-traumatic growth can occur in individuals struggling to overcome the consequences of a traumatic event and describes a psychological adjustment reaction with strengthened positive perceptions of self, others, one's life, and meaning of the event [@Tedeschi1996].
+Although both resilience and post-traumatic growth are salutogenetic concepts their exact link is still under investigation [@Levine2009].
 
-Reports on mental health following accidents during mountain sports are scarce. 
-The available data concern mountain guides, rescuers or avalanche accident survivors [@Leonard2021; @Sommer2004; @Mikutta2022; @Thordardottir2015; @Traber2020]. 
-Considering the large number of mountain accident victims, and heterogeneity of accident and injuries, accurate characteristic of mental health of the affected patients is urgently needed. 
-In particular, tools allowing for identification of patients at risk of later mental health problems during acute medical management of the mountain accident would facilitate the early targeted psychological support and help to preserve the positive mental health effects of physical activity.
-
-Herein, we analyzed a broad palette of mental health readouts such as anxiety, depression, resilience, sense of coherence, quality of life, post-traumatic growth and signs of PTSD in a collective of `r nrow(ptsd$dataset)` patients treated for injuries during mountain sport accidents at the Medical University of Innsbruck, Austria by means of semi-supervised clustering.
-Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain accident with a machine learning approach.
+With the increasing popularity of mountain sports, it is essential to characterize the mental health consequences of accidents during mountain sports. 
+Prevention of PTSD or other mental health problems following an accident is essential to preserve the positive mental health effects of physical activity in an mountain environment despite the accident risk.
+Available data on the consequences of accidents during mountain sports on mental health are scarce. 
+Prior studies were performed on mountain guides, rescuers or avalanche accident survivors and mostly lack clinical data [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004].
+We analyzed mental health readouts such as symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in `r nrow(ptsd$dataset)` individuals treated at the Hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident by means of semi-supervised clustering.
+Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain sport accident with a machine learning approach.
 
@@
 
-Patients treated for a mountain sport accident at the Department for Orthopedics and Traumatology at the Medical University of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 (i.e.
-at least 6 months prior to the start of the study) were screened for participation.
-Patients fulfilling the inclusion criteria: age $\geq$ 18 years, residence in a German-speaking country and proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail.
-Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey and `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S2__).
+Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology at the University Clinic of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation.
+Individuals fulfilling the inclusion criteria: 
+hospital admission at least 6 months prior to the start of the study, 
+age $\geq$ 18 years, residence in a German-speaking country, and 
+proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail. 
+Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey. 
+Surveys of `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (overall response rate: `r signif(nrow(ptsd$dataset)/nrow(ptsd$cleared), 2) * 100`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).
 
@@
 Sociodemographic (e.g.
-age, sex, education, profession, income, prior sport accidents and traumatic events), accident- and recovery-related (e.g.
-rescue, psychological support, persistent physical health consequences) and psychometric variables were recorded with a cross-sectional on-line survey.
+age, sex, education, profession, income, prior mountain sport accidents and prior traumatic events), accident- and recovery-related (e.g.
+mode of rescue, psychological support, persistent physical health consequences) and psychometric variables were recorded with a cross-sectional online survey.
 
 The psychometric battery consisted of German versions of assessment tools for symptoms of 
+PTSD (PCL-5: PTSD checklist for DSM-5) [@Bovin2016], 
 anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
-depression (PHQ: patient health questionnaire, PHQ-9, depression module) [@Lowe2002; @Grafe2004], 
-panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
-persistent somatic symptoms (PHQ-15, somatization module) [@Kroenke2002], 
+depression (PHQ-9: 9-item patient health questionnaire for depressive symptoms) [@Grafe2004], 
+panic (PHQ-panic module) [@Grafe2004], 
+somatization (PHQ-15: 15-item patient health questionnaire of common somatic symptoms as a substitute for somatization) [@Kroenke2002], 
 resilience (RS13: 13-item resilience scale) [@Leppert2008], 
-lack of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
+sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
 quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
-post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
-post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
+and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
 The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).
@@
 depression (PHQ-9 $\geq$ 11) [@Manea2012], 
-persistent somatic symptoms (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
-resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73)[@Leppert2008] 
+somatization (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
+resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73) [@Leppert2008] 
 were defined with published cutoffs.
@@
 Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
-Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].
+Significant PTSD symptoms were assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].
 
-Traumatic events before the accident were assessed with the DIA-X tool [@Maercker2005]. 
-Flashbacks frequency was surveyed as none, > 1/year and > 1/month. 
+Traumatic events prior to the accident were assessed with the DIA-X tool [@Maercker2005]. 
+Flashback frequency was surveyed as none, > 1/year and > 1/month. 
 Smoking was surveyed as a single yes/no question. 
 Problematic alcohol use was investigated with the CAGE tool [@OBrien2008]. 
-Data on the type of the accident mountain sport, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient's records.
+Data on the type of the mountain sport activity during the accident, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and body location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient records.
 
@@
 
-The primary analysis endpoint was assessment of mental health of mountain accident victims by clustering using standardized psychometric tools for symptoms of anxiety, depression, panic, persistent somatic symptoms, quality of life, resilience, sense of coherence, post-traumatic growth and PTSD.
+The primary analysis endpoint was assessment of mental health following a mountain sport accident by clustering using standardized psychometric tools for symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life.
 The secondary analysis endpoints were 
-(1) to compare demographic, socioeconomic and clinical characteristics between the mental health clusters of mountain accident victims and 
-(2) to identify patients at risk of post-traumatic mental disorders based on explanatory variables available during early medical treatment of the mountain sport accident.
+(1) to compare demographic, socioeconomic and clinical characteristics between the different mental health clusters of mountain accident victims and 
+(2) to identify patients at risk of mental disorders based on a set of explanatory variables available during acute medical treatment after mountain sport accidents.
 
@@
 
-The study collective consisted of victims of mountain sport accidents which happened in Tyrol, Austria, a renowned and popular alpine touristic region. 
-In- and outpatient adult survivors of mountain sport accidents and treated at the Department of Orthopedics and Traumatology at Medical University of Innsbruck (Tyrol, Austria) between 1^st^ January 2018 and 31^st^ December 2020 were invited to participate (n = `r nrow(ptsd$cleared)`).
-Among the invited patients, 
-`r length(ptsd$interview_ID)` completed the study survey and 
-`r nrow(ptsd$dataset)` individuals with the complete psychometric data set were analyzed (overall analysis inclusion rate: 
-`r ((nrow(ptsd$dataset)/nrow(ptsd$cleared)) * 100) %>% signif(2)`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__). 
-The median time between the trauma center admission and survey completion was 
+Surveys with complete psychometric data from `r nrow(ptsd$dataset)` individuals were analyzed (__Figure \@ref(fig:fig-consort)__). 
+The median time between the trauma center admission after the accident and the survey completion was 
 `r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
-As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and less surgery diagnoses (__Supplementary Table S4__).
+As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and fewer surgery diagnoses (__Supplementary Table S4__).
 The survey participants excluded from the analysis due to missingness of psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__).
@@
 (`r get_percent(ptsd$dataset, 'employment_status')$percent['employed']`%). 
-Less than 8% of participants worked in a trauma-exposed or mountain sport profession. 
-High annual household incomes of $\geq$ 45000 Euro was reported by 
+Less than 8% of participants worked in a search and rescue or mountain sport profession. 
+Annual household incomes of $\geq$ 45000 Euro were reported by 
 `r get_percent(ptsd$dataset, 'high_income')$percent['yes']`% of participants. 
-Less than 10% of participants were smokers or at risk of problematic alcohol use. 
+Less than 10% of participants were smokers or at risk of problematic alcohol use defined as CAGE $\geq$ 2. 
 Pre-existing physical disorders were reported by 
 `r get_percent(ptsd$dataset, 'somatic_comorbidity')$percent['yes']`% of participants with cardiovascular, neurological and metabolic illness being the most frequent. 
-Four of ten participants had experiences by or witnessed a traumatic event prior to the accident, 
-`r sum(get_percent(ptsd$dataset, 'traumatic_number')$percent[c('2', '3+')])`% participants experienced two or more traumatic events. 
-Mental disorders prior the accident affected 
+Mental disorders diagnosed by a physician before the accident affected 
 `r get_percent(ptsd$dataset, 'psych_comorbidity')$percent['yes']`% of the cohort with affective 
 (`r get_percent(mental$analysis_tbl, 'affective')$percent['yes']`%) and somatoform disorders 
-(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental conditions (__Table \@ref(tab:tab-cohort-demo)__).
+(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental conditions. 
+Four of ten participants had experienced by or witnessed a traumatic event prior to the accident, 
+`r sum(get_percent(ptsd$dataset, 'traumatic_number')$percent[c('2', '3+')])`% participants had been exposed to two or more traumatic events (__Table \@ref(tab:tab-cohort-demo)__).
 
@@
 (`r get_percent(ptsd$dataset, 'sport_type')$percent['biking']`%) and classical mountain sports such as climbing, hiking, mountaineering or back-country skiing 
-(`r get_percent(ptsd$dataset, 'climbing/hiking/mountaineering/skitour')$percent['climbing/hiking/mountaineering/skitour']`%). 
+(`r get_percent(ptsd$dataset, 'sport_type')$percent['climbing/hiking/mountaineering/skitour']`%). 
 One-third of participants were alone during the accident 
-(`r get_percent(ptsd$dataset, 'accident_alone')$percent['yes']`%) and, in most cases, were the sole culprit 
+(`r get_percent(ptsd$dataset, 'accident_alone')$percent['yes']`%) and, in most cases, were the only person responsible to the accident 
 (`r get_percent(ptsd$dataset, 'accident_culprit')$percent['self']`). 
@@
 `r get_percent(ptsd$dataset, 'accident_rescue')$percent['rescue team']`% of the accidents. 
-In `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['2']`% of participants the injury was moderate (AIS 2) and in 
+In `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['2']`% of participants the injury severity was moderate (AIS 2) and in 
 `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['3+']`% severe-to-critical (AIS $\geq$ 3). 
-Limb injuries were the most common followed by head and face (__Supplementary Figure S1A__). 
-Hospitalization and surgery rates following the accident were 
+Limb injuries were the most common followed by injuries of the head and face (__Supplementary Figure S1A__). 
+Hospitalization and surgery rates were 
 `r get_percent(ptsd$dataset, 'hospitalization')$percent['yes']`% and 
@@
 `r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`% individuals. 
-A subset of participants (`get_percent(ptsd$dataset, 'psych_support_need')$percent['yes']`%), who had not received it, declared need for psychological support following the accident. 
-Despite persisting physical consequences of the accident, which affected 
-`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks during mountain sport at the time of the questionnaire completion in 
-`r get_percent(ptsd$dataset, 'unwilling_flashback')$percent['yes']`% of the cohort, 
-most participants returned to the same mountain sport following the accident 
-(`r get_percent(ptsd$dataset, 'same_sport_type_post_accident')$percent['yes']`%). 
-Yet, `r get_percent(ptsd$dataset, 'caution_post_accident')$percent['more cautious']`% described their behavior during mountain sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).
+A subset of participants (`r get_percent(ptsd$dataset, 'psych_support_need')$percent['yes']`%), who had not received psychological or psychiatric support, declared a need for psychological or psychiatric support following the accident.
+Persisting physical consequences of the accident were reported by 
+`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks during mountain sport at the time of the questionnaire completion were observed in 
+`r get_percent(ptsd$dataset, 'unwilling_flashback')$percent['yes']`% of the cohort. 
+Although most individuals returned to the same mountain sport following the accident 
+(`r get_percent(ptsd$dataset, 'same_sport_type_post_accident')$percent['yes']`%), 
+`r get_percent(ptsd$dataset, 'caution_post_accident')$percent['more cautious']`% of all participants described their behavior during mountain sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).
 
-At least one diagnostic criterion of PTSD defined by the PCL-5 tool (B, C, D or E) were met by 
+At least one diagnostic criterion of PTSD defined as positive scoring of the B, C, D or E domains of the PCL-5 tool was met by 
 `r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% of participants. 
 The domain B PTSD symptoms were the most common (`r get_percent(ptsd$dataset, 'dsm5_B_class')$percent['positive']`%) and the domain D symptoms were the least frequent (`r get_percent(ptsd$dataset, 'dsm5_D_class')$percent['positive']`%).
-Solely `r sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)` patients 
+Solely `r digit2string(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0))` patients 
 (`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) were screened positive for all four diagnostic criteria of PTSD (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1BC__).
 Clinically relevant symptoms of 
-anxiety (`r get_percent(ptsd$dataset, 'gad7_total_class')$percent['positive']`), 
+anxiety (`r get_percent(ptsd$dataset, 'gad7_total_class')$percent['positive']`%), 
 depression (`r get_percent(ptsd$dataset, 'phq9_total_class')$percent['positive']`%) 
-and somatic symptoms (`r get_percent(ptsd$dataset, 'phq_events_total_class')$percent['positive']`%) were rare in this cohort consisting of `r get_percent(ptsd$dataset, 'rs13_total_class')$percent['high']`% highly resilient individuals (__Table \@ref(tab:tab-cohort-mental)__).
+and somatization (`r get_percent(ptsd$dataset, 'phq_events_total_class')$percent['positive']`%) were rare in this cohort consisting of `r get_percent(ptsd$dataset, 'rs13_total_class')$percent['high']`% highly resilient individuals (__Table \@ref(tab:tab-cohort-mental)__).
+We could not observe any significant association of values of psychometric scores or frequency of mental health problems with the time between trauma center admission and survey completion (not shown).
 
@@
 test subset (n = `r table(ptsd$dataset$partition)['test']`). 
-Significant differences of weak effect size between these subsets concerned resilience class distribution and scores of quality of life (__Supplementary Table S6__).
+The sole significant differences between these subsets concerned resilience class distribution and scores of quality of life, their effect size was weak (__Supplementary Table S6__).
 
-## Three clusters of mental response in sport accident victims
+## Three clusters of mental health response after mountain sport accidents
 
-To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to partition around medoids clustering in respect to a broad range of psychometry scores [@Drost2018; @Schubert2019] (__Supplementary Table S2__).
-Three mental clusters: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress), named after their key mental characteristic were identified in the training subset (__Supplementary Figure S3__).
-Subsequently, the mental cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
+To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to medoid clustering in respect to a broad range of psychometric scores (__Supplementary Table S2__).
+Three mental health clusters were identified in the training subset of the study cohort: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress) and named after their key mental characteristic (__Supplementary Figure S3__).
+Subsequently, the mental health cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
 (training: `r signif(semi_clust$stats$variance[1], 2)`, test: `r signif(semi_clust$stats$variance[2], 2)`), 
 comparable average silhouette statistics 
-(training: `r signif(semi_clust$stats$mean_sil[1], 2)`, test: `r signif(semi_clust$stats$mean_sil[2], 2)`) [@Rousseeuw1987], 
+(training: `r signif(semi_clust$stats$mean_sil[1], 2)`, test: `r signif(semi_clust$stats$mean_sil[2], 2)`), 
 comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).
 
-The mental clusters were of approximately equal size. 
-The neutral cluster was characterized by low levels of anxiety, depression, panic, somatic symptoms, post-traumatic stress and post-traumatic growth along with high rating of sense of coherence, resilience and quality of life.
+The mental health clusters were of approximately equal size. 
+The neutral cluster was characterized by low scores of PTSD symptoms, anxiety, depression, panic, somatization, and post-traumatic growth along with high ratings of sense of coherence, resilience and quality of life.
 The PTG cluster demonstrated similarly low scores of major mental health disorders, high resilience and sense of coherence. 
 Its key characteristic were the highest levels of post-traumatic growth. 
-The remaining PTS cluster displayed the highest scores of for anxiety, depression, panic disorder, somatic symptoms and post-traumatic stress as well as poor sense of coherence, and low resilience and low quality of life.
+The remaining PTS cluster displayed the highest scores of PTSD symptoms, anxiety, depression, panic disorder, somatization as well as poor sense of coherence, and low resilience and low quality of life.
 Post-traumatic growth scores in the PTS cluster were higher than in neutral but lower than in PTG cluster participants (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Table S7__). 
@@
 anxiety (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'gad7_total_class')$percent['positive']`%), 
-depression (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq9_total_class')$percent['positive']`%) and somatic symptoms (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq_events_total_class')$percent['positive']`%)
+depression (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq9_total_class')$percent['positive']`%) and somatization (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq_events_total_class')$percent['positive']`%)
 were present virtually only in the PTS cluster. 
@@
 moderate resilience classes (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'rs13_total_class')$percent['moderate']`%) peaked in the PTS cluster. 
-Finally, the frequency of flashbacks, PTSD symptoms specified by the B, C, D and E PCL-5 domains and the frequency of patients meeting at least one PCL-5 PTSD diagnostic criterion 
-(PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'dsm5_cluster_class')$percent['positive']`%) were the highest in PTS cluster individuals (__Figure \@ref(fig:fig-clust-symptoms)__, __Supplementary Table S7__).
+Finally, frequency of flashbacks, PTSD symptoms specified by the B, C, D and E PCL-5 domains and frequency of individuals scoring positive for at least one diagnostic domain of PCL-5 
+(PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'dsm5_cluster_class')$percent['positive']`%) were the highest in PTS cluster individuals (__Figure \@ref(fig:fig-clust-symptoms)__, __Supplementary Table S8__).
 
-## Demographic, socioeconomic and clinical background of the mental health clusters
+## Demographic, socioeconomic and clinical characteristic of the mental health clusters
 
 Among 
-`r length(clust_bcg$variables[!stri_detect(clust_bcg$variables, regex = '^(dsm|rs13|phq|gad|flash|unwilling)')])` investigated demographic, socioeconomic, clinical, accident- and recovery-related variables, only
-`r length(clust_bcg$top_factors[!stri_detect(clust_bcg$top_factors, regex = '^(dsm|rs13|phq|gad|flash|unwilling)')])` features (age, pre-existing physical illness, pre-existing mental disorder, psychological support need, persistent physical health consequences of the accident, and caution during sport activity) were found to differ significantly between the mental clusters.
-The effect size of these differences was weak (__Supplementary Table S8__).
+`r length(clust_bcg$bcg_variables)` investigated demographic, socioeconomic, clinical, accident- and recovery-related variables, only 
+`r sum(clust_bcg$top_factors %in% clust_bcg$bcg_variables)` features 
+(age, 
+pre-existing physical illness, 
+pre-existing mental disorder, 
+need for psychological or psychiatric support after the accident, 
+persistent physical health consequences of the accident, 
+and caution during mountain sport following the accident) 
+were found to differ significantly between the mental health clusters. 
+The effect size of these differences was weak (__Supplementary Table S9__).
 
-In more detail, PTS cluster patients were the youngest of all participants and had the highest frequency of pre-existing mental conditions. 
-The neutral cluster was in turn substantially enriched in males, and tertiary education and high income individuals; these effects were, however not significant. 
-Participants suffering from pre-existing somatic conditions were significantly enriched in the PTG and PTS clusters. 
-A similar, yet not significant tendency was observed for frequency of prior traumatic events, which was higher in the PTG and PTS clusters as compared with neutral cluster patients. 
-There were no substantial differences in frequency of participants with trauma-risk professions and employment structure between the clusters (__Figure \@ref(fig:fig-clinic-factors)__, __Supplementary Figure S7__).
-Participants reporting being alone during the accident, the only culprit of the accident, and requiring comrade or professional rescue were enriched in the PTG cluster. 
+In more detail, individuals in the PTS cluster were the youngest of all participants and had the highest frequency of pre-existing mental disorders. 
+The neutral cluster was in turn substantially enriched in males, and tertiary education and high-income individuals; these effects were not significant. 
+Participants suffering from pre-existing physical illnesses were significantly enriched in the PTG and PTS clusters. 
+A similar, yet not significant tendency was observed for the frequency of prior traumatic events, which was higher in the PTG and PTS clusters as compared with neutral cluster individuals. 
+There were no substantial differences in frequency of search and rescue professionals and employment structure between the clusters (__Figure \@ref(fig:fig-clinic-factors)__, __Supplementary Figure S7__).
+Participants reporting being alone during the accident, responsible for the accident, and requiring companion or professional rescue were enriched in the PTG cluster. 
 In turn, frequency of self-rescuers tended to be the highest in the neutral cluster. 
 These differences in accident and rescue circumstances were not significant. 
-We could not observe any substantial differences in accident season, mountain sport type, and numbers of injured persons between the clusters (__Figure \@ref(fig:fig-injury)A__, __Supplementary Figure S7__).
+We could not observe any substantial differences in accident season, mountain sport type, and numbers of injured persons between the clusters (__Figure \@ref(fig:fig-injury)A__, __Supplementary Figure S7__, __Supplementary Table S9__).
 There were no significant differences in severity and injury and its location between the clusters. 
-However, moderate-to-severe injuries (AIS $\geq$ 2), hospitalization and surgery rates were higher in the PTG and PTS clusters as compared with neutral cluster patients (__Figure \@ref(fig:fig-injury)__).
+However, moderate-to-severe injuries (AIS $\geq$ 2), hospitalization and surgery rates were substantially higher in the PTG and PTS clusters as compared with neutral cluster participants (not significant, __Figure \@ref(fig:fig-injury)__, __Supplementary Table S9__).
 
-There were no significant differences in rates of psychological or psychiatric support after the accident between the clusters. 
-Among PTG and PTS cluster individuals, `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'psych_support_need')$percent['yes']`% and `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'psych_support_need')$percent['yes']`%, respectively, reported need for psychological support following the accident as compared with none in the neutral cluster.
+There were no significant differences in rates of participants who received psychological or psychiatric support after the accident between the clusters. 
+Among PTG and PTS cluster individuals, `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'psych_support_need')$percent['yes']`% and `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'psych_support_need')$percent['yes']`%, respectively, reported need for psychological or psychiatric support following the accident as compared with none in the neutral cluster.
 Frequency of participants suffering from persistent physical health consequences of the accident was the highest in the PTS cluster (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'accident_aftermath')$percent['yes']`%,
@@
 neutral: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'neutral'), 'caution_post_accident')$percent['more cautious']`%). 
-By contrast, percentages of participants who returned to the same mountain sport after the accident were comparable between the mental clusters (__Figure \@ref(fig:fig-aftermath)__).
+By contrast, percentages of participants who returned to the same mountain sport after the accident were comparable between the mental health clusters (__Figure \@ref(fig:fig-aftermath)__).
 
-## Prediction of the mental cluster assignment by demographic, socioeconomic and accident-related factors
+## Prediction of the mental health cluster assignment by demographic, socioeconomic and accident-related factors
 
-Finally, we intended to model the mental cluster assignment with  demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident (__Supplementary Table S9__) with seven popular machine learning algorithms: the random forests [@Wright2017; @Breiman2001], neural networks [@Ripley2014], support vector machines [@Weston1998], recursive partitioning [@Therneau2022], discriminant analysis [@Ahdesmaki2010; @Ahdesmaki2022], elastic net regression [@Zou2005; @Friedman2010], and conditional random forests [@Hothorn2006; @Strobl2007; @Hothorn2022] (__Supplementary Table S10__).
+Finally, we intended to model the mental health cluster assignment with demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the patient (__Supplementary Table S10__) with seven popular machine learning algorithms:
+the random forests, 
+neural networks, 
+support vector machines, 
+recursive partitioning, 
+discriminant analysis, 
+elastic net regression, and 
+conditional random forests (__Supplementary Table S11__). 
 Such models would enable early identification of mountain sport accident victims at risk of mental health problems.
@@
 The cluster assignment models employing early predictors demonstrated a moderate-to-excellent prediction performance in the training subset of the study cohort 
-(`r paste0((filter(early_class$overall_stats, dataset == 'train')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`% of correctly assigned observations, 
-Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$kappa), 2) %>% paste(collapse = ' - ')`, Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$bs), 2) %>% paste(collapse = ' - ')`).
+(training accuracy: `r paste0(filter(early_class$overall_stats, dataset == 'train')$correct_rate %>% range %>% signif(2) * 100, '%') %>% paste(collapse = ' - ')`, 
+Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$kappa), 2) %>% paste(collapse = ' - ')`, Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'train')$brier_score), 2) %>% paste(collapse = ' - ')`).
 However, their accuracy in cross-validation and in the test subset was poor 
-(test: `r paste0((filter(early_class$overall_stats, dataset == 'test')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`% of correctly assigned observations, 
+(test accuracy: `r paste0((filter(early_class$overall_stats, dataset == 'test')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`, 
 Cohen's $\kappa$: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$kappa), 2) %>% paste(collapse = ' - ')`, 
-Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$bs), 2) %>% paste(collapse = ' - ')`). 
-Concerning the early identification of the vulnerable PTS cluster patients, the conditional random forest, random forest and support vector machine models were able to differentiate PTS cluster patients from the pooled neutral + PTG cluster subset with the best, at least
-`r min(signif(early_class$clust_stats %>% filter(dataset == 'test', clust_id == 'PTS') %>% top_n(n = 3, sens) %>% .$sens, 2) * 100)`% sensitivity in the test subset of the study cohort (__Figure \@ref(fig:fig-early-class)__, __Supplementary Table S12__).
+Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`). 
+Concerning the early identification of the vulnerable PTS cluster patients, the conditional random forest, random forest and support vector machine models were able to differentiate between the PTS cluster and the pooled neutral and PTG clusters with the best sensitivity (test subset sensitivity $\geq$ `r min(signif(early_class$clust_stats %>% filter(dataset == 'test', clust_id == 'PTS') %>% top_n(n = 3, Se) %>% .$Se, 2) * 100)`%, __Figure \@ref(fig:fig-early-class)__, __Supplementary Table S12__).
 Pre-existing mental disorder or physical illness, hospitalization and surgery, injury severity, age, rescue mode, education, annual income, and traumatic events in the past were among the most important explanatory factors employed by these models (__Supplementary Figure S8__).
-Of note, inclusion of follow-up-related variables such as psychological support, physical health consequences or caution during sport activity in the machine learning models of the mental cluster assignment did not improve their accuracy in cross-validation or the test subset of the study cohort
-(test: `r paste0((filter(full_class$overall_stats, dataset == 'test')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`% of correctly assigned observations, 
+Of note, inclusion of follow-up-related variables such as received psychological or psychiatric support, psychological or psychiatric support need, physical health consequences or caution during mountain sport activity in the machine learning models did not improve their accuracy at prediction of the mental health cluster assignment in cross-validation or the test subset (test accuracy: `r paste0((filter(full_class$overall_stats, dataset == 'test')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`,
 Cohen's $\kappa$: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$kappa), 2) %>% paste(collapse = ' - ')`, 
-Brier score: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$bs), 2) %>% paste(collapse = ' - ')`).
+Brier score: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`, __Supplementary Figures S9__ - __S10__, __Supplementary Table S10__).
+
+\newpage
 
@@
 
-Herein, we cross-sectionally explored mental health in victims of mountain sport accidents treated at a tertiary trauma center at median `r round(median(ptsd$dataset$obs_time)/30.436)` months after the trauma center admission.
-We identified three approximately equally sized subsets of participants with distinct mental health characteristic: 
-(1) the neutral mental health cluster, 
-(2) the PTG cluster characterized by predominantly salutatory reaction to the accident manifested by post-traumatic growth, and 
-(3) the PTS cluster hallmarked by symptoms of PTSD, anxiety, depression, panic disorder, persistent somatic symptoms as well as low resilience, quality of life and sense of coherence.
-PTS cluster patients were characterized by lower age, more frequent pre-existing physical and mental conditions, self-reported need for psychological support, long-term physical health consequences of the accident and cautious behavior during sport.
-The PTS cluster was also substantially enriched in hospitalized and surgery patients. 
-However, even with robust machine learning algorithms we could not establish a reliable model to identify PTS cluster patients at risk of long-term mental health problems with data available during acute medical management of the mountain sport accident.
+We characterized mental health outcomes in individuals treated at a tertiary trauma center for an accident in mountain sport at a median of `r round(median(ptsd$dataset$obs_time)/30.436)` months following the incident.
+We identified three approximately equally sized subsets of participants with distinct mental health characteristic: (1) the neutral mental health cluster, (2) the PTG cluster characterized by a predominantly salutatory reaction to the accident manifested by post-traumatic growth, and (3) the PTS cluster hallmarked by symptoms of PTSD, anxiety, depression, panic disorder and somatization as well as low resilience, quality of life and sense of coherence.
+Individuals in the PTS cluster were characterized by younger age, the highest frequency of pre-existing mental disorders and persisting physical health consequences of the accident compared to individuals in the other two clusters.
+They reported a need for psychological or psychiatric support following the accident and more cautious behavior during mountain sports since the accident.
+However, even with robust machine learning algorithms we could not establish a reliable model to identify PTS cluster patients at risk of long-term mental health problems with data available during acute medical management of the mountain sport accident victim.
 
-To the best of our knowledge this is the first study evaluating a broad palette of mental features including symptoms of anxiety and depression, somatic symptoms, resilience, post-traumatic growth and symptoms of PTSD in survivors of mountain sport accidents.
-This population is of special interest because, on one hand, mountain sport is generally considered beneficial for mental health [@Ower2019; @Niedermeier2017a]. 
-On the other hand, mountain sport is associated with specific individual physical and mental traits, and bears potential of danger [@Faulhaber2017; @Faulhaber2020].
-Studies conducted on mountain guides and mountain rescue personnel, showed that 71% - 78% of them experienced a traumatic event [@Mikutta2022; @Sommer2004]. 
-Intriguingly, the frequency of PTSD symptoms in those populations was estimated for 1% - 5.9% [@Traber2020; @Mikutta2022; @Sommer2004], which was comparable to the general population [@Frans2005; @PTSD2023; @Trautmann2018; @Burri2014] and, particularly, lower than in other potentially vulnerable groups such as emergency workers [@Berger2012].
+This study evaluates a broad palette of mental features including symptoms of anxiety and depression, somatiziation, resilience, sense of coherence and quality of life, post-traumatic growth and symptoms of PTSD in survivors of mountain sport accidents.
+This population is of special interest because, on one hand, mountain sports are generally considered beneficial for mental health [@Ower2019; @Niedermeier2017a]. 
+On the other hand, mountain sports are associated with specific individual physical and mental traits, and bear accident and injury risks [@Faulhaber2017; @Faulhaber2020]. 
+Not many studies have assessed PTSD rates following accidents in mountain sports. 
+Survivors of avalanches were affected by acute and long-term PTSD symptoms such as hyperarousal or sleep disorders at rates between 11% - 16% [@Thordardottir2015; @Leonard2021]. 
+This figure is comparable with `r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% participants of our study screened positive for at least one of of the PTSD symptom domains defined by the PCL-5 tool [@Bovin2016].
+PTSD frequencies of 11.8% - 26% have been reported at 3 - 12 months after acute trauma [@Shalev2019; @Schultebraucks2020; @Joseph2020].
+However, in our cohort, solely `r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`% of participants demonstrated concomitant PCL-5 domain B, C, D and E symptoms suggestive of fully manifest PTSD.
+
+Studies conducted on mountain guides and mountain rescue personnel, showed that 71% - 78% of them experienced a traumatic event [@Mikutta2022; @Sommer2004], while the reported frequency of PTSD was very low (0.9% - 2.7%) [@Mikutta2022; @Sommer2004].
+This is comparable to trauma event estimates for the general population [@PTSD2023; @Trautmann2018; @Burri2014] and, particularly, lower than in other potentially vulnerable groups such as emergency workers [@Berger2012].
 By contrast, a recent Swiss study found a prevalence of PTSD in up to 22% of mountain rescuers once differentiating between clinical interviews and self-reporting of symptoms for the assessment of PTSD [@Traber2020].
-Concerning mountain accidents, survivors of avalanches were affected by acute and long-term PTSD symptoms such as hyperarousal or sleep disorders at rates between 11 - 16% [@Thordardottir2015; @Leonard2021].
-This figure is comparable with `r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% participants of our study screened positive for at least one of of the PTSD symptom domains defined by the PCL-5 tool [@Bovin2016].
-Of note, the frequency of at least one PCL-5 PTSD symptom in our cohort recruited among patients of a trauma department is similar to the 11.8% - 26.6% prevalence reported in literature for emergency department or acute trauma patients [@Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Joseph2020].
-However, in our cohort, solely 
-`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`% of participants demonstrated concomitant PCL-5 domain B, c, D and E symptoms suggestive of fully manifest PTSD.
-Resilience has been proposed as a protective trait against mental health deterioration following trauma [@Ahmed2007; @Bensimon2012] and high resilience levels in mountain rescuers may be explain low prevalence of PTSD in this population despite trauma exposure [@Mikutta2022].
+
+Resilience has been proposed as a protective trait against negative mental health consequences following trauma [@Bensimon2012] and high resilience levels in mountain rescuers may explain low prevalence of PTSD in this population despite trauma exposure [@Mikutta2022].
 Two-thirds of our study participants were classified as highly resilient [@Leppert2008]. 
@@
 (EUROHIS, median: `r get_num_stats(ptsd$dataset, 'eurohis_total', collapse = ', interquartile range: ')`) comparable to the general western European population [@Schmidt2006].
-Post-traumatic growth represents a salutogenic reaction to a traumatic event with positive changes in personality, self-assurance, experience and inter-personal relationships [@Tedeschi1996], which may overlap with resilience and PTSD symptoms [@Liu2017; @Bensimon2012].
-This phenomenon was evident also in our study, where PTG cluster participants were characterized by a pronounced post-traumatic growth and good resilience scoring, which was however paralleled by flashbacks and single PTSD symptoms.
 
-By semi-supervised clustering, we identified mountain sport accident victims with co-existing symptoms of PTSD, depression, anxiety, somatic symptoms, as well as low sense of coherence and quality of life.
-Those patients assigned to the PTS subset may represent a potentially clinically relevant risk group of post-traumatic mental disorders. 
-The psychometric characteristic of the PTS cluster corresponds to the multi-faceted mental symptoms described for PTSD individuals [@Gagnon-Sanschagrin2022; @Brady2000; @Bryant2019; @Howie2019]. 
-Regarding early predictors of mental health problems, the PTS cluster was characterized by a significantly younger age as well as high rates of pre-existing physical and mental disorders. 
-It was also substantially enriched in females, low-to-moderate income patients, individuals requiring comrade or professional rescue, hospitalization and surgery as compared with neutral cluster patients.
+By semi-supervised clustering, we identified mountain sport accident victims with co-existing symptoms of PTSD, depression, anxiety, somatization, as well as low sense of coherence and quality of life.
+Those individuals assigned to the PTS subset may represent a clinically relevant risk group of post-traumatic mental disorders. 
+The psychometric characteristic of the PTS cluster corresponds to the multi-faceted mental symptoms described for 
+individuals following a traumatic event [@Gagnon-Sanschagrin2022; @Brady2000; @Bryant2019]. 
+Regarding early predictors of mental health problems in mountain accident victims, the PTS cluster was characterized by a significantly younger age as well as high rates of pre-existing physical and mental disorders.
+It was also substantially enriched in females, low-to-moderate income participants, individuals requiring companion or professional rescue, hospitalization and surgery as compared with neutral cluster patients.
 Of note, 
-female gender [@Shalev2019; @Christiansen2020; @Seedat2005; @Baecher2018], 
-young age at traumatization [@Schultebraucks2020; @Kongshoej2022], 
+female gender [@Shalev2019], 
+young age at traumatization [@Schultebraucks2020], 
 low income [@Carbone2019] and 
-pre-existing mental disorders [@Papini2018; @Sandweiss2011; @Baecher2018] has been proposed as risk factors of PTSD. 
-The recovery features of the PTS cluster were the highest rates of self-reported need for psychological support, persistent physical health consequences of the accident and the highest percentage of more cautious behavior during sport activity as compared with the neutral or PTG cluster.
-However, differences in the early or follow-up variables between the mental clusters in the study cohort were only of weak effect size, which underlines the lack of strong predictive markers and moderate explanatory power of published models predicting post-traumatic mental problems [@Shalev2019; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018; @Cohen2013; @Rice2005].
-PTSD risk has been previously associated with injury severity, head injuries, hospitalization length and pain [@Baecher2018; @Blanchard1995; @Galatzer-Levy2014]. 
-We could not establish any link between the clusters and, in particular, the PTS cluster with objective injury severity measured with the AIS scale [@Gennarelli2006] or injury location. 
+pre-existing mental disorders [@Papini2018] have been proposed as risk factors of PTSD. 
+PTSD risk has also been previously associated with injury severity, head injuries, hospitalization length and pain [@Baecher2018; @Blanchard1995; @Galatzer-Levy2014]. 
+We could not establish any link between the mental health clusters and, in particular, the PTS cluster with objective injury severity measured with the AIS scale [@Gennarelli2006] or injury location. 
 Intriguingly, subjective injury severity, which was not surveyed in our cohort, was shown to be a stronger predictor of PTSD symptoms than objective injury severity [@Gabert-Quillen2011]. 
-This phenomenon together with the longer survey - treatment time interval than in most other studies and differences in composition of the study sample, may explain the lacking effect of injury on later mental disorders in our collective.
-Importantly, the percentage of participants having received psychological or psychiatric support after the accident was generally low (`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`%) in the study cohort and did not differ between the mental clusters.
+This phenomenon together with the longer time interval between hospital admission and survey than in most other studies and differences in composition of the study sample, may explain the lacking effect of injury on subsequent mental disorders in our collective.
+Importantly, the percentage of participants having received psychological or psychiatric support after the accident was generally low in the study cohort (`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`%) and did not differ between the mental health clusters.
 Hence, it is unlikely that psychological or psychiatric intervention had any impact on classification of the participants.
 
-Early identification of trauma patients at risk of development of PTSD or follow-up mental health disorders has previously been tackled by diverse machine learning algorithms [@Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
-In our hands, none of seven machine learning algorithms employing different mathematical principles, random forest, neural network, support vector machines, recursive partitioning, conditional random forest, discriminant analysis or elastic net regression, was able to reliably predict the mental cluster assignment based on `r length(class_globals$variables$early)` demographic, socioeconomic, clinical and accident-related variables available during acute medical treatment of the mountain sport accident.
-This phenomenon is likely attributed to the lacking strong non-mental explanatory variables differentiating between the mental clusters and, of clinical importance, highly specific markers of post-traumatic stress patients.
+Early identification of trauma patients at risk of developing PTSD or further mental disorders has previously been tackled by diverse machine learning algorithms [@Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
+In our hands, none of seven machine learning procedures using different mathematical principles, random forest, neural network, support vector machines, recursive partitioning, conditional random forest, discriminant analysis or elastic net regression, was able to reliably predict the mental health cluster assignment based on `r length(class_globals$variables$early)` demographic, socioeconomic, clinical and accident-related variables available during acute medical treatment of the patient.
+This phenomenon is likely attributed to the lacking strong non-psychometric explanatory variables differentiating between the mental health clusters and, of clinical importance, highly specific markers of post-traumatic stress patients.
 Interestingly, inclusion of peri-traumatic mental symptoms like PTSD features, flashbacks, hyperarousal, or subjective need for support was shown to be crucial for optimal prediction of long-term mental disorders or PTSD during recovery [@Gabert-Quillen2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
@@
 Our study bears limitations. 
-The overall analysis inclusion rate was low and, as indicated by comparison of the excluded and analyzed participants, the results were potentially confounded by a selection bias towards better patients with higher social status and more severe injuries.
-Although the key of the mental clusters, post-traumatic growth and PTSD, were clearly associated with trauma [@AmericanPsychiatricAssociation2013; @Tedeschi1996; @Bovin2016], the mental cluster assignment could have been affected by other events than the sport accident.
-Furthermore, our data set misses important explanatory factors which may facilitate distinction between the mental clusters such as rehabilitation need or ability to work. 
-Analogically, due to the cross-sectional design of the study, we were not able to assess peri-traumatic mental disorder symptoms, which has been proven crucial for identification of vulnerable patients  [@Gabert-Quillen2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
-Finally, a longitudinal survey may have helped to assess individual trajectories of mental health reaction to the mountain sport accident.
+The overall analysis inclusion rate was low and, as indicated by comparison of the excluded and analyzed participants, the results were potentially affected by a selection bias towards individuals with higher social status and more severe injuries.
+Furthermore, the study survey did not recorded potentially important explanatory factors for distinction between the mental health clusters such as rehabilitation need or ability to work. 
+Due to the cross-sectional design of the study, we were not able to assess peri-traumatic mental health, which has been proven crucial for identification of vulnerable patients  [@Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Papini2018].
+A longitudinal survey may have helped to assess individual trajectories of mental health reaction to the mountain sport accident. 
+Finally, this analysis only included individuals who were treated for a mountain sport accident at the local trauma department. 
+Individuals without contact to the medical system or bystanders, who can also be affected by PTSD, were not examined.
 
@@
 
-Our study represents the first attempt to characterize post-traumatic stress and further facets of mental health following mountain sport accidents. 
-We identified symptoms of mental health impairment in roughly one third of the victims. 
-However, we could not identify robust predictors for post-traumatic stress among variables available during early medical treatment.
-Hence, considering the increasing popularity of mountain sports, low-threshold information and psychological support for trauma department patients treated seems the most appropriate strategy for successful multi-disciplinary medial management of mountain sport accidents.
-Such information campaign can include 
-informative leaflets, 
-additional information in the discharge letter, 
-as well as websites or mobile applications explaining frequent warning signs and symptoms of post-traumatic stress disorder and providing contact data for points of psychological support. 
+Our study represents the first attempt to characterize symptoms of PTSD and further facets of mental health following mountain sport accidents. 
+We identified symptoms of mental health impairment in roughly one third of the investigated individuals. 
+However, we could not identify robust predictors for impaired mental health at follow up among variables available during early medical treatment. 
+Hence, considering the increasing popularity of mountain sports, psychoeducational information and low threshold access to psychological support currently seems the most appropriate strategy for successful interdisciplinary management of individuals following accidents in mountain sport.
+Psychoeducational information can include informative leaflets, information in the discharge letter, as well as access to websites or mobile applications outlining the normal reaction to a traumatic event as well as warning signs indicating the need for professional support.
+Additionally low threshold contact information for psychological support are helpful. 
 We are currently implementing such measures at our hospital.
@@
 The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.
+
+# Acknowledgements
+
+We thank the participants for sharing their experiences with us. 
+This study was supported by a grant of the “Deutsche Gesellschaft für Berg- und Expeditionsmedizin”,  German Association for Mountain and Expedition Medicine to H.S. and K.H.
+
+# Conflict of interests
+
+The authors declare no conflict of interest.
 
@@
 flextable::flextable(tables$cohort_demo) %>% 
-  set_widths(c(5.5, 7.5)) %>% 
+  set_widths(c(7.5, 7.5)) %>% 
   footnote(1, 1, 
@@
 flextable::flextable(tables$cohort_accident) %>% 
-  set_widths(c(7.5, 7.5)) %>% 
+  set_widths(c(8.5, 7.5)) %>% 
   footnote(2, 1, 
@@
   flextable %>% 
-  set_widths(c(5.5, 7.5)) %>% 
+  set_widths(c(7.5, 7.5)) %>% 
   footnote(1, 1, 
-           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder.'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization, EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder.'),
            part = 'header', 
@@
 
-```{r fig-clust-psych, fig.width = figur::convert(figures$clust_psych, to = 'in')$w, fig.height = figur::convert(figures$clust_psych, to = 'in')$h, fig.cap = 'Scores of mental disorder symptoms, sense of coherence, resilience, quality of life, post-traumatic growth, and post-traumatic stress disorder in the mental clusters.'}
+```{r fig-clust-psych, fig.width = figur::convert(figures$clust_psych, to = 'in')$w, fig.height = figur::convert(figures$clust_psych, to = 'in')$h, fig.cap = 'Cluster-defining scores of symptoms of mental disorders, sense of coherence, resilience, quality of life, post-traumatic growth, and post-traumatic stress disorder in the mental health clusters.'}
 
@@
 __Figure \@ref(fig:fig-clust-psych).
-Scores of mental disorder symptoms, sense of coherence, resilience, quality of life, post-traumatic growth, and post-traumatic stress disorder in the mental clusters.__
+Cluster-defining scores of symptoms of mental disorders, sense of coherence, resilience, quality of life, post-traumatic growth, and post-traumatic stress disorder in the mental health clusters.__
 
-_Study participants in the training subset of the cohort were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) mental clusters defined in respect to psychometric scores by the partition around medoids (PAM) algorithm.
-Study participants in the test subset of the cohort were assigned to the mental clusters by the inverse distance weighted 7-nearest neighbors classifier.
+_Study participants in the training subset of the cohort were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) mental health clusters defined in respect to psychometric scores by the partition around medoids (PAM) algorithm.
+Study participants in the test subset of the cohort were assigned to the mental health clusters by the inverse distance weighted 27-nearest neighbor classifier.
 Differences in the cluster-defining psychometric scores between the clusters were assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
 P values were adjusted for multiple testing with the false discovery rate method.
-Mean normalized scores of major mental health readouts, sense of coherence and resilience (A), quality of life (B), post-traumatic growth (C) and post-traumatic stress disorder (PTSD, D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines.
+Mean normalized scores of major mental health readouts, sense of coherence and resilience (A), quality of life (B), post-traumatic growth (C) and post-traumatic stress disorder (PTSD, D) in the mental health clusters of the training and test subsets of the study cohort are depicted as solid lines.
 Tinted regions represent 2 $\times$ SEM (standard error of the mean).
@@
 Significant effects are highlighted in bold._
-_GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; som.
-sympt.: somatic symptoms; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
+
+_GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
 
@@
 
-```{r fig-clust-symptoms, fig.width = figur::convert(figures$clust_symptoms, to = 'in')$w, fig.height = figur::convert(figures$clust_symptoms, to = 'in')$h, fig.cap = 'Symptoms of mental disorders and resilience classes in the mental clusters.'}
+```{r fig-clust-symptoms, fig.width = figur::convert(figures$clust_symptoms, to = 'in')$w, fig.height = figur::convert(figures$clust_symptoms, to = 'in')$h, fig.cap = 'Symptoms of mental disorders and resilience classes in the mental health clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-clust-symptoms). Symptoms of mental disorders and resilience classes in the mental clusters.__
+__Figure \@ref(fig:fig-clust-symptoms). Symptoms of mental disorders and resilience classes in the mental health clusters.__
 
-_Differences in frequency of mental disorder symptoms and flashbacks, and distribution of resilience classes between the mental clusters (neutral, post-traumatic growth [PTG] and post-traumatic stress [PTS]) were investigated in the entire study cohort by $\chi^2$ test with Cramer's V effect size statistic.
+_Differences in frequency of symptoms of mental disorder and flashbacks, and distribution of resilience classes between the mental health clusters (neutral, post-traumatic growth [PTG] and post-traumatic stress [PTS]) were investigated in the entire study cohort by $\chi^2$ test with Cramer's V effect size statistic.
 P values were adjusted for multiple testing with the false discovery rate method._
 
-_(A) Symptoms of mental disorders, presence of flashbacks during sport activity and symptoms of post-traumatic stress disorder (PTSD).
-Percentages of affected individuals in the mental clusters are presented in bar plots.
+_(A) Symptoms of mental disorders, presence of flashbacks during mountain sport activity and symptoms of post-traumatic stress disorder (PTSD).
+Percentages of affected individuals in the mental health clusters are presented in bar plots.
 Effect sizes and p values are indicated in the Y axes.
@@
 
-_(B) Resilience classes and frequency of flashbacks during sport activity in the mental clusters presented in stack plots.
+_(B) Resilience classes and frequency of flashbacks during sport activity in the mental health clusters presented in stack plots.
 Effect sizes and p values are displayed in the plot captions.
@@
 
-_GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder._
+_GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder._
 
@@
 
-```{r fig-clinic-factors, fig.width = figur::convert(figures$clinic_factors, to = 'in')$w, fig.height = figur::convert(figures$clinic_factors, to = 'in')$h, fig.cap = 'Sociodemographic and medical history characteristic of the mental clusters of accident victims.'}
+```{r fig-clinic-factors, fig.width = figur::convert(figures$clinic_factors, to = 'in')$w, fig.height = figur::convert(figures$clinic_factors, to = 'in')$h, fig.cap = 'Sociodemographic and medical history characteristic of the mental health clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-clinic-factors). Sociodemographic and medical history characteristic of the mental clusters of accident victims.__
+__Figure \@ref(fig:fig-clinic-factors). Sociodemographic and medical history characteristic of the mental health clusters.__
 
-_Differences in age during the accident between the mental clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were assessed in the entire cohort by Kruskal-Wallis test with $\eta^2$ effect size statistic.
-Differences in genders, education, annual income classes (K: 1000 Euro), percentages of participants with trauma risk profession, traumatic events in the past (DIA-X: Diagnostic Expert System, traumatic event score), as well as pre-existing physical illness and pre-existing professionally diagnosed mental disorders between the mental clusters were investigated in the entire cohort by $\chi^2$ test with Cramer's V effect size statistic.
+_Differences in age at the time of the accident between the mental health clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were assessed in the entire cohort by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Differences in genders, education, annual income classes (K: 1000 Euro), percentages of participants with a search and rescue profession, traumatic events in the past (DIA-X: Diagnostic Expert System, traumatic event score), as well as pre-existing physical illness and pre-existing mental disorders diagnosed by a physician between the mental health clusters were investigated in the entire cohort by $\chi^2$ test with Cramer's V effect size statistic.
 P values were adjusted for multiple testing  with the false discovery rate method.
@@
 
-```{r fig-injury, fig.width = figur::convert(figures$injury, to = 'in')$w, fig.height = figur::convert(figures$injury, to = 'in')$h, fig.cap = 'Accident culprits, accident rescue, injury severity and injured body parts in the mental clusters.'}
+```{r fig-injury, fig.width = figur::convert(figures$injury, to = 'in')$w, fig.height = figur::convert(figures$injury, to = 'in')$h, fig.cap = 'Responsibility for the accident, accident rescue, injury severity and injured body parts in the mental health clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-injury). Accident culprits, accident rescue, injury severity and injured body parts in the mental clusters.__
+__Figure \@ref(fig:fig-injury). Responsibility for the accident, accident rescue, injury severity and injured body parts in the mental health clusters..__
 
-_Differences in accident culprit, rescue mode, injury severity, hospitalization and surgery rates, and injured body regions between the mental clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were investigated in the entire cohort by $\chi^2$ test with Cramer's V effect size statistic.
+_Differences in responsibility for the accident, rescue mode, injury severity, hospitalization and surgery rates, and injured body regions between the mental health clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were investigated in the entire cohort by $\chi^2$ test with Cramer's V effect size statistic.
 P values were adjusted for multiple testing with the false discovery rate method._
 
-_(A, B) Percentages of self and non-self accident culprit, distribution of the rescue modes and injury severity classes (AIS: abbreviated injury scale), rates of hospitalization and surgery in the mental clusters were presented in stack plots.
+_(A, B) Accident responsibility, distribution of the rescue modes and injury severity classes (AIS: abbreviated injury scale), rates of hospitalization and surgery expressed as percentages of the mental health clusters were presented in stack plots.
 Effect sizes and p values are displayed in the plot captions.
@@
 
-```{r fig-aftermath, fig.width = figur::convert(figures$aftermath, to = 'in')$w, fig.height = figur::convert(figures$aftermath, to = 'in')$h, fig.cap = 'Psychological support and consequences of the accident in the mental clusters.'}
+```{r fig-aftermath, fig.width = figur::convert(figures$aftermath, to = 'in')$w, fig.height = figur::convert(figures$aftermath, to = 'in')$h, fig.cap = 'Psychological support and consequences of the accident in the mental health clusters.'}
 
@@
 
-__Figure \@ref(fig:fig-aftermath). Psychological support and consequences of the accident in the mental clusters.__
+__Figure \@ref(fig:fig-aftermath). Psychological support and consequences of the accident in the mental health clusters.__
 
-_Differences in rates of received psychological or psychiatric support, self-reported need for psychological support, persistent physical health consequences of the accident, cautious behavior during mountain sport activity and percentages of participants having returned to the same mountain sport activity between the mental clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic.
+_Differences in rates of received psychological or psychiatric support, self-reported need for psychological or psychiatric support support after the accident, persistent physical health consequences of the accident, cautious behavior during mountain sport and percentages of participants having returned to the same mountain sport between the mental health clusters (neutral, post-traumatic growth [PTG], post-traumatic stress [PTS]) were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic.
 P values were adjusted for multiple testing with the false discovery rate method.
@@
 
-```{r fig-early-class, fig.width = figur::convert(figures$early_class, to = 'in')$w, fig.height = figur::convert(figures$early_class, to = 'in')$h, fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.'}
+```{r fig-early-class, fig.width = figur::convert(figures$early_class, to = 'in')$w, fig.height = figur::convert(figures$early_class, to = 'in')$h, fig.cap = 'Assignment of accident victims to the mental health clusters based on explanatory factors available during acute medical management of the patient.'}
 
@@
 
-__Figure \@ref(fig:fig-early-class). Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.__
+__Figure \@ref(fig:fig-early-class). Assignment of accident victims to the mental health clusters based on explanatory factors available during acute medical management of the patient.__
 
-_The mental cluster assignment was modeled with demographic, medical history and accident-related explanatory factors available during acute medical management of the accident.
-Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set._
+_The mental health cluster assignment was modeled with demographic, medical history and accident-related explanatory factors available during acute medical management of the patient.
+Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes, presence and frequency of flashbacks and variables concerning recovery and long term consequences of the accident were excluded from the explanatory factor set._
 
@@
 _(B) Sensitivity (Se) and specificity (Sp) of detection of participants assigned to the PTS cluster (post-traumatic stress) investigated by receiver-operating characteristic in the training and test subset of the study cohort.
-Sensitivity and specificity values are indicted in the plots.
+Sensitivity and specificity values are indicated in the plots.
 Line color codes for the modeling algorithm.

### 2023-11-07 paper/markdown/supplementary_material.Rmd
Feedback, October 2023, markdown and figures

@@
 
-For semi-supervised clustering, diagnostic and performance of the clustering analysis, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized.
+For semi-supervised clustering, diagnostic and performance testing of the clustering analysis, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized.
 UMAP (uniform manifold approximation and projection) was done with the _umap_ and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) packages [@Konopka2022; @McInnes2018]. 
@@
 panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
-persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
+common persistent somatic symptoms as a substitute for somatization (PHQ-15) [@Kroenke2002], 
 resilience (RS13: 13-item resilience scale) [@Leppert2008], 
@@
 post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
-post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016].
+post-traumatic stress disorder (PCL-5: PTSD checklist for DSM-5) [@Bovin2016].
 
-Clinically relevant symptoms of anxiety were defined as GAD-7 $geq$ 10, 
+Clinically relevant symptoms of anxiety were defined as GAD-7 $\geq$ 10, 
 clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], 
@@
 
-Items of the EUROHIS QOL 8 tool were scored as with 1 - 5 likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing).
+Items of the EUROHIS QOL 8 tool were scored as with 1 - 5 Likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing).
 The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].
@@
 
-Separate scores were calculated for domains B, C, D and E of the PCL-5 DSM-5 tool along with the total score being the sum of all items. 
-Each PCL-5 DSM-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
+Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
+Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
 Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
 Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
-Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 DSM-5 domains [@Bovin2016].
+Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].
 
-Traumatic events prior to the sport accident were assessed with the DIA-X tool (Diagnostic Expert System) [@Maercker2005]. 
+Traumatic events prior to the mountain sport accident were assessed with the DIA-X tool (Diagnostic Expert System) [@Maercker2005]. 
 Direct personal experience or being a witness of a traumatic event specified by the DIA-X questionnaire or by an additional yes/no item ('other traumatic events') was scored as 1. 
 Prior traumatic event was assumed with at least one DIA-X item scored with 1. 
-Prior sport accidents, 
+Pre-existing mental disorder diagnosed by a physician, 
+prior mountain sport accidents, 
 flashbacks during mountain sport activity, 
 confusion during mountain sport activity, 
-self-reported need for psychological support, 
+self-reported need for psychological support following the accident, 
 psychological/psychiatric support/therapy after the accident 
-and presence of persistent physical health consequences after the accident 
+and presence of persistent physical health consequences related to the accident 
 were surveyed as single yes/no items. 
@@
 
-* Cohen's $\kappa$ reliability: none: < 0.2, minimal: 0.2 - 0.4, weak: 0.4 - 0.6, moderate: 0.6 - 0.8, strong: $\geq$ 0.8
+* Cohen's $\kappa$ inter-rater reliability statistic: none: < 0.2, minimal: 0.2 - 0.4, weak: 0.4 - 0.6, moderate: 0.6 - 0.8, strong: $\geq$ 0.8
 
@@
 The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
-The choice of the clustering algorithm was motivated by its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares, method `var()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier, method `cv()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__.
+The choice of the clustering algorithm was motivated by 
+its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares), 
+good separability between the clusters measured by mean silhouette width [@Rousseeuw1987], 
+and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__.
+The explained variance, mean silhouette width and cross-validated accuracy statistics were computed with the methods `var()`, `silhouette()` and `cv()` from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
 The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020].
-By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).
+By this means, three mental health clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).
 
-Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
+Assignment of the training subset observations to the mental health clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
 This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
@@
 Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
-An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental clusters (i.e.
+An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental health clusters (i.e.
 neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g.
 neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__).
-Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).
+Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables between the mental health clusters in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).
 
-Differences in frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical, accident- and recovery-related factors between the mental clusters were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic and by Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Supplementary Table S\@ref(tab:tab-clust-bcg)__).
+Differences in frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical, accident- and recovery-related factors between the mental health clusters were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic and by Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Supplementary Table S\@ref(tab:tab-clust-symptoms)__ and __S\@ref(tab:tab-clust-bcg)__).
 
@@
 
-Two types of multi-parameter machine learning classifiers of the mental cluster assignment were developed in the training subset:
+Two types of multi-parameter machine learning classifiers of the mental health cluster assignment were developed in the training subset:
 
@@
 models employing candidate early predictors of cluster assignment, i.e.
-demographic, socioeconomic, medical history and accident-related explanatory factors available during acute medical management of the accident
+demographic, socioeconomic, medical history and accident-related explanatory factors available during acute medical management of the accident victim
 
@@
 
-Of note, psychometric variables used for definition of the mental clusters as well as mental disorder symptoms, presence and frequency of flashbacks were excluded from the explanatory variable sets. 
+Of note, psychometric variables used for definition of the mental health clusters as well as symptoms of mental disorders, presence and frequency of flashbacks were excluded from the explanatory variable sets.
 The explanatory variables are listed in __Supplementary Table S\@ref(tab:tab-mod-variables)__.
@@
 
-```{r tab-analysis-vars, tab.cap = 'Variables used in the analysis pipeline.'}
+```{r tab-analysis-vars, tab.cap = 'Variables used in the analysis pipeline. The table is available as a supplementary Excel file.'}
 
@@
   footnote(1, 2:3, 
-           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; PHQ-panic: patient health questionnaire module for panic disorders; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
            part = 'header', 
@@
 flextable::flextable(suppl_tables$consistency) %>% 
-  set_widths(c(3.5, 3, 3)) %>% 
+  set_widths(c(4.5, 3, 3)) %>% 
    footnote(1, 1, 
-           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; PHQ-panic: patient health questionnaire module for panic disorders; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'),
            part = 'header', 
@@
    footnote(1, 1, 
-           value = as_paragraph('AIS: abbreviated injury scale'), 
+           value = as_paragraph('AIS: abbreviated injury scale; ICD-10: 10th version of the International Statistical Classification of Diseases and Related Health Problems.'), 
            part = 'header', 
@@
 
-```{r tab-clust-fct, tab.cap = 'Differences in psychometric clustering factors between the mental clusters.
+```{r tab-clust-fct, tab.cap = 'Differences in psychometric clustering factors between the mental health clusters.
 Numeric variables are presented as medians with interquartile ranges (IQR).
@@
 
-```{r tab-clust-bcg, tab.cap = 'Significant differences in demographic, socioeconomic, clinical, accident- and recovery-related factors, and mental disorder symptoms between the mental clusters in the entire cohort.
+```{r tab-clust-symptoms, tab.cap = 'Frequency of mental disorder symptoms in the mental health clusters in the entire cohort.
+Categorical variables are presented as percentages and counts within the clusters.'}
+
+flextable::flextable(suppl_tables$clust_symptoms) %>% 
+  set_widths(c(4, rep(4.3, 3), 2.5, 2.3)) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; PHQ-panic: patient health questionnaire module for panic disorders; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 5:6, 
+           value = as_paragraph('\u03C7\u00B2 test with Cramer V effect size statistic. P values were corrected for multiple testing with the false discovery rate method.'), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-clust-bcg, tab.cap = 'Differences in demographic, socioeconomic, clinical, accident- and recovery-related factors, and between the mental clusters in the entire cohort.
+Significant effects are presented, the full table is available as a supplementary Excel file.
 Numeric variables are presented as medians with interquartile ranges (IQR).
-Categorical variables are presented as percentages and counts within the complete observation set.'}
+Categorical variables are presented as percentages and counts within the clusters.'}
 
-flextable::flextable(suppl_tables$clust_bcg) %>% 
+suppl_tables$clust_bcg %>%
+  filter(!stri_detect(Significance, regex = '^ns')) %>% 
+  flextable %>% 
   set_widths(c(3, rep(4.3, 3), 2.5, 2.3)) %>% 
   footnote(1, 1, 
-           value = as_paragraph('AIS: abbreviated injury score; PTSD: post-traumatic stress disorder; RS13: 13-item resilience scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire.'),
+           value = as_paragraph('psychological support need: subjective need for psychological or psychiatric support following the accident; physical health consequences: physical health consequences related to the accident.'),
            part = 'header', 
@@
 
-```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors used for modeling of the mental cluster assignment.'}
+```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors used for modeling of the mental health cluster assignment.'}
 
@@
   footnote(1, 1, 
-           value = as_paragraph('early predictors: variables available during acute medical management of the accident; full predictor set: variables available during acute medical management of the accident and during follow-up.'),
+           value = as_paragraph('early predictors: variables available during acute medical management of the patient; full predictor set: variables available during acute medical management of the patient and during follow-up.'),
            part = 'header', 
@@
 
-```{r tab-early-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental cluster assignment.
-Models employing early predictors available during acute medical management of the accident.'}
+```{r tab-early-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental health cluster assignment.
+Models employing early predictors available during acute medical management of the patient.'}
 
@@
 
-```{r tab-full-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental cluster assignment.
-Models employing the full predictor set available during acute medical management of the accident and follow-up.'}
+```{r tab-full-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental health cluster assignment.
+Models employing the full predictor set available during acute medical management of the patient and follow-up.'}
 
@@
 
-```{r fig-clust-dev, fig.width = figur::convert(suppl_figures$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_figures$clust_dev, to = 'in')$h, fig.cap = 'Definition of the mental clusters in the training subset of the study cohort.'}
+```{r fig-clust-dev, fig.width = figur::convert(suppl_figures$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_figures$clust_dev, to = 'in')$h, fig.cap = 'Definition of the mental health clusters in the training subset of the study cohort.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-clust-dev). Definition of the mental clusters in the training subset of the study cohort.__
+__Supplementary Figure S\@ref(fig:fig-clust-dev). Definition of the mental health clusters in the training subset of the study cohort.__
 
-_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations._
+_The mental health clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations._
 
@@
 
-_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations.
-Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 27-nearest neighbors classifier.
-By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress).
-Numbers of observations in the mental clusters are shown in the plot legend in (A)._
+_The mental health clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations.
+Assignment of the test subset observations to the mental health clusters was done with the inverse distance weighted 27-nearest neighbors classifier.
+By this procedure, three mental health clusters were identified: neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress).
+Numbers of observations in the mental health clusters are shown in the plot legend in (A)._
 
@@
 
-_(B) Pairwise cosine distances between observations in the mental clusters of the training and test subsets presented in heat maps._
+_(B) Pairwise cosine distances between observations in the mental health clusters of the training and test subsets presented in heat maps._
 
@@
 
-```{r fig-cosine, fig.width = figur::convert(suppl_figures$cosine, to = 'in')$w, fig.height = figur::convert(suppl_figures$cosine, to = 'in')$h, fig.cap = 'Distribution of the mental clusters and cosine distances between the mental clusters in the training and test subset of the study cohort.'}
+```{r fig-cosine, fig.width = figur::convert(suppl_figures$cosine, to = 'in')$w, fig.height = figur::convert(suppl_figures$cosine, to = 'in')$h, fig.cap = 'Distribution of the mental health clusters and cosine distances between the mental health clusters in the training and test subset of the study cohort.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-cosine). Distribution of the mental clusters and cosine distances between the mental clusters in the training and test subset of the study cohort.__
+__Supplementary Figure S\@ref(fig:fig-cosine). Distribution of the mental health clusters and cosine distances between the mental health clusters in the training and test subset of the study cohort.__
 
-_Study cohorts observations were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) clusters by semi-supervised clustering as presented in Supplementary Figure S\@ref(fig:fig-semi-clust)._
+_Study cohorts observations were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) mental health clusters by semi-supervised clustering as presented in Supplementary Figure S\@ref(fig:fig-semi-clust)._
 
-_(A) Percentages of observations in the training and test subsets of the study cohort in the mental clusters. Numbers of complete observations are indicated in the Y axis._
+_(A) Percentages of observations in the training and test subsets of the study cohort in the mental health clusters. Numbers of complete observations are indicated in the Y axis._
 
-_(B) Cosine cross-distances between the mental clusters in the training and test subset of the study dataset visualized as a heat map. Cross-distances are indicated in the tiles._
+_(B) Cosine cross-distances between the mental health clusters in the training and test subsets of the study dataset visualized as a heat map.
+Average cross-distances with 95% percentile range are indicated in the tiles._
 
@@
 
-```{r fig-feat-hm, fig.width = figur::convert(suppl_figures$feat_hm, to = 'in')$w, fig.height = figur::convert(suppl_figures$feat_hm, to = 'in')$h, fig.cap = 'Levels of psychometric scores used for the cluster definition in the mental clusters.'}
+```{r fig-feat-hm, fig.width = figur::convert(suppl_figures$feat_hm, to = 'in')$w, fig.height = figur::convert(suppl_figures$feat_hm, to = 'in')$h, fig.cap = 'Levels of psychometric scores used for the cluster definition in the mental health clusters.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-feat-hm). Levels of psychometric scores used for the cluster definition in the mental clusters.__
+__Supplementary Figure S\@ref(fig:fig-feat-hm). Levels of psychometric scores used for the cluster definition in the mental health clusters.__
 
-_Levels of psychometric scores used in clustering in the mental clusters.
+_Levels of psychometric scores used in clustering in the mental health clusters.
 Normalized, median-centered score values (Z scores) are presented as heat maps.
 Numbers of observations in the clusters are displayed in the plot captions.
-PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
+PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
 
@@
 
-```{r fig-demo, fig.width = figur::convert(suppl_figures$demo, to = 'in')$w, fig.height = figur::convert(suppl_figures$demo, to = 'in')$h, fig.cap = 'Employment status, prior sport accidents, and accident details in the mental clusters.'}
+```{r fig-demo, fig.width = figur::convert(suppl_figures$demo, to = 'in')$w, fig.height = figur::convert(suppl_figures$demo, to = 'in')$h, fig.cap = 'Employment status, prior mountain sport accidents, and accident details in the mental health clusters.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-demo). Employment status, prior sport accidents, and accident details in the mental clusters.__
+__Supplementary Figure S\@ref(fig:fig-demo). Employment status, prior mountain sport accidents, and accident details in the mental health clusters.__
 
-_Distribution of employment status, frequency of sport accidents in the past, distribution of the accidents in the seasons, accident sport types, frequency of being alone during the accident and number of injured persons in the mental clusters.
+_Distribution of employment status, frequency of mountain sport accidents in the past, distribution of the accidents in the seasons, accident sport types, frequency of being alone during the accident and number of injured persons in the mental health clusters.
 Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
@@
 ```{r fig-importance-early, fig.width = figur::convert(suppl_figures$importance_early, to = 'in')$w, fig.height = figur::convert(suppl_figures$importance_early, to = 'in')$h, fig.cap = 'Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
-Predictors available during acute medical management of the accident.'}
+Predictors available during acute medical management of the patient.'}
 
@@
 Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
-Predictors available during acute medical management of the accident.__
+Predictors available during acute medical management of the patient.__
 
-_The cluster assignment was modeled with demographic, medical history and accident-related explanatory factors available during acute medical management of the accident.
-Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set.
-Variable importance metrics of machine learning algorithms with the best performance at predicting the mental cluster assignment in the test subset of the study cohorts were computed.
+_Assignment to the mental health clusters was modeled with demographic, medical history and accident-related explanatory factors available during acute medical management of the patient. 
+Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes, presence and frequency of flashbacks and variables concerning recovery and long term consequences of the accident were excluded from the explanatory factor set.
+Variable importance metrics of machine learning algorithms with the best performance at predicting the mental health cluster assignment in the test subset of the study cohorts were computed.
 Importance metrics for the top 20 most important variables are presented as bar plots._
@@
 
-```{r fig-full-class, fig.width = figur::convert(suppl_figures$full_class, to = 'in')$w, fig.height = figur::convert(suppl_figures$full_class, to = 'in')$h, fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident and long-term follow-up.'}
+```{r fig-full-class, fig.width = figur::convert(suppl_figures$full_class, to = 'in')$w, fig.height = figur::convert(suppl_figures$full_class, to = 'in')$h, fig.cap = 'Assignment of accident victims to the mental health clusters based on explanatory factors available during acute medical management of the patient and long-term follow-up.'}
 
@@
 __Supplementary Figure S\@ref(fig:fig-full-class).
-Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident and long-term follow-up.__
+Assignment of accident victims to the mental health clusters based on explanatory factors available during acute medical management of the patient and long-term follow-up.__
 
-_The cluster assignment was modeled with demographic, medical history, accident- and recovery-related explanatory factors available during acute medical management of the accident and follow-up.
-Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set._
+_Assignment to the mental health clusters was modeled with demographic, medical history, accident- and recovery-related explanatory factors available during acute medical management of the patient and follow-up.
+Psychometric variables used for cluster definition, symptoms of mental disorders, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set._
 
-_(A) Accuracy of the predicted Cluster assignment and predictive performance of the modeling algorithms was assessed by overall cluster assignment accuracy, Cohen's $\kappa$ inter-rater accuracy metric (high values indicate good accuracy) and Brier score (low values indicate good performance) in the training and test subsets of the study cohort.
+_(A) Accuracy of the predicted Cluster assignment and predictive performance of the modeling algorithms was assessed by overall accuracy, Cohen's $\kappa$ inter-rater accuracy metric (high values indicate good accuracy) and Brier score (low values indicate good performance) in the training and test subsets of the study cohort.
 Performance metrics are presented in scatter plots.
@@
 ```{r fig-importance-full, fig.width = figur::convert(suppl_figures$importance_full, to = 'in')$w, fig.height = figur::convert(suppl_figures$importance_full, to = 'in')$h, fig.cap = 'Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
-Predictors available during acute medical management of the accident and during the follow-up.'}
+Predictors available during acute medical management of the patient and follow-up.'}
 
@@
 Variable importance metrics for the random forest, support vector machine, discriminant analysis, and conditional random forest algorithms.
-Predictors available during acute medical management of the accident and during the follow-up.__
+Predictors available during acute medical management of the patient and follow-up.__
 
-_The cluster assignment was modeled with demographic, medical history, accident- and recovery-related explanatory factors available during acute medical management of the accident and follow-up.
-Psychometric variables used for cluster definition, mental disorder symptoms, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set.
-Variable importance metrics of machine learning algorithms with the best performance at predicting the mental cluster assignment in the test subset of the study cohorts were computed.
+_Assignment to the mental health clusters was modeled with demographic, medical history, accident- and recovery-related explanatory factors available during acute medical management of the patient and follow-up.
+Psychometric variables used for cluster definition, symptoms of mental disorders, resilience classes as well as presence and frequency of flashbacks were excluded from the explanatory factor set.
+Variable importance metrics of machine learning algorithms with the best performance at predicting the mental health cluster assignment in the test subset of the study cohorts were computed.
 Importance metrics for the top 20 most important variables are presented as bar plots._

### 2023-11-17 paper/markdown/manuscript.Rmd
Compatibility with new clustering tools
Changes in code for compatibility with the latest version of `clustTools`. Minor graphical changes in figures. No effects on results.

@@
 Subsequently, the mental health cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
-(training: `r signif(semi_clust$stats$variance[1], 2)`, test: `r signif(semi_clust$stats$variance[2], 2)`), 
+(training: `r signif(semi_clust$stats$frac_var[1], 2)`, test: `r signif(semi_clust$stats$frac_var[2], 2)`), 
 comparable average silhouette statistics 
-(training: `r signif(semi_clust$stats$mean_sil[1], 2)`, test: `r signif(semi_clust$stats$mean_sil[2], 2)`), 
+(training: `r signif(semi_clust$stats$sil_width[1], 2)`, test: `r signif(semi_clust$stats$sil_width[2], 2)`), 
 comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).

### 2023-11-17 paper/markdown/supplementary_material.Rmd
Compatibility with new clustering tools
Changes in code for compatibility with the latest version of `clustTools`. Minor graphical changes in figures. No effects on results.

@@
 This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
-training (variance = `r signif(semi_clust$stats$variance[1], 2)`, 
-average silhouette = `r signif(semi_clust$stats$mean_sil[1], 2)`) and 
-test subset (variance = `r signif(semi_clust$stats$variance[2], 2)`, 
-average silhouette = `r signif(semi_clust$stats$mean_sil[2], 2)`). 
+training (variance = `r signif(semi_clust$stats$frac_var[1], 2)`, 
+average silhouette = `r signif(semi_clust$stats$sil_width[1], 2)`) and 
+test subset (variance = `r signif(semi_clust$stats$frac_var[2], 2)`, 
+average silhouette = `r signif(semi_clust$stats$sil_width[2], 2)`). 
 Comparably good separation of the clusters could be discerned in the training and test subsets by 

### 2023-11-17 clustering scripts/clustering.R

@@ -55,7 +55,7 @@
   
 # Cluster assignment ------
   
-  insert_msg('Clusetr assignment tables')
+  insert_msg('Cluster assignment tables')
   
   semi_clust$assignment <- semi_clust$clust_obj %>% 
     map(~.x$clust_assignment) %>% 
@@ -65,36 +65,20 @@
   
   insert_msg('Clustering variances and silhouettes')
   
-  ## variance
-  
-  semi_clust$stats$variance <- semi_clust$clust_obj %>% 
-    map(var) %>% 
-    map(~.x$frac_var) %>% 
-    compress(names_to = 'partition', 
-             values_to = 'variance') %>% 
-    map_dfc(unlist)
-  
+  ## explained clustering variance
   ## average silhouette and percentage of observations with negative 
   ## metric values indicative of possible miss-classification
   
-  semi_clust$stats$silhouette <- semi_clust$clust_obj %>% 
-    map(silhouette) %>% 
+  semi_clust$stats <- semi_clust$clust_obj %>% 
     map(summary) %>% 
-    map(filter, clust_id == 'global') %>% 
-    compress(names_to = 'partition') %>% 
-    select(partition, perc_negative, mean) %>% 
-    set_names(c('partition', 
-                'perc_negative_sil', 
-                'mean_sil'))
-  
-  semi_clust$stats <- reduce(semi_clust$stats, left_join, by = 'partition')
+    compress(names_to = 'partition')
 
 # Plots of variances and average silhouettes -------
   
   insert_msg('Plots of variances and silhouettes')
   
   semi_clust$stat_plot <- semi_clust$stats %>% 
-    pivot_longer(cols = all_of(c('variance', 'mean_sil')), 
+    pivot_longer(cols = all_of(c('frac_var', 'sil_width')), 
                  names_to = 'statistic', 
                  values_to = 'value') %>%  
     ggplot(aes(x = value, 
@@ -112,8 +96,8 @@
                                  test = 'coral4'), 
                       name = 'Subset') +
     scale_x_continuous(limits = c(0, 1)) + 
-    scale_y_discrete(labels = c(mean_sil = 'avg. silhouette', 
-                                variance = 'variance')) + 
+    scale_y_discrete(labels = c(sil_width = 'avg. silhouette', 
+                                frac_var = 'variance')) + 
     globals$common_theme + 
     theme(axis.title.y = element_blank()) + 
     labs(title = 'Semi-supervised clustering', 
[…]

### 2023-11-17 clustering scripts/development.R

@@ -6,6 +6,12 @@
   
   clust_devel <- list()
   
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
 # globals -----
   
   insert_msg('Analysis globals')
@@ -50,36 +56,21 @@
         k = 3, 
         clust_fun = 'pam') 
 
-# Clustering variances ------
+# Silhouettes, fractions of misclassified, variance and neighborhood -----
   
-  insert_msg('Clustering variances')
+  insert_msg('Quality stats')
   
-  clust_devel$variance <- clust_devel$algos %>% 
-    map(var) %>% 
-    map(~.x$frac_var) %>% 
-    compress(names_to = 'method', 
-             values_to = 'variance') %>% 
-    mutate(variance = unlist(variance))
+  ## done for the entire training cohort
   
-# Average silhouette widths -------
-  
-  insert_msg('Average silhouette widths')
-  
-  clust_devel$silhouette <- clust_devel$algos %>% 
-    map(silhouette) %>% 
+  clust_devel$train_stats <- clust_devel$algos %>% 
     map(summary) %>% 
-    map(filter, clust_id == 'global') %>% 
-    compress(names_to = 'method') %>% 
-    select(method, perc_negative, mean) %>% 
-    set_names(c('method', 'perc_negative_sil', 'mean_sil'))
-  
+    compress(names_to = 'method')
+
 # Cross-validation ----
   
   insert_msg('Cross-validation')
   
-  plan('multisession')
-  
-  clust_devel$cv <- clust_devel$algos %>% 
+  clust_devel$cv_stats <- clust_devel$algos %>% 
     future_map(cv, 
                nfolds = 10, 
                kNN = 27, 
@@ -90,19 +81,17 @@
                                         packages = c('clustTools', 
                                                      'somKernels'))) 
   
-  plan('sequential')
-  
-  clust_devel$cv <- clust_devel$cv %>% 
+  clust_devel$cv_stats <- clust_devel$cv %>% 
     map(summary) %>% 
     compress(names_to = 'method') %>% 
-    select(method, accuracy_mean)
+    select(method, ends_with('mean'))
     
 # Common results table and result visualization ------
   
   insert_msg('Common result table and result visualization')
   
   clust_devel$result_tbl <- 
-    clust_devel[c("variance", "silhouette", "cv")] %>% 
+    clust_devel[c("train_stats", "cv_stats")] %>% 
[…]

### 2023-11-17 exploration.R

@@ -68,19 +68,10 @@
   
   ## cached result of the power estimation analysis
   
-  if(file.exists('./cache/power.RData')) {
-    
-    insert_msg('Loading cached power estimation results')
-    
-    load('./cache/power.RData')
-    
-  } else {
-    
-    source_all('./exploration scripts/power.R', 
-               message = TRUE, crash = TRUE)
-    
-  }
-  
+  access_cache(cache_path = './cache/power.RData', 
+               script_path = './exploration scripts/power.R', 
+               message = 'Loading cached power estimation results')
+
   ## observation time dependency of psychometric variables
   
   c('./exploration scripts/time_psycho.R', 

### 2023-11-17 import.R

@@ -1097,18 +1097,10 @@
   
   ## randomization be numeric psychometric outcomes
   
-  if(file.exists('./cache/randomization.RData')) {
-    
-    insert_msg('Loading cached randomization results')
-    
-    load('./cache/randomization.RData')
-    
-  } else {
-    
-    source_all('./import scripts/partition.R')
-    
-  }
-  
+  access_cache(cache_path = './cache/randomization.RData', 
+               script_path = './import scripts/partition.R', 
+               message = 'Loading cached randomization results')
+
   ptsd$rand_scheme <- part$best_scheme
   
   ptsd$dataset <- left_join(ptsd$dataset, 

### 2023-11-17 paper scripts/supplementary.R

@@ -103,7 +103,7 @@
   suppl_figures$semi_clust$upper <- 
     list(x = semi_clust$data_umap, 
          y = paste('UMAP,', c('training', 'test')), 
-         z = semi_clust$stats$variance) %>% 
+         z = semi_clust$stats$frac_var) %>% 
     pmap(function(x, y, z) x + 
            labs(title = y, 
                 subtitle = paste0(x$labels$subtitle, 
@@ -188,6 +188,9 @@
   suppl_figures$feat_hm <- feat_clust$clust_hm %>% 
     map(~.x + 
           theme(legend.position = 'none', 
+                axis.text.x = element_blank(),
+                axis.ticks.x = element_blank(), 
+                axis.line.x = element_blank(), 
                 plot.tag = element_blank(), 
                 strip.background.y = element_blank(), 
                 strip.text.y = element_blank(), 

### 2024-02-08 paper/markdown/review_response.Rmd
P2P reply, supplementary material
Changes in supplementary figures and tables required by the reviewer's comments

---
title: "Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center"
subtitle: "Point-to-point response"
author: "Psychiatry Study Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: ptsd_biblio.bib
csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 9)

```

\newpage

# Editor

Dear Editor,

Thank you very much for the positive and very thoughtful review of our manuscript. We appreciate a lot the reviewers’ comments and suggestions to improve our manuscript. 
We incorporated them in the revised text.
Please find below the point-to-point response to the reviewers’ issues. 
Since the reviewers requested more information and literature evidence, we also deleted few less important references from the initial submission to keep the bibliography within reasonable limits. 
Additionally, some minor copyediting was done. 
Changes in the main manuscript text are highlighted in red. 
We hope that the manuscript is now suitable for publication in _European Archives of Psychiatry and Clinical Neuroscience_.

Best regards,

Katharina Hüfner

\newpage

# Reviewer 1

## Point 1

Summary: the study provides the first data on mental healthy consequences of alpine sport accidents. 
The topic is of high interest specifically of the medical specialists of the alpine region countries. 
The study design is sound and the methods a highly innovative. 
Overall I recommend publishing this interesting data.

## Response 1

Thank you for the constructive feedback and the overall positive assessment of our study.

## Point 2

__Abstract__: The abstract is well structured. 
Probably it might be better not to make the first sentence about a topic which the study does not cover (positive effects of mountain sports).

## Response 2

We agree that the first sentence of __Abstract__ does not specifically relate to the subject of our research, which concerns primarily negative psycho-pathological burden of mountain sport accidents.
We changed this sentence of __Abstract__ attempting to accommodate the reviewer´s concern but did not completely remove it because we feel that especially “mountain sports” are generally believed to bolster mental health. 
Additionally, individuals performing mountain sports are often portrayed as resilient personalities protected from mental disorders, which we can in part support with our data (__Supplementary Figures S4__ and __S5__). 
In our clinical practice, these beliefs are a common source of stigma in members of the mountaineering community. 
So I hope this makes it more clear why we did not remove the first sentence completely and hopefully the reviewer can understand our intentions.

## Point 3

__Introduction__: The introduction is well written. 
At the end I think it would help to make the hypothesis more clear. 
I know that with SVM methods a hypothesis free approach is common, however the authors intend to (1) categorise the sample (2) find predictors.

## Response 3

Thank you for this suggestion. 
In the last paragraph of __Introduction__, we stressed the exploratory character of the study and the general strategy, i.e. classification of participants by their mental health measures followed by a search for predictors of mental health classes/clusters.

## Point 4

__Methods__: The methods are highly interesting and - as far as I can judge - are sound. 
Sample: The authors clearly state the problem of the low response rate of about 6%. 
There is a clear bias, however studies indicate that in mental health normally prevalences are under reported not overestimated.

## Response 4

We absolutely agree with your comment that the low response rate is a source of bias. 
Unfortunately this is a challenge of many studies tackling with mental disorders such as PTSD, anxiety and depression.

To assess the potential effects of the low response rate on the cohort composition we included an extensive comparison of the analysis cohort with invited individuals and participants excluded due to incomplete psychometric data in the initial submission. 
In the revised manuscript we investigated the potential selection bias in more detail, by comparing sociodemographic features of the cohort with the general Austrian population and by comparing characteristic of the mountain sport accident in the study cohort with a nation-wide collective of Austrian mountain accident victims.

As presented in __Supplementary Table S6__ and discussed in __Results/Characteristic of the study cohort__, the study sample was characterized by significantly higher frequencies of males, persons with tertiary education grade and actively employed as compared with the general Austrian population. 
In turn, rates of smokers and individuals suffering from chronic physical or mental disorders was significantly lower. 
Effect sizes of those differences were, however, very small (Cramer's V 0.00078 to 0.0025). 
Results of the comparison of the accident characteristic are presented in __Supplementary Figure 1__ and described in __Results/Characteristic of the study cohort__. 
They indicate significantly higher frequency of winter sport accidents in the study cohorts than in the Austrian data set. 
Yet the effect size of the difference was again small (Cramer’s V ≤ 0.14). 
Taken together, these analyses suggest that there is indeed a weak but significant selection bias towards men, well educated, professionally active and healthy individuals, and victims of winter sport accidents. 
We point this out in the limitation section of __Discussion__ of the revised manuscript.

To address the possible under- or overestimation (see: __Point 10__, __Reviewer 2__) of mental disorder symptoms in the study cohort, we compared frequency of traumatic events, manifest PTSD, symptoms of anxiety and depression in the study cohort with large survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017], as well as data of the German mental health monitoring [@Junker2023] and an Austrian general health microcensus study [@StatistikAustria2019a]. 
Frequency of traumatic events and manifest PTSD in our sample was lower or comparable with the estimated prevalence in the general German population reported by Hauffa and colleagues [@Hauffa2011] (__Supplementary Figure S4A__). 
Interestingly, frequency of manifest PTSD and frequency of the PCL-5 domain B, C, D, and E PTSD symptoms did not differ significantly from figures published by Mikutta et al. for Swiss mountain rescuers [@Mikutta2022] (__Supplementary Figure S4B__). 
Furthermore, frequency of 
anxious symptoms (`r lit_plots$depr_anx_stats %>% filter(disorder == 'anxiety', source == 'cohort') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`) assessed by the GAD-2 instrument was significantly higher than in the general German population within the time frame of the study survey 
(`r lit_plots$depr_anx_stats %>% filter(disorder == 'anxiety', source == 'desurv') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`). 
Depressive symptoms measured by the PHQ-2 tool were, in turn, much less prevalent 
in our cohort (`r lit_plots$depr_anx_stats %>% filter(disorder == 'depression', source == 'cohort') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`) than 
estimates for the general German population (`r lit_plots$depr_anx_stats %>% filter(disorder == 'depression', source == 'desurv') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`). 
Finally, frequency of moderate-to-severe symptoms of depression (PHQ-8 $\geq$ 10) was comparable in the study samples and the Austrian microcensus results (__Supplementary Figure S5AB__). 
These results let us reason that, despite the mountain accident, the overall mental health status of the study cohort was comparable to the general population of German-speaking countries and we elaborate on this in __Results/Characteristic of the study cohort__ and __Discussion__.

## Point 5

__Discussion__: The discussion is well structured: Probably it would help if you’d point out your own findings at the beginning of each paragraph.

## Response 5

Thank you for the suggestion. 
We summarize our most important findings in the first paragraph of __Discussion__. 
Where appropriate we have now also added additional summaries at the beginning of the Discussion's paragraphs.

## Point 6

Line 51: The think the study was performed at the University of Grenoble which is in France not Switzerland.

## Response 6

We are sorry for the confusion. 
This was corrected in the revised manuscript.

## Point 7

Throughout the discussion: Please find a clear terminology of the PTSD: sometimes you talk about manifest PTSD sometimes about PTSD symptoms. 
As you mostly do refer to symptoms of PTSD I’d stick with that.

## Response 7

Thank you for raising this important issue. 
We agree that the distinction between symptoms of PTSD and manifest PTSD was not completely coherent in the initial text. 
We define symptoms of PTSD as positivity for at least one domain of the PCL-5 tool. 
Manifest PTSD is considered for participants screened positive for all four domains of the PCL-5 instrument [@Bovin2016]. 
This distinction is described now in __Methods__ and __Supplementary Methods__.

Of note, while single PTSD symptoms affected nearly one of five study participants, manifest PTSD was discerned in four individuals only. 
We corrected the ambiguous passages and describe frequency of manifest PTSD solely in __Results/Characteristic of the study cohort__. 
We use the term 'symptoms of PTSD elsewhere in the revised text.

## Point 8

Figures: The figures a very clear and have a nice layout. Congrats.

## Response 8

Thank you very much.

\newpage

# Reviewer 2

## Point 9

This paper aims at characterizing mental health in persons after mountain sport accidents. 
This is an interesting question approached with a very transparent and accessibly documented analysis.

## Response 9

Thank you for the constructive feedback and appreciation of our work!

## Point 10

I would almost certainly expect there to be a strong self-selection bias in the obtained responses to your survey (i.e., those still dealing with the consequences of their accident), likely overestimating the prevalence of mental health-related impairment in these individuals. 
In addition, the supplementary table S4 shows that respondents had more severe injuries requiring hospitalization and surgery compared to non-respondents, also reflecting self-selection of particularly affected individuals among those who experienced a mountaineering accident. Does that change the implication(s) of your results?

## Response 10

We absolutely agree that there is a selection bias in this study. 
This was already pointed out in the limitations section of __Discussion__ of the initial submission. 
As described in more detail in __Response 4__ we took now additional measures to better characterize the potential selection bias. 
From comparisons of our sample with the general Austrian population (__Supplementary Table S4__) and the collective of mountain accident victims in Austria in 2023 (__Supplementary Figure S1__), we infer that the study cohort consisted of significantly more males, highly educated and healthy individuals and victims of winter sport accidents as expected for the general population. 
Effect size of those differences was weak (Cramer's V effect size statistic $\leq$ 0.1)
We elaborate on those findings in __Results/Characteristic of the study cohort__ and summarize these findings on the selection bias in the limitation part of __Discussion__. 
Of note, the selection bias may explain low frequency of traumatic events prior to the mountain accident and rare cases of manifest PTSD in our cohort, as e.g. men and tertiary school graduates are less likely to be exposed to traumatic events and develop PTSD in the lifetime [@Darves-Bornoz2008] (__Supplementary Figure S1__). 
We describe this possibility in __Discussion__ of the revised manuscript.

Next, we addressed the issue of under- or over-reporting of mental health symptoms in the study sample, which may be attributed to the selection bias of the sample. 
To this end, we compared frequency of traumatic events, manifest PTSD, anxious and depressive symptoms in the study cohort with estimates for the general population (__Supplementary Figure S4__). 
As outlined in __Response 4__ in more detail, frequency of traumatic events and manifest PTSD was at the lower tail of the reported prevalence and comparable with figures for the German population published by Hauffa and colleagues [@Hauffa2011]. 
Anxious symptoms measured with the GAD-2 instrument were more common in the study cohort than in the German population during the survey [@Junker2023]. 
In turn, depending on the tool and reference population, overall depressive symptoms and symptoms of moderate-to-severe depression were lower or comparable with the general German or Austrian population (__Supplementary Figure S5__). 
We discuss in __Results/Characteristic of the study cohort__ and in __Discussion__, that the overall frequency of mental disorder symptoms in our cohort is, in spite of the accident, comparable with the general population. 
It is hence possible that bias of the sample towards protective characteristic (education, male sex, economic status) compensates the negative effects of the accident and injury on mental health. 
Additionally, in light of the comparison of the mental disorder symptoms in the cohort and estimates for the general population, over-reporting of mental disorder symptoms in the study cohort is unlikely.

Finally, we investigated the impact of factors related to the selection bias of our sample, i.e. different income structure, more frequent hospitalization, surgery and more severe injury in non-responders or excluded survey participants than in the analysis cohort, on definition of the mental health clusters. 
This analysis was motivated the general concept of one-at-time sensitivity analysis [@Thabane2013]. 
In brief, we developed the mental health clusters with the PAM/cosine distance procedure for the genuine analysis cohort and the following:

* the analysis cohort appended with observations with incomplete psychometry data. The missing values of the mental health assessment battery were imputed with the 9-nearest neighbor algorithm [@Hastie2023].

* the analysis cohort without, respectively, participants with no, low, intermediate and high annual income

* the analysis cohort without participants with, respectively, mild, moderate and severe-to-critical injuries

* separately for hospitalized and non-hospitalized study participants

* referring to your issue of effects of the accident - survey time on mental health status, we developed the clusters for the analysis cohort without, respectively, accidents in 2018, 2019 and 2020

* referring to your issue of multiple mountain sport accidents, we also defined the mental health clusters for participants with and without prior mountain sport accidents

Due to the low number of participants subjected to surgery (n = `r get_percent(ptsd$dataset, 'surgery_done')$count['yes']`), we were not able to define meaningful clusters for this subset. 
Subsequently, we compared fractions of explained clustering variances and cluster sizes between the mental clusters of the genuine analysis cohort and the mental clusters defined for the extended or depleted data sets. 
As shown in __Supplementary Figure S10__, inclusion of the missing observations or removal of subsets of the analysis cohort had minimal effects on the clustering quality measured by the explained variance statistic (analysis cohort: V = `r summary(se_globals$clust_obj)$frac_var %>% signif(2)`, extended/depleted data sets: V = `r filter(se_summary$stats, split_factor %in% c('missing', 'household_income_class', 'prior_accident', 'hospitalization', 'injury_sev_strata'))$frac_var %>% range %>% signif(2) %>% paste(collapse = ' to ')`), or on the cluster size distribution. 
Hence, the effects of exclusion of participants with incomplete psychometric data, changes in the income structure, injury severity, hospitalization status, prior accidents or particular accident years had little effect on clustering of the study cohort by its mental health features.

Interestingly, participants excluded from the analysis due to incomplete psychometric data, were likely characterized by low scoring of post-traumatic growth, low resilience and quality of life and high scores of panic as compared with participants with complete psychometric data and were hence predominantly assigned to the PTS cluster (__Supplementary Figure S11__). 
This observation suggests that by exclusion of survey participants with incomplete mental health data, rates of poor mental health and life quality are rather underestimated in our analysis. 
We elaborate on this new data in __Results/Three clusters of mental health response after mountain sport accidents__ and __Discussion__.

## Point 11

Respondents completed the survey after about 3.5 years (median) after their accident. 
This seems like a long time, how certain are you that any mental health-related burden is still due to the accident? 
Did you screen for any events that might have happened in the meantime?

## Response 11

We agree that the accident - survey interval may be problematic. 
Our initial intention was to analyze persistent, long term alterations of mental health in victims of mountain sport accidents, i.e. at least six months after the accident - we state this point clearly in the final paragraph of __Introduction__ and __Methods/Analysis endpoints__.

During the exploratory data analysis, we have not found any effects of the accident-survey time on the psychometric scores and for effects of the accident year on the cluster assignment (__Review Figure R\@ref(fig:fig-time)__). 
Results of this analysis are available for an interested reader in the [GitHub code repository of the project](https://github.com/PiotrTymoszuk/mental_accident/tree/main/paper). 
Furthermore, we were not able to identify any substantial effects on accidents in 2018, 2019 and 2020 on development of the mental clusters in a sensitivity analysis described in __Response 10__ (__Supplementary Figure S10__). 
In light of those data, we are pretty confident that the time interval between the accident or medical treatment and the survey did not have significant effects on the mental health status or clustering results as a standalone confounder.

Still, as you suggest, we can not rule out that the survey results could have been influenced by events after the accident of interest. 
However, in case of the instruments used for quantification of PTSD symptoms, post-traumatic growth, and flashback frequency, our survey indicated that the questions refer to the mountain sport accident. 
For this reason, we assume that the PTS and PTG mental health clusters hallmarked by, respectively, PTSD symptoms and post-traumatic growth represent a long term reaction to the accident. 
We included this information in __Methods__ and __Supplementary Methods__ of the revised manuscript, and we point it also out in __Results/characteristic of the study cohort__ and describe in __Discussion__. 
The possibility that the investigated mental health features are not related to the accident, is stated in the limitation paragraph of __Discussion__.

```{r fig-time, fig.width = figur::convert(rev_figures$time, to = 'in')$w, fig.height = figur::convert(rev_figures$time, to = 'in')$h, fig.cap = 'Effects of ward admission-survey time and of accident year on the major psychometric readouts and the mental cluster assignment.'}

rev_figures$time$plot

```

__Review Figure R\@ref(fig:fig-time). Effects of ward admission-survey time and of accident year on the major psychometric readouts and the mental cluster assignment.__

_(A) Differences in the major scores of mental health and quality of life as a function of the ward admission - survey time (PTSD: PCL-5 score, sum of scores for the B, C, D, and E domains; post-traumatic growth: PTGI, sum of all dimensions; anxiety: GAD-7, depression: PHQ-9; resilient coping: RS13; quality of life: EUROHIS-QOL 8, arithmetic mean of the the domain scores). Statistical significance was determined by Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Score values are presented in violin plots. Single observations are visualized as points. Red diamonds with whiskers represent medians with interquartile ranges. Effect sizes and p values are displayed in the plot captions. Numbers of complete observations are displayed in the X axes._

_(B) Frequency of accidents in 2018, 2019 and 2020 was compared between the mental health clusters by $\chi^2$ test with Cramer's V effect size statistic. Percentages of accidents in particular years in the clusters are presented in a stack plot. The effect size and p value are shown in the plot caption. Numbers of observations in the clusters are presented in the X axis._

## Point 12

The addition of a control group (without a mountain sport accident) may be considered - could you point out how the PTSD symptom rates obtained in this study compare to those in the general population (page 13, lines 4-9?), taking into account the self-selection bias mentioned above? 
What does this mean for the interpretation of your results?

## Response 12

We agree, that inclusion of an adequately large control group without a mountain sport accident would enable us to assess if and how the observed mental health alterations are attributed to the accident. 
Unfortunately, this is not possible due to the time span of the revision.  
To address this problem at least in part we compared frequency of traumatic events, manifest PTSD, as well as symptoms of anxiety and depression in the study cohort with estimates for the general population. 
Please refer to __Responses 4__ and __10__, __Supplementary Table S6__, and __Supplementary Figures S1__ and  __S4__ for details and interpretation.

## Point 13

According to your statement on page 13 (lines 4-9), it seems that your sample was especially resilient compared to the general population. 
Wouldn't this perhaps imply a protective factor of (mountain) sports overall, despite accident risk?

## Response 13

Absolutely! 
We make this more clear now in __Discussion__ section to emphasize this point. 
To support this notion with hard figures, we compared average scores with samples from the general German [@Leppert2008] and Tyrolean/West Austrian population [@Chernova2021], and with resilience scores published by Mikutta and colleagues for a cohort of Swiss mountain rescue service workers [@Mikutta2022]. 
Supporting the common belief of positive effects of mountain activity on mental strength, resilience scoring of our cohort (despite the accident) and of the Swiss mountain rescue sample (despite high prevalence of traumatic events) was substantially higher than in the general population (__Supplementary Figure S5__). 
This again, may at least in part explain low frequency of manifest PTSD and PTSD symptoms in those two collectives (__Supplementary Figure S4B__). 
We elaborate on this new data in more detail in __Results/Characteristic of the study cohort__ and in __Discussion__.

## Point 14

Please describe more clearly which results were significant and which were not. 
Non-significant results should be clearly labelled as such and not paraphrased as "enriched in" or a "tendency". 
In lines 33-56 (page 10) and lines 4-16 (page 11), it was very difficult to discern which group differences were significant and which were not (see also minor points below).

## Response 14

We are sorry for this confusion. 
In the revised text, we abstained from the non-scientific term 'tendency' or 'enrichment' and point out the significant effects in the text by the adjective 'significant', p value and effect size.

## Point 15

What new insights does this study offer? 
Pre-existing mental disorders and persistent somatic symptoms are well-known risk factors for the development of psychiatric disorders. 
Indeed, a reliable model identifying at-risk individuals would have been interesting, but this study could not provide one.

## Response 15

Alone in Austria with 9 million inhabitants, there are over 900000 members of diverse alpine clubs [@VAVOE2024], which amounts to an enormous community of mountain sport enthusiasts, professional and free-time athletes. 
The dark side of popularity of mountain sports are accidents whose number exceeded `r table(lit_kurasi$data$sport_type$subset)['Austria'] %>% signif(2)` in 2023 in the nation-wide scale, according to the official counts by the safety board and police [@OsterreichischesKuratoriumfurAlpineSicherheit2024]. 
By comparison, this number corresponds approximately to the yearly sum incidence of prostate and breast cancer in Austria [@StatistikAustria2023a]. 
Despite these huge figures and the burden of mountain sport accidents on health system and economy of the alpine region, only a handful of papers has been published, which tackle single, isolated aspects of mental health of mountain sportspersons and victims of mountain sport accidents [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004]. 
Our study provides an integrated and 'holistic' characteristic of mental health after a mountain sport accident, by assessment of multiple dimensions of mental readouts with semi-supervised clustering and machine learning. 
This is the prime point of novelty of our work.  
Despite the positive effects of mountain sports on physical and mental health and the common perception of sportspersons as mentally strong individuals, our data suggest that a large subset of mountain accident victims, the PTS cluster, is prone to long-term mental health deterioration. 
This is another novel finding of out work. 
As pointed out by the reviewer, there are many reports on markers and diagnostic tools of post-traumatic stress [@Sandweiss2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021]. 
Yet, we were not able to establish a reliable set of predictors of mental health disorders after the mountain sport accident. 
This negative finding, which has not been reported so far, has profound clinical implications and underscores the need of widely accessible psychological support in management of this specific accidents. 
We stress there three novel findings of our work in the first paragraph of __Discussion__ and the conclusion part of the revised manuscript.

## Point 16

Why was residence in a German-speaking country required, when German-speaking proficiency was already an inclusion criterion?

## Response 16

We are grateful for this point. 
First of all, we spotted an error in the initial version of the manuscript. 
Instead of 'German-speaking country', the term 'German-speaking region' is appropriate, as our screening strategy encompassed also South Tyrol/Alto Adige province of Italy, which has a large community of German speakers. 
We apologize for this flaw and corrected it in the revised text.

These both study inclusion criteria had to do with the multilingual structure of the alpine region and the profile of patients of our clinical center. 
First, we genuinely designed the study survey in German. 
Yet, in some parts of the region of interest, other languages are widely spoken, like Italian and Ladin in South Tyrol/Alto Adige or French and Italian in Switzerland. 
Such Italian or French native speakers are not necessarily enough proficient in German to complete the survey. 
Second, tourists, especially in the winter season, make a large part of patients of the study center, but were not considered the target population of our project. 
Hence, with the combination of the language and residency criteria, we intended to sample the population of the alpine region residents that is likely to perform mountain sports on a regular basis.

## Point 17

Why do you think did 20% of respondents not complete the survey? 
Why did they send back an incomplete battery?

## Response 17

This is a good point, thank you! 
As a part of exploratory analysis of the study data set, we routinely investigate missingness of variables. 
As presented in __Reviewer Figure R\@ref(fig:fig-miss-mental)__, the highest percentage of missing observations among the mental health assessment battery was observed for the resilience, depression, panic, quality of life, and anxiety variables. 
Those tools were missing in all or almost all of the `r table(miss_mental$analysis_tbl$status)['excluded']` respondents excluded from the analysis due to incompleteness of the psychometric data. 
In turn, for the PCL-5 instrument measuring PTSD symptoms, the missingness rates were the lowest (< 50% of excluded respondents). 
We can hence only speculate, that the excluded survey participants found questions of the questionnaire concerning their fears, concerns and life satisfaction too private.

Additionally, from our experience with other similar studies, a drop-out rate of 20 - 30% is quite common in longer surveys or questionnaires covering difficult topics such as mental health.

```{r fig-miss-mental, fig.width = figur::convert(rev_figures$miss_mental, to = 'in')$w, fig.height = figur::convert(rev_figures$miss_mental, to = 'in')$h, fig.cap = 'Percentages of missing observations for the mental health assessment battery variables.'}

rev_figures$miss_mental$plot

```

__Reviewer Figure R\@ref(fig:fig-miss-mental). Percentages of missing observations for the mental health assessment battery variables.__

_Frequencies of missing observations per mental health assessment battery are expressed as percentages of the survey respondents with incomplete psychometric data and hence excluded from the analysis.__

## Point 18

You report that 38% of respondents already had another mountain sport accident in the past. 
This number seems rather high at first sight- is there any data of the typical accident prevalence in this sport available? 
Are these high risk-taking individuals?

## Response 18

Thank you for raising this important point. 
Unfortunately, we can not answer it with published data. 
A query in the data base of mountain accidents in Austria (Austrian Board of Mountain Safety) did not help either, because the records are fully anonymized, and assignment of multiple accidents to the same victim is not possible.

Based on the comparison of the accident sports in the study cohort and the nation-wide collective of accident victims in 2023 (__Supplementary Figure S1__), we can conclude that the global difference in sport frequency distribution was significant but weak ($\chi^2$ test with Cramer's V effect size statistic: `r filter(lit_kurasi$test, variable == 'sport_detail')$plot_cap`). 
In particular, we could not observe higher percentages of sports generally perceived as risky such as ski touring, ice climbing, climbing, mountaineering or paragliding in the analyzed cohort as compared with the Austrian data set. 
This let us assume, that there are no large subsets of the cohort exposed to particularly high risk of injury.

Additionally, we did not ask about severity or time frame of mountain sport accidents in the past, so it is hard to judge if an how they affected mental health at the time point of the survey. 
As inferred from results of the sensitivity analysis presented in __Supplementary Figure S10__, prior mountain sport accidents had little impact on clustering of the study participants by their mental health features. 
They frequency also did not differ between the mental health clusters (__Supplementary Figure S12__). 
For these reasons it is unlikely that mountain accidents in the past had an substantial effect on the outcome of our study. 
We touch upon this point in __Results/Three clusters of mental health response after mountain sport accidents__ and __Results/Demographic, socioeconomic and clinical characteristic of the mental health clusters__.

## Point 19

Page 10, lines 35-37 "substantially enriched in… […]; these effects were not significant" and page 10, lines 39-41 (frequency of higher traumatic events): I would not point out any apparent differences that are not statistically significant. 
Page 10, line 43 - 48: Do not describe mean differences as "enriched in" if they are not statistically significant, as this could confuse the reader and possibly convey false information on group differences.

## Response 19

We agree and stick strictly to the significant/not significant distinction in the revised text.

## Point 20

Will the data be made openly accessible? This would be desirable.

## Response 20

Unfortunately, the initial study protocol and patient's consent form did not include an open data strategy, and the data set still contains information which may enable identification of an individual such as localization of injury, accident and medical treatment date. 
For this reason, our ethics committee did not agree on publication of the data set. 
Its anonymous parts will be made available at request.

\newpage

# References

### 2024-02-08 paper/markdown/manuscript.Rmd
P2P reply, supplementary material
Changes in supplementary figures and tables required by the reviewer's comments

@@
 hospital admission at least 6 months prior to the start of the study, 
-age $\geq$ 18 years, residence in a German-speaking country, and 
-proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail. 
+age $\geq$ 18 years, 
+residence in a <span custom-style = 'revision'>German-speaking regions (Germany, Austria, Switzerland and South Tyrol/Alto Adige in Italy)</span>, and 
+proficiency in German (n = `r nrow(ptsd$cleared)`) 
+were invited to participate in the online study survey via conventional mail. 
 Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey. 
@@
 anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
-depression (PHQ-9: 9-item patient health questionnaire for depressive symptoms) [@Grafe2004], 
+depression (PHQ-9, <span custom-style = 'revision'>PHQ-2 and PHQ-8: 9-, 8- and 2-item patient health questionnaires</span> for depressive symptoms) [@Grafe2004], 
 panic (PHQ-panic module) [@Grafe2004], 
@@
 and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
+<span custom-style = 'revision'>Items of the PTGI and PTSD instruments were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').</span>
 The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).
@@
 Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
-Significant PTSD symptoms were assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].
+Significant PTSD symptoms were assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains. 
+<span custom-style = 'revision'>Manifest PTSD was considered for participants positive for all four PCL-5 domains [@Bovin2016].</span>
 
 Traumatic events prior to the accident were assessed with the DIA-X tool [@Maercker2005]. 
-Flashback frequency was surveyed as none, > 1/year and > 1/month. 
+<span custom-style = 'revision'>Flashbacks of the accident of interest during mountain sport activity were surveyed as</span> none, > 1/year and > 1/month. 
 Smoking was surveyed as a single yes/no question. 
@@
 As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and fewer surgery diagnoses (__Supplementary Table S4__).
-The survey participants excluded from the analysis due to missingness of psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__).
+The survey participants excluded from the analysis due to incomplete psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__).
 Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).

### 2024-02-08 paper/markdown/supplementary_material.Rmd
P2P reply, supplementary material
Changes in supplementary figures and tables required by the reviewer's comments

@@
 Data analysis and transformation was accomplished with R version 4.2.3. 
-The study dataset was imported from an SPSS file with the _foreign_ package [@RCoreTeam]. 
+The study data set was imported from an SPSS file with the _foreign_ package [@RCoreTeam]. 
 Tabular data were handled with the _tidyverse_ package bundle [@Wickham2019], and the packages _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
 Text variables were handled with _stringi_ [@Gagolewski2021]. 
+Imputation of missing values of psychometric variables was done with the package _impute_ [@Hastie2023].
 
@@
 The psychometric battery (__Supplementary Table S\@ref(tab:tab-mental-battery)__) consisted of German versions of assessment tools for 
-anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
-depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
+anxiety (GAD-7 and GAD-2: 7- and 2-item general anxiety disorder scale) [@Spitzer2006; @Junker2023], 
+depression (PHQ: patient health questionnaire, PHQ-9, PHQ-2 and PHQ-8) [@Lowe2002; @Grafe2004; @Kroenke2002a; @Junker2023], 
 panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
@@
 post-traumatic stress disorder (PCL-5: PTSD checklist for DSM-5) [@Bovin2016]. 
+In the study survey, items of the PTGI and PTSD instruments were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').
 
@@
 Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008]. 
+For a comparison of frequency of anxious and depressive symptoms in the study cohort with the estimates for the German population [@Junker2023], presence of anxious and depressive symptoms was considered for GAD-2 $\geq$ 3 and PHQ-2 $\geq$ 3 points, respectively.
+For a comparison of frequency and severity of depressive symptoms in the study cohort with microcensus data for the general Austrian population, mild depressive symptoms were defined as 5 - 9, moderate symptoms as 10 - 14, moderately severe symptoms as 15 - 19, and severe symptoms as $\geq$ 20 points of the PHQ-8 score [@Kroenke2002a].
 
@@
 Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
-Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].
+Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains. 
+Manifest PTSD was considered for participants positive for all four PCL-5 domains [@Bovin2016].
 
@@
 prior mountain sport accidents, 
-flashbacks during mountain sport activity, 
+flashbacks of the surveyed accident during mountain sport activity, 
 confusion during mountain sport activity, 
@@
 
+Information on  
+genders ($\geq$ 18 years, in 2023) [@StatistikAustria2023], 
+age ($\geq$ 18 years, in 2023) [@StatistikAustria2023], 
+education (15 - 64 years old, in 2020) [@StatistikAustria2020]
+and employment structure ($\geq$ 15 years, in 2021) [@StatistikAustria2021], 
+smoking status ($\geq$ 15 years, in 2019) [@StatistikAustria2019], 
+and self-reported chronic physical or mental illness ($\geq$ 15 years, in 2019) [@StatistikAustria2019a] 
+for the Austrian population were extracted from publications of Statistik Austria, the national statistical bureau.
+
+Counts of mountain accident victims in 2023 in Austria stratified by mountain sport type, age class and month of the accident were extracted from the annual report of the Austrian Board of Mountain Safety [@OsterreichischesKuratoriumfurAlpineSicherheit2024].
+Of note, we were not able to consistently eliminate deaths (n = 266), accidents with tourists from non-German-speaking countries, non-adult victims (nation-wide, $\geq$ 20 years old: n = `r 496 + 2074`) and non-sportive accidents (n = 648) from this published nation-wide data set.
+Yet these subsets of mountain accident victims constituted a minority of all nation-wide mountain accidents in 2023 (n = 13681).
+
+Lifetime estimates of prevalence of traumatic events and PTSD in the general population were obtained from representative survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017].
+Frequency of traumatic events, manifest PTSD and PTSD symptoms subsumed under the B, C, D, and E domains of the PCL-5 instrument in Swiss mountain rescuers were extracted from the report by Mikutta et al.
+[@Mikutta2022].
+
+Monthly estimates of population frequency of symptoms of anxiety and depression in the general German population during the study survey (`r ptsd$dataset$survey_date %>% range %>% paste(collapse = ' to ')`) were derived from the [data repository of the German mental health monitoring](https://github.com/robert-koch-institut/Hochfrequente_Mental_Health_Surveillance) [@Junker2023].
+Data on frequency and severity of depressive symptoms measured with the PHQ-8 instrument [@Kroenke2002a] in a nation-wide Austrian microcensus study were obtained from a publication by Statistik Austria [@StatistikAustria2019a].
+
+Mean scores of resilient comping with their standard deviations (SD) obtained with the RS-13 tool were extracted from reports with samples of the German [@Leppert2008] and Tyrolean [@Chernova2021] populations, and from a cohort of Swiss mountain rescuers [@Mikutta2022].
+
 ## Consistency of psychometric tools and power analysis
@@
 
-To find the optimal size of a training subset of the study cohort, random subsamples of the study dataset of varying sizes were investigated (`r paste(pow$sizes, collapse = ', ')` observations, 50 random draws per subsample size).
+To find the optimal size of a training subset of the study cohort, random subsamples of the study data set of varying sizes were investigated (`r paste(pow$sizes, collapse = ', ')` observations, 50 random draws per subsample size).
 For each random subsample, clustering tendency was assessed by Hopkins statistic (function `get_clust_tendency()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Hopkins1954]. 
@@
 (`r signif(pca$clust_tendency$data$hopkins_stat, 2)`). 
-For these reasons, 3/4 of the entire dataset or n = 
+For these reasons, 3/4 of the entire data set or n = 
 `r count(ptsd$dataset, partition)$n[1]` was considered as the adequate size of the training subset of the study cohort for a reproducible clustering analysis.
@@
 The study participants were assigned to the training and test subsets with the 3:1 size ratio (__Figure 1__). 
-To this end, `r nrow(part$dists)` random splits of the datasets were generated. 
+To this end, `r nrow(part$dists)` random splits of the data sets were generated. 
 The subset assignment scheme was chosen with the possibly smallest differences in sociodemographic, medical history, clinical and accident- and injury-related variables assessed by Gower distance between the training and test subsets (function `dist()`, package _proxy_) [@Meyer2022].
@@
 Significant (p < 0.05) differences between the included/excluded participants are presented in __Supplementary Table S\@ref(tab:tab-no-response)__ and __S\@ref(tab:tab-missing)__.
+
+## Comparison of socioeconomic, accident and mental health features with population estimates and literature reports
+
+Differences in age, gender, education, professional activity, and frequency of chronic physical and mental disorders between the study cohort and the Austrian population [@StatistikAustria2023; @StatistikAustria2023; @StatistikAustria2020; @StatistikAustria2021; @StatistikAustria2019] were investigated by Mann-Whitney U test with r effect size statistic (age as a numeric variable) or $\chi^2$ test with Cramer's V effect size statistic (remaining categorical variables).
+Global differences in distribution of mountain sport types, age and accident months between the analyzed cohort and the nation-wide collective of Austrian mountain accident victims [@OsterreichischesKuratoriumfurAlpineSicherheit2024] were compared with $\chi^2$ test with Cramer's V effect size statistic.
+In the comparison of sport types, nation-wide counts of non-sportive mountain accidents (e.g. during forestry or hunting) were excluded. 
+In the comparison of age distributions, only individuals older than 20 years were included. 
+Post-hoc comparisons of the study cohort and nation-wide Austrian counts were done with Holm-corrected $\chi^2$ test. 
+The statistical hypothesis testing was done with the function `compare_variables()` from the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package. 
+The comparison results are presented in __Supplementary Table S\@ref(tab:tab-aut-population)__ and __Supplementary Figure S\@ref(fig:fig-accident)__.
+
+Frequency of traumatic events, manifest PTSD and PTSD symptoms in the current cohort was compared with original reports providing estimates of frequency of traumatic events and PTSD in the general population [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017] and in a cohort of Swiss mountain rescuers [@Mikutta2022].
+Frequencies of symptoms of anxiety and depression measured with the GAD-2 and PHQ-2 instruments, respectively, were compared with estimates for the general German population gauged by the German mental health monitoring during the study survey [@Junker2023].
+Frequency and severity of depressive symptoms measured by the GAD-8 tools in the study cohort [@Kroenke2002a] were compared with the Austrian microcensus estimates. 
+In those analyses, bootstrap with B = 1000 resamples was used for computation of the bias-corrected and accelerated (BCA) 95% confidence intervals (function `bmap()`, package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat)).
+Significant differences were assumed for non-overlapping 95% confidence intervals of the mental symptom frequency estimates.
 
@@
 Differences in frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical, accident- and recovery-related factors between the mental health clusters were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic and by Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Supplementary Table S\@ref(tab:tab-clust-symptoms)__ and __S\@ref(tab:tab-clust-bcg)__).
+
+## Assessment of effects of missing observations, annual income, prior mountain sport accidents, injury severity, hospitalization, and accident year on mental health cluster definition
+
+To investigate effects observations excluded from the analysis due to an incomplete psychometric data set on definition of the mental health clusters, the excluded observations with less than 80% of missing psychometric entries were subjected to imputation with the 9-nearest neighbor algorithm implemented by the `impute.knn()` function from the R package _impute_ [@Hastie2023].
+Subsequently, the mental health clusters were developed for the analysis cohort and for the analysis cohort appended with the imputed observations. 
+Analogically, to explore effects of prior mountain sport accidents, the mental health clusters were developed for the entire cohort, the individuals having experienced a mountain accident in the past, and participants without prior mountain sport accidents.
+To evaluate effects of annual household income, the mental clusters were developed for the analysis cohort and the analysis cohort without, respectively, participants with no income, yearly income < 30000 Euro, yearly income ranging from 30000 to 45000 Euro and yearly income > 45000 Euro.
+To assess the impact of injury severity, the mental clusters were developed for the analysis cohort and the analysis cohort depleted of, respectively, individuals with mild (AIS 1), moderate (AIS 2), and severe-to-critical injuries (AIS $\geq$ 3).
+To investigate the effect of hospitalization, the mental health clusters were developed separately for the analysis cohort, non-hospitalized and hospitalized participants. 
+To assess the impact of accident year, the mental clusters were developed for the analysis cohort without, respectively, accidents in 2018, 2019 and 2022. 
+In each case, the unsupervised PAM/cosine distance clustering algorithm was used, i.e. the procedure employed for definition of the genuine mental health clusters. 
+The effects of extending or depletion of the data set on the quality of the clustering structure was evaluated by comparing fractions of the explained clustering variance between the clustering solutions for the particular data set and for the analysis cohort [@Thabane2013].
+Additionally, distribution of sizes of the neutral, PTG and PTS clusters expressed as percentage of all clustered observations was compared (__Supplementary Figure S\@ref(fig:fig-cluster-se)__ and __S\@ref(fig:fig-cluster-miss)__).
 
@@
 
+```{r tab-aut-population, tab.cap = 'Comparison of sociodemographic features of the study cohort and estimates for the general Austrian population.
+Numeric variables are presented as medians with interquartile ranges (IQR).
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+flextable::flextable(suppl_tables$aut_population) %>% 
+  set_widths(c(3.5, rep(5, 2), 2.8, 2.8)) %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic, categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
+           ref_symbols = 'a', 
+           part = 'header') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
 ```{r tab-train-test, tab.cap = 'Significant differences between the training and test subset of the study cohort.
@@
 flextable::flextable(suppl_tables$clust_symptoms) %>% 
-  set_widths(c(4, rep(4.3, 3), 2.5, 2.3)) %>% 
+  set_widths(c(4, rep(4, 3), 2.5, 2.3)) %>% 
   footnote(1, 1, 
@@
 
+```{r fig-accident, fig.width = figur::convert(suppl_figures$accident, to = 'in')$w, fig.height = figur::convert(suppl_figures$accident, to = 'in')$h, fig.cap = 'Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.'}
+
+suppl_figures$accident$plot
+
+```
+
+__Supplementary Figure \@ref(fig:fig-accident). Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.__
+
+_Mountain sport types (A), accident months (B) and age class distribution (C) were compared between the study cohort and victims of mountain sport accidents in Austria in 2023.
+Frequencies for the nation-wide data set of mountain accidents were extracted from the annual report of the Austrian Board of Mountain Safety.
+In the comparison of sport types, non-sportive mountain accidents (e.g.
+during hunting or forestry) were excluded and sport types in the study cohort re-classified according to the scheme of the Austrian Board of Mountain Safety.
+In the comparison of the age class distribution, only individuals older than 20 years were included.
+Statistical significance of the global differences was determined by $\chi^2$ test with Cramer's V effect size statistic.
+Pairwise differences for particular sport types, months and age classes were assessed by $\chi^2$ test corrected for multiple testing with the Holm method.
+Percentages of categories within the complete observation sets are presented as bar plots.
+Bars are labeled with the corresponding percent values.
+Total observation numbers are displayed in the plot legends.
+Global effect sizes and p values are presented in the plot captions.
+Significant results of post-hoc tests are highlighted with bold X axis labels._
+
+_alpine skiing: alpine skiing and snowboarding in a ski resort and lift accidents; ski touring: ski touring and freeride skiing; air: air sports and paragliding; water: water sports._
+
+\newpage
+
 ```{r fig-parts-ptsd, fig.width = figur::convert(suppl_figures$parts_ptsd, to = 'in')$w, fig.height = figur::convert(suppl_figures$parts_ptsd, to = 'in')$h, fig.cap = 'Injured body regions and symptoms of PTSD in the study cohort.'}
@@
 The number of complete observations is indicated in the plot caption._
+
+\newpage
+
+```{r fig-ptsd-literature, fig.width = figur::convert(suppl_figures$ptsd_literature, to = 'in')$w, fig.height = figur::convert(suppl_figures$ptsd_literature, to = 'in')$h, fig.cap = 'Comparison of frequency of traumatic events, manifest PTSD, and symptoms of  PTSD in the study cohort and literature.'}
+
+suppl_figures$ptsd_literature$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-ptsd-literature). Comparison of frequency of traumatic events, manifest PTSD, and symptoms of  PTSD in the study cohort and literature.__
+
+_Frequency estimates of traumatic events, manifest PTSD and PTSD symptoms with bias-corrected and accelerated (BCA) 95% confidence computed by bootstrap are depicted in Forest plots._
+
+_(A) Comparison of frequencies of traumatic events and manifest PTSD in the study cohort with estimates of population prevalence of lifetime traumatic events and manifest PTSD in published surveys.
+Tools used for identification of traumatic events and PTSD symptoms are indicated in the Y axis (CIDI: composite international diagnostic interview, DSM-5: diagnostic and statistical manual of mental disorders, 5^th^ edition, PDS: post-traumatic stress diagnostic scale)._
+
+_(B) Comparison of frequencies of traumatic events, manifest PTSD and PTSD symptoms in the study cohort and the cohort of Swiss mountain rescuers reported by Mikutta and colleagues.
+PTSD symptoms were assessed by the PCL-5 tool and are subsumed under the B, C, B, and E domains.
+Manifest PTSD is defined as positivity for all domains of the PCL-5 instrument._
+
+\newpage
+
+```{r fig-mental-literature, fig.width = figur::convert(suppl_figures$mental_literature, to = 'in')$w, fig.height = figur::convert(suppl_figures$mental_literature, to = 'in')$h, fig.cap = 'Comparison of frequency of anxious and depressive symptoms and of resilience scoring in the study cohort and literature.'}
+
+suppl_figures$mental_literature$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-mental-literature). Comparison of frequency of anxious and depressive symptoms and of resilience scoring in the study cohort and literature.__
+
+_Frequency estimates of mental disorder symptoms with bias-corrected and accelerated (BCA) 95% confidence computed by bootstrap are depicted in Forest plots.
+Mean resilience scores obtained with the RS13 tool with standard deviations (SD) are shown in a Forest plot._
+
+_(A) Comparison of frequency of anxious and depressive symptoms measured by the GAD-2 and PHQ-2 instruments in the study cohort with frequency of anxious and depressive symptoms obtained for the general German population by monthly mental health monitoring._
+
+_(B) Comparison of frequency and severity of depressive symptoms measured by the PHQ-8 tool in the study cohort with 2019 estimates for the general Austrian population._
+
+_(C) Comparison of resilient comping scores in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta at al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.)_
 
@@
 
-_(B) Cosine cross-distances between the mental health clusters in the training and test subsets of the study dataset visualized as a heat map.
+_(B) Cosine cross-distances between the mental health clusters in the training and test subsets of the study data set visualized as a heat map.
 Average cross-distances with 95% percentile range are indicated in the tiles._
@@
 PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
+
+\newpage
+
+```{r fig-cluster-se, fig.width = figur::convert(suppl_figures$cluster_se, to = 'in')$w, fig.height = figur::convert(suppl_figures$cluster_se, to = 'in')$h, fig.cap = 'Effects of incomplete psychometry data, annual income, prior mountain sport accidents, injury severity, and hospitalization on development of the mental health clusters.'}
+
+suppl_figures$cluster_se$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-cluster-se).
+Effects of incomplete psychometry data, annual income, prior mountain sport accidents, injury severity, and hospitalization on development of the mental health clusters.__
+
+_The mental health clusters were developed by PAM/cosine distance clustering for the analyzed study cohort, for the analyzed study cohort appended with participants with missing and imputed psychometric variables, participants with and without prior mountain sport accidents, hospitalized and non-hospitalized participants, the analysis cohort without each of no, low, intermediate and high annual income, the analysis cohort without each of 2018, 2019 and 2020 accidents, as well as for the analysis cohort with exclusion of participants with minor (abbreviated injury scale [AIS] 1), moderate (AIS2) and severe-to-critical injuries (AIS $\geq$ 3).
+Fractions of explained clustering variance (ratio of total between-cluster sum of squares to total sum of squares) were compared between the mental health cluster assignment of the entire cohort and the mental clusters of collectives described above._
+
+_(A) Percent differences in fractions of explained variance between the mental health clusters of the analysis cohort and the mental health clusters in the analysis cohort appended with or depleted of the subsets indicated in the Y axis.
+Total numbers of observations are displayed in the Y axis._
+
+_(B) Distribution of the mental health clusters sizes expressed as percentages of all observations presented in a stack plot. Total observation numbers are displayed in the Y axis._
+
+\newpage
+
+```{r fig-cluster-miss, fig.width = figur::convert(suppl_figures$cluster_miss, to = 'in')$w, fig.height = figur::convert(suppl_figures$cluster_miss, to = 'in')$h, fig.cap = 'Mental health cluster assignment of participants excluded from the analysis due incomplete psychometry data.'}
+
+suppl_figures$cluster_miss$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-cluster-miss). Mental health cluster assignment of participants excluded from the analysis due incomplete psychometry data.__
+
+_Incomplete psychometry data were imputed for participants with < 80% missing answers in the mental health battery by a 9-nearest neighbors regressor.
+The mental health clusters were developed for separately the analysis cohort (A) and the analysis cohort appended with the imputed psychometry data (B).
+Normalized levels of the clustering variables (Z-scores) in the mental health clusters are visualized as heat maps.
+Assignment of the imputed observations to the mental health clusters is depicted in the color bar below the heat map in (B).
+Numbers of observations in the clusters are displayed in the plot captions._
 

### 2024-02-28 paper/markdown/review_response.Rmd
Revised figures, tables, manuscript and supplement

@@
 Since the reviewers requested more information and literature evidence, we also deleted few less important references from the initial submission to keep the bibliography within reasonable limits. 
+We also introduced three major changes to the manuscript text and data independently of the reviewers' requests:
+
+1.
+We re-classified accident sport types according to the official scheme of the Austrian Board of Mountain Safety [@OsterreichischesKuratoriumfurAlpineSicherheit2024].
+This allowed us to compare the profile of accident sports in our cohort with the national registry data.
+Please note, that the re-classification of the sport types requires us to re-train machine learning models of the mental health cluster assignment.
+For this reason, minor differences in performance of the models were observed as compared with the initial submission (e.g.
+in __Figure 7__).
+Still, they did not affected the general conclusions.
+
+2.
+We identified clinically relevant symptoms of anxiety and depression with the uniform $\geq$ 10 point cutoff of, respectively, the GAD-7 and PHQ-9 tools in accordance with the work by Humer et al.
+This change allowed us for a direct comparison of frequency of clinically relevant anxious and depressive symptoms in our manuscript and the estimates for the general Austrian population at the time of our study survey [@Humer2022].
+
+3.
+We provided additional numeric measures of quality of the mental clusters, misclassification rate [@Rousseeuw1987] and neighborhood preservation [@Venna2001], which were used by us to argument for the excellent reproducibility of the mental clusters (__Supplementary Figure S8__).
+
 Additionally, some minor copyediting was done. 
@@
 We agree that the first sentence of __Abstract__ does not specifically relate to the subject of our research, which concerns primarily negative psycho-pathological burden of mountain sport accidents.
-We changed this sentence of __Abstract__ attempting to accommodate the reviewer´s concern but did not completely remove it because we feel that especially “mountain sports” are generally believed to bolster mental health.
+We changed this sentence of __Abstract__ attempting to accommodate the reviewer's concern but did not completely remove it because we feel that especially “mountain sports” are generally believed to bolster mental health.
 Additionally, individuals performing mountain sports are often portrayed as resilient personalities protected from mental disorders, which we can in part support with our data (__Supplementary Figures S4__ and __S5__).
@@
 classification of participants by their mental health measures followed by a search for predictors of mental health classes/clusters.
+We also re-wrote __Methods/Analysis endpoints accordingly__.
 
@@
 
-To address the possible under- or overestimation (see: __Point 10__, __Reviewer 2__) of mental disorder symptoms in the study cohort, we compared frequency of traumatic events, manifest PTSD, symptoms of anxiety and depression in the study cohort with large survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017], as well as data of the German mental health monitoring [@Junker2023] and an Austrian general health microcensus study [@StatistikAustria2019a].
+To address the possible under- or overestimation (see: __Point 10__, __Reviewer 2__) of mental disorder symptoms in the study cohort, we compared frequency of traumatic events, manifest PTSD, symptoms of anxiety and depression in the study cohort with large survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017; @Humer2022], as well as an Austrian 2019 general health microcensus study [@StatistikAustria2019a].
 Frequency of traumatic events and manifest PTSD in our sample was lower or comparable with the estimated prevalence in the general German population reported by Hauffa and colleagues [@Hauffa2011] (__Supplementary Figure S4A__).
@@
 for Swiss mountain rescuers [@Mikutta2022] (__Supplementary Figure S4B__).
-Furthermore, frequency of 
-anxious symptoms (`r lit_plots$depr_anx_stats %>% filter(disorder == 'anxiety', source == 'cohort') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`) assessed by the GAD-2 instrument was significantly higher than in the general German population within the time frame of the study survey
-(`r lit_plots$depr_anx_stats %>% filter(disorder == 'anxiety', source == 'desurv') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`).
-Depressive symptoms measured by the PHQ-2 tool were, in turn, much less prevalent 
-in our cohort (`r lit_plots$depr_anx_stats %>% filter(disorder == 'depression', source == 'cohort') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`) than
-estimates for the general German population (`r lit_plots$depr_anx_stats %>% filter(disorder == 'depression', source == 'desurv') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`).
-Finally, frequency of moderate-to-severe symptoms of depression (PHQ-8 $\geq$ 10) was comparable in the study samples and the Austrian microcensus results (__Supplementary Figure S5AB__). 
-These results let us reason that, despite the mountain accident, the overall mental health status of the study cohort was comparable to the general population of German-speaking countries and we elaborate on this in __Results/Characteristic of the study cohort__ and __Discussion__.
+Furthermore, frequency of clinically relevant anxious (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'anxiety')`) and depressive symptoms (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'depression')`) assessed, respectively by the GAD-7 and PHQ-9 instruments, was significantly lower in the study cohort as compared with the 2022 estimates for the general Austrian population reported by Humer et al.
+(anxiety: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'anxiety')`, depression: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'depression')`, __Supplementary Figure S5A__) [@Humer2022].
+These results let us reason that, despite the mountain accident, the overall mental health status of the study cohort was comparable to or better than the general population of German-speaking countries and we elaborate on this in __Results/Characteristic of the study cohort__ and __Discussion__.
 
@@
 Of note, while single PTSD symptoms affected nearly one of five study participants, manifest PTSD was discerned in four individuals only. 
-We corrected the ambiguous passages and describe frequency of manifest PTSD solely in __Results/Characteristic of the study cohort__. 
-We use the term 'symptoms of PTSD elsewhere in the revised text.
+In the revised text, we defined this distinction in __Methods__ and __Supplementary Methods__ and use consistently the terms 'manifest PTSD' and 'symptoms of PTSD' throughout the manuscript.
 
@@
 As described in more detail in __Response 4__ we took now additional measures to better characterize the potential selection bias. 
-From comparisons of our sample with the general Austrian population (__Supplementary Table S4__) and the collective of mountain accident victims in Austria in 2023 (__Supplementary Figure S1__), we infer that the study cohort consisted of significantly more males, highly educated and healthy individuals and victims of winter sport accidents as expected for the general population.
-Effect size of those differences was weak (Cramer's V effect size statistic $\leq$ 0.1)
+From comparisons of our sample with the general Austrian population (__Supplementary Table S4__) and the collective of mountain accident victims in Austria in 2023 (__Supplementary Figure S1__), we infer that the study cohort consisted of significantly more males, highly educated and healthy individuals, and victims of winter sport accidents that expected for the general population.
+Effect size of those differences was weak (Cramer's V effect size statistic $\leq$ 0.3)
 We elaborate on those findings in __Results/Characteristic of the study cohort__ and summarize these findings on the selection bias in the limitation part of __Discussion__. 
@@
 As outlined in __Response 4__ in more detail, frequency of traumatic events and manifest PTSD was at the lower tail of the reported prevalence and comparable with figures for the German population published by Hauffa and colleagues [@Hauffa2011].
-Anxious symptoms measured with the GAD-2 instrument were more common in the study cohort than in the German population during the survey [@Junker2023]. 
-In turn, depending on the tool and reference population, overall depressive symptoms and symptoms of moderate-to-severe depression were lower or comparable with the general German or Austrian population (__Supplementary Figure S5__).
-We discuss in __Results/Characteristic of the study cohort__ and in __Discussion__, that the overall frequency of mental disorder symptoms in our cohort is, in spite of the accident, comparable with the general population.
+Anxious symptoms measured with the GAD-7 instrument were less common in the study cohort than in the Austrian population during the survey [@Humer2022]. 
+In turn, depending on the tool and time point of survey, overall depressive symptoms and symptoms of moderate-to-severe depression were lower or comparable with the general Austrian population (__Supplementary Figure S5__).
+We discuss in __Results/Characteristic of the study cohort__ and in __Discussion__, that the overall frequency of mental disorder symptoms in our cohort is, in spite of the accident, comparable with or event lower than in the general population.
 It is hence possible that bias of the sample towards protective characteristic (education, male sex, economic status) compensates the negative effects of the accident and injury on mental health. 
@@
 
-_(A) Differences in the major scores of mental health and quality of life as a function of the ward admission - survey time (PTSD: PCL-5 score, sum of scores for the B, C, D, and E domains; post-traumatic growth: PTGI, sum of all dimensions; anxiety: GAD-7, depression: PHQ-9; resilient coping: RS13; quality of life: EUROHIS-QOL 8, arithmetic mean of the the domain scores).
+_(A) Differences in the major scores of mental health and quality of life as a function of the ward admission - survey time (PTSD: PCL-5 score, sum of scores for the B, C, D, and E domains; post-traumatic growth: PTGI, sum of all dimensions; anxiety: GAD-7, depression: PHQ-9; resilient coping: RS13; quality of life: EUROHIS-QOL 8, arithmetic mean of the domain scores).
 Statistical significance was determined by Kruskal-Wallis test with $\eta^2$ effect size statistic.
@@
 We make this more clear now in __Discussion__ section to emphasize this point. 
-To support this notion with hard figures, we compared average scores with samples from the general German [@Leppert2008] and Tyrolean/West Austrian population [@Chernova2021], and with resilience scores published by Mikutta and colleagues for a cohort of Swiss mountain rescue service workers [@Mikutta2022].
+To support this notion with hard figures, we compared average scores with samples from the general German [@Leppert2008] and Tyrolean/Austrian population [@Chernova2021], and with resilience scores published by Mikutta and colleagues for a cohort of Swiss mountain rescue service workers [@Mikutta2022].
 Supporting the common belief of positive effects of mountain activity on mental strength, resilience scoring of our cohort (despite the accident) and of the Swiss mountain rescue sample (despite high prevalence of traumatic events) was substantially higher than in the general population (__Supplementary Figure S5__).
-This again, may at least in part explain low frequency of manifest PTSD and PTSD symptoms in those two collectives (__Supplementary Figure S4B__). 
+This may at least in part explain low frequency of manifest PTSD and PTSD symptoms in those two collectives (__Supplementary Figure S4B__). 
 We elaborate on this new data in more detail in __Results/Characteristic of the study cohort__ and in __Discussion__.
@@
 Alone in Austria with 9 million inhabitants, there are over 900000 members of diverse alpine clubs [@VAVOE2024], which amounts to an enormous community of mountain sport enthusiasts, professional and free-time athletes.
-The dark side of popularity of mountain sports are accidents whose number exceeded `r table(lit_kurasi$data$sport_type$subset)['Austria'] %>% signif(2)` in 2023 in the nation-wide scale, according to the official counts by the safety board and police [@OsterreichischesKuratoriumfurAlpineSicherheit2024].
-By comparison, this number corresponds approximately to the yearly sum incidence of prostate and breast cancer in Austria [@StatistikAustria2023a]. 
+The dark side of popularity of mountain sports are accidents whose number exceeded `r table(lit_kurasi$data$sport_detail$subset)['Austria'] %>% signif(2)` in 2023 in the nation-wide scale, according to the official counts by the mountain safety board [@OsterreichischesKuratoriumfurAlpineSicherheit2024].
+By comparison, this number corresponds approximately to the cumulative yearly incidence of prostate and breast cancer in Austria [@StatistikAustria2023a]. 
 Despite these huge figures and the burden of mountain sport accidents on health system and economy of the alpine region, only a handful of papers has been published, which tackle single, isolated aspects of mental health of mountain sportspersons and victims of mountain sport accidents [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004].
@@
 In turn, for the PCL-5 instrument measuring PTSD symptoms, the missingness rates were the lowest (< 50% of excluded respondents). 
-We can hence only speculate, that the excluded survey participants found questions of the questionnaire concerning their fears, concerns and life satisfaction too private.
+We can hence only speculate, that the excluded survey participants found questions of the questionnaire concerning their fears, concerns, life satisfaction, and mental health problems too private. 
+This notion was also supported by results of the sensitivity analysis of the mental health clusters, where the majority of 'imputable' participants with missing psychometric data were assigned to the PTS cluster (__Supplementary Figure S11__).
 
@@
 
-Additionally, we did not ask about severity or time frame of mountain sport accidents in the past, so it is hard to judge if an how they affected mental health at the time point of the survey. 
+Additionally, we did not ask about severity or time frame of mountain sport accidents in the past, so it is hard to judge if and how they affected mental health at the time point of the survey. 
 As inferred from results of the sensitivity analysis presented in __Supplementary Figure S10__, prior mountain sport accidents had little impact on clustering of the study participants by their mental health features.
 They frequency also did not differ between the mental health clusters (__Supplementary Figure S12__). 
-For these reasons it is unlikely that mountain accidents in the past had an substantial effect on the outcome of our study. 
+For these reasons it is unlikely that mountain accidents in the past had a substantial effect on the outcome of our study. 
 We touch upon this point in __Results/Three clusters of mental health response after mountain sport accidents__ and __Results/Demographic, socioeconomic and clinical characteristic of the mental health clusters__.

### 2024-02-28 paper/markdown/manuscript.Rmd
Revised figures, tables, manuscript and supplement

@@
 
-__Background:__ The restorative effect of physical activity in natural alpine environments on mental and physical health is increasingly recognized. 
-However, mountain sports are also linked to a risk of accidents. 
+__Background:__ <span custom-style = 'revision'>The restorative effect of physical activity in alpine environments on mental and physical health is recognized.
+However, risk of accidents and post-accident mental health problems is inherent to every sport, including mountain sport activity.</span>
 We aimed to characterize mental health in individuals following mountain sport accidents requiring professional medical management.
@@
 
-Diagnostic symptoms of PTSD  include re-experiencing (criterion B), avoidance (criterion C), negative alterations in cognition and mood (criterion D) and hyper-arousal (criterion E) [@AmericanPsychiatricAssociation2013], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence.
-PTSD may co-occur with symptoms of depression and anxiety and substance abuse disorders [@Gagnon-Sanschagrin2022; @Brady2000]. 
+Diagnostic symptoms of PTSD  include 
+re-experiencing (such as involuntary, intrusive memories, dreams or flashbacks of the traumatic event;, criterion B), 
+avoidance (such as avoiding memories, thoughts or external reminders related to the traumatic event, criterion C), 
+negative alterations in cognition and mood (criterion D) and
+hyper-arousal (such as irritability, problems with sleep or concentration, criterion E) [@AmericanPsychiatricAssociation2013], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence.
+PTSD may co-occur with symptoms of depression and anxiety and substance <span custom-style = 'revision'>use</span> disorders [@Brady2000]. 
 Although most individuals have experienced at least one traumatic event during their lifetime [@Benjet2016], the prevalence of PTSD in the general European population has been estimated between 0.38 to 6.67% [@Burri2014].
@@
 So far it has proven difficult to reliably identify early predictors of PTSD development in accident victims. 
-While some recent studies on specific populations have identified risk factors, such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020].
+While some recent studies on specific populations have identified risk factors, such as pre-existing psychological distress <span custom-style = 'revision'>or mental disorders</span> [@Ziobrowski2021], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020].
 These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019]. 
+<span custom-style = 'revision'>Overall, combination of multiple demographic, clinical and mental factors seem to yield more accurate PTSD predictions compared to single variables [@Papini2018].</span>
 
 Resilience, i.e.
-the ability to recover from adversities can protect an individual from developing PTSD and can help to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012].
+the ability to recover from adversities can protect an individual from developing PTSD and can help to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007].
 Generally individuals who perform mountain sports display high resilience values [@Ower2019]. 
 Post-traumatic growth can occur in individuals struggling to overcome the consequences of a traumatic event and describes a psychological adjustment reaction with strengthened positive perceptions of self, others, one's life, and meaning of the event [@Tedeschi1996].
-Although both resilience and post-traumatic growth are salutogenetic concepts their exact link is still under investigation [@Levine2009].
+Although both resilience and post-traumatic growth are salutogenic concepts their exact link is still under investigation [@Levine2009].
 
-With the increasing popularity of mountain sports, it is essential to characterize the mental health consequences of accidents during mountain sports. 
-Prevention of PTSD or other mental health problems following an accident is essential to preserve the positive mental health effects of physical activity in an mountain environment despite the accident risk.
+With the increasing popularity of mountain sports, it is <span custom-style = 'revision'>key</span> to characterize the mental health consequences of accidents during mountain sports. 
+Prevention of PTSD or other mental health problems following an accident is essential to preserve the positive mental health effects of physical activity in a mountain environment despite the <span custom-style = 'revision'>inherent</span> accident risk.
 Available data on the consequences of accidents during mountain sports on mental health are scarce. 
-Prior studies were performed on mountain guides, rescuers or avalanche accident survivors and mostly lack clinical data [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004].
-We analyzed mental health readouts such as symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in `r nrow(ptsd$dataset)` individuals treated at the Hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident by means of semi-supervised clustering.
-Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain sport accident with a machine learning approach.
+Prior studies were performed on mountain guides, rescuers or avalanche accident survivors, mostly lack clinical data<span custom-style = 'revision'>, and focus on single aspects of mental health such as trauma exposure or PTSD</span> [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004].
+<span custom-style = 'revision'>A characteristic of multiple facets of mental health following a mountain sport accident, including pathological and salutogenic features, is not clear at the moment.</span>
+<span custom-style = 'revision'>Additionally, risk factors of mental health problems specific for mountain sport accidents in a non-professional setting have not been reported so far.</span> 
+<span custom-style = 'revision'>Identification of such risk-factors, and especially predictors available during acute medical management of the accident or a combination thereof, would greatly facilitate management of mountain sport accident by early, targeted psychological and psychiatric interventions.</span>
+
+<span custom-style = 'revision'>Here, we aimed to identify long-term, persistent patterns of mental health in a sample of `r nrow(ptsd$dataset)` individuals treated at the hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident.
+Such patterns were defined by symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in a semi-supervised clustering analysis.</span>
+<span custom-style = 'revision'>Subsequently, by statistical hypothesis testing and machine learning, we searched for early sociodemographic and clinical factors that characterize the newly defined mental health subsets of patients with a special focus on individuals prone to mental health impairment.</span>
 
@@
 
-Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology at the University Clinic of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation.
+Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology <span custom-style = 'revision'>of the hospital of the Medical University of Innsbruck</span> between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation.
 Individuals fulfilling the inclusion criteria: 
@@
 anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
-depression (PHQ-9, <span custom-style = 'revision'>PHQ-2 and PHQ-8: 9-, 8- and 2-item patient health questionnaires</span> for depressive symptoms) [@Grafe2004], 
+depression (PHQ-9, <span custom-style = 'revision'> and PHQ-8: 9- and 8-item patient health questionnaires</span> for depressive symptoms) [@Grafe2004; @Kroenke2002a], 
 panic (PHQ-panic module) [@Grafe2004], 
@@
 and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
-<span custom-style = 'revision'>Items of the PTGI and PTSD instruments were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').</span>
+<span custom-style = 'revision'>Items of the PTGI and PTSD instruments, and questions concerning flashbacks were provided with captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your mountain sport accident').</span>
 The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).
@@
 Clinically significant symptoms of anxiety (GAD-7 $\geq$ 10), 
-depression (PHQ-9 $\geq$ 11) [@Manea2012], 
+<span custom-style = 'revision'>depression (PHQ-9 $\geq$ 10) [@Humer2022],</span> 
 somatization (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
@@
 
-The primary analysis endpoint was assessment of mental health following a mountain sport accident by clustering using standardized psychometric tools for symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life.
-The secondary analysis endpoints were 
-(1) to compare demographic, socioeconomic and clinical characteristics between the different mental health clusters of mountain accident victims and 
-(2) to identify patients at risk of mental disorders based on a set of explanatory variables available during acute medical treatment after mountain sport accidents.
+<span custom-style = 'revision'>The primary analysis endpoint was identification of subsets of patients differing in long-term symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life following a mountain sport accident.
+This endpoint was pursued by semi-supervised clustering analysis.</span>
+<span custom-style = 'revision'>The secondary analysis endpoints were (1) comparison of demographic, socioeconomic and clinical characteristics of these newly defined subsets in a search for predictors of mental health disorders following a mountain sport accident, and (2) identification of</span> patients at risk of mental disorders based on a set of explanatory variables available during acute medical treatment after mountain sport accidents.
+<span custom-style = 'revision'>The secondary analysis endpoints were addressed by statistical hypothesis testing and machine learning.</span>
 
@@
 `r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
-As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and fewer surgery diagnoses (__Supplementary Table S4__).
-The survey participants excluded from the analysis due to incomplete psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__).
-Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).
+As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents and lower rate of alpine ski/snowboard accidents (<span custom-style = 'revision'>significant, `r filter(excl$test$no_response, variable == 'sport_type')$plot_cap`</span>).
+<span custom-style = 'revision'>As compared with the analyzed participants participants, the non-responders were also characterized by significantly more frequent minor and less frequent severe injuries (`r filter(excl$test$no_response, variable == 'injury_sev_strata')$plot_cap`), overall lower injury grade (`r filter(excl$test$no_response, variable == 'injury_severity_ais')$plot_cap`), significantly lower rates of hospitalization (`r filter(excl$test$no_response, variable == 'hospitalization')$plot_cap`) and surgery (`r filter(excl$test$no_response, variable == 'surgery_done')$plot_cap`), as well as significantly lower surgery complexity reflected by surgical diagnosis numbers (`r filter(excl$test$no_response, variable == 'surgery_complexity')$plot_cap`).</span>
+<span custom-style = 'revision'>Effect sizes of those differences were small (__Supplementary Table S4__).</span> 
+The survey participants excluded from the analysis due to incomplete psychometric data had significantly 
+lower income <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'household_income_class')$plot_cap`)</span>, 
+less severe injury <span custom-style = 'revision'>grade (`r filter(excl$test$missing, variable == 'injury_severity_ais')$plot_cap`)</span>, 
+more frequent injuries of the upper limbs <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'injury_upper_limbs')$plot_cap`)</span>, 
+less frequently required hospitalization <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'hospitalization')$plot_cap`)</span> or 
+surgery <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'surgery_done')$plot_cap`)</span>, and 
+suffered less frequently from persistent physical health consequences of the accident than the analyzed participants <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'accident_aftermath')$plot_cap`)</span>.
+Effect sizes of the differences between the included and excluded participants were weak <span custom-style = 'revision'>(__Supplementary Table S5__).</span>
 
@@
 `r get_percent(ptsd$dataset, 'somatic_comorbidity')$percent['yes']`% of participants with cardiovascular, neurological and metabolic illness being the most frequent. 
-Mental disorders diagnosed by a physician before the accident affected 
-`r get_percent(ptsd$dataset, 'psych_comorbidity')$percent['yes']`% of the cohort with affective 
+Mental disorders diagnosed by a <span custom-style = 'revision'>health care professional prior to</span> the accident affected 
+`r get_percent(ptsd$dataset, 'psych_comorbidity')$percent['yes']`% of the cohort, with affective 
 (`r get_percent(mental$analysis_tbl, 'affective')$percent['yes']`%) and somatoform disorders 
-(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental conditions. 
+(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental <span custom-style = 'revision'>disorders</span>. 
 Four of ten participants had experienced by or witnessed a traumatic event prior to the accident, 
@@
 
+<span custom-style = 'revision'>The age structure was similar, while males were significantly more frequent in the study cohort as compared with the general Austrian population (`r filter(aut_stats$test, variable == 'sex')$plot_cap`) [@StatistikAustria2023].</span>
+<span custom-style = 'revision'>Individuals with secondary and tertiary education (`r filter(aut_stats$test, variable == 'education')$plot_cap`), as well as professionally active people and students (`r filter(aut_stats$test, variable == 'employment_status')$plot_cap`) were more frequent in the investigated sample than in the general Austrian population [@StatistikAustria2020; @StatistikAustria2021].</span>
+<span custom-style = 'revision'>In turn, rates of active smokers (`r filter(aut_stats$test, variable == 'smoking_status')$plot_cap`), and physical or mental illness were significantly lower in the study cohort (`r filter(aut_stats$test, variable == 'illness')$plot_cap`).</span>
+<span custom-style = 'revision'>Effect sizes of these significant differences were small (__Supplementary Table S6__).</span>
+
 Mountain sport accidents in the past were reported by `r get_percent(ptsd$dataset, 'prior_accident')$percent['yes']`% of participants. 
-Most of the investigated accidents occurred during alpine skiing, snowboarding or cross-country skiing 
-(`r get_percent(ptsd$dataset, 'sport_type')$percent['ski/snowboard/cross-country']`%) followed by biking 
-(`r get_percent(ptsd$dataset, 'sport_type')$percent['biking']`%) and classical mountain sports such as climbing, hiking, mountaineering or back-country skiing 
-(`r get_percent(ptsd$dataset, 'sport_type')$percent['climbing/hiking/mountaineering/skitour']`%). 
+Most of the investigated accidents occurred 
+<span custom-style = 'revision'>during alpine skiing or snowboarding in secured areas of ski resorts (`r get_percent(ptsd$dataset, 'sport_type')$percent['alpine skiing/snowboarding']`%)</span> followed by biking
+(`r get_percent(ptsd$dataset, 'sport_type')$percent['biking']`%), 
+<span custom-style = 'revision'>hiking (`r get_percent(ptsd$dataset, 'sport_type')$percent['hiking']`%), cross-country skiing (`r get_percent(ptsd$dataset, 'sport_type')$percent['cross-country skiing']`%), and sledding (`r get_percent(ptsd$dataset, 'sport_type')$percent['sledding']`%).</span>
+<span custom-style = 'revision'>Mountain sports perceived as particularly risk-associated such as alpine skiing/snowboarding in non-secured areas (`r get_percent(ptsd$dataset, 'sport_type')$percent['ski touring/freeride']`%), ice-climbing (`r get_percent(ptsd$dataset, 'sport_type')$percent['ice climbing']`%), climbing (`r get_percent(ptsd$dataset, 'sport_type')$percent['ski touring/freeride']`%), mountaineering (`r get_percent(ptsd$dataset, 'sport_type')$percent['mountaineering']`%) and paragliding (`r get_percent(ptsd$dataset, 'sport_type')$percent['air sport']`%) comprised a minute fraction of all accident sports in the study cohort (__Table \@ref(tab:tab-cohort-accident)__)</span>
+<span custom-style = 'revision'>The profile of accident sports in the analyzed cohort differed significantly with weak effect size from distribution of sports in the Austrian statistic of mountain accidents in 2023 [@OsterreichischesKuratoriumfurAlpineSicherheit2024] (`r filter(lit_kurasi$test, variable == 'sport_detail')$plot_cap`).</span>
+<span custom-style = 'revision'>Frequency of alpine skiing/snowboarding, cross-country skiing, biking, and water sport accidents was significantly higher in the analyzed sample as compared with the national registry.</span>
+<span custom-style = 'revision'>Percentages of ski/snowboard accidents in non-secured areas and hiking accidents were in turn significantly lower in the study cohort than in the national registry.</span>
+<span custom-style = 'revision'>Additionally, accidents in the winter season were more common  while accidents in the summer season were less frequent in the study cohort than in the national statistic (small effect size, `r filter(lit_kurasi$test, variable == 'accident_month')$plot_cap`).</span>
+<span custom-style = 'revision'>The age of study sample deviated significantly from the 2023 national registry data with weak effect size (`r filter(lit_kurasi$test, variable == 'age_class')$plot_cap`).</span>
+<span custom-style = 'revision'>In particular, accident victims aged 51 - 60 years were more frequent in the study cohort (__Supplementary Figure S1__).</span>
+
 One-third of participants were alone during the accident 
@@
 `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['3+']`% severe-to-critical (AIS $\geq$ 3). 
-Limb injuries were the most common followed by injuries of the head and face (__Supplementary Figure S1A__). 
+Limb injuries were the most common followed by injuries of the head and face (__Supplementary Figure S2A__). 
 Hospitalization and surgery rates were 
@@
 Persisting physical consequences of the accident were reported by 
-`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks during mountain sport at the time of the questionnaire completion were observed in 
+`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks <span custom-style = 'revision'>of the accident of interest</span> during mountain sport at the time of the questionnaire completion were observed in
 `r get_percent(ptsd$dataset, 'unwilling_flashback')$percent['yes']`% of the cohort. 
@@
 
-At least one diagnostic criterion of PTSD defined as positive scoring of the B, C, D or E domains of the PCL-5 tool was met by 
-`r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% of participants. 
-The domain B PTSD symptoms were the most common (`r get_percent(ptsd$dataset, 'dsm5_B_class')$percent['positive']`%) and the domain D symptoms were the least frequent (`r get_percent(ptsd$dataset, 'dsm5_D_class')$percent['positive']`%).
+<span custom-style = 'revision'>PTSD symptoms related to the mountain sport accident defined as positive scoring of at least one of the B, C, D or E domains of the PCL-5 tool were observed in</span> `r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% of participants.
+<span custom-style = 'revision'>Concerning particular PCL-5 domains, positive scoring for the domain B was the most common (`r get_percent(ptsd$dataset, 'dsm5_B_class')$percent['positive']`%), and frequency of participants screened positive for the domain D was the lowest (`r get_percent(ptsd$dataset, 'dsm5_D_class')$percent['positive']`%).</span>
 Solely `r digit2string(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0))` patients 
-(`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) were screened positive for all four diagnostic criteria of PTSD (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1BC__).
+(`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) were screened positive for all four diagnostic criteria <span custom-style = 'revision'>suggestive of manifest PTSD</span> (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S2BC__).
 Clinically relevant symptoms of 
@@
 and somatization (`r get_percent(ptsd$dataset, 'phq_events_total_class')$percent['positive']`%) were rare in this cohort consisting of `r get_percent(ptsd$dataset, 'rs13_total_class')$percent['high']`% highly resilient individuals (__Table \@ref(tab:tab-cohort-mental)__).
+<span custom-style = 'revision'>Mean scoring of quality of life assessed by the EUROHIS tool in the cohort was `r mean(ptsd$dataset$eurohis_total) %>% signif(2)` and was significantly higher that the mean of 4.08 reported for the comparable general German population (one-sample Wilcoxon test: r = `r ptsd$dataset %>% wilcox_effsize(eurohis_total ~ 1, mu = 4.08) %>% .$effsize %>% signif(2)`, p < 0.001) [@Schmidt2006].</span>
+<span custom-style = 'revision'>The general quality of life and housing domains were assigned the highest scores, while esteem was rated the lowest among the study participants.</span> 
+<span custom-style = 'revision'>The median total score of accident-related post-traumatic growth was `r median(ptsd$dataset$ptgi_total)` (interquartile range: `r quantile(ptsd$dataset$ptgi_total, c(0.25, 0.75)) %>% paste(collapse = ' to ')`), with the relations domain with the highest rating and the spiritual domain with the lowest scores (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S3__).</span>
 We could not observe any significant association of values of psychometric scores or frequency of mental health problems with the time between trauma center admission and survey completion (not shown).
+
+<span custom-style = 'revision'>As compared with general population estimates [@Hauffa2011; @Kilpatrick2013; @Koenen2017; @Kessler2017], frequency of traumatic events (`r lit_data$stats %>% get_est_ci(source == 'cohort', variable == 'traumatic_event', level == 'yes')`) and manifest PTSD in the study sample defined as positivity for all PCL-5 domains (`r lit_data$stats %>% get_est_ci(source == 'cohort', variable == 'dsm5_all_class', level == 'positive')`) was low but comparable to figures reported for the general German population by Hauffa and colleagues (traumatic events, CIDI tool: `r lit_data$stats %>% get_est_ci(source == 'hauffa_2011', variable == 'traumatic_event', level == 'yes')`, manifest PTSD, PDS tool: `r lit_data$stats %>% get_est_ci(source == 'hauffa_2011', variable == 'ptsd_lifetime', level == 'positive')`, __Supplementary Figure S4A__) [@Hauffa2011].</span>
+<span custom-style = 'revision'>Furthermore, frequency of traumatic events in the study cohort was significantly lower than in a sample of Swiss mountain rescuers reported by Mikutta et al.
+(`r lit_data$stats %>% get_est_ci(source == 'mikutta_2022', variable == 'traumatic_event', level == 'yes')`) [@Mikutta2022].</span>
+<span custom-style = 'revision'>In turn, frequency of PTSD symptoms and of manifest PTSD measured with the PCL-5 instrument were similar in both cohorts (__Supplementary Figure S4B__).</span>
+
+<span custom-style = 'revision'>Our sample was characterized by significantly lower percentages of anxious (GAD-7 $\geq$ 10, `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'anxiety')`) and depressive symptoms (PHQ-9 $\geq$ 10, `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'depression')`) as compared with estimates for the general Austrian population in 2022 by Humer et al.
+(anxiety, GAD-7 $\geq$ 10: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'anxiety')`, depression, PHQ-9 $\geq$ 10: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'depression')`, __Supplementary Figure S5A__) [@Humer2022].</span>
+<span custom-style = 'revision'>As assessed by the PHQ-8 instrument, the frequency of moderate-to-severe depressive symptoms in the study sample was similar to the frequency in the pre-pandemic general Austrian population estimated by a 2019 microcensus study [@StatistikAustria2019a] (__Supplementary Figure S5B__).</span>
+<span custom-style = 'revision'>The resilience scores in our study collective (mean: `r lit_plots$rs_stats %>% filter(source == 'cohort') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'cohort') %>% .$sd %>% signif(2)`) were comparable with resilience levels in the Swiss mountain rescuer sample (mean: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$sd %>% signif(2)`) [@Mikutta2022], and higher as compared with the general population estimates obtained by Chernova et al.
+with the same RS-13 tool (mean: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$sd %>% signif(2)`, __Supplementary Figure S5C__) [@Chernova2021].</span>
 
@@
 test subset (n = `r table(ptsd$dataset$partition)['test']`). 
-The sole significant differences between these subsets concerned resilience class distribution and scores of quality of life, their effect size was weak (__Supplementary Table S6__).
+The sole significant differences between these subsets concerned resilience class distribution and scores of quality of life, their effect size was weak <span custom-style = 'revision'>(__Supplementary Table S7__).</span>
 
@@
 
-To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to medoid clustering in respect to a broad range of psychometric scores (__Supplementary Table S2__).
-Three mental health clusters were identified in the training subset of the study cohort: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress) and named after their key mental characteristic (__Supplementary Figure S3__).
+<span custom-style = 'revision'>To identify patterns</span> of mental health following mountain sport accidents, we subjected the participants to medoid clustering in respect to a broad range of psychometric scores (__Supplementary Table S2__).
+Three mental health clusters were identified in the training subset of the study cohort: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress) and named after their key mental characteristic <span custom-style = 'revision'>(__Supplementary Figure S6__)</span>.
 Subsequently, the mental health cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
@@
 (training: `r signif(semi_clust$stats$sil_width[1], 2)`, test: `r signif(semi_clust$stats$sil_width[2], 2)`), 
-comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).
+<span custom-style = 'revision'>similarly low frequency of potentially misclassified observations (fraction of observations with negative silhouette widths, training: `r signif(semi_clust$stats$frac_misclassified[1], 2)`, test: `r signif(semi_clust$stats$frac_misclassified[2], 2)`), and good preservation of the nearest neighborhood (fraction of 5-nearest neighbors in the same cluster, training: `r signif(semi_clust$stats$frac_np[1], 2)`, test: `r signif(semi_clust$stats$frac_np[2], 2)`).</span>
+<span custom-style = 'revision'>Furthermore, classification of the training and test individuals yielded similar distribution of cluster sizes,</span> good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets <span custom-style = 'revision'>(__Supplementary Figures S7 - S9__).</span>
 
-The mental health clusters were of approximately equal size. 
-The neutral cluster was characterized by low scores of PTSD symptoms, anxiety, depression, panic, somatization, and post-traumatic growth along with high ratings of sense of coherence, resilience and quality of life.
+The mental health clusters were of approximately equal size <span custom-style = 'revision'>but differed significantly in psychometric characteristic.</span>
+The neutral cluster was characterized by <span custom-style = 'revision'>low scores of PCL-5 domains assessing PTSD symptoms as well as low ratings of</span> anxiety, depression, panic, somatization, and post-traumatic growth along with high ratings of sense of coherence, resilience and quality of life.
 The PTG cluster demonstrated similarly low scores of major mental health disorders, high resilience and sense of coherence. 
 Its key characteristic were the highest levels of post-traumatic growth. 
-The remaining PTS cluster displayed the highest scores of PTSD symptoms, anxiety, depression, panic disorder, somatization as well as poor sense of coherence, and low resilience and low quality of life.
-Post-traumatic growth scores in the PTS cluster were higher than in neutral but lower than in PTG cluster participants (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Table S7__). 
-Consequently, clinically relevant symptoms of 
-anxiety (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'gad7_total_class')$percent['positive']`%), 
-depression (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq9_total_class')$percent['positive']`%) and somatization (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq_events_total_class')$percent['positive']`%)
-were present virtually only in the PTS cluster. 
+The remaining PTS cluster displayed <span custom-style = 'revision'>the highest scores of PTSD symptoms measured by domains of the PCL-5 instrument, as well as the peak ratings of</span> anxiety, depression, somatization, poor sense of coherence, low resilience and low quality of life.
+Post-traumatic growth scores in the PTS cluster were higher than in neutral but lower than in PTG cluster participants (__Figure \@ref(fig:fig-clust-psych)__, <span custom-style = 'revision'>__Supplementary Figure S9__</span>).
+
+Clinically relevant symptoms of 
+anxiety (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'gad7_total_class')$percent['positive']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'gad7_total_class')$plot_cap`</span>),
+depression (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq9_total_class')$percent['positive']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'phq9_total_class')$plot_cap`</span>) and
+somatization (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'phq_events_total_class')$percent['positive']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'phq_events_total_class')$plot_cap`</span>)
+were <span custom-style = 'revision'>significantly more frequent in the PTS cluster as compared with the PTG or neutral clusters.</span>
 Furthermore, frequencies of 
 low (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'rs13_total_class')$percent['low']`%) and 
-moderate resilience classes (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'rs13_total_class')$percent['moderate']`%) peaked in the PTS cluster. 
-Finally, frequency of flashbacks, PTSD symptoms specified by the B, C, D and E PCL-5 domains and frequency of individuals scoring positive for at least one diagnostic domain of PCL-5 
-(PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'dsm5_cluster_class')$percent['positive']`%) were the highest in PTS cluster individuals (__Figure \@ref(fig:fig-clust-symptoms)__, __Supplementary Table S8__).
+moderate resilience classes (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'rs13_total_class')$percent['moderate']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'rs13_total_class')$plot_cap`</span>) peaked in the PTS cluster.
+<span custom-style = 'revision'>Concerning PTSD symptoms, frequency of flashbacks as well as percentages of individuals screened positive for each of the B, C, D, and E PCL-5 domains were significantly higher in PTS cluster patients as compared with the remaining individuals.</span>
+<span custom-style = 'revision'>Analogically, the frequency of participants with PTSD symptoms defined by at least one positive PCL-5 domain was</span> the highest in <span custom-style = 'revision'>the</span> PTS cluster (`r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'dsm5_cluster_class')$percent['positive']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'dsm5_cluster_class')$plot_cap`</span>, __Figure \@ref(fig:fig-clust-symptoms)__, <span custom-style = 'revision'>__Supplementary Table S9__</span>).
+
+<span custom-style = 'revision'>Next, we investigated, how the clustering of study participants is affected by missing observations, time after the accident, and factors that may influence mental health after the accident such as annual income, prior mountain sport accidents, injury severity, and hospitalization.</span>
+<span custom-style = 'revision'>To this end, we re-created the clustering structure for the whole cohort, the study cohort appended with imputed missing observations, and the cohort data set depleted of categories of the variables listed above (e.g.
+the analysis cohort without high income participants or severe injuries) with the medoid clustering algorithm.</span>
+<span custom-style = 'revision'>Next, we compared fractions of explained clustering variances between the clustering structures in the genuine cohort and the modified data sets in a kind of sensitivity analysis [@Thabane2013].</span>
+<span custom-style = 'revision'>As presented in __Supplementary Figure S10A__, differences in explained variance between the original clustering structure and clustering structures developed for the modified data sets were minor (genuine data set: `r se_summary$stats %>% filter(split_factor == 'global') %>% .$frac_var %>% signif(2)`, modified: `r se_summary$stats %>% filter(split_factor %in% c('missing', 'household_income_class', 'prior_accident', 'injury_sev_strata', 'hospitalization', 'accident_year')) %>% .$frac_var %>% range %>% signif(2) %>% paste(collapse = ' to ')`).</span>
+<span custom-style = 'revision'>There were no significant differences in distribution of the cluster sizes for the clustering schemes developed in the analyzed cohort and the modified data sets (__Supplementary Figure S10B__).</span>
+<span custom-style = 'revision'>Collectively, this lets us conclude that neither missing observations, differences in income, accidents in the past, time after accident, injury severity, nor hospitalization significantly affected quality and distribution of the mental health clusters.</span>
+<span custom-style = 'revision'>In the classification scheme developed for the data set appended with incomplete observations, the participants with missing psychometric data which had been successfully imputed were preferentially assigned to the PTS cluster (imputed observations in the PTS cluster: n = `r se_summary$missing_plots$color_bar_data %>% filter(clust_id == 'PTS', subset == 'missing') %>% nrow`, total imputed: n = `r table(se_summary$missing_plots$color_bar_data$subset)['missing']`, __Supplementary Figure S11__).</span>
 
@@
 were found to differ significantly between the mental health clusters. 
-The effect size of these differences was weak (__Supplementary Table S9__).
+The effect size of these differences was weak <span custom-style = 'revision'>(__Supplementary Table S10__).</span>
 
-In more detail, individuals in the PTS cluster were the youngest of all participants and had the highest frequency of pre-existing mental disorders. 
-The neutral cluster was in turn substantially enriched in males, and tertiary education and high-income individuals; these effects were not significant. 
-Participants suffering from pre-existing physical illnesses were significantly enriched in the PTG and PTS clusters. 
-A similar, yet not significant tendency was observed for the frequency of prior traumatic events, which was higher in the PTG and PTS clusters as compared with neutral cluster individuals. 
-There were no substantial differences in frequency of search and rescue professionals and employment structure between the clusters (__Figure \@ref(fig:fig-clinic-factors)__, __Supplementary Figure S7__).
-Participants reporting being alone during the accident, responsible for the accident, and requiring companion or professional rescue were enriched in the PTG cluster. 
-In turn, frequency of self-rescuers tended to be the highest in the neutral cluster. 
-These differences in accident and rescue circumstances were not significant. 
-We could not observe any substantial differences in accident season, mountain sport type, and numbers of injured persons between the clusters (__Figure \@ref(fig:fig-injury)A__, __Supplementary Figure S7__, __Supplementary Table S9__).
-There were no significant differences in severity and injury and its location between the clusters. 
-However, moderate-to-severe injuries (AIS $\geq$ 2), hospitalization and surgery rates were substantially higher in the PTG and PTS clusters as compared with neutral cluster participants (not significant, __Figure \@ref(fig:fig-injury)__, __Supplementary Table S9__).
+In more detail, individuals in the PTS cluster were the youngest of all participants <span custom-style = 'revision'>(global difference between the clusters: `r filter(clust_bcg$test, variable == 'age')$plot_cap`)</span> and had the highest frequency of pre-existing mental disorders <span custom-style = 'revision'>(`r filter(clust_bcg$test, variable == 'psych_comorbidity')$plot_cap`)</span>.
+<span custom-style = 'revision'>Frequency of pre-existing physical disorders was significantly higher in the PTG and PTS that in the neutral cluster (`r filter(clust_bcg$test, variable == 'somatic_comorbidity')$plot_cap`).</span>
+<span custom-style = 'revision'>Differences in gender, income, employment and education, in frequency of search and rescue professionals and prior traumatic events between the clusters were not significant.</span>
+<span custom-style = 'revision'>We could not observe significant differences in frequency of being alone during the accident, or self-rescue, companion or professional rescue between the clusters.</span>
+<span custom-style = 'revision'>No significant differences in accident season, mountain sport type, and numbers of injured persons between the clusters were detected.</span>
+<span custom-style = 'revision'>Analogically, there were no significant differences in severity and injury and its location between the clusters (__Figures \@ref(fig:fig-clinic-factors)__ - __\@ref(fig:fig-injury)__, __Supplementary Figure S12__ - __S13__, __Supplementary Table S10__).</span>
 
 There were no significant differences in rates of participants who received psychological or psychiatric support after the accident between the clusters. 
-Among PTG and PTS cluster individuals, `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'psych_support_need')$percent['yes']`% and `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'psych_support_need')$percent['yes']`%, respectively, reported need for psychological or psychiatric support following the accident as compared with none in the neutral cluster.
+Among PTG and PTS cluster individuals, `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'psych_support_need')$percent['yes']`% and `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'psych_support_need')$percent['yes']`%, respectively, reported need for psychological or psychiatric support following the accident as compared with none in the neutral cluster <span custom-style = 'revision'>(`r filter(clust_bcg$test, variable == 'psych_support_need')$plot_cap`).</span>
 Frequency of participants suffering from persistent physical health consequences of the accident was the highest in the PTS cluster (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'accident_aftermath')$percent['yes']`%,
 PTG: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'accident_aftermath')$percent['yes']`%, 
-neutral: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'neutral'), 'accident_aftermath')$percent['yes']`%). 
+neutral: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'neutral'), 'accident_aftermath')$percent['yes']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'accident_aftermath')$plot_cap`</span>).
 Analogically, PTS cluster patients reported the highest rates of cautious behavior during mountain sport following the accident 
@@
 PTG: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTG'), 'caution_post_accident')$percent['more cautious']`%, 
-neutral: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'neutral'), 'caution_post_accident')$percent['more cautious']`%). 
+neutral: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'neutral'), 'caution_post_accident')$percent['more cautious']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'caution_post_accident')$plot_cap`</span>).
 By contrast, percentages of participants who returned to the same mountain sport after the accident were comparable between the mental health clusters (__Figure \@ref(fig:fig-aftermath)__).
@@
 
-Finally, we intended to model the mental health cluster assignment with demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the patient (__Supplementary Table S10__) with seven popular machine learning algorithms:
+Finally, we intended to model the mental health cluster assignment with demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the patient <span custom-style = 'revision'>(__Supplementary Table S11__)</span> with seven popular machine learning algorithms:
 the random forests, 
@@
 elastic net regression, and 
-conditional random forests (__Supplementary Table S11__). 
+conditional random forests <span custom-style = 'revision'>(__Supplementary Table S12__)</span>. 
 Such models would enable early identification of mountain sport accident victims at risk of mental health problems.
@@
 Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`). 
-Concerning the early identification of the vulnerable PTS cluster patients, the conditional random forest, random forest and support vector machine models were able to differentiate between the PTS cluster and the pooled neutral and PTG clusters with the best sensitivity (test subset sensitivity $\geq$ `r min(signif(early_class$clust_stats %>% filter(dataset == 'test', clust_id == 'PTS') %>% top_n(n = 3, Se) %>% .$Se, 2) * 100)`%, __Figure \@ref(fig:fig-early-class)__, __Supplementary Table S12__).
-Pre-existing mental disorder or physical illness, hospitalization and surgery, injury severity, age, rescue mode, education, annual income, and traumatic events in the past were among the most important explanatory factors employed by these models (__Supplementary Figure S8__).
+Concerning the early identification of the vulnerable PTS cluster patients, the conditional random forest, random forest and support vector machine models were able to differentiate between the PTS cluster and the pooled neutral and PTG clusters with the best sensitivity (test subset sensitivity $\geq$ `r min(signif(early_class$clust_stats %>% filter(dataset == 'test', clust_id == 'PTS') %>% top_n(n = 3, Se) %>% .$Se, 2) * 100)`%, __Figure \@ref(fig:fig-early-class)__, <span custom-style = 'revision'>(__Supplementary Table S13__)</span>).
+Pre-existing mental disorder or physical illness, hospitalization and surgery, injury severity, age, rescue mode, education, annual income, and traumatic events in the past were among the most important explanatory factors employed by these models <span custom-style = 'revision'>(__Supplementary Figure S14__).</span>
 Of note, inclusion of follow-up-related variables such as received psychological or psychiatric support, psychological or psychiatric support need, physical health consequences or caution during mountain sport activity in the machine learning models did not improve their accuracy at prediction of the mental health cluster assignment in cross-validation or the test subset (test accuracy: `r paste0((filter(full_class$overall_stats, dataset == 'test')$correct_rate %>% range %>% signif(2)) * 100, '%') %>% paste(collapse = ' - ')`,
 Cohen's $\kappa$: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$kappa), 2) %>% paste(collapse = ' - ')`, 
-Brier score: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`, __Supplementary Figures S9__ - __S10__, __Supplementary Table S10__).
+Brier score: `r signif(range(filter(full_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`, <span custom-style = 'revision'>__Supplementary Figures S15__ - __S16__, __Supplementary Table S14__</span>).
 
@@
 We characterized mental health outcomes in individuals treated at a tertiary trauma center for an accident in mountain sport at a median of `r round(median(ptsd$dataset$obs_time)/30.436)` months following the incident.
-We identified three approximately equally sized subsets of participants with distinct mental health characteristic: (1) the neutral mental health cluster, (2) the PTG cluster characterized by a predominantly salutatory reaction to the accident manifested by post-traumatic growth, and (3) the PTS cluster hallmarked by symptoms of PTSD, anxiety, depression, panic disorder and somatization as well as low resilience, quality of life and sense of coherence.
+We identified three approximately equally sized subsets of participants with distinct mental health characteristic: (1) the neutral mental health cluster, (2) the PTG cluster characterized by a predominantly salutatory reaction to the accident manifested by post-traumatic growth, and (3) the PTS cluster hallmarked by symptoms of PTSD, anxiety, depression, and somatization as well as low resilience, quality of life and sense of coherence.
 Individuals in the PTS cluster were characterized by younger age, the highest frequency of pre-existing mental disorders and persisting physical health consequences of the accident compared to individuals in the other two clusters.
@@
 
-This study evaluates a broad palette of mental features including symptoms of anxiety and depression, somatiziation, resilience, sense of coherence and quality of life, post-traumatic growth and symptoms of PTSD in survivors of mountain sport accidents.
+<span custom-style = 'revision'>The analyzed cohort was significantly enriched with males, secondary/tertiary graduates, professionally active, and physically healthier individuals as compared with the general Austrian population.</span>
+<span custom-style = 'revision'>We speculate that this reflects rather the overall characteristic of physically active inhabitants of the alpine region than a selection bias.</span>
+<span custom-style = 'revision'>Differences in the accident sport type, accident season, and age structure between the study cohort and the collective of mountain accident victims in the Austrian registry were of weak effect size.</span>
+<span custom-style = 'revision'>Within the analyzed sample, winter sport accidents in general, and, in particular, alpine skiing/snowboarding accidents in secured areas of ski resorts were significantly more frequent.</span>
+<span custom-style = 'revision'>Notably, we did not observe higher percentages of high-risk mountain sport types such as climbing, ski touring/freeride, paragliding or high mountain sport in the cohort as compared with the national registry.</span>
+<span custom-style = 'revision'>Analogically, there were no significant differences in frequency of such sports between the analyzed sample and non-responders to the study invitation.</span>
+<span custom-style = 'revision'>This lets us conclude that the study was unlikely biased towards risk-seeking individuals.</span>
+<span custom-style = 'revision'>As compared with the non-responders, injury severity, and hospitalization and surgery rates were significantly higher in the analyzed cohort, effect size of those differences was, however, small.</span>
+<span custom-style = 'revision'>Collectively, despite the low response rate, our study cohort can be regarded as reasonably representative sample of the entire collective of mountain sport accident victims in the alpine region.</span>
+
+This study evaluates a broad palette of <span custom-style = 'revision'>features of mental health</span> including symptoms of anxiety and depression, somatiziation, resilience, sense of coherence and quality of life, post-traumatic growth, and symptoms of PTSD <span custom-style = 'revision'>defined by positive screening of at least one PCL-5 domain</span> in survivors of mountain sport accidents.
 This population is of special interest because, on one hand, mountain sports are generally considered beneficial for mental health [@Ower2019; @Niedermeier2017a]. 
 On the other hand, mountain sports are associated with specific individual physical and mental traits, and bear accident and injury risks [@Faulhaber2017; @Faulhaber2020]. 
-Not many studies have assessed PTSD rates following accidents in mountain sports. 
-Survivors of avalanches were affected by acute and long-term PTSD symptoms such as hyperarousal or sleep disorders at rates between 11% - 16% [@Thordardottir2015; @Leonard2021]. 
-This figure is comparable with `r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% participants of our study screened positive for at least one of of the PTSD symptom domains defined by the PCL-5 tool [@Bovin2016].
+<span custom-style = 'revision'>We compared rates of traumatic events measured by DIA-X tool [@Maercker2005], and of manifest PTSD defined as positive screening for all of the B, C, D and E of the PCL-5 instrument [@Bovin2016] in our cohort with published estimates for the general population.</span>
+<span custom-style = 'revision'>In such analysis, the frequency of both traumatic events (`r get_percent(ptsd$dataset, variable = 'traumatic_event')$percent['yes']`%) and manifest PTSD (`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) in the cohort was significantly lower than in most multi-population surveys [@Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017], and comparable with estimates provided for the German population by Hauffa and colleagues (traumatic events: `r lit_data$stats %>% filter(source == 'hauffa_2011', variable == 'traumatic_event', level == 'yes') %>% .$percent %>% signif(2)`%, manifest lifetime PTSD: `r lit_data$stats %>% filter(source == 'hauffa_2011', variable == 'ptsd_lifetime', level == 'positive') %>% .$percent %>% signif(2)`%) [@Hauffa2011].</span>
+<span custom-style = 'revision'>Despite the substantial burden of mountain sport accidents on medical system of alpine countries, with more than 13000 accidents in 2023 in Austria alone [@OsterreichischesKuratoriumfurAlpineSicherheit2024], not many studies assessed rates of PTSD and other mental disorders in survivors of mountain sport accidents.</span>.
+<span custom-style = 'revision'>Survivors of avalanches reported long-term symptoms suggestive of PTSD</span> at rates between 11% - 16% [@Thordardottir2015; @Leonard2021]. 
+<span custom-style = 'revision'>Those estimates are</span> comparable with `r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% participants of our study screened positive for at least one of of the PTSD symptom domains defined by the PCL-5 tool [@Bovin2016].
 PTSD frequencies of 11.8% - 26% have been reported at 3 - 12 months after acute trauma [@Shalev2019; @Schultebraucks2020; @Joseph2020]. 
@@
 
-Studies conducted on mountain guides and mountain rescue personnel, showed that 71% - 78% of them experienced a traumatic event [@Mikutta2022; @Sommer2004], while the reported frequency of PTSD was very low (0.9% - 2.7%) [@Mikutta2022; @Sommer2004].
-This is comparable to trauma event estimates for the general population [@PTSD2023; @Trautmann2018; @Burri2014] and, particularly, lower than in other potentially vulnerable groups such as emergency workers [@Berger2012].
-By contrast, a recent Swiss study found a prevalence of PTSD in up to 22% of mountain rescuers once differentiating between clinical interviews and self-reporting of symptoms for the assessment of PTSD [@Traber2020].
+<span custom-style = 'revision'>Our study did not specifically address professional mountain rescuers or mountain sport professionals, and only `r get_percent(ptsd$dataset, 'trauma_risk_profession')$percent['yes']`% of the sample reported a search and rescue profession and `r get_percent(ptsd$dataset, 'sport_profession')$percent['yes']`% a mountain sport profession.</span>
+Studies conducted on mountain guides and mountain rescue personnel, showed that 71% - 78% of them experienced a traumatic event, while the reported frequency of PTSD was very low (0.9% - 2.7%) [@Mikutta2022; @Sommer2004].
+This is comparable to estimates <span custom-style = 'revision'>of traumatic events and manifest PTSD in</span> the general population [@Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017] and, particularly, lower than in other potentially vulnerable groups such as emergency workers [@Berger2012].
+By contrast, a recent <span custom-style = 'revision'>French</span> study found PTSD in up to 22% of mountain rescuers once differentiating between clinical interviews and self-reporting of symptoms for the assessment of PTSD [@Traber2020].
+<span custom-style = 'revision'>A similarly low frequency of manifest PTSD and an analogical profile of single PTSD symptoms were observed in the study cohort and the cohort of mountain rescue workers investigated with the PCL-5 tool by Mikutta et al.
+(manifest PTSD: `r lit_data$stats %>% filter(source == 'mikutta_2022', variable == 'dsm5_all_class', level == 'positive') %>% .$percent %>% signif(2)`) [@Mikutta2022].</span>
+Resilience has been proposed as a protective trait against negative mental health consequences following trauma [@Bensimon2012]. 
+Two-thirds of our study participants were classified as highly resilient [@Leppert2008]. 
+<span custom-style = 'revision'>Of note, both in the study cohort and the mountain rescuer collective by Mikutta et al.
+displayed substantially increased levels of resilience as compared with the general population [@Chernova2021; @Mikutta2022; @Leppert2008], which can explain low prevalence of manifest PTSD in individuals performing mountain sports in a free-time or professional setting.</span>
+<span custom-style = 'revision'>It is likely that this high resilience is attributed to the protective effect of physical activity in an alpine environment [@Ower2019].</span>
+<span custom-style = 'revision'>The elevated resilience scores may also</span> explain lower frequencies of clinically relevant symptoms of anxiety and depression in our cohort than estimated for the general Austrian population <span custom-style = 'revision'>at the time of our survey</span> [@Humer2022].
+The high resilience in the study cohort may also pertain to the rating of quality of life <span custom-style = 'revision'>which was significantly higher than in the general western European population [@Schmidt2006].</span>
 
-Resilience has been proposed as a protective trait against negative mental health consequences following trauma [@Bensimon2012] and high resilience levels in mountain rescuers may explain low prevalence of PTSD in this population despite trauma exposure [@Mikutta2022].
-Two-thirds of our study participants were classified as highly resilient [@Leppert2008]. 
-This observation may hence explain the low rate of manifest PTSD as well as lower frequencies of clinically relevant symptoms of anxiety and depression in our cohort than estimated for the general Austrian population [@Humer2022].
-The high resilience in the study cohort may also pertain to the high rating of quality of life 
-(EUROHIS, median: `r get_num_stats(ptsd$dataset, 'eurohis_total', collapse = ', interquartile range: ')`) comparable to the general western European population [@Schmidt2006].
-
+<span custom-style = 'revision'>To our best knowledge, our study provides the first multi-faceted characteristic of long-term consequences of a mountain sport accidents for mental health.</span>
 By semi-supervised clustering, we identified mountain sport accident victims with co-existing symptoms of PTSD, depression, anxiety, somatization, as well as low sense of coherence and quality of life.
@@
 The psychometric characteristic of the PTS cluster corresponds to the multi-faceted mental symptoms described for 
-individuals following a traumatic event [@Gagnon-Sanschagrin2022; @Brady2000; @Bryant2019]. 
+individuals following a traumatic event [@Brady2000; @Bryant2019]. 
 Regarding early predictors of mental health problems in mountain accident victims, the PTS cluster was characterized by a significantly younger age as well as high rates of pre-existing physical and mental disorders.
-It was also substantially enriched in females, low-to-moderate income participants, individuals requiring companion or professional rescue, hospitalization and surgery as compared with neutral cluster patients.
 Of note, 
@@
 PTSD risk has also been previously associated with injury severity, head injuries, hospitalization length and pain [@Baecher2018; @Blanchard1995; @Galatzer-Levy2014]. 
-We could not establish any link between the mental health clusters and, in particular, the PTS cluster with objective injury severity measured with the AIS scale [@Gennarelli2006] or injury location. 
+<span custom-style = 'revision'>In our study,</span> we could not establish any link between the mental health clusters and, in particular, the PTS cluster with objective injury severity measured with the AIS scale [@Gennarelli2006] or injury location.
 Intriguingly, subjective injury severity, which was not surveyed in our cohort, was shown to be a stronger predictor of PTSD symptoms than objective injury severity [@Gabert-Quillen2011]. 
-This phenomenon together with the longer time interval between hospital admission and survey than in most other studies and differences in composition of the study sample, may explain the lacking effect of injury on subsequent mental disorders in our collective.
+This phenomenon together with the longer time interval between hospital admission and survey than in most other studies, and differences in <span custom-style = 'revision'>investigated populations</span>, may explain <span custom-style = 'revision'>the lack of</span> effect of injury <span custom-style = 'revision'>location or severity</span> on subsequent mental disorders in our collective.
 Importantly, the percentage of participants having received psychological or psychiatric support after the accident was generally low in the study cohort (`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`%) and did not differ between the mental health clusters.
@@
 
-Early identification of trauma patients at risk of developing PTSD or further mental disorders has previously been tackled by diverse machine learning algorithms [@Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
 In our hands, none of seven machine learning procedures using different mathematical principles, random forest, neural network, support vector machines, recursive partitioning, conditional random forest, discriminant analysis or elastic net regression, was able to reliably predict the mental health cluster assignment based on `r length(class_globals$variables$early)` demographic, socioeconomic, clinical and accident-related variables available during acute medical treatment of the patient.
 This phenomenon is likely attributed to the lacking strong non-psychometric explanatory variables differentiating between the mental health clusters and, of clinical importance, highly specific markers of post-traumatic stress patients.
-Interestingly, inclusion of peri-traumatic mental symptoms like PTSD features, flashbacks, hyperarousal, or subjective need for support was shown to be crucial for optimal prediction of long-term mental disorders or PTSD during recovery [@Gabert-Quillen2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
+<span custom-style = 'revision'>Early identification of trauma patients at risk of developing PTSD or further mental disorders has previously been tackled by diverse machine learning algorithms [@Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].</span>
+Interestingly, inclusion of peri-traumatic mental symptoms like PTSD features, flashbacks, hyperarousal, or subjective need for support was shown to be crucial for optimal prediction of long-term mental disorders or PTSD during recovery [@Shalev2019; @Schultebraucks2020; @Ziobrowski2021; @Galatzer-Levy2014; @Papini2018].
 Unfortunately, such variables were not available in our cohort investigated at least six months after the accident. 
+<span custom-style = 'revision'>On one hand side, the difficulty of prediction of long-term mental health impairment in our study cohort motivates the search for new, better predictors of mental disorders in survivors of mountain sport accidents.</span>
+<span custom-style = 'revision'>On the other hand side, it underlines the importance of a broad access to psychological and psychiatric support granted to such patients as a part of multi-disciplinary treatment and care [@AmericanCollegeofSurgeons2023].</span>
 
@@
 The overall analysis inclusion rate was low and, as indicated by comparison of the excluded and analyzed participants, the results were potentially affected by a selection bias towards individuals with higher social status and more severe injuries.
+<span custom-style = 'revision'>Low response rates are unfortunately a problem of many studies tackling PTSD and mental disorders [@Ophuis2018] and may lead to under- or over-reporting of symptoms [@Angerpointner2020; @Merckelbach2019; @Takayanagi2014].</span>
+<span custom-style = 'revision'>Our comparison of demographic and accident-related features in the study cohort, Austrian population, and the national registry of alpine accidents indicates that the analyzed collective is a fairly representative sample of the general physically active population.</span>
+<span custom-style = 'revision'>However, we can not exclude that the low frequency of manifest PTSD, and clinically relevant symptoms of anxiety and depression is in part attributed to the low response rate and under-reporting of mental health problems.</span>
+<span custom-style = 'revision'>In fact, assignment of the survey participants with incomplete mental health battery to the PTS cluster suggests that mental health problems were one of the reasons of data missingness.</span>
+<span custom-style = 'revision'>Additionally, enrichment of the study sample with men and individuals of high social status, which are considered protective factors of mental health disorders, may have contributed to the low frequency of symptoms of PTSD, anxiety and depression.</span>
+<span custom-style = 'revision'>Therefore, our results may underestimate frequency of mental disorder symptoms in survivors of mountain sport accidents and have to be interpreted with caution.</span>
+<span custom-style = 'revision'>The aim of our study was to identity long-term patterns or alterations of mental health in survivors of mountain sports accidents.</span>
+<span custom-style = 'revision'>Yet, we cannot exclude the possibility, that mental disorder symptom frequency and clustering results were affected by other events than the accident of interest.</span>
+<span custom-style = 'revision'>This was in part addressed by the design of the survey questions which pointed to the mountain sport accident.</span>
+<span custom-style = 'revision'>As assessed by sensitivity analysis, our clustering structure was hardly affected by the time of between the accident and survey or prior mountain sport accidents.</span>
+<span custom-style = 'revision'>This suggests that the mental health clusters described by us reflect accident-specific alterations of mental health.</span>
+<span custom-style = 'revision'>However, to finally clarify the issue of the accident specificity of mental health alterations, follow-up studies with longitudinal design or a control group are required.</span>
 Furthermore, the study survey did not recorded potentially important explanatory factors for distinction between the mental health clusters such as rehabilitation need or ability to work. 
@@
 
-Our study represents the first attempt to characterize symptoms of PTSD and further facets of mental health following mountain sport accidents. 
+Our study characterizes symptoms of PTSD and <span custom-style = 'revision'>other</span> facets of mental health following mountain sport accidents. 
 We identified symptoms of mental health impairment in roughly one third of the investigated individuals. 
-However, we could not identify robust predictors for impaired mental health at follow up among variables available during early medical treatment. 
-Hence, considering the increasing popularity of mountain sports, psychoeducational information and low threshold access to psychological support currently seems the most appropriate strategy for successful interdisciplinary management of individuals following accidents in mountain sport.
-Psychoeducational information can include informative leaflets, information in the discharge letter, as well as access to websites or mobile applications outlining the normal reaction to a traumatic event as well as warning signs indicating the need for professional support.
-Additionally low threshold contact information for psychological support are helpful. 
-We are currently implementing such measures at our hospital.
+<span custom-style = 'revision'>This finding is significant because people performing mountain sports are generally regarded as robust in terms of physical and mental health.</span>
+<span custom-style = 'revision'>However, despite high levels of resilience, these individuals are not free of risk of symptoms of PTSD, manifest PTSD, and other mental disorders.</span>
+<span custom-style = 'revision'>Destigmatization of such mental health problems is essential in the mountaineering community.</span>
+
+<span custom-style = 'revision'>In the current analysis,</span> we could not identify robust predictors <span custom-style = 'revision'>of</span> impaired mental health at follow up among variables available during early <span custom-style = 'revision'>routine</span> medical treatment.
+<span custom-style = 'revision'>Hence, until standardized robust predictor variables and screening protocols are available,</span> low threshold access to psychological support <span custom-style = 'revision'>as well as to psychoeducational measures</span> seems the most appropriate strategy for successful interdisciplinary management of individuals following accidents in mountain sport.
 
@@
 This study was supported by a grant of the “Deutsche Gesellschaft für Berg- und Expeditionsmedizin”,  German Association for Mountain and Expedition Medicine to H.S. and K.H. 
+The project was conducted during the diploma thesis of H.S. The results of this study has been and will in the future be presented at scientific meetings.
 
@@
 The authors declare no conflict of interest.
+
+# Authors’ contributions
+
+Conceptualization: HS, MS, PP, HB, MF, BS-U, KH; 
+Data Curation: HS, PT, MS, NK, KH; 
+Formal Analysis: HS, PT; 
+Funding Acquisition: HS, KH; 
+Investigation: HS, NK, KH; 
+Methodology: PT, KH; 
+Project Administration: BS-U, KH; 
+Resources: HS, NK; 
+Software: PT; 
+Supervision: MS, PP, HB, MF, BS-U, KH;
+Validation: all authors; 
+Visualization: HS, PT, KH; Writing – Original Draft Preparation: HS, PT, KH;  
+Writing – Review & Editing: all authors
 
@@
   footnote(2, 1, 
-           value = as_paragraph('ski/snowdoard/cross-country: alpine skiing, snowboarding and cross-country skiing\nsledding: sledding or bobsled\nclimbing/hiking/mountaineering: hiking, rock and ice climbing, mountaineering, skitouring\nbiking: mountainbike, tour and road cycling.'),
+           value = as_paragraph('alpine skiing/snowboarding: alpine skiing or smowboarding in secured areas of ski resorts; aki touring/freeride: skiing or snowboarding besides secured areas of ski resorts; sledding: sledding and bob sledding; ice climbing: ice and mixed climbing; climbing: bouldering, sport and rock climbing; mountaineering: high mountain and glecier tours; biking: mountain, cross-country and road cycling; air sport: paragliding.'),
            part = 'body', 

### 2024-02-28 paper/markdown/supplementary_material.Rmd
Revised figures, tables, manuscript and supplement

@@
 Statistical hypothesis testing and correlation analyses were accomplished with [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and _rstatix_ [@Kassambara2021]. 
+Bootstrapped estimates of prevalence of mental disorder symptoms and traumatic events with their 95% confidence intervals were computed with the development package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat).
 
@@
 The psychometric battery (__Supplementary Table S\@ref(tab:tab-mental-battery)__) consisted of German versions of assessment tools for 
-anxiety (GAD-7 and GAD-2: 7- and 2-item general anxiety disorder scale) [@Spitzer2006; @Junker2023], 
-depression (PHQ: patient health questionnaire, PHQ-9, PHQ-2 and PHQ-8) [@Lowe2002; @Grafe2004; @Kroenke2002a; @Junker2023], 
+anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
+depression (PHQ: patient health questionnaire, PHQ-9 and PHQ-8) [@Lowe2002; @Grafe2004; @Kroenke2002a], 
 panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
@@
 post-traumatic stress disorder (PCL-5: PTSD checklist for DSM-5) [@Bovin2016]. 
-In the study survey, items of the PTGI and PTSD instruments were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').
+In the study survey, items of the PTGI and PTSD instruments, and questions concerning flashbacks were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').
 
 Clinically relevant symptoms of anxiety were defined as GAD-7 $\geq$ 10, 
-clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], 
+clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 10 [@Humer2022], 
 significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
 Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008]. 
-For a comparison of frequency of anxious and depressive symptoms in the study cohort with the estimates for the German population [@Junker2023], presence of anxious and depressive symptoms was considered for GAD-2 $\geq$ 3 and PHQ-2 $\geq$ 3 points, respectively.
 For a comparison of frequency and severity of depressive symptoms in the study cohort with microcensus data for the general Austrian population, mild depressive symptoms were defined as 5 - 9, moderate symptoms as 10 - 14, moderately severe symptoms as 15 - 19, and severe symptoms as $\geq$ 20 points of the PHQ-8 score [@Kroenke2002a].
@@
 
+Accident sport types were classified in accordance with the annual accident report of the Austrian Board of Mountain Safety [@OsterreichischesKuratoriumfurAlpineSicherheit2024] as follows: 
+_alpine skiing/snowboarding_ (all types of alpine ski and snowboard sports in secured areas of ski resorts: pistes, secured slopes, lifts and cable cars), 
+_ski touring/freeride_ (all types of alpine ski and snowboard sports performed in non-secured areas of ski reports, e.g. ski touring, freeride), 
+_cross-country skiing_, 
+_sledding_ (all types of sledding sports such as sledding and bob-sledding), 
+_ice climbing_ (ice and mixed climbing in the winter season), 
+_hiking_, 
+_climbing_ (sport and rock climbing, bouldering), 
+_mountaineering_ (high mountain and glacier tours), 
+_biking_ (all types of biking including road, cross-country and mountain biking), 
+_air sport_ (only paragliding present in our cohort), 
+_water sport_, and 
+_other_ (non-classifiable sport types such as ski jumping). 
+Alpine skiing/snowboarding, ski touring/freeride, cross-country skiing, sledding, and ice climbing were classified as winter sports. 
+Hiking, climbing, mountaineering, biking, air sport, and water sport were classified as summer sports. 
 Additional information on study variables and their stratification schemes are presented in __Supplementary Table S\@ref(tab:tab-analysis-vars)__.
@@
 
-Monthly estimates of population frequency of symptoms of anxiety and depression in the general German population during the study survey (`r ptsd$dataset$survey_date %>% range %>% paste(collapse = ' to ')`) were derived from the [data repository of the German mental health monitoring](https://github.com/robert-koch-institut/Hochfrequente_Mental_Health_Surveillance) [@Junker2023].
-Data on frequency and severity of depressive symptoms measured with the PHQ-8 instrument [@Kroenke2002a] in a nation-wide Austrian microcensus study were obtained from a publication by Statistik Austria [@StatistikAustria2019a].
+Data on frequency of clinically relevant anxious (GAD-7 $\geq$ 10) and depressive (PHQ-9 $\geq$ 10) symptoms in Austria in 2022 were obtained from a publication by Humer et al. [@Humer2022]. 
+Data on frequency and severity of depressive symptoms measured with the PHQ-8 instrument [@Kroenke2002a] in a pre-pandemic nation-wide Austrian 2019 microcensus study were obtained from a publication by Statistik Austria [@StatistikAustria2019a].
 
@@
 Frequency of traumatic events, manifest PTSD and PTSD symptoms in the current cohort was compared with original reports providing estimates of frequency of traumatic events and PTSD in the general population [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017] and in a cohort of Swiss mountain rescuers [@Mikutta2022].
-Frequencies of symptoms of anxiety and depression measured with the GAD-2 and PHQ-2 instruments, respectively, were compared with estimates for the general German population gauged by the German mental health monitoring during the study survey [@Junker2023].
-Frequency and severity of depressive symptoms measured by the GAD-8 tools in the study cohort [@Kroenke2002a] were compared with the Austrian microcensus estimates. 
+Frequencies of clinically relevant symptoms of anxiety and depression measured, respectively, with GAD-7 and PHQ-9, were compared with a representative sample of the general Austrian population reported by Humer et al.
+[@Humer2022].
+Frequency and severity of depressive symptoms measured by the PHQ-8 tool in the study cohort [@Kroenke2002a] were compared with the Austrian 2019 microcensus estimates [@StatistikAustria2019a]. 
 In those analyses, bootstrap with B = 1000 resamples was used for computation of the bias-corrected and accelerated (BCA) 95% confidence intervals (function `bmap()`, package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat)).
@@
 The choice of the clustering algorithm was motivated by 
-its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares), 
+its good explanatory performance measured by the fraction of explained clustering variance (ratio of the total between-cluster sum of squares to the total sum of squares), 
 good separability between the clusters measured by mean silhouette width [@Rousseeuw1987], 
 and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__.
-The explained variance, mean silhouette width and cross-validated accuracy statistics were computed with the methods `var()`, `silhouette()` and `cv()` from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
-The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020].
+The explained variance, mean silhouette width and cross-validated accuracy statistics were computed with the methods `summary()` and `cv()` from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
+The number of clusters was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020].
 By this means, three mental health clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).
@@
 Assignment of the training subset observations to the mental health clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
-This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
-training (variance = `r signif(semi_clust$stats$frac_var[1], 2)`, 
-average silhouette = `r signif(semi_clust$stats$sil_width[1], 2)`) and 
-test subset (variance = `r signif(semi_clust$stats$frac_var[2], 2)`, 
-average silhouette = `r signif(semi_clust$stats$sil_width[2], 2)`). 
 Comparably good separation of the clusters could be discerned in the training and test subsets by 
@@
 pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
-Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
+Additionally, the reproducibility of the clustering structure was assessed by 
+mean silhouette width as a measure of general cluster separation (training: `r signif(semi_clust$stats$sil_width[1], 2)`, test: `r signif(semi_clust$stats$sil_width[2], 2)`) [@Rousseeuw1987], 
+fraction of observations with negative silhouette widths as a metric of misclassification rate (training: `r signif(semi_clust$stats$frac_misclassified[1], 2)`, test: `r signif(semi_clust$stats$frac_misclassified[2], 2)`) [@Rousseeuw1987],
+fraction of explained clustering variance defined as ratio of the total between-cluster sum of squares to the total sum of squares (training: `r signif(semi_clust$stats$frac_var[1], 2)`, test: `r signif(semi_clust$stats$frac_var[2], 2)`),
+and fraction of 5-nearest neighbors placed in the same cluster as a measure of neighborhood preservation (training: `r signif(semi_clust$stats$frac_np[1], 2)`, test: `r signif(semi_clust$stats$frac_np[2], 2)`) [@Venna2001].
+These statistics were computed with the `summary()` method from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package. 
+The results of the comparison of the numeric statistics are presented in __Supplementary Figure S\@ref(fig:fig-cosine)A__. 
+Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)B__). 
 An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental health clusters (i.e.
 neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g.
-neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__).
+neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)C__).
 Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables between the mental health clusters in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).
@@
 The effects of extending or depletion of the data set on the quality of the clustering structure was evaluated by comparing fractions of the explained clustering variance between the clustering solutions for the particular data set and for the analysis cohort [@Thabane2013].
-Additionally, distribution of sizes of the neutral, PTG and PTS clusters expressed as percentage of all clustered observations was compared (__Supplementary Figure S\@ref(fig:fig-cluster-se)__ and __S\@ref(fig:fig-cluster-miss)__).
+Additionally, sizes of the neutral, PTG and PTS clusters expressed as percentage of all clustered observations were compared (__Supplementary Figure S\@ref(fig:fig-cluster-se)__ and __S\@ref(fig:fig-cluster-miss)__).
 
@@
 flextable::flextable(suppl_tables$no_response) %>% 
-  set_widths(c(3.2, 5.8, 5.8, 2.5, 2.2)) %>% 
+  set_widths(c(2.8, 6.3, 6.3, 2.5, 2.2)) %>% 
    footnote(1, 1, 
@@
 
-__Supplementary Figure \@ref(fig:fig-accident). Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.__
+__Supplementary Figure S\@ref(fig:fig-accident). Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.__
 
@@
 In the comparison of sport types, non-sportive mountain accidents (e.g.
-during hunting or forestry) were excluded and sport types in the study cohort re-classified according to the scheme of the Austrian Board of Mountain Safety.
+during hunting or forestry) were excluded from the national statistic.
+Non-classifiable sport types (category 'other') in the study cohort were excluded from the analysis presented in (C).
 In the comparison of the age class distribution, only individuals older than 20 years were included.
@@
 
-_alpine skiing: alpine skiing and snowboarding in a ski resort and lift accidents; ski touring: ski touring and freeride skiing; air: air sports and paragliding; water: water sports._
+_alpine skiing: alpine skiing and snowboarding in secured areas of ski resorts; ski touring/freeride: alpine skiing and snowboarding performed besides secured areas of ski resorts; sledding: sledding and bob sledding; ice climbing: ice and mixed climbing; climbing: bouldering, sport and rock climbing; mountaineering: high mountain and glacier tours; biking: road, cross-country and mountain biking._
 
@@
 
-_(A) Comparison of frequency of anxious and depressive symptoms measured by the GAD-2 and PHQ-2 instruments in the study cohort with frequency of anxious and depressive symptoms obtained for the general German population by monthly mental health monitoring._
+_(A) Comparison of frequency of anxious and depressive symptoms measured by the GAD-7 and PHQ-9 instruments in the study cohort with frequency of anxious and depressive symptoms reported in a representative sample of Austrian population in 2022 by Humer and colleagues._
 
@@
 
-_(C) Comparison of resilient comping scores in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta at al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.)_
+_(C) Comparison of resilient comping scores in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta et al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.)_
 
@@
 
-_(A) Percentages of observations in the training and test subsets of the study cohort in the mental health clusters. Numbers of complete observations are indicated in the Y axis._
+_(A) Comparison of numeric statistics of quality of clustering structures between the training and test subset of the study cohort.
+Mean silhouette width served as a measure of cluster separation.
+Fractions of observations with negative values of the silhouette width statistics were used to gauge misclassification rates.
+Explained clustering variance was defined as ratio of the total between-cluster sum of squares to the total sum of squares.
+Neighborhood preservation was measured by a fraction of the 5-nearest neighbors placed in the same cluster.
+SS: sum of squares, NN: nearest neighbors._
 
-_(B) Cosine cross-distances between the mental health clusters in the training and test subsets of the study data set visualized as a heat map.
+_(B) Percentages of observations in the training and test subsets of the study cohort in the mental health clusters. Numbers of complete observations are indicated in the Y axis._
+
+_(C) Cosine cross-distances between the mental health clusters in the training and test subsets of the study data set visualized as a heat map.
 Average cross-distances with 95% percentile range are indicated in the tiles._
@@
 
-_Distribution of employment status, frequency of mountain sport accidents in the past, distribution of the accidents in the seasons, accident sport types, frequency of being alone during the accident and number of injured persons in the mental health clusters.
+_Distribution of employment status, frequency of mountain sport accidents in the past, distribution of the accidents in the seasons, frequency of being alone during the accident and number of injured persons in the mental health clusters.
 Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic.
@@
 Numbers of observations in the clusters are presented in the X axes._
+
+\newpage
+
+```{r fig-cluster-sport, fig.width = figur::convert(suppl_figures$cluster_sport, to = 'in')$w, fig.height = figur::convert(suppl_figures$cluster_sport, to = 'in')$h, fig.cap = 'Mountain sport types in the clusters.'}
+
+suppl_figures$cluster_sport$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-cluster-sport). Mountain sport types in the clusters.__
+
+_The global difference in distribution of accident sport types between the mental health clusters was compared by $\chi^2$ test with Cramer's V effect size statistic.
+Post-hoc testing for differences in frequency of particular sport types between the clusters was done with Holm-corrected $\chi^2$ test with Cramer's V effect size statistic.
+Percentages of sport types within the clusters are presented in bar plots.
+Effect size statistic and p value for the global difference are indicated in the plot caption.
+Results of post-hoc tests are displayed in the Y axis.
+Numbers of observations in the clusters are shown in the plot legend._
 

### 2024-03-04 import.R
Revision and Author's Feedback

@@ -437,7 +437,7 @@
                                     'skating' = 'other'; 
                                     'mountaineering' = 'mountaineering'; 
                                     'skitouring' = 'ski touring/freeride'; 
-                                    'paragliding' = 'air sport'; 
+                                    'paragliding' = 'air sports'; 
                                     'sport climbing/bouldering' = 'climbing'; 
                                     'snowboarding' = 'alpine skiing/snowboarding'; 
                                     'ice climbing' = 'ice climbing'; 
@@ -446,7 +446,7 @@
                                     'figeln' = 'ski touring/freeride'; 
                                     'ski jumping' = 'other'; 
                                     'other winter' = 'other'; 
-                                    'other water' = 'water sport'"), 
+                                    'other water' = 'water sports'"), 
            sport_type = factor(sport_type, 
                                c('alpine skiing/snowboarding', 
                                  'ski touring/freeride', 
@@ -457,8 +457,8 @@
                                  'climbing', 
                                  'mountaineering', 
                                  'biking', 
-                                 'air sport', 
-                                 'water sport', 
+                                 'air sports', 
+                                 'water sports', 
                                  'other')))
   
 # Counts of injured body regions ------
@@ -795,7 +795,7 @@
   
   ptsd$cleared <- ptsd$cleared %>% 
     mutate(phq_events_total_class = cut(phq_events_total, 
-                                        c(-Inf, 10, Inf), 
+                                        c(-Inf, 9, Inf), 
                                         c('negative', 'positive')))
   
 # PHQD psychical satisfaction, depression and anxiety ------

### 2024-03-04 literature scripts/kurasi.R

@@ -86,26 +86,16 @@
   lit_kurasi$age_class$own_data <- ptsd$dataset %>% 
     filter(age > 20) %>% 
     mutate(age_class = cut(age, 
-                           c(-Inf, 30, 40, 50, 60, 70, 80, 90, Inf), 
+                           c(-Inf, 30, 60, Inf), 
                            c('21 - 30', 
-                             '31 - 40', 
-                             '41 - 50', 
-                             '51 - 60', 
-                             '61 - 70', 
-                             '71 - 80', 
-                             '81 - 90', 
-                             '91+'))) %>% 
+                             '31 - 60', 
+                             '60+'))) %>% 
     .$age_class
   
   lit_kurasi$age_class$kurasi_data <- 
     c(rep('21 - 30', 2033), 
-      rep('31 - 40', 1691), 
-      rep('41 - 50', 1890), 
-      rep('51 - 60', 2386), 
-      rep('61 - 70', 1559), 
-      rep('71 - 80', 640), 
-      rep('81 - 90', 223), 
-      rep('91+', 10))
+      rep('31 - 60', 1691 + 1890 + 2386), 
+      rep('60+', 1559 + 640 + 223 + 10))
   
   lit_kurasi$age_class$data <- 
     map2_dfr(c('cohort', 'Austria'), 
@@ -114,13 +104,8 @@
                      age_class = .y)) %>% 
     mutate(age_class = factor(age_class,
                               c('21 - 30', 
-                                '31 - 40', 
-                                '41 - 50', 
-                                '51 - 60', 
-                                '61 - 70', 
-                                '71 - 80', 
-                                '81 - 90', 
-                                '91+')))
+                                '31 - 60', 
+                                '60+')))
   
 # Detailed sport type -------
   
@@ -134,11 +119,11 @@
   
   lit_kurasi$sport_detail$kurasi_data[['climbing']] <- 544 + 11 + 4
   
-  lit_kurasi$sport_detail$kurasi_data[['air sport']] <- 295
+  lit_kurasi$sport_detail$kurasi_data[['air sports']] <- 295
   
   lit_kurasi$sport_detail$kurasi_data[['mountaineering']] <- 90
   
-  lit_kurasi$sport_detail$kurasi_data[['water sport']] <- 45
+  lit_kurasi$sport_detail$kurasi_data[['water sports']] <- 45
   
[…]

### 2024-03-04 paper scripts/supplementary.R

@@ -11,9 +11,8 @@
   insert_msg('Figure S1: accident characteristic')
   
   suppl_figures$accident <- lit_kurasi$panels[c("sport_detail", 
-                                                "accident_month", 
-                                                "age_class")] %>% 
-    map2(., c(68, 25, 35), 
+                                                "accident_month")] %>% 
+    map2(., c(68, 25), 
          ~.x + 
            expand_limits(y = .y) + 
            theme(axis.title.x = element_blank()))
@@ -23,16 +22,22 @@
   
   suppl_figures$accident <- suppl_figures$accident %>% 
     plot_grid(plotlist = ., 
-              nrow = 3, 
+              nrow = 2, 
               axis = 'tblr',
-              rel_heights = c(1.2, 1, 1), 
-              labels = LETTERS, 
+              rel_heights = c(1.3, 1), 
+              labels = c('B', 'C'), 
               label_size = 10) %>%
+    plot_grid(plot_grid(lit_kurasi$panels$age_class + 
+                          expand_limits(y = 65), 
+                        ncol = 2, 
+                        rel_widths = c(0.65, 0.35)), ., 
+              nrow = 2, 
+              rel_heights = c(1, 2.3)) %>% 
     as_figure(label = 'figure_s1_accident_characteristic', 
               ref_name = 'accident', 
-              caption = paste('Mountain sport type, accident month, and age of', 
-                              'the mountain accident victim in the study cohort', 
-                              'and Austria.'), 
+              caption = paste('Age, mountain sport type, and accident month', 
+                              'in the study cohort and Austrian national', 
+                              'registry of mountain accidents.'), 
               w = 180, 
               h = 230)
 
@@ -97,13 +102,10 @@
     list(lit_plots$trauma_plots$general,
          lit_plots$ptsd_plot$general + 
            scale_x_continuous(limits = c(-0.5, 10)), 
-         plot_grid(lit_plots$trauma_plots$alpine, 
-                   get_legend(lit_plots$ptsd_symptom_plot), 
-                   nrow = 2, 
-                   rel_heights = c(2, 1)), 
          lit_plots$ptsd_symptom_plot + 
            theme(legend.position = 'none') + 
-           scale_x_continuous(limits = c(-0.5, 25))) %>% 
+           scale_x_continuous(limits = c(-0.5, 25)), 
+         get_legend(lit_plots$ptsd_symptom_plot)) %>% 
[…]

### 2024-03-04 paper/markdown/review_response.Rmd
Figure and Table update

@@
 
-Thank you very much for the positive and very thoughtful review of our manuscript. We appreciate a lot the reviewers’ comments and suggestions to improve our manuscript. 
+Thank you very much for the positive and very thoughtful review of our manuscript. 
+We appreciate a lot the reviewers’ comments and suggestions to improve our manuscript. 
 We incorporated them in the revised text.
 Please find below the point-to-point response to the reviewers’ issues. 
-Since the reviewers requested more information and literature evidence, we also deleted few less important references from the initial submission to keep the bibliography within reasonable limits. 
-We also introduced three major changes to the manuscript text and data independently of the reviewers' requests:
+Since the reviewers requested more information in some points, we also deleted few less important references from the initial submission to keep the bibliography within reasonable limits. 
+We also included additional data and performed the following additional analyses:
 
@@
 Please note, that the re-classification of the sport types requires us to re-train machine learning models of the mental health cluster assignment.
-For this reason, minor differences in performance of the models were observed as compared with the initial submission (e.g.
-in __Figure 7__).
-Still, they did not affected the general conclusions.
+Minor differences in performance of the models were observed as compared with the initial submission (e.g.
+in __Figure 7__), which did not affected the general conclusions.
 
 2.
-We identified clinically relevant symptoms of anxiety and depression with the uniform $\geq$ 10 point cutoff of, respectively, the GAD-7 and PHQ-9 tools in accordance with the work by Humer et al.
+We now identify clinically relevant symptoms of each of anxiety, depression and somatization with the uniform $\geq$ 10 point cutoff of, respectively, the GAD-7, PHQ-9 and PHQ-15 tools in accordance with the work by Humer et al.
+and Kroenke et al.
+[@Humer2022; @Kroenke2002].
 This change allowed us for a direct comparison of frequency of clinically relevant anxious and depressive symptoms in our manuscript and the estimates for the general Austrian population at the time of our study survey [@Humer2022].
@@
 Unfortunately this is a challenge of many studies tackling with mental disorders such as PTSD, anxiety and depression. 
+Also, as stated now more clearly in the last paragraph of __Introduction__ and __Methods/Analysis endpoints__, the goal of the study was not to estimate prevalence of mental disorder in a representative sample of survivors of mountain sport accidents, but to explore phenotypes of mental health in this population in a _qualitative manner_.
+We now provide this information at several points in the text.
 
-To assess the potential effects of the low response rate on the cohort composition we included an extensive comparison of the analysis cohort with invited individuals and participants excluded due to incomplete psychometric data in the initial submission.
+Nevertheless for better interpretation of our data and to put them into the broader context of the literature, we performed an extensive comparison of the analysis cohort with invited individuals and participants excluded due to incomplete psychometric data in the initial submission.
 In the revised manuscript we investigated the potential selection bias in more detail, by comparing sociodemographic features of the cohort with the general Austrian population and by comparing characteristic of the mountain sport accident in the study cohort with a nation-wide collective of Austrian mountain accident victims.
@@
 Yet the effect size of the difference was again small (Cramer’s V ≤ 0.14). 
-Taken together, these analyses suggest that there is indeed a weak but significant selection bias towards men, well educated, professionally active and healthy individuals, and victims of winter sport accidents.
+Taken together, these analyses suggest that there is indeed a small effect size but significant selection bias towards men, well educated, professionally active and healthy individuals, and victims of winter sport accidents.
 We point this out in the limitation section of __Discussion__ of the revised manuscript.
@@
 for Swiss mountain rescuers [@Mikutta2022] (__Supplementary Figure S4B__).
-Furthermore, frequency of clinically relevant anxious (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'anxiety')`) and depressive symptoms (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'depression')`) assessed, respectively by the GAD-7 and PHQ-9 instruments, was significantly lower in the study cohort as compared with the 2022 estimates for the general Austrian population reported by Humer et al.
+Furthermore, frequency of clinically relevant anxiety (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'anxiety')`) and depressive symptoms (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'depression')`) assessed, respectively by the GAD-7 and PHQ-9 instruments, was significantly lower in the study cohort as compared with the 2022 estimates for the general Austrian population reported by Humer et al.
 (anxiety: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'anxiety')`, depression: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'depression')`, __Supplementary Figure S5A__) [@Humer2022].
@@
 We absolutely agree that there is a selection bias in this study. 
+Of note, the primary goal of the study was the _qualitative_ characteristic of patterns of mental health in survivor of mountain sport accidents (see: last paragraph of __Introduction__ and __Methods/Analysis endpoints__).
+We did not aim at determining prevalences of mental disorder symptoms in a representative sample of mountain accident victims in the Alps. 
 This was already pointed out in the limitations section of __Discussion__ of the initial submission. 
-As described in more detail in __Response 4__ we took now additional measures to better characterize the potential selection bias. 
+As described in more detail in __Response 4__ we took now additional measures to better characterize the potential selection bias, in order to allow the reader to put our data better into perspective of the existing data and literature.
 From comparisons of our sample with the general Austrian population (__Supplementary Table S4__) and the collective of mountain accident victims in Austria in 2023 (__Supplementary Figure S1__), we infer that the study cohort consisted of significantly more males, highly educated and healthy individuals, and victims of winter sport accidents that expected for the general population.
-Effect size of those differences was weak (Cramer's V effect size statistic $\leq$ 0.3)
+Effect size of those differences was small (Cramer's V effect size statistic $\leq$ 0.3)
 We elaborate on those findings in __Results/Characteristic of the study cohort__ and summarize these findings on the selection bias in the limitation part of __Discussion__. 
@@
 Next, we addressed the issue of under- or over-reporting of mental health symptoms in the study sample, which may be attributed to the selection bias of the sample. 
-To this end, we compared frequency of traumatic events, manifest PTSD, anxious and depressive symptoms in the study cohort with estimates for the general population (__Supplementary Figure S4__). 
+To this end, we compared frequency of traumatic events, manifest PTSD, anxiety and depressive symptoms in the study cohort with estimates for the general population (__Supplementary Figure S4__). 
 As outlined in __Response 4__ in more detail, frequency of traumatic events and manifest PTSD was at the lower tail of the reported prevalence and comparable with figures for the German population published by Hauffa and colleagues [@Hauffa2011].
-Anxious symptoms measured with the GAD-7 instrument were less common in the study cohort than in the Austrian population during the survey [@Humer2022]. 
-In turn, depending on the tool and time point of survey, overall depressive symptoms and symptoms of moderate-to-severe depression were lower or comparable with the general Austrian population (__Supplementary Figure S5__).
+Symptoms of anxiety and depression were less common in the study cohort than in the Austrian population during the survey [@Humer2022] (__Supplementary Figure S5__). 
 We discuss in __Results/Characteristic of the study cohort__ and in __Discussion__, that the overall frequency of mental disorder symptoms in our cohort is, in spite of the accident, comparable with or event lower than in the general population.
@@
 Alone in Austria with 9 million inhabitants, there are over 900000 members of diverse alpine clubs [@VAVOE2024], which amounts to an enormous community of mountain sport enthusiasts, professional and free-time athletes.
-The dark side of popularity of mountain sports are accidents whose number exceeded `r table(lit_kurasi$data$sport_detail$subset)['Austria'] %>% signif(2)` in 2023 in the nation-wide scale, according to the official counts by the mountain safety board [@OsterreichischesKuratoriumfurAlpineSicherheit2024].
+The flip side of popularity of mountain sports are accidents whose number exceeded `r table(lit_kurasi$data$sport_detail$subset)['Austria'] %>% signif(2)` in 2023 in the nation-wide scale, according to the official counts by the mountain safety board [@OsterreichischesKuratoriumfurAlpineSicherheit2024].
 By comparison, this number corresponds approximately to the cumulative yearly incidence of prostate and breast cancer in Austria [@StatistikAustria2023a]. 
@@
 This is the prime point of novelty of our work.  
-Despite the positive effects of mountain sports on physical and mental health and the common perception of sportspersons as mentally strong individuals, our data suggest that a large subset of mountain accident victims, the PTS cluster, is prone to long-term mental health deterioration.
+Despite the positive effects of mountain sports on physical and mental health and the common perception of sportspersons as mentally strong individuals, our data suggest that a subset of mountain accident victims, the PTS cluster, is prone to long-term mental health consequences.
 This is another novel finding of out work. 
-As pointed out by the reviewer, there are many reports on markers and diagnostic tools of post-traumatic stress [@Sandweiss2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021]. 
-Yet, we were not able to establish a reliable set of predictors of mental health disorders after the mountain sport accident. 
-This negative finding, which has not been reported so far, has profound clinical implications and underscores the need of widely accessible psychological support in management of this specific accidents.
-We stress there three novel findings of our work in the first paragraph of __Discussion__ and the conclusion part of the revised manuscript.
+As pointed out by the reviewer, there are many reports on markers and diagnostic tools of post-traumatic stress [@Sandweiss2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021], however, the early identification of individuals at risk of PTSD remains still challenging and no universally valid prediction models are yet available.
+In the current sample, we were not able to establish a reliable set of predictors of mental health disorders after the mountain sport accident. 
+This negative finding, which has not been reported so far, underlines the importance of easily accessible psychological and psychiatric support in management of these specific accidents. 
+Additionally, the new guidelines by the American College of Surgeons concerning psychological and psychiatric management of acute trauma patients [@AmericanCollegeofSurgeons2023] could be applied and tested in this patient group in future studies.
+
+We describe the gap of of the existing knowledge on mental health in the context of mountain sports in the final part of __Introduction__ and stress the three novel findings of our work in the first paragraph of __Discussion__ and the conclusion part of the revised manuscript.
 
@@
 In turn, for the PCL-5 instrument measuring PTSD symptoms, the missingness rates were the lowest (< 50% of excluded respondents). 
-We can hence only speculate, that the excluded survey participants found questions of the questionnaire concerning their fears, concerns, life satisfaction, and mental health problems too private. 
+We can hence only speculate, that the excluded survey participants found questions of the questionnaire concerning their fears, concerns, life satisfaction, and mental health problems too distressful.
 This notion was also supported by results of the sensitivity analysis of the mental health clusters, where the majority of 'imputable' participants with missing psychometric data were assigned to the PTS cluster (__Supplementary Figure S11__).
@@
 
-_Frequencies of missing observations per mental health assessment battery are expressed as percentages of the survey respondents with incomplete psychometric data and hence excluded from the analysis.__
+_Frequencies of missing observations per mental health assessment battery are expressed as percentages of the survey respondents with incomplete psychometric data and hence excluded from the analysis._
 
@@
 
-Based on the comparison of the accident sports in the study cohort and the nation-wide collective of accident victims in 2023 (__Supplementary Figure S1__), we can conclude that the global difference in sport frequency distribution was significant but weak ($\chi^2$ test with Cramer's V effect size statistic: `r filter(lit_kurasi$test, variable == 'sport_detail')$plot_cap`).
+Based on the comparison of the accident sports in the study cohort and the nation-wide collective of accident victims in 2023 (__Supplementary Figure S1__), we can conclude that the global difference in sport frequency distribution was significant but small ($\chi^2$ test with Cramer's V effect size statistic: `r filter(lit_kurasi$test, variable == 'sport_detail')$plot_cap`).
 In particular, we could not observe higher percentages of sports generally perceived as risky such as ski touring, ice climbing, climbing, mountaineering or paragliding in the analyzed cohort as compared with the Austrian data set.
 This let us assume, that there are no large subsets of the cohort exposed to particularly high risk of injury. 
+However, we need to stress again that our cohort was not generated in a way to be a representative sample of mountain accident victims.
 
@@
 
-Unfortunately, the initial study protocol and patient's consent form did not include an open data strategy, and the data set still contains information which may enable identification of an individual such as localization of injury, accident and medical treatment date.
+Unfortunately, the initial study protocol and patient's consent form did not include an open data strategy, and the data set still contains contains very sensitive mental health data in conjunction with information which may potentially enable identification of an individual such as localization of injury, accident and medical treatment date.
 For this reason, our ethics committee did not agree on publication of the data set. 
-Its anonymous parts will be made available at request.
+Its anonymous parts will be made available to individual researchers at request.
 

### 2024-03-04 paper/markdown/manuscript.Rmd
Figure and Table update

@@
 __Background:__ <span custom-style = 'revision'>The restorative effect of physical activity in alpine environments on mental and physical health is recognized.
-However, risk of accidents and post-accident mental health problems is inherent to every sport, including mountain sport activity.</span>
+However, risk of accidents and post-accident mental health problems is inherent to every sport, including mountain sports.</span>
 We aimed to characterize mental health in individuals following mountain sport accidents requiring professional medical management.
 
-__Methods:__ Adult victims of mountain sport accidents treated at the hospital of the Medical University of Innsbruck (Austria) between 2018 and 2020 completed a cross-sectional survey at a median `r round(median(ptsd$dataset$obs_time)/30.436)` months following the admission (n = `r nrow(ptsd$dataset)`).
+__Methods:__ Adult victims of mountain sport accidents treated at the hospital of the Medical University of Innsbruck (Austria) between 2018 and 2020 completed a cross-sectional survey at least six months following the admission (median `r round(median(ptsd$dataset$obs_time)/30.436)` months, n = `r nrow(ptsd$dataset)`).
 Symptoms of post-traumatic stress disorder (PTSD, PCL-5), anxiety, depression, and somatization (PHQ), resilience (RS-13), sense of coherence (SOC-9L), post-traumatic growth (PTGI) and quality of life (EUROHIS-QOL), as well as sociodemographic and clinical information were obtained from an online survey and extracted from electronic health records.
@@
 Available data on the consequences of accidents during mountain sports on mental health are scarce. 
-Prior studies were performed on mountain guides, rescuers or avalanche accident survivors, mostly lack clinical data<span custom-style = 'revision'>, and focus on single aspects of mental health such as trauma exposure or PTSD</span> [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004].
+Prior studies were performed on mountain guides, rescuers or avalanche accident survivors, mostly lack clinical data<span custom-style = 'revision'>, and focus predominantly on single aspects of mental health such as PTSD</span> [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004].
 <span custom-style = 'revision'>A characteristic of multiple facets of mental health following a mountain sport accident, including pathological and salutogenic features, is not clear at the moment.</span>
@@
 
-<span custom-style = 'revision'>Here, we aimed to identify long-term, persistent patterns of mental health in a sample of `r nrow(ptsd$dataset)` individuals treated at the hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident.
-Such patterns were defined by symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in a semi-supervised clustering analysis.</span>
-<span custom-style = 'revision'>Subsequently, by statistical hypothesis testing and machine learning, we searched for early sociodemographic and clinical factors that characterize the newly defined mental health subsets of patients with a special focus on individuals prone to mental health impairment.</span>
+<span custom-style = 'revision'>Here, we aimed to identify long-term patterns of mental health in a sample of `r nrow(ptsd$dataset)` individuals treated at the hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident.
+Such patterns were investigated at least 6 months following the accident and defined by symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in a semi-supervised clustering analysis.</span>
+<span custom-style = 'revision'>Subsequently, by statistical hypothesis testing and machine learning, we searched for early sociodemographic and clinical predictors that characterize these health subsets.</span>
 
@@
 Individuals fulfilling the inclusion criteria: 
-hospital admission at least 6 months prior to the start of the study, 
+hospital admission at least six months prior to the start of the study, 
 age $\geq$ 18 years, 
@@
 anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
-depression (PHQ-9, <span custom-style = 'revision'> and PHQ-8: 9- and 8-item patient health questionnaires</span> for depressive symptoms) [@Grafe2004; @Kroenke2002a], 
+depression (PHQ-9: 9-item patient health questionnaire for depressive symptoms) [@Grafe2004; @Kroenke2002a], 
 panic (PHQ-panic module) [@Grafe2004], 
@@
 <span custom-style = 'revision'>Items of the PTGI and PTSD instruments, and questions concerning flashbacks were provided with captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your mountain sport accident').</span>
-The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).
+The tools displayed good-to-excellent consistency (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).
 
@@
 <span custom-style = 'revision'>depression (PHQ-9 $\geq$ 10) [@Humer2022],</span> 
-somatization (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
+somatization (PHQ-15 $\geq$ 10) [@Kroenke2002] and 
 resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73) [@Leppert2008] 
@@
 The training and test participant subsets (3:1 size ratio) were obtained by random splitting which minimized differences in in sociodemographic, medical history, clinical and accident- and injury-related variables assessed by Gower distance (__Figure \@ref(fig:fig-consort)__).
-The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as assessed by Hopkins metrics of `r signif(pca$clust_tendency$data$hopkins_stat, 2)` indicative of substantial spontaneous clustering tendency [@Kassambara2020].
+The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as assessed by Hopkins metrics of `r signif(pca$clust_tendency$data$hopkins_stat, 2)` indicative of substantial spontaneous clustering tendency.
 
@@
 
-<span custom-style = 'revision'>The primary analysis endpoint was identification of subsets of patients differing in long-term symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life following a mountain sport accident.
+<span custom-style = 'revision'>The primary analysis endpoint was identification of subsets of patients differing in symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life at least six months following a mountain sport accident.
 This endpoint was pursued by semi-supervised clustering analysis.</span>
-<span custom-style = 'revision'>The secondary analysis endpoints were (1) comparison of demographic, socioeconomic and clinical characteristics of these newly defined subsets in a search for predictors of mental health disorders following a mountain sport accident, and (2) identification of</span> patients at risk of mental disorders based on a set of explanatory variables available during acute medical treatment after mountain sport accidents.
+<span custom-style = 'revision'>The secondary analysis endpoint was comparison of demographic, socioeconomic and clinical characteristics of these newly defined subsets in a search for predictors of mental health disorders following a mountain sport accident.</span>
 <span custom-style = 'revision'>The secondary analysis endpoints were addressed by statistical hypothesis testing and machine learning.</span>
@@
 Differences in numeric variables were analyzed by Mann-Whitney or Kruskal-Wallis test with r or $\eta^2$ effect size statistics. 
-Differences in categorical variable distribution were assessed by $\chi^2$ test with Cramer V effect size statistic [@Kassambara2021].
+Differences in categorical variable distribution were assessed by $\chi^2$ test with Cramer V effect size statistic.
 
-The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
+The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Schubert2019]. 
 This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares), separation between clusters (average silhouette width) [@Rousseeuw1987] and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical clustering and KMEANS algorithms.
-The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
+The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic. 
 The training subset observations were assigned to the clusters with an inverse distance-weighted 27-nearest neighbor classifier.
@@
 Multi-parameter machine learning classifiers of the mental health cluster assignment were trained with 
-the canonical random forest algorithm [@Wright2017; @Breiman2001], 
+the canonical random forest algorithm [@Breiman2001], 
 regularized neural network with a single hidden layer [@Ripley2014], 
-support vector machine algorithm with radial kernel [@Karatzoglou2004; @Weston1998], 
-recursive partitioning [@Therneau2022; @Breiman2017], 
-multinomial elastic net regression [@Zou2005; @Friedman2010], 
-and conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022]. 
-The cluster assignment was predicted for the test subset and the classifiers' performance at predicting the cluster assignment was assessed by the accuracy and $\kappa$ statistics, Brier score as well as sensitivity and specificity of predicted assignment to the post-traumatic stress cluster [@Cohen1960; @Brier1950; @Kuhn2008].
+support vector machine algorithm with radial kernel [@Weston1998], 
+recursive partitioning [@Breiman2017], 
+multinomial elastic net regression [@Zou2005], 
+and conditional random forest algorithm [@Hothorn2006]. 
+The cluster assignment was predicted for the test subset and the classifiers' performance at predicting the cluster assignment was assessed by the accuracy and $\kappa$ statistics, Brier score as well as sensitivity and specificity of predicted assignment to the post-traumatic stress cluster.
 
@@
 suffered less frequently from persistent physical health consequences of the accident than the analyzed participants <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'accident_aftermath')$plot_cap`)</span>.
-Effect sizes of the differences between the included and excluded participants were weak <span custom-style = 'revision'>(__Supplementary Table S5__).</span>
+Effect sizes of the differences between the included and excluded participants were <span custom-style = 'revision'>small (__Supplementary Table S5__).</span>
 
@@
 `r sum(get_percent(ptsd$dataset, 'traumatic_number')$percent[c('2', '3+')])`% participants had been exposed to two or more traumatic events (__Table \@ref(tab:tab-cohort-demo)__). 
-
-<span custom-style = 'revision'>The age structure was similar, while males were significantly more frequent in the study cohort as compared with the general Austrian population (`r filter(aut_stats$test, variable == 'sex')$plot_cap`) [@StatistikAustria2023].</span>
-<span custom-style = 'revision'>Individuals with secondary and tertiary education (`r filter(aut_stats$test, variable == 'education')$plot_cap`), as well as professionally active people and students (`r filter(aut_stats$test, variable == 'employment_status')$plot_cap`) were more frequent in the investigated sample than in the general Austrian population [@StatistikAustria2020; @StatistikAustria2021].</span>
-<span custom-style = 'revision'>In turn, rates of active smokers (`r filter(aut_stats$test, variable == 'smoking_status')$plot_cap`), and physical or mental illness were significantly lower in the study cohort (`r filter(aut_stats$test, variable == 'illness')$plot_cap`).</span>
+<span custom-style = 'revision'>The age structure was similar to the general Austrian population [@StatistikAustria2023].</span> 
+<span custom-style = 'revision'>Males (`r filter(aut_stats$test, variable == 'sex')$plot_cap`), individuals with secondary and tertiary education (`r filter(aut_stats$test, variable == 'education')$plot_cap`), as well as professionally active people and students (`r filter(aut_stats$test, variable == 'employment_status')$plot_cap`) were more frequent in the investigated sample than in the general Austrian population [@StatistikAustria2020; @StatistikAustria2021].</span>
+<span custom-style = 'revision'>Active smoking (`r filter(aut_stats$test, variable == 'smoking_status')$plot_cap`), and physical or mental illness were significantly less frequent in the study cohort (`r filter(aut_stats$test, variable == 'illness')$plot_cap`).</span>
 <span custom-style = 'revision'>Effect sizes of these significant differences were small (__Supplementary Table S6__).</span>
@@
 <span custom-style = 'revision'>Mountain sports perceived as particularly risk-associated such as alpine skiing/snowboarding in non-secured areas (`r get_percent(ptsd$dataset, 'sport_type')$percent['ski touring/freeride']`%), ice-climbing (`r get_percent(ptsd$dataset, 'sport_type')$percent['ice climbing']`%), climbing (`r get_percent(ptsd$dataset, 'sport_type')$percent['ski touring/freeride']`%), mountaineering (`r get_percent(ptsd$dataset, 'sport_type')$percent['mountaineering']`%) and paragliding (`r get_percent(ptsd$dataset, 'sport_type')$percent['air sport']`%) comprised a minute fraction of all accident sports in the study cohort (__Table \@ref(tab:tab-cohort-accident)__)</span>
-<span custom-style = 'revision'>The profile of accident sports in the analyzed cohort differed significantly with weak effect size from distribution of sports in the Austrian statistic of mountain accidents in 2023 [@OsterreichischesKuratoriumfurAlpineSicherheit2024] (`r filter(lit_kurasi$test, variable == 'sport_detail')$plot_cap`).</span>
+<span custom-style = 'revision'>As compared with the Austrian statistic of mountain accidents in 2023 [@OsterreichischesKuratoriumfurAlpineSicherheit2024], young adults aged $\leq$ 30 years were significantly less frequent in the study cohort (`r filter(lit_kurasi$test, variable == 'age_class')$plot_cap`).</span>
 <span custom-style = 'revision'>Frequency of alpine skiing/snowboarding, cross-country skiing, biking, and water sport accidents was significantly higher in the analyzed sample as compared with the national registry.</span>
-<span custom-style = 'revision'>Percentages of ski/snowboard accidents in non-secured areas and hiking accidents were in turn significantly lower in the study cohort than in the national registry.</span>
-<span custom-style = 'revision'>Additionally, accidents in the winter season were more common  while accidents in the summer season were less frequent in the study cohort than in the national statistic (small effect size, `r filter(lit_kurasi$test, variable == 'accident_month')$plot_cap`).</span>
-<span custom-style = 'revision'>The age of study sample deviated significantly from the 2023 national registry data with weak effect size (`r filter(lit_kurasi$test, variable == 'age_class')$plot_cap`).</span>
-<span custom-style = 'revision'>In particular, accident victims aged 51 - 60 years were more frequent in the study cohort (__Supplementary Figure S1__).</span>
+<span custom-style = 'revision'>Percentages of ski/snowboard accidents in non-secured areas and hiking accidents were in turn significantly lower in the study cohort than in the national registry (global difference: `r filter(lit_kurasi$test, variable == 'sport_detail')$plot_cap`).</span>
+<span custom-style = 'revision'>Effect sizes of the differences between the study sample and the national registry were small (__Supplementary Figure S1__).</span>
 
@@
 (`r get_percent(ptsd$dataset, 'accident_alone')$percent['yes']`%) and, in most cases, were the only person responsible to the accident 
-(`r get_percent(ptsd$dataset, 'accident_culprit')$percent['self']`). 
+(`r get_percent(ptsd$dataset, 'accident_culprit')$percent['self']`%). 
 Professional rescue service was involved in 
@@
 and somatization (`r get_percent(ptsd$dataset, 'phq_events_total_class')$percent['positive']`%) were rare in this cohort consisting of `r get_percent(ptsd$dataset, 'rs13_total_class')$percent['high']`% highly resilient individuals (__Table \@ref(tab:tab-cohort-mental)__).
-<span custom-style = 'revision'>Mean scoring of quality of life assessed by the EUROHIS tool in the cohort was `r mean(ptsd$dataset$eurohis_total) %>% signif(2)` and was significantly higher that the mean of 4.08 reported for the comparable general German population (one-sample Wilcoxon test: r = `r ptsd$dataset %>% wilcox_effsize(eurohis_total ~ 1, mu = 4.08) %>% .$effsize %>% signif(2)`, p < 0.001) [@Schmidt2006].</span>
-<span custom-style = 'revision'>The general quality of life and housing domains were assigned the highest scores, while esteem was rated the lowest among the study participants.</span> 
+<span custom-style = 'revision'>Mean scoring of quality of life in the cohort was significantly higher than in the comparable general German population (cohort: `r mean(ptsd$dataset$eurohis_total) %>% signif(2)`, population estimate: 4.08, one-sample Wilcoxon test: r = `r ptsd$dataset %>% wilcox_effsize(eurohis_total ~ 1, mu = 4.08) %>% .$effsize %>% signif(2)`, p < 0.001) [@Schmidt2006].</span>
 <span custom-style = 'revision'>The median total score of accident-related post-traumatic growth was `r median(ptsd$dataset$ptgi_total)` (interquartile range: `r quantile(ptsd$dataset$ptgi_total, c(0.25, 0.75)) %>% paste(collapse = ' to ')`), with the relations domain with the highest rating and the spiritual domain with the lowest scores (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S3__).</span>
@@
 
-<span custom-style = 'revision'>As compared with general population estimates [@Hauffa2011; @Kilpatrick2013; @Koenen2017; @Kessler2017], frequency of traumatic events (`r lit_data$stats %>% get_est_ci(source == 'cohort', variable == 'traumatic_event', level == 'yes')`) and manifest PTSD in the study sample defined as positivity for all PCL-5 domains (`r lit_data$stats %>% get_est_ci(source == 'cohort', variable == 'dsm5_all_class', level == 'positive')`) was low but comparable to figures reported for the general German population by Hauffa and colleagues (traumatic events, CIDI tool: `r lit_data$stats %>% get_est_ci(source == 'hauffa_2011', variable == 'traumatic_event', level == 'yes')`, manifest PTSD, PDS tool: `r lit_data$stats %>% get_est_ci(source == 'hauffa_2011', variable == 'ptsd_lifetime', level == 'positive')`, __Supplementary Figure S4A__) [@Hauffa2011].</span>
-<span custom-style = 'revision'>Furthermore, frequency of traumatic events in the study cohort was significantly lower than in a sample of Swiss mountain rescuers reported by Mikutta et al.
-(`r lit_data$stats %>% get_est_ci(source == 'mikutta_2022', variable == 'traumatic_event', level == 'yes')`) [@Mikutta2022].</span>
-<span custom-style = 'revision'>In turn, frequency of PTSD symptoms and of manifest PTSD measured with the PCL-5 instrument were similar in both cohorts (__Supplementary Figure S4B__).</span>
-
-<span custom-style = 'revision'>Our sample was characterized by significantly lower percentages of anxious (GAD-7 $\geq$ 10, `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'anxiety')`) and depressive symptoms (PHQ-9 $\geq$ 10, `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'depression')`) as compared with estimates for the general Austrian population in 2022 by Humer et al.
-(anxiety, GAD-7 $\geq$ 10: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'anxiety')`, depression, PHQ-9 $\geq$ 10: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'depression')`, __Supplementary Figure S5A__) [@Humer2022].</span>
-<span custom-style = 'revision'>As assessed by the PHQ-8 instrument, the frequency of moderate-to-severe depressive symptoms in the study sample was similar to the frequency in the pre-pandemic general Austrian population estimated by a 2019 microcensus study [@StatistikAustria2019a] (__Supplementary Figure S5B__).</span>
-<span custom-style = 'revision'>The resilience scores in our study collective (mean: `r lit_plots$rs_stats %>% filter(source == 'cohort') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'cohort') %>% .$sd %>% signif(2)`) were comparable with resilience levels in the Swiss mountain rescuer sample (mean: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$sd %>% signif(2)`) [@Mikutta2022], and higher as compared with the general population estimates obtained by Chernova et al.
-with the same RS-13 tool (mean: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$sd %>% signif(2)`, __Supplementary Figure S5C__) [@Chernova2021].</span>
+<span custom-style = 'revision'>As compared with general population estimates [@Hauffa2011; @Kilpatrick2013; @Koenen2017; @Kessler2017], frequency of traumatic events and manifest PTSD in the study sample defined as positivity for all PCL-5 domains was low.</span>
+<span custom-style = 'revision'>Notably, frequency of PTSD symptoms and of manifest PTSD measured with the PCL-5 instrument were similar in the study cohort and in a sample pf Swiss mountain rescuers [@Mikutta2022] (at least one PCL-5 domain positive, cohort: `r lit_plots$ptsd_symptom_stats %>% get_est_ci(source == 'cohort', variable == 'dsm5_cluster_class')`, Mikutta et al.: `r lit_plots$ptsd_symptom_stats %>% get_est_ci(source == 'mikutta_2022', variable == 'dsm5_cluster_class')`; all domains positive, cohort: `r lit_plots$ptsd_symptom_stats %>% get_est_ci(source == 'cohort', variable == 'dsm5_all_class')`, Mikutta et al.: `r lit_plots$ptsd_symptom_stats %>% get_est_ci(source == 'mikutta_2022', variable == 'dsm5_all_class')`, __Supplementary Figure S4__).</span>
+<span custom-style = 'revision'>Our sample was characterized by significantly lower percentages of clinically relevant symptoms of anxiety (GAD-7 $\geq$ 10, `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'anxiety')`) and depression (PHQ-9 $\geq$ 10, `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'depression')`) as compared with estimates for the general Austrian population in 2022 (anxiety, GAD-7 $\geq$ 10: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'anxiety')`, depression, PHQ-9 $\geq$ 10: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'depression')`, __Supplementary Figure S5A__) [@Humer2022].</span>
+<span custom-style = 'revision'>The resilience scores in our study collective (mean: `r lit_plots$rs_stats %>% filter(source == 'cohort') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'cohort') %>% .$sd %>% signif(2)`) were comparable with resilience levels in the Swiss mountain rescuer sample [@Mikutta2022], and higher as compared with the general population estimates obtained by Chernova et al.
+with the same RS-13 tool (mean: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$mean %>% signif(2)`, SD: `r lit_plots$rs_stats %>% filter(source == 'chernova_2021') %>% .$sd %>% signif(2)`, __Supplementary Figure S5B__) [@Chernova2021].</span>
 
@@
 test subset (n = `r table(ptsd$dataset$partition)['test']`). 
-The sole significant differences between these subsets concerned resilience class distribution and scores of quality of life, their effect size was weak <span custom-style = 'revision'>(__Supplementary Table S7__).</span>
+The sole significant differences between these subsets concerned resilience class distribution and scores of quality of life, their effect size was small <span custom-style = 'revision'>(__Supplementary Table S7__).</span>
 
@@
 moderate resilience classes (PTS: `r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'rs13_total_class')$percent['moderate']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'rs13_total_class')$plot_cap`</span>) peaked in the PTS cluster.
-<span custom-style = 'revision'>Concerning PTSD symptoms, frequency of flashbacks as well as percentages of individuals screened positive for each of the B, C, D, and E PCL-5 domains were significantly higher in PTS cluster patients as compared with the remaining individuals.</span>
+<span custom-style = 'revision'>Concerning PTSD symptoms, frequency of flashbacks as well as percentages of individuals screened positive for each of the B, C, D, and E PCL-5 domains were significantly higher in the PTS as compared with the remaining clusters.</span>
 <span custom-style = 'revision'>Analogically, the frequency of participants with PTSD symptoms defined by at least one positive PCL-5 domain was</span> the highest in <span custom-style = 'revision'>the</span> PTS cluster (`r get_percent(filter(clust_bcg$analysis_tbl, clust_id == 'PTS'), 'dsm5_cluster_class')$percent['positive']`%, <span custom-style = 'revision'>`r filter(clust_bcg$test, variable == 'dsm5_cluster_class')$plot_cap`</span>, __Figure \@ref(fig:fig-clust-symptoms)__, <span custom-style = 'revision'>__Supplementary Table S9__</span>).
 
-<span custom-style = 'revision'>Next, we investigated, how the clustering of study participants is affected by missing observations, time after the accident, and factors that may influence mental health after the accident such as annual income, prior mountain sport accidents, injury severity, and hospitalization.</span>
-<span custom-style = 'revision'>To this end, we re-created the clustering structure for the whole cohort, the study cohort appended with imputed missing observations, and the cohort data set depleted of categories of the variables listed above (e.g.
-the analysis cohort without high income participants or severe injuries) with the medoid clustering algorithm.</span>
-<span custom-style = 'revision'>Next, we compared fractions of explained clustering variances between the clustering structures in the genuine cohort and the modified data sets in a kind of sensitivity analysis [@Thabane2013].</span>
-<span custom-style = 'revision'>As presented in __Supplementary Figure S10A__, differences in explained variance between the original clustering structure and clustering structures developed for the modified data sets were minor (genuine data set: `r se_summary$stats %>% filter(split_factor == 'global') %>% .$frac_var %>% signif(2)`, modified: `r se_summary$stats %>% filter(split_factor %in% c('missing', 'household_income_class', 'prior_accident', 'injury_sev_strata', 'hospitalization', 'accident_year')) %>% .$frac_var %>% range %>% signif(2) %>% paste(collapse = ' to ')`).</span>
-<span custom-style = 'revision'>There were no significant differences in distribution of the cluster sizes for the clustering schemes developed in the analyzed cohort and the modified data sets (__Supplementary Figure S10B__).</span>
-<span custom-style = 'revision'>Collectively, this lets us conclude that neither missing observations, differences in income, accidents in the past, time after accident, injury severity, nor hospitalization significantly affected quality and distribution of the mental health clusters.</span>
-<span custom-style = 'revision'>In the classification scheme developed for the data set appended with incomplete observations, the participants with missing psychometric data which had been successfully imputed were preferentially assigned to the PTS cluster (imputed observations in the PTS cluster: n = `r se_summary$missing_plots$color_bar_data %>% filter(clust_id == 'PTS', subset == 'missing') %>% nrow`, total imputed: n = `r table(se_summary$missing_plots$color_bar_data$subset)['missing']`, __Supplementary Figure S11__).</span>
+<span custom-style = 'revision'>Effects of missing observations, year of the accident as a readout of the accident - survey time, as well as factors of potential importance for mental health such as annual income, prior mountain sport accidents, injury severity, and hospitalization were investigated by sensitivity analysis (__Supplementary Methods__) [@Thabane2013].</span>
+<span custom-style = 'revision'>Those factors had only a minor influence on clustering as assessed by fractions of explained clustering variance (genuine data set: `r se_summary$stats %>% filter(split_factor == 'global') %>% .$frac_var %>% signif(2)`, sensitivity analysis: `r se_summary$stats %>% filter(split_factor %in% c('missing', 'household_income_class', 'prior_accident', 'injury_sev_strata', 'hospitalization', 'accident_year')) %>% .$frac_var %>% range %>% signif(2) %>% paste(collapse = ' to ')`) or cluster size distribution (__Supplementary Figure S10__).</span>
+<span custom-style = 'revision'>In the clustering of the study cohort appended with incomplete observations, the participants with missing psychometric data which had been successfully imputed were preferentially assigned to the PTS cluster (imputed observations in the PTS cluster: n = `r se_summary$missing_plots$color_bar_data %>% filter(clust_id == 'PTS', subset == 'missing') %>% nrow`, total imputed: n = `r table(se_summary$missing_plots$color_bar_data$subset)['missing']`, __Supplementary Figure S11__).</span>
 
@@
 were found to differ significantly between the mental health clusters. 
-The effect size of these differences was weak <span custom-style = 'revision'>(__Supplementary Table S10__).</span>
+The effect size of these differences was small <span custom-style = 'revision'>(__Supplementary Table S10__).</span>
 
@@
 Brier score: `r signif(range(filter(early_class$overall_stats, dataset == 'test')$brier_score), 2) %>% paste(collapse = ' - ')`). 
-Concerning the early identification of the vulnerable PTS cluster patients, the conditional random forest, random forest and support vector machine models were able to differentiate between the PTS cluster and the pooled neutral and PTG clusters with the best sensitivity (test subset sensitivity $\geq$ `r min(signif(early_class$clust_stats %>% filter(dataset == 'test', clust_id == 'PTS') %>% top_n(n = 3, Se) %>% .$Se, 2) * 100)`%, __Figure \@ref(fig:fig-early-class)__, <span custom-style = 'revision'>(__Supplementary Table S13__)</span>).
+<span custom-style = 'revision'>Accordingly, sensitivity of discrimination of  the PTS cluster from the pooled neutral and PTG clusters was unsatisfactory in the cross-validation and test data (test, sensitivity: `r early_class$clust_stats %>% filter(.outcome == 'PTS', dataset == 'test') %>% .$Se %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-early-class)__, __Supplementary Table S13__).</span>
 Pre-existing mental disorder or physical illness, hospitalization and surgery, injury severity, age, rescue mode, education, annual income, and traumatic events in the past were among the most important explanatory factors employed by these models <span custom-style = 'revision'>(__Supplementary Figure S14__).</span>
@@
 
-<span custom-style = 'revision'>The analyzed cohort was significantly enriched with males, secondary/tertiary graduates, professionally active, and physically healthier individuals as compared with the general Austrian population.</span>
+<span custom-style = 'revision'>Whereas our study cohort was clearly not a representative sample of the alpine region population, we compared the analyzed sample with existing literature to put our results in a broader context of mental health in mountain sports and accident survivors.</span>
+<span custom-style = 'revision'>The analyzed cohort contained significantly more males, secondary/tertiary graduates, professionally active, and physically healthier individuals as compared with the general Austrian population.</span>
 <span custom-style = 'revision'>We speculate that this reflects rather the overall characteristic of physically active inhabitants of the alpine region than a selection bias.</span>
-<span custom-style = 'revision'>Differences in the accident sport type, accident season, and age structure between the study cohort and the collective of mountain accident victims in the Austrian registry were of weak effect size.</span>
+<span custom-style = 'revision'>Differences in the accident sport type, accident season, and age structure between the study cohort and the collective of mountain accident victims in the Austrian registry were of small effect size.</span>
 <span custom-style = 'revision'>Within the analyzed sample, winter sport accidents in general, and, in particular, alpine skiing/snowboarding accidents in secured areas of ski resorts were significantly more frequent.</span>
@@
 <span custom-style = 'revision'>As compared with the non-responders, injury severity, and hospitalization and surgery rates were significantly higher in the analyzed cohort, effect size of those differences was, however, small.</span>
-<span custom-style = 'revision'>Collectively, despite the low response rate, our study cohort can be regarded as reasonably representative sample of the entire collective of mountain sport accident victims in the alpine region.</span>
+<span custom-style = 'revision'>Individuals with incomplete mental battery data were preferentially assigned to the PTS cluster in the sensitivity analysis.</span>
+<span custom-style = 'revision'>This suggests mental distress or disorders as one of the reasons for not completing of the study survey.</span>
+<span custom-style = 'revision'>Additionally, this may indicate that symptoms of mental health disorders are under-reported in the study cohort.</span>
 
@@
 <span custom-style = 'revision'>We compared rates of traumatic events measured by DIA-X tool [@Maercker2005], and of manifest PTSD defined as positive screening for all of the B, C, D and E of the PCL-5 instrument [@Bovin2016] in our cohort with published estimates for the general population.</span>
-<span custom-style = 'revision'>In such analysis, the frequency of both traumatic events (`r get_percent(ptsd$dataset, variable = 'traumatic_event')$percent['yes']`%) and manifest PTSD (`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) in the cohort was significantly lower than in most multi-population surveys [@Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017], and comparable with estimates provided for the German population by Hauffa and colleagues (traumatic events: `r lit_data$stats %>% filter(source == 'hauffa_2011', variable == 'traumatic_event', level == 'yes') %>% .$percent %>% signif(2)`%, manifest lifetime PTSD: `r lit_data$stats %>% filter(source == 'hauffa_2011', variable == 'ptsd_lifetime', level == 'positive') %>% .$percent %>% signif(2)`%) [@Hauffa2011].</span>
+<span custom-style = 'revision'>It is important to note that we specifically asked participants to rate symptoms of PTSD in relation to their mountain sport accident.</span>
+<span custom-style = 'revision'>In such analysis, the frequency of both traumatic events prior to the accidents and accident-related manifest PTSD in the cohort was significantly lower than in most multi-population surveys [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017].</span>
 <span custom-style = 'revision'>Despite the substantial burden of mountain sport accidents on medical system of alpine countries, with more than 13000 accidents in 2023 in Austria alone [@OsterreichischesKuratoriumfurAlpineSicherheit2024], not many studies assessed rates of PTSD and other mental disorders in survivors of mountain sport accidents.</span>.
@@
 <span custom-style = 'revision'>A similarly low frequency of manifest PTSD and an analogical profile of single PTSD symptoms were observed in the study cohort and the cohort of mountain rescue workers investigated with the PCL-5 tool by Mikutta et al.
-(manifest PTSD: `r lit_data$stats %>% filter(source == 'mikutta_2022', variable == 'dsm5_all_class', level == 'positive') %>% .$percent %>% signif(2)`) [@Mikutta2022].</span>
+[@Mikutta2022].</span>
 Resilience has been proposed as a protective trait against negative mental health consequences following trauma [@Bensimon2012]. 
@@
 Our study bears limitations. 
+<span custom-style = 'revision'>Our cohort is not a representative sample of the general population.</span>
 The overall analysis inclusion rate was low and, as indicated by comparison of the excluded and analyzed participants, the results were potentially affected by a selection bias towards individuals with higher social status and more severe injuries.
-<span custom-style = 'revision'>Low response rates are unfortunately a problem of many studies tackling PTSD and mental disorders [@Ophuis2018] and may lead to under- or over-reporting of symptoms [@Angerpointner2020; @Merckelbach2019; @Takayanagi2014].</span>
-<span custom-style = 'revision'>Our comparison of demographic and accident-related features in the study cohort, Austrian population, and the national registry of alpine accidents indicates that the analyzed collective is a fairly representative sample of the general physically active population.</span>
-<span custom-style = 'revision'>However, we can not exclude that the low frequency of manifest PTSD, and clinically relevant symptoms of anxiety and depression is in part attributed to the low response rate and under-reporting of mental health problems.</span>
-<span custom-style = 'revision'>In fact, assignment of the survey participants with incomplete mental health battery to the PTS cluster suggests that mental health problems were one of the reasons of data missingness.</span>
-<span custom-style = 'revision'>Additionally, enrichment of the study sample with men and individuals of high social status, which are considered protective factors of mental health disorders, may have contributed to the low frequency of symptoms of PTSD, anxiety and depression.</span>
+<span custom-style = 'revision'>Our comparison of demographic and accident-related features in the study cohort, Austrian population, and the national registry of alpine accidents indicates that the analyzed collective is a sample of the general physically active population.</span>
+<span custom-style = 'revision'>Low response rates are unfortunately a problem of many studies tackling PTSD and mental disorders [@Ophuis2018].</span>
+<span custom-style = 'revision'>Mental health problems were likely one of the reasons of data missingness in our analysis.</span>
 <span custom-style = 'revision'>Therefore, our results may underestimate frequency of mental disorder symptoms in survivors of mountain sport accidents and have to be interpreted with caution.</span>
@@
 <span custom-style = 'revision'>As assessed by sensitivity analysis, our clustering structure was hardly affected by the time of between the accident and survey or prior mountain sport accidents.</span>
-<span custom-style = 'revision'>This suggests that the mental health clusters described by us reflect accident-specific alterations of mental health.</span>
 <span custom-style = 'revision'>However, to finally clarify the issue of the accident specificity of mental health alterations, follow-up studies with longitudinal design or a control group are required.</span>
@@
 <span custom-style = 'revision'>In the current analysis,</span> we could not identify robust predictors <span custom-style = 'revision'>of</span> impaired mental health at follow up among variables available during early <span custom-style = 'revision'>routine</span> medical treatment.
-<span custom-style = 'revision'>Hence, until standardized robust predictor variables and screening protocols are available,</span> low threshold access to psychological support <span custom-style = 'revision'>as well as to psychoeducational measures</span> seems the most appropriate strategy for successful interdisciplinary management of individuals following accidents in mountain sport.
+<span custom-style = 'revision'>Hence, until standardized robust predictor variables are available and screening protocols assessing mental health outcomes are widely implemented,</span> low threshold access to psychological support <span custom-style = 'revision'>as well as psychoeducational measures</span> seems the most appropriate strategy for successful interdisciplinary management of individuals following accidents in mountain sport.
 
@@
 
-An R data (RData) file with anonymized patient data will be made available upon request to the corresponding author. 
+Anonymized patient data will be made available upon request to the corresponding author. 
 The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.

### 2024-03-04 paper/markdown/supplementary_material.Rmd
Figure and Table update

@@
 clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 10 [@Humer2022], 
-significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
+significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 10 [@Kroenke2002]. 
 Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008]. 
@@
 [@Mikutta2022].
-
 Data on frequency of clinically relevant anxious (GAD-7 $\geq$ 10) and depressive (PHQ-9 $\geq$ 10) symptoms in Austria in 2022 were obtained from a publication by Humer et al. [@Humer2022]. 
-Data on frequency and severity of depressive symptoms measured with the PHQ-8 instrument [@Kroenke2002a] in a pre-pandemic nation-wide Austrian 2019 microcensus study were obtained from a publication by Statistik Austria [@StatistikAustria2019a].
-
 Mean scores of resilient comping with their standard deviations (SD) obtained with the RS-13 tool were extracted from reports with samples of the German [@Leppert2008] and Tyrolean [@Chernova2021] populations, and from a cohort of Swiss mountain rescuers [@Mikutta2022].
@@
 
-* r statistic: weak: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+* r statistic: small: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
 
-* Cramer's V statistic: weak: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+* Cramer's V statistic: small: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
 
-* $\eta^2$ statistic: weak: < 0.06, moderate: 0.06 - 0.14, large: $\geq$ 0.14
+* $\eta^2$ statistic: small: < 0.06, moderate: 0.06 - 0.14, large: $\geq$ 0.14
 
-* Cohen's $\kappa$ inter-rater reliability statistic: none: < 0.2, minimal: 0.2 - 0.4, weak: 0.4 - 0.6, moderate: 0.6 - 0.8, strong: $\geq$ 0.8
+* Cohen's $\kappa$ inter-rater reliability statistic: none: < 0.2, minimal: 0.2 - 0.4, small: 0.4 - 0.6, moderate: 0.6 - 0.8, large: $\geq$ 0.8
 
@@
 
-An RData file with anonymized patient data will be made available upon request to the corresponding author. 
+Anonymized patient data will be made available upon request to the corresponding author. 
 The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.
@@
   set_widths(c(3.5, rep(5, 2), 2.8, 2.8)) %>% 
+  footnote(7, 1, 
+           value = as_paragraph('In the study survey: participants reporting a chronic pre-existing physical illness or a mental disorder diagnosed by a medical professional.'), 
+           ref_symbols = 'a', 
+           part = 'body') %>%
   footnote(1, 4:5, 
            value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic, categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
-           ref_symbols = 'a', 
+           ref_symbols = 'b', 
            part = 'header') %>% 
@@
 
-```{r fig-accident, fig.width = figur::convert(suppl_figures$accident, to = 'in')$w, fig.height = figur::convert(suppl_figures$accident, to = 'in')$h, fig.cap = 'Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.'}
+```{r fig-accident, fig.width = figur::convert(suppl_figures$accident, to = 'in')$w, fig.height = figur::convert(suppl_figures$accident, to = 'in')$h, fig.cap = 'Age, mountain sport type, and accident month in the study cohort and Austrian national registry of mountain accidents.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-accident). Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.__
+__Supplementary Figure S\@ref(fig:fig-accident). Age, mountain sport type, and accident month in the study cohort and Austrian national registry of mountain accidents.__
 
-_Mountain sport types (A), accident months (B) and age class distribution (C) were compared between the study cohort and victims of mountain sport accidents in Austria in 2023.
+_Age class distribution (A), mountain sport types (B), and accident months (C) were compared between the study cohort and victims of mountain sport accidents in Austria in 2023.
 Frequencies for the nation-wide data set of mountain accidents were extracted from the annual report of the Austrian Board of Mountain Safety.
+In the comparison of the age class distribution, only individuals older than 20 years were included.
 In the comparison of sport types, non-sportive mountain accidents (e.g.
@@
 Non-classifiable sport types (category 'other') in the study cohort were excluded from the analysis presented in (C).
-In the comparison of the age class distribution, only individuals older than 20 years were included.
 Statistical significance of the global differences was determined by $\chi^2$ test with Cramer's V effect size statistic.
@@
 
-_(B) Comparison of frequencies of traumatic events, manifest PTSD and PTSD symptoms in the study cohort and the cohort of Swiss mountain rescuers reported by Mikutta and colleagues.
+_(B) Comparison of manifest PTSD and PTSD symptoms in the study cohort and the cohort of Swiss mountain rescuers reported by Mikutta and colleagues.
 PTSD symptoms were assessed by the PCL-5 tool and are subsumed under the B, C, B, and E domains.
@@
 
-```{r fig-mental-literature, fig.width = figur::convert(suppl_figures$mental_literature, to = 'in')$w, fig.height = figur::convert(suppl_figures$mental_literature, to = 'in')$h, fig.cap = 'Comparison of frequency of anxious and depressive symptoms and of resilience scoring in the study cohort and literature.'}
+```{r fig-mental-literature, fig.width = figur::convert(suppl_figures$mental_literature, to = 'in')$w, fig.height = figur::convert(suppl_figures$mental_literature, to = 'in')$h, fig.cap = 'Comparison of frequency of clinically relevant symptoms of anxiety and depression, and of resilience scoring in the study cohort and literature.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-mental-literature). Comparison of frequency of anxious and depressive symptoms and of resilience scoring in the study cohort and literature.__
+__Supplementary Figure S\@ref(fig:fig-mental-literature).
+Comparison of frequency of clinically relevant symptoms of anxiety and depression, and of resilience scoring in the study cohort and literature.__
 
@@
 
-_(B) Comparison of frequency and severity of depressive symptoms measured by the PHQ-8 tool in the study cohort with 2019 estimates for the general Austrian population._
-
-_(C) Comparison of resilient comping scores in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta et al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.)_
+_(C) Comparison of scores of resilient coping in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta et al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.) investigated with the RS13 instrument._
 
